---
title: "Biomass Pipeline — Live Explorer"
subtitle: "A Shiny-powered companion to pipeline_documentation.qmd: pick a year range and a set of GSAs, and the figures below actually re-run against real per-sample MEDITS data — not just a generated config block."
author: "Daniel Vilas"
format: html
server: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

suppressWarnings(suppressMessages({
  library(shiny)
  library(data.table)
  library(leaflet)
  library(ggplot2)
  library(sf)
}))

## Resolve every relative path below against THIS DOCUMENT's own directory,
## not whatever the current working directory happens to be when the R
## process starts. quarto/knitr USUALLY sets the working directory to match
## already, but "usually" isn't "always" - it depends on `execute.dir` in a
## _quarto.yml (if this file is part of a project), and on exactly where/how
## `quarto serve`/`quarto preview` was invoked from. A relative path silently
## resolving against the wrong directory is exactly the kind of failure that
## surfaces as a LATER, unrelated-looking error (e.g. deep in a Shiny
## reactive) rather than a clear "file not found" right here - anchoring
## explicitly removes that ambiguity, and every error message below prints
## the full resolved path so a real "it's genuinely not there" case is easy
## to tell apart from a "wrong base directory" case.
DOC_DIR <- tryCatch({
  ci <- knitr::current_input(dir = TRUE)
  if (is.null(ci) || !nzchar(ci)) stop("knitr::current_input() returned nothing")
  dirname(normalizePath(ci, mustWork = TRUE))
}, error = function(e) {
  message("[pipeline_documentation_shiny] Could not resolve this document's own directory via ",
          "knitr::current_input() (", conditionMessage(e), ") - falling back to the current ",
          "working directory instead. If LIB_PATH/CACHE_PATH below can't be found, try running ",
          "quarto from this .qmd's own folder, or set LIB_PATH/CACHE_PATH to absolute paths.")
  getwd()
})
message("[pipeline_documentation_shiny] Resolving paths relative to: ", DOC_DIR)

## This document needs the SAME library file every numbered pipeline script
## sources, since the reactive code below calls its real aggregation/plotting
## functions (compute_fg_densities_by_stratum(), weight_by_strata(),
## weight_by_area(), plot_fg_timeseries_by_area(), plot_fg_timeseries_regional(),
## plot_strata_profile(), plot_sample_map()) - NOT a reimplementation of them.
## Adjust this path if lib_survey_fg_density_functions.R lives somewhere else
## relative to this .qmd (an absolute path works too - DOC_DIR is only
## prepended to a RELATIVE path, via file.path() below).
LIB_PATH <- file.path(DOC_DIR, "lib_survey_fg_density_functions.R")
if (!file.exists(LIB_PATH)) {
  stop("Can't find '", LIB_PATH, "' (resolved from this document's own directory, '", DOC_DIR,
       "') - set LIB_PATH above to wherever lib_survey_fg_density_functions.R actually is, ",
       "relative to this file or as an absolute path.")
}
source(LIB_PATH)

## The data cache 01_biomass.R optionally writes when SAVE_SHINY_CACHE_PATH is
## set (see that variable's own comment in 01_biomass.R's Configuration
## section) - the final per-sample MEDITS table (FG-matched, area/stratum-
## assigned, catchability-corrected, outliers already resolved), plus the GSA
## shapefile and strata definition needed to re-aggregate it. Building this
## cache is a one-time step, done by running 01_biomass.R itself once (ideally
## with FILTER_AREAS/YEAR_ECOPATH/TS_YEARS set to their full available extent,
## so this app has the widest possible range to explore) - see the callout
## below for the exact steps.
##
## Missing cache is NOT a stop() here on purpose - unlike LIB_PATH above
## (genuinely fatal; there's no app at all without the shared functions),
## a missing cache is the normal state the FIRST time anyone opens this
## document, before they've run 01_biomass.R once. CACHE_AVAILABLE gates
## every reactive/render block below via shiny::validate(need(...)), which
## Shiny displays as a clean inline message in place of that output - not
## an uncaught error that kills the whole `quarto serve` process. The app
## still starts and shows the "Before you start" instructions either way.
CACHE_PATH <- file.path(DOC_DIR, "shiny_cache/survey_density_raw.rds")
CACHE_AVAILABLE <- file.exists(CACHE_PATH)
CACHE_MISSING_MSG <- paste0(
  "Data cache not found at '", CACHE_PATH, "' (resolved from this document's own directory, '",
  DOC_DIR, "'). See \"Before you start\" above for how to build it by running 01_biomass.R once ",
  "with SAVE_SHINY_CACHE_PATH set, then reload this page. If you DID build it and it's still not ",
  "found here, check it actually landed in a 'shiny_cache/' folder next to THIS .qmd file, not ",
  "next to 01_biomass.R's own out_dir - those are not the same directory unless you put this ",
  "document there too."
)

## AUTO_BUILD_CACHE: if TRUE and no cache exists yet, this app builds one
## itself automatically, by launching 01_biomass.R (with
## SAVE_SHINY_CACHE_PATH/STOP_AFTER_SHINY_CACHE set) as a genuinely
## SEPARATE background process (Rscript, via system2(..., wait = FALSE)) -
## NOT source()'d in-process the way an earlier version of this did. That
## distinction is exactly what makes this safe to default to TRUE: this
## process returns immediately, so the page still paints straight away
## (into the synthetic demo dataset below, since the real cache obviously
## doesn't exist yet at this exact instant) instead of blocking on a
## multi-minute pipeline run the way the old in-process approach did. A
## background poll (see the cache-state chunk further down) then picks up
## the real cache and swaps every control/figure over to it automatically
## - no page reload needed - once that background process finishes (a few
## minutes: GSA shapefile + bathymetry downloads, MEDITS loading, FG
## matching). A grey "demo data" banner stays visible the whole time.
##
## Set AUTO_BUILD_CACHE <- FALSE here only if you specifically don't want
## this (e.g. no pCloud access at all from wherever quarto serve/preview
## is running) - the app then just stays on demo data until you build a
## real cache by hand (see "Before you start" below) and reload the page.
AUTO_BUILD_CACHE <- TRUE

BUILD_LOG_PATH  <- file.path(DOC_DIR, "shiny_cache", "auto_build.log")
BUILD_LOCK_PATH <- file.path(DOC_DIR, "shiny_cache", "auto_build.lock")

if (!CACHE_AVAILABLE && AUTO_BUILD_CACHE) {
  BIOMASS_SCRIPT_PATH <- file.path(DOC_DIR, "01_biomass.R")
  ## A recent lock file means a background build is already in flight -
  ## e.g. the user reloaded the page while an earlier build was still
  ## running. Without this check, EVERY reload during that window would
  ## spawn ANOTHER competing Rscript process running the full pipeline in
  ## parallel. 30 minutes is generous slack for a crashed/killed process
  ## that never got to remove its own lock (see the finally= below) -
  ## past that, a fresh attempt is allowed again rather than staying
  ## stuck forever on a stale lock.
  lock_is_stale <- !file.exists(BUILD_LOCK_PATH) ||
    difftime(Sys.time(), file.info(BUILD_LOCK_PATH)$mtime, units = "mins") > 30

  if (!file.exists(BIOMASS_SCRIPT_PATH)) {
    message("[pipeline_documentation_shiny] AUTO_BUILD_CACHE is TRUE but '", BIOMASS_SCRIPT_PATH,
            "' isn't next to this document, so there's nothing to auto-run - staying on demo data",
            " until a real cache is built by hand (see \"Before you start\" below).")
  } else if (!lock_is_stale) {
    message("[pipeline_documentation_shiny] A background cache build already appears to be in",
            " progress (lock file '", BUILD_LOCK_PATH, "' is recent) - not starting another one.",
            " Showing demo data until it finishes; progress/errors: '", BUILD_LOG_PATH, "'.")
  } else {
    dir.create(dirname(CACHE_PATH), recursive = TRUE, showWarnings = FALSE)
    writeLines(as.character(Sys.time()), BUILD_LOCK_PATH)
    ## Passed as a single -e expression (not a temp .R file) to keep this
    ## self-contained in one chunk. finally= removes the lock whether the
    ## build succeeds or fails, so a later restart isn't stuck behind it.
    build_expr <- paste0(
      "SAVE_SHINY_CACHE_PATH <- ", deparse(CACHE_PATH), "; ",
      "STOP_AFTER_SHINY_CACHE <- TRUE; ",
      "tryCatch({ source(", deparse(BIOMASS_SCRIPT_PATH), ") }, ",
      "shinyCacheStop = function(e) invisible(NULL), ",
      "error = function(e) cat('AUTO BUILD FAILED: ', conditionMessage(e), '\\n'), ",
      "finally = unlink(", deparse(BUILD_LOCK_PATH), "))"
    )
    spawn_ok <- tryCatch({
      system2("Rscript", args = c("-e", shQuote(build_expr)),
              stdout = BUILD_LOG_PATH, stderr = BUILD_LOG_PATH, wait = FALSE)
      TRUE
    }, error = function(e) {
      message("[pipeline_documentation_shiny] Could not start the background build (",
              conditionMessage(e), ") - is Rscript on this machine's PATH? Staying on demo data;",
              " you can still build the cache by hand (see \"Before you start\" below).")
      unlink(BUILD_LOCK_PATH)
      FALSE
    })
    if (isTRUE(spawn_ok)) {
      message("[pipeline_documentation_shiny] No cache found - building one automatically in the",
              " background (a separate Rscript process, not blocking this app). This app already",
              " shows demo data below and will switch over to your real results automatically,",
              " with no reload needed, once that finishes. Progress/errors: '", BUILD_LOG_PATH, "'.")
    }
  }
}

## build_demo_cache(): a small, entirely synthetic stand-in for a real
## SAVE_SHINY_CACHE_PATH cache - same list shape 01_biomass.R saves
## (dt/area_shp/area_id_col/strata_def), but with made-up polygons,
## strata and FGs instead of real MEDITS/GSA data. Its only purpose is
## to give this app SOMETHING to show - populated sliders, a real map,
## real (if meaningless) figures - the moment it's opened, even before
## any real cache exists or an automatic build has had a chance to
## succeed. Every value in it is fabricated; IS_DEMO_DATA (set where
## this is used, below) is what tells the UI to say so.
##
## Deliberately does NOT call compute_strata_area_by_area() - that
## function downloads REAL NOAA bathymetry for whatever polygon it's
## given, which would either fail outright or return nonsense for a
## hand-drawn square that isn't really seafloor. Its output shape
## (AreaID/Stratum/area_km2/prop) is instead hand-built directly here.
build_demo_cache <- function() {
  set.seed(42)

  ## Two small square "GSA-like" polygons, far enough apart to look
  ## like distinct areas on the map, both with an explicit CRS (plot_
  ## sample_map() stop()s without one - see its own comment).
  poly1 <- sf::st_polygon(list(rbind(c(3, 40), c(4, 40), c(4, 41), c(3, 41), c(3, 40))))
  poly2 <- sf::st_polygon(list(rbind(c(5.5, 39.5), c(6.5, 39.5), c(6.5, 40.5), c(5.5, 40.5), c(5.5, 39.5))))
  demo_area_shp <- sf::st_sf(AreaID = c(1, 2), geometry = sf::st_sfc(poly1, poly2, crs = 4326))

  demo_strata_def <- data.table(stratum_num = 1:3, depth_min = c(0, 50, 100), depth_max = c(50, 100, 200))
  demo_fgs <- data.table(FG_num = 1:3, FG_name = c("Demo hake-like FG", "Demo shrimp-like FG", "Demo mullet-like FG"))
  demo_years <- 2015:2023
  n_samples_per_area_year <- 25

  samples <- rbindlist(lapply(list(list(id = 1, poly = poly1), list(id = 2, poly = poly2)), function(a) {
    bb <- sf::st_bbox(a$poly)
    rbindlist(lapply(demo_years, function(yr) {
      n <- n_samples_per_area_year
      data.table(
        SampleID = paste0("DEMO_", a$id, "_", yr, "_", seq_len(n)),
        AreaID   = a$id,
        Year     = yr,
        Stratum  = sample(1:3, n, replace = TRUE, prob = c(0.5, 0.3, 0.2)),
        Lon      = stats::runif(n, bb["xmin"], bb["xmax"]),
        Lat      = stats::runif(n, bb["ymin"], bb["ymax"])
      )
    }))
  }))

  ## Cross each sample with every demo FG (one density observation per
  ## sample/FG, some genuinely zero - a real haul rarely catches every
  ## FG), with a mild upward trend over the demo years for FG 1 only,
  ## so plot_fg_timeseries_by_area()/plot_fg_timeseries_regional() have
  ## something visibly non-flat to draw.
  demo_dt <- samples[rep(seq_len(.N), each = nrow(demo_fgs))]
  demo_dt[, FG_num := rep(demo_fgs$FG_num, times = nrow(samples))]
  demo_dt <- merge(demo_dt, demo_fgs, by = "FG_num")
  demo_dt[, Density := round(stats::rexp(.N, rate = 1 / 4) * (stats::runif(.N) > 0.25), 2)]
  demo_dt[FG_num == 1, Density := round(Density * (1 + 0.06 * (Year - min(demo_years))), 2)]

  ## Hand-built strata areas (km2) - fabricated but internally
  ## consistent (prop sums to 1 within each AreaID), exactly the shape
  ## weight_by_strata()/weight_by_area() expect from a real
  ## compute_strata_area_by_area() call.
  demo_strata_area_by_area <- data.table(
    AreaID    = rep(c(1, 2), each = 3),
    Stratum   = rep(1:3, times = 2),
    area_km2  = c(900, 500, 260, 700, 480, 220)
  )
  demo_strata_area_by_area[, prop := area_km2 / sum(area_km2), by = AreaID]

  list(
    dt                         = demo_dt,
    area_shp                   = demo_area_shp,
    area_id_col                = "AreaID",
    strata_def                 = demo_strata_def,
    species_taxonomy           = NULL,
    cached_at                  = Sys.time(),
    filter_areas_at_cache_time = c(1, 2),
    year_range_at_cache_time   = range(demo_years),
    strata_area_by_area_demo   = demo_strata_area_by_area
  )
}

IS_DEMO_DATA <- FALSE

if (CACHE_AVAILABLE) {
  cache <- readRDS(CACHE_PATH)

  ## Bathymetry-derived strata areas are the slow part (one bathymetry
  ## download per GSA) and depend ONLY on which areas exist in the cache,
  ## never on the year range or GSA selection the user picks interactively -
  ## computed once here at startup, not inside a reactive(), so changing the
  ## year slider or the GSA checkboxes never re-triggers a bathymetry
  ## download.
  ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
  STRATA_AREA_BY_AREA_FULL <- compute_strata_area_by_area(
    area_ids   = ALL_AREA_IDS,
    area_shp   = cache$area_shp,
    area_id_col = cache$area_id_col,
    strata_def = cache$strata_def
  )

  YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
  YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
} else {
  ## Final fallback: no real cache exists (yet). If AUTO_BUILD_CACHE just
  ## kicked off a background build above, it's still running at this exact
  ## instant (it takes minutes; this line runs in milliseconds) - the
  ## cache-state chunk further down polls for it and swaps rv$cache over
  ## automatically once it's ready, no reload needed. Rather than showing
  ## an empty "no data yet" state in the meantime, fall back to a small
  ## synthetic demo dataset (built purely in-memory, in a few milliseconds
  ## - no downloads, no pCloud access) so every control and figure below
  ## has SOMETHING real to react to immediately. CACHE_AVAILABLE is set
  ## TRUE here on purpose: every validate(need(CACHE_AVAILABLE, ...)) guard
  ## downstream means "is there ANY data to show" (real or demo) -
  ## IS_DEMO_DATA is the separate flag the UI uses to say which one it is.
  message("[pipeline_documentation_shiny] ", CACHE_MISSING_MSG,
          " Falling back to a small synthetic demo dataset so this app shows",
          " something immediately - see the banner in the app itself once it",
          " loads. It'll switch over automatically once a real cache is ready",
          " (from the background build, if AUTO_BUILD_CACHE is on, or from",
          " building one by hand - see \"Before you start\" above).")
  demo_cache <- tryCatch(build_demo_cache(), error = function(e) {
    message("[pipeline_documentation_shiny] Building the synthetic demo dataset also failed (",
            conditionMessage(e), ") - falling back to the plain \"no data yet\" message instead.")
    NULL
  })

  if (!is.null(demo_cache)) {
    cache <- demo_cache
    IS_DEMO_DATA <- TRUE
    CACHE_AVAILABLE <- TRUE

    ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
    STRATA_AREA_BY_AREA_FULL <- cache$strata_area_by_area_demo
    YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
    YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
  } else {
    cache <- NULL
    ALL_AREA_IDS <- integer(0)
    STRATA_AREA_BY_AREA_FULL <- NULL
    YEAR_MIN <- 1990L
    YEAR_MAX <- 2023L
  }
}
```

```{r cache-state, context="server"}
## Per-session reactive mirror of the plain globals set in the setup
## chunk above (cache/CACHE_AVAILABLE/ALL_AREA_IDS/STRATA_AREA_BY_AREA_FULL/
## YEAR_MIN/YEAR_MAX/IS_DEMO_DATA). This exists specifically so the UI
## (rendered via renderUI() below, not a static chunk) and every reactive/
## output elsewhere can actually SEE these values change later - a plain
## global can drive what the page shows on first load, but writing to it
## again after that load does nothing to an already-rendered page; only a
## reactiveValues() write triggers Shiny to re-render whatever depends on
## it, which is what makes the automatic real-data swap below possible.
rv <- reactiveValues(
  cache                    = cache,
  CACHE_AVAILABLE          = CACHE_AVAILABLE,
  ALL_AREA_IDS             = ALL_AREA_IDS,
  STRATA_AREA_BY_AREA_FULL = STRATA_AREA_BY_AREA_FULL,
  YEAR_MIN                 = YEAR_MIN,
  YEAR_MAX                 = YEAR_MAX,
  IS_DEMO_DATA             = IS_DEMO_DATA
)

## Polls every 4 seconds for the background build (see setup chunk above,
## AUTO_BUILD_CACHE) finishing - but ONLY while still on demo data
## (req(isTRUE(rv$IS_DEMO_DATA)) - once swapped over, or if this was never
## on demo data to begin with, this observer's own check is a no-op file.
## exists() call, not the full pipeline. The FIRST poll (in this or any
## other open session/tab) to see the file appear does the one genuinely
## slow step here - compute_strata_area_by_area(), one bathymetry
## download per GSA - and swaps THIS session's rv over; if several tabs
## are open at once, each does this independently the next time its own
## poll fires, which costs a few redundant bathymetry downloads but keeps
## this simple and avoids a shared-state race between sessions.
observe({
  invalidateLater(4000)
  req(isTRUE(rv$IS_DEMO_DATA))
  if (!file.exists(CACHE_PATH)) return(NULL)

  ## readRDS() can fail if the file exists but saveRDS() hasn't finished
  ## writing it yet (a brief window) - tryCatch here just means "try
  ## again next tick" rather than crashing this session's reactive graph.
  real_cache <- tryCatch(readRDS(CACHE_PATH), error = function(e) NULL)
  if (is.null(real_cache)) return(NULL)

  message("[pipeline_documentation_shiny] Background build finished - switching this session over",
          " to the real cache now (no reload needed).")
  real_area_ids <- sort(unique(real_cache$dt$AreaID))
  rv$cache                    <- real_cache
  rv$ALL_AREA_IDS             <- real_area_ids
  rv$STRATA_AREA_BY_AREA_FULL <- compute_strata_area_by_area(
    area_ids    = real_area_ids,
    area_shp    = real_cache$area_shp,
    area_id_col = real_cache$area_id_col,
    strata_def  = real_cache$strata_def
  )
  rv$YEAR_MIN     <- min(real_cache$dt$Year, na.rm = TRUE)
  rv$YEAR_MAX     <- max(real_cache$dt$Year, na.rm = TRUE)
  rv$IS_DEMO_DATA <- FALSE
})
```

:::: {.callout-warning}
## Before you start: this needs a live R session, and a one-time data cache

Unlike `pipeline_documentation.qmd`, this file is **not** a static document
you can just open in a browser. `server: shiny` in the YAML above means it
needs an actual R process running behind it - open it with `quarto preview
pipeline_documentation_shiny.qmd` (or `quarto serve`) from a machine with R,
Shiny, and this pipeline's packages installed, or deploy it to
shinyapps.io / Posit Connect / your own Shiny Server. Double-clicking a
rendered `.html` file will not work - there's no R behind a plain HTML file.

It also needs a **one-time data cache** before it can show your real
MEDITS results - but you should never see a blank page waiting on that,
and by default you don't have to lift a finger to get the real thing
either. **The moment you open this app with no real cache yet, two things
happen at once:** it immediately shows a small synthetic demo dataset -
fabricated GSAs, strata, functional groups and samples, built in-memory
in milliseconds, no network or pCloud access needed - so every control
and figure below is already populated and interactive on first load; and,
in the background, it launches `01_biomass.R` as a **separate** process
(not inside this app itself) to build the real cache. A grey banner above
the controls says plainly when you're looking at demo data - and that
banner, along with every control and figure, **switches over to your real
results automatically, with no page reload**, the moment that background
build finishes (a few minutes: GSA shapefile + bathymetry downloads,
MEDITS loading, FG matching). Progress and any errors from that
background run land in `shiny_cache/auto_build.log`, next to this `.qmd`
file.

Running that build as a genuinely separate process (rather than inside
this document's own startup code) is what makes it safe to have on by
default: an earlier version of this ran it in-process, which blocked the
page from rendering anything at all until it finished - looking
identical to the app being broken if pCloud wasn't reachable from
wherever `quarto serve`/`quarto preview` was running, or the run was
simply slow. If you'd rather skip the automatic background build entirely
(e.g. no pCloud access at all from this machine), set `AUTO_BUILD_CACHE
<- FALSE` in this document's setup chunk - the app then just stays on
demo data until you build a real cache by hand, using the manual steps
below.

The manual steps to build a real cache yourself:

1. In `01_biomass.R`'s Configuration section, temporarily set
   `FILTER_AREAS` to every GSA you have MEDITS data for (not just your usual
   region) and `YEAR_ECOPATH`/`TS_YEARS` to your full available year range -
   the widest extent you want this app to be able to explore later.
2. Set `SAVE_SHINY_CACHE_PATH <- "shiny_cache/survey_density_raw.rds"`
   (relative to wherever `01_biomass.R` writes `out_dir`) and run the script
   once, in full.
3. Copy (or symlink) the resulting `shiny_cache/survey_density_raw.rds` to
   sit next to this `.qmd` file - `CACHE_PATH` in the setup chunk above
   expects it there.
4. Revert `FILTER_AREAS`/`YEAR_ECOPATH`/`TS_YEARS`/`SAVE_SHINY_CACHE_PATH`
   back to their normal values for `01_biomass.R`'s regular use - this cache
   is a frozen snapshot for THIS document only, not something the numbered
   pipeline scripts read back.

The cache is a snapshot either way: it does not update itself if the
underlying MEDITS data changes. Repeat steps 1-3 by hand, or (with
`AUTO_BUILD_CACHE <- TRUE`, the default) just delete
`shiny_cache/survey_density_raw.rds` and reload the page to trigger
another automatic background build, whenever you want it reflecting
newer data.
::::

## What's actually live here, and what isn't

Changing the **year range** or the **GSAs included** below genuinely
re-runs the pipeline's own strata-weighting cascade
(`compute_fg_densities_by_stratum()` → `weight_by_strata()` →
`weight_by_area()`) against the cached per-sample data, and the four figures
re-render from that real result - not a lookup of a pre-rendered image.

Three of the static form's other options don't carry over as live toggles,
for reasons worth being upfront about rather than silently shipping a
control that doesn't really do anything:

* **`DROP_OUTLIERS`** - outlier removal already ran once, at cache-build
  time (`01_biomass.R`'s own Step, before this document's cache is saved).
  Toggling it here would need the raw pre-outlier-removal samples cached
  too, which this cache deliberately doesn't keep (it's already large
  enough as the full per-sample extent).
* **`STRATA`** - `weight_by_area()` (needed for the region-wide figure)
  only implements the strata-weighted path; the unstratified
  (`simple_area_density()`) path has no region-wide aggregation wired up in
  the shared library today. This app always uses `STRATA = TRUE`
  to keep the region-wide figure working; the by-area figure would be the
  only one affected by `STRATA = FALSE` in any case.
* **Custom bounding box** - re-deriving a custom polygon's strata *areas*
  live would mean a fresh bathymetry download per input change, which is
  far too slow for an interactive control. GSA selection is live because
  `STRATA_AREA_BY_AREA_FULL` above is precomputed once for every GSA in the
  cache; a live bbox would need that recomputed on the fly instead.

The **"Generate config block"** panel at the bottom still works exactly like
the static version - it always reflects your current picks above, ready to
paste into `01_biomass.R`, `DROP_OUTLIERS`/`STRATA`/bbox included, even
though only year range and GSA selection also drive the figures live.

```{r ui-inputs}
## Bare uiOutput() placeholder - NOT the controls/banner content itself.
## This chunk has no context="server" on purpose (it only needs to exist
## as one static HTML tag), and it's entirely re-populated by the
## controls-render chunk below (context="server", driven by rv$...) - a
## plain static chunk here couldn't react to rv$IS_DEMO_DATA flipping
## FALSE after the page has already loaded, which is the whole point of
## the automatic real-data swap.
uiOutput("controls_and_banner")
```

```{r controls-render, context="server"}
## Rebuilds the whole controls panel (and the demo banner) whenever
## rv$CACHE_AVAILABLE/rv$IS_DEMO_DATA/rv$YEAR_MIN/rv$YEAR_MAX/rv$
## ALL_AREA_IDS change - in particular, right after cache-state's observe()
## swaps from demo to real data, with no page reload. sliderInput/
## checkboxGroupInput/leafletOutput keep the SAME ids ("year_range",
## "filter_areas", "gsa_map") across a rebuild, which is what lets every
## reactive/output elsewhere (reactive-pipeline, plot-render, map-render,
## config-download) keep referencing input$year_range etc. unchanged - a
## rebuild with new min/max/choices just looks like a normal input-value
## change to everything downstream.
##
## DROP_OUTLIERS/NORMALIZE_TS/STRATA/the two Ecopath-base-year numbers
## aren't tied to rv (they're config-block-only - see the note above this
## section), so a swap resets them to their hardcoded defaults rather than
## preserving whatever the user had picked against the demo data; a minor,
## acceptable side effect of the whole panel being rebuilt around a
## completely different dataset.
output$controls_and_banner <- renderUI({
  if (!isTRUE(rv$CACHE_AVAILABLE)) {
    return(div(
      style = paste(
        "padding: 1em 1.25em; margin: 0.5em 0 1.5em; border: 1px solid #f0ad4e;",
        "border-radius: 6px; background: #fcf8e3; color: #664d03;"
      ),
      strong("This app has no data to show yet."), br(),
      CACHE_MISSING_MSG
    ))
  }

  tagList(
    if (isTRUE(rv$IS_DEMO_DATA)) {
      div(
        style = paste(
          "padding: 0.75em 1.25em; margin: 0.5em 0 1em; border: 1px solid #6c757d;",
          "border-radius: 6px; background: #f1f1f1; color: #333;"
        ),
        strong("You're viewing synthetic demo data, not real MEDITS data."), br(),
        "A real cache hasn't finished building yet - so every control and figure below is ",
        "running against a small, made-up dataset (fabricated GSAs, strata, functional groups ",
        "and samples) just so you can see how this app works while you wait. See \"Before you ",
        "start\" above for how that build works. This banner disappears automatically, with no ",
        "reload needed, the moment your real MEDITS-derived cache is ready."
      )
    },
    fluidRow(
      column(
        width = 4,
        sliderInput("year_range", "Ecosim time-series years (TS_YEARS)",
                    min = rv$YEAR_MIN, max = rv$YEAR_MAX,
                    value = c(rv$YEAR_MIN, rv$YEAR_MAX), step = 1, sep = ""),
        checkboxGroupInput("filter_areas", "GSA numbers to include (FILTER_AREAS)",
                            choices = rv$ALL_AREA_IDS, selected = rv$ALL_AREA_IDS, inline = TRUE),
        hr(),
        checkboxInput("drop_outliers", "DROP_OUTLIERS (config block only - see note above)", value = TRUE),
        checkboxInput("normalize_ts", "NORMALIZE_TS (config block only)", value = TRUE),
        checkboxInput("strata_config", "STRATA (config block only - figures always use TRUE)", value = TRUE),
        hr(),
        numericInput("year_ecopath_start", "Ecopath base year, start", value = 1994),
        numericInput("year_ecopath_end", "Ecopath base year, end", value = 1996),
        hr(),
        downloadButton("download_config", "Download config block (.R)")
      ),
      column(
        width = 8,
        leafletOutput("gsa_map", height = 320)
      )
    )
  )
})
```

```{r reactive-pipeline, context="server"}
## context="server" is required here - confirmed the hard way: without it,
## this chunk's reactives (selected_areas, filtered_dt, fg_index, ...) get
## created in Quarto's static-UI-build pass, a DIFFERENT environment from
## the one context="server" chunks (map-render, config-download, and now
## plot-render below) actually run in per session. The visible symptom was
## "Error in selected_areas: could not find function 'selected_areas'"
## from inside output$gsa_map's renderLeaflet() and its observe() block -
## the map/plot code could see cache/CACHE_AVAILABLE (those are plain
## objects from the setup chunk, same environment either way) but not
## these reactive()-wrapped functions, because they simply didn't exist in
## the environment those context="server" chunks execute in.
##
## req(rv$CACHE_AVAILABLE) first in every reactive here - without it, a
## reactive further down the chain (e.g. filtered_dt(), which touches
## rv$cache$dt) would be the first thing to notice the cache is missing,
## with a less clear error than the validate(need(...)) message each
## render*() block below already shows. This just makes the failure
## point explicit and identical everywhere. Reading from rv$... (not the
## plain cache/CACHE_AVAILABLE globals) is what makes these reactives
## automatically pick up the real data after cache-state's swap, with no
## page reload - a plain global wouldn't re-trigger anything downstream.
selected_areas <- reactive({
  req(rv$CACHE_AVAILABLE, input$filter_areas)
  as.numeric(input$filter_areas)
})

filtered_dt <- reactive({
  rv$cache$dt[Year >= input$year_range[1] & Year <= input$year_range[2] &
             AreaID %in% selected_areas()]
})

per_group_fg <- reactive({
  compute_fg_densities_by_stratum(filtered_dt(), strata = TRUE)
})

n_samples_by_stratum <- reactive({
  filtered_dt()[!is.na(Stratum), .(n_samples = uniqueN(SampleID)), by = .(AreaID, Year, Stratum)]
})

strata_area_selected <- reactive({
  rv$STRATA_AREA_BY_AREA_FULL[AreaID %in% selected_areas()]
})

fg_index <- reactive({
  weight_by_strata(per_group_fg(), n_samples_by_stratum(), strata_area_selected())
})

fg_index_regional <- reactive({
  weight_by_area(fg_index())
})
```

## Sample coverage

```{r plot-sample-map-ui}
## Bare plotOutput() (not a renderPlot() call) - a plain UI placeholder tag,
## the same way you'd declare it in ui.R. This chunk has NO context="server"
## on purpose: it only needs to exist as static HTML, not run per-session -
## the actual renderPlot() logic lives in plot-render below, in
## context="server", where the reactives from reactive-pipeline are
## visible. Splitting UI-placeholder from server-logic this way avoids
## relying on Quarto's implicit "a bare renderPlot() becomes an output"
## behavior, which is exactly what silently failed to update here before.
plotOutput("plot_sample_map")
```

## FG density by area

```{r plot-by-area-ui}
plotOutput("plot_by_area")
```

## FG density, region-wide

```{r plot-regional-ui}
plotOutput("plot_regional")
```

## Density by depth stratum

```{r plot-strata-ui}
plotOutput("plot_strata")
```

```{r plot-render, context="server"}
## All four figures' actual rendering logic, together in one
## context="server" chunk so they share the same live environment as
## reactive-pipeline above (selected_areas(), filtered_dt(), fg_index(),
## fg_index_regional() all need to be reachable here) - each output feeds
## the plotOutput() placeholder of the matching id declared just above its
## own section heading.
output$plot_sample_map <- renderPlot({
  validate(need(rv$CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_sample_map(filtered_dt(), rv$cache$area_shp, rv$cache$area_id_col,
                   selected_areas = selected_areas())
})

output$plot_by_area <- renderPlot({
  validate(need(rv$CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_by_area(fg_index())
})

output$plot_regional <- renderPlot({
  validate(need(rv$CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_regional(fg_index_regional())
})

output$plot_strata <- renderPlot({
  validate(need(rv$CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_strata_profile(fg_index(), rv$cache$strata_def)
})
```

```{r map-render, context="server"}
## context="server" is required here, not optional - this is the actual
## fix for "Error: object 'output' not found" at this chunk. Quarto's
## Shiny engine evaluates every chunk WITHOUT context="server" during an
## initial render pass that only builds the static UI, where `output`
## (and `session`) don't exist yet - that pass is what raised the error.
## reactive()/renderPlot() calls elsewhere in this document survive that
## pass because they just build a promise object to return into the UI;
## an explicit `output$x <- ...` assignment needs `output` to actually
## exist, which only happens inside the live server function - i.e. in a
## context="server" chunk. (config-download below needs the same fix,
## for the same reason - it also assigns into `output`.)
##
## Also depends on reactive-pipeline above ALSO being context="server" -
## without that, selected_areas() below would fail with "could not find
## function 'selected_areas'" (confirmed live) even with this chunk's own
## context="server" fix in place, since a reactive created in a non-server
## chunk isn't visible from a context="server" chunk's environment.
##
## Reads rv$cache/rv$CACHE_AVAILABLE (not the plain globals) so this map
## redraws with the real GSA polygons automatically once cache-state's
## swap happens - both blocks below already depend on rv$cache via the
## body they evaluate, so no extra req()/observe() wiring is needed for
## that; leafletOutput("gsa_map") also gets freshly recreated by
## controls-render's renderUI() on that same swap, which is what makes
## renderLeaflet() below actually re-run for the new container.
##
## No leafletOutput("gsa_map") even exists in the UI when !CACHE_AVAILABLE
## (see controls-render above), so neither block below ever actually runs
## in that case - guarded anyway so this chunk fails the same explicit way
## as everything else if that ever changes, rather than silently.
output$gsa_map <- renderLeaflet({
  validate(need(rv$CACHE_AVAILABLE, CACHE_MISSING_MSG))
  shp <- rv$cache$area_shp
  leaflet(shp) |>
    addTiles() |>
    addPolygons(
      fillColor = ~ifelse(get(rv$cache$area_id_col) %in% isolate(selected_areas()), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(rv$cache$area_id_col))
    )
})

observe({
  req(rv$CACHE_AVAILABLE)
  shp <- rv$cache$area_shp
  leafletProxy("gsa_map", data = shp) |>
    clearShapes() |>
    addPolygons(
      fillColor = ~ifelse(get(rv$cache$area_id_col) %in% selected_areas(), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(rv$cache$area_id_col))
    )
})
```

```{r config-download, context="server"}
## context="server" required here too - this chunk assigns into
## output$download_config, same reason as map-render above.
##
## downloadButton("download_config") only exists in the UI when
## rv$CACHE_AVAILABLE (see controls-render above), so this is unreachable
## otherwise - req() guard kept for the same reason as map-render.
config_text <- reactive({
  req(rv$CACHE_AVAILABLE)
  paste(
    '## --- generated by "Live Explorer" - 01_biomass.R ---',
    paste0("FILTER_AREAS  <- c(", paste(selected_areas(), collapse = ", "), ")"),
    paste0("STRATA        <- ", if (input$strata_config) "TRUE" else "FALSE"),
    paste0("YEAR_ECOPATH  <- ", input$year_ecopath_start, ":", input$year_ecopath_end),
    paste0("TS_YEARS      <- ", input$year_range[1], ":", input$year_range[2]),
    paste0("DROP_OUTLIERS <- ", if (input$drop_outliers) "TRUE" else "FALSE"),
    paste0("NORMALIZE_TS  <- ", if (input$normalize_ts) "TRUE" else "FALSE"),
    sep = "\n"
  )
})

output$download_config <- downloadHandler(
  filename = function() "biomass_config_block.R",
  content = function(file) writeLines(config_text(), file)
)
```