---
title: "Survey-to-Ecopath/Ecosim Pipeline"
subtitle: "Code documentation: scripts, functions, inputs and outputs — for any GFCM GSA subset or user-defined region in the Mediterranean"
author: "Documentation generated with Claude"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, tidy = "styler")
# figure chunks below override eval/echo individually so this document can
# show real pipeline output when it's available; everything else in the
# document stays non-executed (this is documentation, not a live report).

# safe_include_graphics(): shows the real figure if it's been copied into
# figures/ next to this .Rmd (see "Populating the real figures" below);
# otherwise prints a plain placeholder message instead of erroring, so this
# document always knits cleanly whether or not the pipeline has been run
# yet against the real WestMed EwE pCloud drive data.
safe_include_graphics <- function(path, script, out_dir_note) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(
      "*Figure not yet available.* Run **", script, "**",
      " against the real WestMed EwE pCloud drive data, then copy\n`",
      out_dir_note, "`\ninto a `figures/` folder next to this `.Rmd` ",
      "(as `", basename(path), "`) and re-knit.",
      sep = ""
    )
  }
}
```

```{css, echo=FALSE}
body {
  max-width: 900px;
}
/* keep the whole page from ever scrolling sideways */
html, body {
  overflow-x: hidden;
}
/* wide tables: scroll horizontally inside their own box instead of
   pushing the page wider than the viewport */
.table-responsive,
table {
  display: block;
  max-width: 100%;
  overflow-x: auto;
}
table {
  width: auto !important;
}
table td, table th {
  white-space: normal;
  word-break: break-word;
}
/* code blocks (the ASCII pipeline diagram, chunk output, etc.):
   wrap long lines instead of overflowing, but still allow a
   horizontal scrollbar as a fallback for anything that truly can't wrap */
pre, code {
  white-space: pre-wrap;
  word-break: break-word;
  overflow-x: auto;
  max-width: 100%;
}
/* mermaid diagrams (see medbs_pipeline_methodology.qmd for the same look):
   don't let the pre-wrap rule above touch them - mermaid.js replaces the
   div's contents with an SVG at render time */
div.mermaid {
  overflow-x: auto;
  max-width: 100%;
  text-align: center;
}
```

<script src="https://cdn.jsdelivr.net/npm/mermaid@10.9.8/dist/mermaid.min.js"></script>
<script>
  mermaid.initialize({ startOnLoad: true, theme: "neutral",
                        flowchart: { padding: 20, nodeSpacing: 60, rankSpacing: 90 } });
</script>

# Purpose of this document

This document describes a general-purpose R pipeline for turning MEDITS/
MEDIAS trawl-survey and catch data into input sheets for an Ecopath-with-
Ecosim (EwE) model, for **any region in the Mediterranean Sea** — a single
GFCM Geographical Sub-Area (GSA), any subset or the full set of GSAs, or a
completely user-defined region (a bounding box or an arbitrary shapefile
that doesn't line up with GSA lines at all, e.g. a bay, an MPA, or a
footprint spanning parts of several GSAs). It covers every script provided,
what each one reads and writes, and every function in the shared library
those scripts call.

The scripts as delivered are configured for the **Western Mediterranean**
(GSA 1–11) as a **worked example** — this is what `01_survey_density_westmed.R`,
`export_gsa_boundaries.R` and `fig_WMed_basemap.R` currently point at, and
what a few reference tables (e.g. the MEDIAS Handbook area crosswalk) are
currently populated for. The underlying mechanism — GSA-based selection via
`FILTER_AREAS`, or an arbitrary region via `01_survey_density_custom.R` — is
not Western-Med-specific; see [Adapting this pipeline to a different region]
below for exactly what to change to point the same pipeline at another part
of the Mediterranean.

All scripts read from and write to **one shared workbook**,
`output/ecopath_ecosim_inputs.xlsx`, via two library functions designed for
exactly this purpose:

* `upsert_workbook_sheets()` — adds/replaces named sheets, leaves every other
  sheet in the workbook untouched. Safe to call from any script, in any order,
  any number of times.
* `finalize_workbook_sheet_order()` — run **last**, after every upstream
  script has run at least once. Renames a couple of sheets (`FG_lookup` →
  `FG`, `References` → `PB_QB_References_`) and reorders every sheet into the
  final layout Ecopath/Ecosim expects.

Because of this design, the four numbered pipeline scripts are
**order-independent with respect to the shared workbook** — except where a
script's own calculations literally need a file another script produced
first (documented per-script below).

# Data and code locations

* **Code** — every script and library file documented here lives in the
  **WestMed EwE GitHub repository** (`WMed_EwE`, cloned locally to what the
  scripts call `git_dir`). All shared-library `source()` calls
  (`lib_survey_fg_density_functions.R`, `lib_worms_taxonomy_lookup.R`,
  `lib_build_species_df_from_survey.R`) resolve against this repository's
  `scripts/` folder.
* **Input data** — every raw survey, catch and reference file each script
  reads (`FG_WMed.xlsx`, `TM_list_(April_2019).xlsx`, MEDITS `TA.csv`/
  `TB.csv`, MEDIAS `abundance.csv`/`biomass.csv`, `ASFIS_sp_2026.1.csv`,
  the catchability CSV, GFCM capture/reference CSVs) is read from the
  **WestMed EwE pCloud drive** (`EwE Western Med 2026`, mounted/synced
  locally to what the scripts call `pcloud_dir`), **not** from the GitHub
  repository — the repository holds code only, no raw data.
* **Outputs** — every CSV, plot, and the shared `ecopath_ecosim_inputs.xlsx`
  workbook are written to a local `out_dir` (not synced back to pCloud or
  GitHub by the scripts themselves).

Throughout this document, `<pcloud_dir>` means *the WestMed EwE pCloud
drive* and `<git_dir>` means *the WestMed EwE GitHub repository* — both are
plain local paths the top of each script resolves (hardcoded for one user,
or picked interactively via `rstudioapi::selectDirectory()` for anyone
else), not fixed locations baked into the code. Retargeting the pipeline to
a different region (see [Adapting this pipeline to a different region])
does not require moving off pCloud/GitHub — only pointing `pcloud_dir`/
`git_dir` at that region's own drive/repository, or adding that region's
data alongside the Western Med data in the same drive/repository.

**A note on the plots and column tables in this document:** this
documentation was written without direct access to the WestMed EwE pCloud
drive's actual data files. The **column-reference tables** throughout are
not guesses — they list exactly the column names each script's own code
reads (`select()`/`merge()`/`setnames()` calls, etc.), so they're accurate
regardless of the actual data values. The **figures**, however, are the
pipeline's own real output and only appear once you've generated them — see
[Populating the real figures] below.

## Populating the real figures

Several sections below embed a figure via a small `safe_include_graphics()`
helper (defined in this document's setup chunk). Each one checks for a file
under `figures/` (a folder that must sit next to this `.Rmd`) and shows the
real image if present, or a plain "not yet available" message with the
exact script/output path to copy from if not — so this document always
knits cleanly either way. To populate them for real:

1. Run the relevant script (named in each placeholder message) against the
   real WestMed EwE pCloud drive data.
2. Copy the named `.png` from that script's `<out_dir>/plots/` into a
   `figures/` folder next to this `.Rmd`, keeping the same filename.
3. Re-knit.

Four **other** diagrams in this document (the pipeline flow, the
FG-matching cascade, the workbook sheet assembly, and the PB/QB taxon
dispatch diagrams) are a different kind of figure: **Mermaid flowcharts**,
in the same style as `medbs_pipeline_methodology.qmd`'s own diagrams
(same `flowchart LR` layout, the same gray/purple/teal/coral
`classDef` palette). They show the code's own logic — which function
calls which, which sheet gets written by what, which fallback branch runs
when — built directly from the script and library contents described
here, not from any survey or model output. Because they're plain text
(Mermaid syntax rendered in-browser via `mermaid.js`, loaded from a CDN
in this document's setup), they contain no data values and no image
files to go missing — unlike the four `safe_include_graphics()`
placeholders above, they always render, with or without an internet
connection to the pCloud drive. They do need an internet connection *to
the Mermaid CDN* at view time (not render time) to draw the SVG in the
browser; if you'd rather they render standalone with no external
dependency, `rmarkdown::html_document(self_contained = TRUE)` still
inlines the two `<script>` tags below the CSS chunk, so this only
matters if that CDN is unreachable from wherever the HTML is opened.

# Pipeline overview

The diagram below is a schematic of the actual script/function call
structure documented in this document — which script produces which
intermediate file, and which workbook sheets each one writes. It is not a
plot of any survey data or model output (same diagram style as
`medbs_pipeline_methodology.qmd`'s "Pipeline flow" section).

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["Step 1a/1b&lt;br/&gt;01_survey_density_*.R"] --> B1["species_density_regional_combined.csv&lt;br/&gt;strata_area_by_area.csv"]
    B1 --> C1["xlsx: FG_spp_Ecopath, Ecopath,&lt;br/&gt;Ecosim, FG_spp_Ecosim, FG, traits_ewe"]
    B1 --> A2["Step 2&lt;br/&gt;02_fao_catches.R"]
    A2 --> C2["xlsx: Catches_Ecopath,&lt;br/&gt;Catches_Ecosim"]
    A3["Step 3 (independent)&lt;br/&gt;03_ecobase_query.R"] --> B3["ecobase_literature_pb_qb_simple.csv&lt;br/&gt;(optional)"]
    B1 --> A4["Step 4 (run last)&lt;br/&gt;04_pbqb_calc.R"]
    B3 -.-> A4
    A2 -.-> A4
    A4 --> C4["xlsx: PB_QB, PB_QB_spp, Ecobase, References"]
    C1 --> F["finalize_workbook_sheet_order()&lt;br/&gt;rename + reorder"]
    C2 --> F
    C4 --> F
    F --> O["output/ecopath_ecosim_inputs.xlsx"]

    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class A1,A2,A3,A4 purple
    class B1,B3 gray
    class C1,C2,C4 teal
    class F,O coral
</div>
```

**Step 1 — pick ONE:**

* `01_survey_density_westmed.R`
  (MEDITS + MEDIAS, GFCM GSA areas), **or**
* `01_survey_density_custom.R`
  (MEDITS only, custom bbox/shapefile)

&#8595; produces `species_density_regional_combined.csv`,
`strata_area_by_area.csv`, and sheets `FG_spp_Ecopath`, `Ecopath`, `Ecosim`,
`FG_spp_Ecosim`, `FG` (renamed from `FG_lookup`), `traits_ewe`.

**Step 2:**

* `02_fao_catches.R`
  (reads Step 1's `strata_area_by_area.csv`)

&#8595; produces sheets `Catches_Ecopath`, `Catches_Ecosim`.

**Step 3 — independent, can run anytime before Step 4:**

* `03_ecobase_query.R`

&#8595; produces `ecobase_literature_pb_qb_simple.csv` (optional Step 4 input).

**Step 4 — run last:**

* `04_pbqb_calc.R`
  (needs Step 1's `species_density_regional_combined.csv`; optionally reads
  Step 2's `fg_catch_timeseries_*.csv` and Step 3's
  `ecobase_literature_pb_qb_simple.csv`)

&#8595; produces sheets `PB_QB`, `PB_QB_spp`, `Ecobase`, `References` (renamed
to `PB_QB_References_`).

**Finally**, `finalize_workbook_sheet_order()` — called from whichever
script runs last — reorders every sheet into the final layout:

&#8595; `output/ecopath_ecosim_inputs.xlsx`, final, ordered, ready for
Ecopath import.

Two more files are **not** part of the numbered pipeline, but support it:

* `lib_survey_fg_density_functions.R` and `lib_worms_taxonomy_lookup.R` and
  `lib_build_species_df_from_survey.R` — shared library files, `source()`'d
  by the numbered scripts, no top-level driver code of their own.
* `export_gsa_boundaries.R` and `fig_WMed_basemap.R` — standalone cartography
  utilities, not required by the numbered pipeline. `export_gsa_boundaries.R`
  in particular produces the GeoJSON embedded in `medbs_pipeline_methodology.qmd`'s
  interactive map.

# Run order and dependencies

| Step | Script | Depends on | Must run before |
|---|---|---|---|
| 1 | `01_survey_density_westmed.R` **or** `01_survey_density_custom.R` (pick one) | nothing | 2, 4 |
| 2 | `02_fao_catches.R` | Step 1's `strata_area_by_area.csv` | — |
| 3 | `03_ecobase_query.R` | nothing (independent web query) | 4 (optional input) |
| 4 | `04_pbqb_calc.R` | Step 1's `species_density_regional_combined.csv`; optionally Step 2's `fg_catch_timeseries_*.csv` and Step 3's `ecobase_literature_pb_qb_simple.csv` | — |

`finalize_workbook_sheet_order()` should be called at the end of whichever
script ran *last* in a given session (each of the numbered scripts already
does this itself at its own final step).

---

# Script 1a — `01_survey_density_westmed.R`

**Role:** Pipeline Step 1 of 4, GFCM-GSA-based version — the general
mechanism for **any GSA subset anywhere in the Mediterranean**, not just the
Western Med. Run this **or** `01_survey_density_custom.R`, not both, as Step
1. No dependency on the other numbered scripts; `02_fao_catches.R` and
`04_pbqb_calc.R` both require this step's output to exist first.

Runs two independent surveys over the same GSAs/FG scheme (MEDITS bottom
trawl + MEDIAS acoustic) and combines both into one set of Ecopath/Ecosim
sheets, rather than producing separate MEDIAS sheets. As delivered it's
configured for the Western Med (GSA 1–11, via `FILTER_AREAS <- 1:11`) as a
worked example; retargeting it to a different GSA subset elsewhere in the
Mediterranean — the Adriatic, the Aegean, the whole basin — is a matter of
changing that one line and the region-specific reference tables listed in
[Adapting this pipeline to a different region].

## Inputs

| File | Format | Role |
|---|---|---|
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → Functional Group reference (`dataframe2`: `ESPECIE`→`ScientificName`, `GF`→`FG_num`, `FG_name`) |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet `traits_ewe` | xlsx | Per-species life-history/ecology traits |
| `<pcloud_dir>/.../TM_list_(April_2019).xlsx` | xlsx | MEDITS species-code → scientific-name lookup |
| `<pcloud_dir>/.../Demersal/TA.csv` | csv | MEDITS haul/sample metadata (position, depth, gear, swept-area inputs) |
| `<pcloud_dir>/.../Demersal/TB.csv` | csv | MEDITS catch-per-haul data |
| `<pcloud_dir>/.../Acoustic/abundance.csv`, `biomass.csv` | csv | MEDIAS acoustic-survey raw records |
| `<pcloud_dir>/data/ASFIS_sp_2026.1.csv` | csv | FAO ASFIS species code list (MEDIAS species-code → scientific name) |
| `<pcloud_dir>/data/catchability_factors_ecotrans_medits_2021_spp.csv` | csv (optional) | Per-species catchability correction factors (q); skipped with a message if absent |
| GFCM GSA shapefile | auto-downloaded zip from `gfcmsitestorage.blob.core.windows.net/website/5.Data/ArcGIS/GFCM_GSA.zip` | Study-area polygons (`area_shp`); `SMU_CODE` 111/112 (W/E Sardinia) merged into GSA 11 |
| NOAA bathymetry | auto-downloaded via `marmap::getNOAA.bathy()` inside `compute_strata_area_by_area()` | Depth-stratum area calculation |

`FILTER_AREAS <- 1:11` (this example's Western Med GSAs — set to any GSA
number or set of GSA numbers, anywhere in the GFCM area, to retarget the
whole script to a different region), `YEAR_ECOPATH <- 1994:1996`,
`TS_YEARS <- 1995:2023`, `STRATA <- TRUE`, `DROP_OUTLIERS <- FALSE`,
`NORMALIZE_TS <- TRUE`.

### Column reference — key input files

Columns as actually referenced by the code (from the `<pcloud_dir>` files
above). Not every column in the real file is necessarily listed — only the
ones the script reads.

**`FG_WMed.xlsx`, sheet 4** (→ `dataframe2`, the species→FG master):

| Column | Renamed to | Meaning |
|---|---|---|
| `ESPECIE` | `ScientificName` | Species scientific name |
| `GF` | `FG_num` | Functional Group number |
| `FG_name` | `FG_name` | Functional Group name |

**`FG_WMed.xlsx`, sheet `traits_ewe`** (alternating "N: Group Name" header
rows / species data rows — see Script 1a Step 10 below):

| Column (original header) | Renamed to |
|---|---|
| *(unnamed index column)* | `row_index` |
| `Species` | `Species_col` |
| `Organism` | `Organism` |
| `Ecology` | `Ecology` |
| `Occurrence status` | `Occurrence_status` |
| `Biomass contribution` | `Biomass_contribution` |
| `Catch contribution` | `Catch_contribution` |
| `IUCN conservation status` | `IUCN_conservation_status` |
| `Exploitation status` | `Exploitation_status` |
| `Vulnerability index (?)` | `Vulnerability_index` |
| `Mean length (?)` | `Mean_length` |
| `Max length (?)` | `Max_length` |
| `Mean weight (?)` | `Mean_weight` |
| `Mean life span (year)` | `Mean_lifespan_years` |

**`TM_list_(April_2019).xlsx`, sheet 1 (skip first 2 rows)** — the code
locates its two needed columns by regex rather than a fixed name, since the
exact header text has drifted between file versions: a column matching
`/MEDITS/i` (the MEDITS species code) and a column matching `/Scientific/i`
(the scientific name).

**MEDITS `TA.csv`** (haul/sample metadata):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID` (concatenated, `_`-separated) |
| `distance` | Trawl distance towed → swept area |
| `wing_opening` | Trawl wing opening (decimetres) → swept area |
| `shooting_depth`, `hauling_depth` | Averaged into `mean_depth` |
| `shooting_latitude`, `shooting_longitude` | DDMM.mmm format — converted to decimal degrees |
| `shooting_quadrant` | Present but **not used** — flagged in the code as an unresolved caveat around hemisphere sign |

**MEDITS `TB.csv`** (catch-per-haul data):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID`, joined against `TA.csv` |
| `genus`, `species` | Concatenated into `species_code`, joined against `TM_list` |
| `ptot` | Total catch weight (grams; negative values treated as `NA`, converted to tonnes) |

**MEDIAS `abundance.csv` / `biomass.csv`** (acoustic survey raw records):

| Column | Used for |
|---|---|
| `country`, `year`, `species` | Grouping key for aggregation |
| `area` | Text field containing a GSA number, extracted via regex (`\d+`) |
| `sex` | Filtered to `"C"` (combined sexes) when present |
| `lengthclass*` (many columns) | Summed row-wise into a single `total_value` (negative values treated as missing) |

**`ASFIS_sp_2026.1.csv`** (FAO species reference list):

| Column | Used for |
|---|---|
| `Alpha3_Code` | MEDIAS species code → joined against `abundance`/`biomass` |
| `Scientific_Name` | Resolved scientific name |
| *(also present: `ISSCAAP_Group`, `Taxonomic_Code`, `English_name`, …)* | Not read by this pipeline |

**Catchability CSV** (`catchability_factors_ecotrans_medits_2021_spp.csv`,
optional):

| Column | Renamed to | Meaning |
|---|---|---|
| `species` | `ScientificName` | Species scientific name |
| `q_FACTOR` | `q` | Catchability correction factor |

```{r sample-coverage-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_sample_map() output — MEDITS sample coverage by area."}
safe_include_graphics(
  "figures/survey_sample_coverage_map.png",
  script = "01_survey_density_westmed.R (or 01_survey_density_custom.R)",
  out_dir_note = "<out_dir>/plots/survey_sample_coverage_map.png"
)
```

## Processing steps

1. **Config** — resolve `out_dir`/`pcloud_dir`/`git_dir`, download/cache the GSA
   shapefile, source `lib_survey_fg_density_functions.R` and
   `lib_worms_taxonomy_lookup.R`.
2. **Load MEDITS data** — read `TA.csv`/`TB.csv`, compute swept area
   (`distance × wing_opening_m / 1e6`, wing opening in decimetres), convert
   DDMM.mmm lat/lon to decimal degrees, fix the Alboran Sea (GSA 1–3)
   longitude sign — a Western-Med-specific correction, harmless (a no-op) if
   your `FILTER_AREAS` doesn't include GSA 1–3 — reshape into the
   standardized `dataframe1`/`dataframe2` format. Runs `validate_survey_data()`
   and `plot_sample_map()` (→ `survey_sample_coverage_map.png`,
   `survey_sample_coverage_map_by_year.png`).
3. **FG matching (Steps 3–4)** — `prepare_fg_lookup()` →
   `match_species_to_fg()` → `match_nominate_subspecies_fg()` → manual
   overrides (species codes absent from the MEDITS taxonomic list, e.g.
   `ARGRACU`→*Argyropelecus aculeatus*) → `fallback_match_fg_by_taxonomy()`
   (Genus→Family→Order→Class→Phylum) → `apply_seed_fg_rules()` (FGs with zero
   reference species, e.g. Holothuroidea→"Sea cucumbers") →
   `summarize_unresolved_species()` → `survey_unmatched_for_manual_review.csv`.
4. **Densities (Step 5)** — `match_samples_to_area()`,
   `assign_depth_stratum()`, `compute_sample_densities()`, restrict to
   `FILTER_AREAS`, `apply_catchability_correction()` (if the catchability CSV
   exists), `remove_sample_outliers()` (→ `survey_outliers_flagged.csv` if
   any flagged), `compute_fg_densities_by_stratum()`.
5. **Strata/area weighting (Steps 6–7)** — `compute_strata_area_by_area()`
   (cached to `strata_area_by_area.csv`), `weight_by_strata()`,
   `weight_by_area()`, `compute_species_densities_by_stratum()`,
   `weight_species_by_area()`.
6. **MEDITS plots (Step 8)** — `plot_fg_timeseries_by_area()`,
   `plot_fg_timeseries_regional()`, `plot_strata_profile()` →
   `survey_fg_density_timeseries.png`,
   `survey_fg_density_timeseries_regional.png`,
   `survey_fg_depth_strata_profile.png`.

   ```{r fg-timeseries-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_fg_timeseries_by_area() output — MEDITS trawl survey density by FG and area (strata-weighted)."}
   safe_include_graphics(
     "figures/survey_fg_density_timeseries.png",
     script = "01_survey_density_westmed.R (or 01_survey_density_custom.R)",
     out_dir_note = "<out_dir>/plots/survey_fg_density_timeseries.png"
   )
   ```

   ```{r strata-profile-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_strata_profile() output — raw per-sample density across depth strata, by FG."}
   safe_include_graphics(
     "figures/survey_fg_depth_strata_profile.png",
     script = "01_survey_density_westmed.R (or 01_survey_density_custom.R)",
     out_dir_note = "<out_dir>/plots/survey_fg_depth_strata_profile.png"
   )
   ```
7. **MEDIAS acoustic survey (Step 9, sub-steps 9a–9g)** — load and collapse
   raw abundance/biomass records, map species codes via ASFIS, match to FG
   the same way as MEDITS, attach area (MEDIAS Handbook Table 1, hand-keyed
   `MEDIAS_HANDBOOK_AREAS` — as delivered this table only has Western Med
   entries, since that's this example's scope; extending to another region
   means adding that region's own Handbook rows, with a bathymetry-derived
   10–200 m fallback via `compute_strata_area_by_area()` covering any
   country/GSA combination the table doesn't have a row for), compute
   region-wide FG and species density, then
   **combine MEDITS + MEDIAS** into one FG-level and one species-level table
   (averaged on Year/FG or Year/species overlap, overlaps flagged/printed).
   Outputs: `medias_fg_annual_density_by_area.csv`,
   `medias_fg_annual_density_regional.csv`,
   `medias_fg_biomass_density_timeseries_regional.png`,
   `medias_fg_abundance_density_timeseries_regional.png`,
   `species_density_regional_combined.csv` (**contract file** — read by
   `lib_build_species_df_from_survey.R` for `04_pbqb_calc.R`).
8. **Excel export (Step 10)** — `survey_fg_annual_index.csv`,
   `survey_fg_annual_index_regional_combined.csv`, then
   `export_ecopath_ecosim_excel()` writes sheets `FG_spp_Ecopath`, `Ecopath`,
   `Ecosim`, `FG_spp_Ecosim` into `ecopath_ecosim_inputs.xlsx`.
9. **Complete `FG_spp_Ecopath`/`FG_lookup` (Step 11)** — union of observed +
   reference-catalog species, zero-fill species with no base-year density,
   attach taxonomy, defensive check for FGs with no species at all. Written
   via `upsert_workbook_sheets()`.
10. **`traits_ewe` sheet (Step 12)** — parse `FG_WMed.xlsx`'s `traits_ewe`
    sheet (alternating "N: Group Name" header rows / species rows,
    fill-down logic), reconcile against `dataframe2`, flag naming-variant and
    missing-traits species. Written via `upsert_workbook_sheets()`.
11. **Finalize sheet order (Step 13)** — `finalize_workbook_sheet_order()`.

## Outputs

* **CSVs** (in `out_dir`): `survey_sample_coverage_map.png` and other plot
  PNGs (in `out_dir/plots`), `survey_unmatched_for_manual_review.csv`,
  `survey_outliers_flagged.csv`, `strata_area_by_area.csv`,
  `medias_fg_annual_density_by_area.csv`,
  `medias_fg_annual_density_regional.csv`,
  `species_density_regional_combined.csv`, `survey_fg_annual_index.csv`,
  `survey_fg_annual_index_regional_combined.csv`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written/updated:
  `FG_spp_Ecopath`, `Ecopath`, `Ecosim`, `FG_spp_Ecosim`, `FG_lookup`
  (renamed to `FG`), `traits_ewe`.

---

# Script 1b — `01_survey_density_custom.R`

**Role:** Alternative Step 1, for **any study area anywhere in the
Mediterranean that doesn't line up with GSA boundaries** — a single custom
bounding box or shapefile (e.g. one EwE model's own footprint, a bay, an
MPA, or a region spanning parts of several GSAs, in any part of the basin).
This is the general-purpose route for a fully user-defined region; Script 1a
above is the general-purpose route for any GSA-aligned region instead. Run
this **instead of** `01_survey_density_westmed.R`, never both. MEDITS-only —
does not run the MEDIAS acoustic survey.

Structurally identical to `01_survey_density_westmed.R` for the MEDITS
loading and FG-matching stages (Steps 1–4 use the exact same shared-library
functions). What's genuinely different:

* **`AREA_NAME`** — a user-supplied label for this custom study area (typed
  interactively via `rstudioapi::showPrompt()`, or hardcoded for the
  `daniel` user). Used to build a **dedicated output subfolder**
  (`out_dir/<AREA_NAME>/`) and to suffix every output filename via helper
  functions `af()`/`pf()`, so two different custom-region runs (different
  `AREA_NAME`s) never collide or overwrite each other's outputs, and never
  collide with `01_survey_density_westmed.R`'s own `out_dir`.
* **Study-area definition (Step 1b)** — `CUSTOM_AREA_TYPE <- "bbox"` (a
  rectangle defined by `CUSTOM_BBOX <- c(xmin, xmax, ymin, ymax)`) or
  `"shapefile"` (an arbitrary `sf` polygon read via `st_read()`). Either way
  the result is a single-feature `area_shp` with `AREA_ID_COL = "area_id"`,
  always `area_id = 1` — there is no GSA sub-division inside a custom region
  unless the user builds one.
* **`AreaID` is derived spatially**, not read off MEDITS' own `area` (GSA)
  column — every sample gets `AreaID = 1`, then `filter_samples_by_area()`
  drops everything outside the custom boundary.
* **Strata-area computation (Step 6)** — `compute_strata_area_by_area()`
  called with `area_ids = 1` against the single custom polygon rather than a
  set of ~30 named GSAs; the function itself is agnostic to this distinction.

**Exception to the `AREA_NAME` filename-suffixing rule:** three files stay at
plain, unsuffixed names on purpose, because `02_fao_catches.R` and
`04_pbqb_calc.R` read them by that exact fixed name regardless of which
Step-1 script produced them: `species_density_regional_combined.csv`,
`ecopath_ecosim_inputs.xlsx`, `strata_area_by_area.csv`. The dedicated
`AREA_NAME` subfolder is what keeps these three from colliding with another
run's copies.

## Inputs

Same as `01_survey_density_westmed.R` minus the MEDIAS/ASFIS/GSA-shapefile
inputs, plus (for `CUSTOM_AREA_TYPE = "shapefile"`) a user-supplied boundary
shapefile.

## Processing steps

Steps 1–5 mirror `01_survey_density_westmed.R` (see above) using the custom
`area_shp`. Steps 6–12 are the same Excel-export / `FG_spp_Ecopath` /
`traits_ewe` / sheet-order-finalization logic, applied to a single-area
result. (Since this script skips MEDIAS, its `species_density_regional`
table is MEDITS-only — no MEDITS+MEDIAS averaging step.)

## Outputs

Same file set and workbook sheets as `01_survey_density_westmed.R`
(see above), written to `out_dir/<AREA_NAME>/` with most filenames suffixed
`_<AREA_NAME>`, except the three contract files listed above which stay
unsuffixed.

---

# Script 2 — `02_fao_catches.R`

**Role:** Pipeline Step 2 of 4. **Requires** Step 1 to have already run — it
reads `strata_area_by_area.csv` (written by Step 1) to convert catch totals
into a t/km² density comparable to Biomass. Running this before Step 1 fails
with a clear "file not found" error pointing back to Step 1.

Config-driven catch-data loader: one `DATASETS` registry entry per source
format, so `GFCM_2025` and (a currently-commented-out) `FAO_2026` share one
processing path (`format = "gfcm_regional"`), while `FAO_2020`
(`format = "legacy_excel"`) is handled separately as a genuinely different
file layout. `DATASET_VERSION <- "GFCM_2025"` is the active, confirmed,
working source; `FAO_2020` is present but its file location is
**unconfirmed** on disk.

## Inputs

| File | Format | Role |
|---|---|---|
| `<git_dir>/data/raw/FI_Regional_2025.1.0/GFCM_Capture_Quantity.csv` | csv | Raw GFCM catch quantities |
| `.../CL_FI_SPECIES_GROUPS.csv` | csv | Species code → name reference (also auto-downloaded from FAO's open-data catalog if not found under `pcloud_dir`) |
| `.../CL_FI_COUNTRY_GROUPS.csv` | csv | Country code → name reference |
| `.../CL_FI_WATERAREA_DIVISION.csv` | csv | FAO/GFCM division code → name reference; catches filtered to divisions `37.1.1`/`37.1.2`/`37.1.3` |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → FG reference, same as Script 1 |
| `strata_area_by_area.csv` | csv | **From Step 1.** Total study-area km² for the density conversion |

### Column reference — key input files

**`GFCM_Capture_Quantity.csv`** (raw catch, `<git_dir>/data/raw/FI_Regional_2025.1.0/`):

| Column | Used for |
|---|---|
| `SPECIES.ALPHA_3_CODE` | Joined against `CL_FI_SPECIES_GROUPS.csv` |
| `COUNTRY.UN_CODE` | Joined against `CL_FI_COUNTRY_GROUPS.csv` |
| `DIVISION.CODE` (`area_code_col`) | Joined against `CL_FI_WATERAREA_DIVISION.csv`; filtered to `37.1.1`/`37.1.2`/`37.1.3` (Western Med divisions — change for a different region) |
| `MEASURE` | Filtered to `"Q_tlw"` (live-weight tonnes) |
| `PERIOD` | Year |
| `VALUE` | Catch quantity |

**Reference CSVs** (same folder):

| File | Columns used |
|---|---|
| `CL_FI_SPECIES_GROUPS.csv` | `3A_Code` (species code) → `Name_En` (species name) |
| `CL_FI_COUNTRY_GROUPS.csv` | `UN_Code` → `Name_En` (country name) |
| `CL_FI_WATERAREA_DIVISION.csv` | `Code` (`area_join_col`) → `Name_En` (division name) |

**`FAO_2020` legacy Excel** (`format = "legacy_excel"`, alternate/unconfirmed
source):

| Column | Renamed to |
|---|---|
| `Species (scientific name)` | `Species` |
| `Area (FAO division)` | `Division` |
| `Area (FAO subarea)` | `Subarea` — filtered to `"Western Med (37.1)"` |
| `Year`, `Quantity`, `Country` | used as-is |

## Processing steps

1. **Config + dataset loading** — resolve directories, `load_gfcm_regional()`
   merges capture/species/country/area reference CSVs, filters to Western
   Med divisions, `MEASURE == "Q_tlw"`, `PERIOD >= 1995`. Builds
   `country_ts`/`species_ts` (Year × Country/Species × Division × Catch).
2. **Plots + audit CSVs** — country- and species-level catch time series
   (`country_catch_timeseries.png`, `species_catch_timeseries.png`),
   `unmatched_species_FG_assignment.csv`, `westmed_species_timeseries.csv`.
3. **Species → FG matching pipeline** (inline, Steps 0–4, same safety
   philosophy as Script 1's taxonomy fallback — a query is only
   auto-resolved if it maps to exactly **one** distinct FG, otherwise flagged
   ambiguous and never guessed):
   - Step 1: direct match (catch-record species text literally equals an
     `FG_name`).
   - Step 2: common name → scientific name via `rfishbase::common_names()`
     (queried **for** the known species list, not searched from all of
     FishBase — avoids cross-matching an unrelated species that happens to
     share a common name).
   - Step 3: FAO genus-level fallback for NEI/aggregate categories (e.g.
     "Sardinellas nei" → genus *Sardinella*).
   - Step 4: word-containment fallback (full token containment only).
   - Manual overrides (e.g. "Turbot" → *Scophthalmus maximus*, correcting a
     wrong FishBase common-name entry).
   - A catch-total sanity check confirms no species record is double-counted
     across matching stages.
   - Output: `species_fg_matched.csv` (one row per species, `status` =
     resolved/unresolved, `match_method`).
4. **FG-level Year timeseries** — joins `species_ts` (has Year, no FG) against
   `species_fg_matched.csv`'s Species→FG mapping, re-aggregates to Year × FG.
   Output: `fg_catch_timeseries_<DATASET_VERSION>.csv` (landings-only, **not**
   confirmed to include discards — read by `04_pbqb_calc.R`'s optional
   `FG_YIELD_SOURCE = "fg_catch_csv"` path).
5. **Add `Catches_Ecopath`/`Catches_Ecosim` to the workbook** — reads
   `strata_area_by_area.csv`, sums to a total study-area km², calls
   `add_catches_to_ecopath_workbook()`. **Known, documented limitation:**
   GFCM Divisions 37.1.1–37.1.3 cover GSA 1–11 **plus** GSA 12 (Northern
   Tunisia), which the survey's `FILTER_AREAS` does not include — the catch
   density is therefore a slight overestimate relative to the Biomass area
   (GFCM's Division-resolution reporting can't isolate GSA 12's specific
   contribution to subtract it out).

**Not implemented — no fleet or gear disaggregation anywhere in this
script.** `GFCM_Capture_Quantity.csv` (the actual, confirmed source above)
resolves only to Country × Species × Division × Year — this is GFCM's
STATLANT-derived aggregate capture-production product, not its
Operational-Unit-level data. GFCM *does* collect catch/effort broken down
by **Operational Unit** (fleet segment × ISSCFG gear code × GSA), via the
DCRF's Task 2 (Catch)/Task 4 (Fleet)/Task 5 (Effort) and the Regional
Fleet Register, submitted by national correspondents — but that's a
genuinely different GFCM product from this file, and isn't confirmed to
have the same kind of open bulk-download `GFCM_Capture_Quantity.csv` does
(it feeds GFCM's own stock-assessment working groups rather than being
published as an open dataset the way STATLANT capture-production is).
Getting real fleet/gear resolution into this pipeline means sourcing that
Operational-Unit data through GFCM's DCRF/SAC channels, or — for the
EU-flagged portion of the fleet specifically — STECF's FDI (Fisheries
Dependent Information) database, which is the same source already
flagged elsewhere in this pipeline as the likely next step for a
discard-inclusive catch figure. Neither is currently plugged in anywhere;
`04_pbqb_calc.R`'s `Fmort_species`/`Fmort_FG` (below) inherit this same
limitation, since they're computed downstream of this file.

**Single fleet by default; multi-fleet via an optional `fleet_structure`
input.** `add_catches_to_ecopath_workbook()` (`lib_survey_fg_density_
functions.R`) now takes an optional `fleet_structure` argument: a table of
`FG_num`, `Fleet`, `prop_catch` (the share, 0–1, of each FG's total catch
attributed to each fleet). Left at its default (`NULL`, and that is still
this script's own default), the workbook is exactly what it always was —
one implicit fleet, `Catches_Ecopath`'s combined `Catch_<year>` columns,
`Catches_Ecosim`'s meta-rows with no `Fleet` field, no `Fleet_Structure`
sheet. Point `FLEET_STRUCTURE_PATH` (near the top of this script) at a CSV
in that `FG_num,Fleet,prop_catch` shape to split catches by fleet instead —
any FG *not* listed there still defaults to a single implicit `"Fleet_1"`,
so a partial table (only the FGs you actually have a fleet split for) is
fine. Once any FG resolves to more than one fleet, **every** FG row in
`Catches_Ecopath` switches to fleet-suffixed columns
(`Catch_<year>_<Fleet>`), `Catches_Ecosim` gets one column-set per FG ×
Fleet, and the resolved fleet table itself is written out as a new
`Fleet_Structure` sheet for traceability.

This only splits catch *after the fact* — it does not source fleet/gear
data itself (see the "not implemented" note above: `GFCM_Capture_
Quantity.csv` still carries no fleet/gear column), so the `fleet_structure`
CSV has to come from somewhere else (GFCM Operational Units, STECF FDI, or
a hand-built allocation). `04_pbqb_calc.R`'s `Fmort_species`/`Fmort_FG`
remain unaffected either way — they're computed from the same un-split
`fg_catch_timeseries_*.csv` regardless of whether `Catches_Ecopath`/
`Catches_Ecosim` end up split by fleet.

## Outputs

* **CSVs:** `unmatched_species_FG_assignment.csv`,
  `westmed_species_timeseries.csv`, `species_fg_matched.csv`,
  `fg_catch_timeseries_<DATASET_VERSION>.csv`.
* **Plots:** `country_catch_timeseries.png`, `species_catch_timeseries.png`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written: `Catches_Ecopath`,
  `Catches_Ecosim`, plus `Fleet_Structure` if `FLEET_STRUCTURE_PATH` was set
  and resolved to more than one fleet.

---

# Script 3 — `03_ecobase_query.R`

**Role:** Pipeline Step 3 of 4. Independent of Steps 1/2 — can run anytime
before Step 4. **Optional** input to Step 4 for literature-based PB/QB
gap-filling and comparison against the empirical formula estimates.

Queries **EcoBase** (ecobase.ecopath.org), an open-access repository of
published Ecopath models, via its SOAP/XML web service — the official
EcoBase R example ported from RCurl/XML to httr/xml2. The script author
notes this was **not tested against the live service** from within the
authoring sandbox (no network access to `sirs.agrocampus-ouest.fr`) — the
diagnostics throughout are there specifically to surface an endpoint change,
not just fail silently.

## Inputs (network)

| Endpoint | Purpose |
|---|---|
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client_3.php` | Full model list |
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client.php?no_model=<id>` | Per-model input values (Biomass/PB/QB per group) |

`pcloud_dir`/`git_dir` are resolved for consistency with the other scripts
but not actually used — this script is self-contained.

## Processing steps

1. Fetch the full EcoBase model list; save `ecobase_model_list_full.csv`.
2. Filter geographically to the Mediterranean, using each model's
   `geographic_extent` bounding box (not text/keyword matching on
   country/ecosystem names — more robust). Save
   `ecobase_mediterranean_models.csv`.
3. Filter further to `dissemination_allow == "true"` — EcoBase's own flag for
   which models actually have downloadable input data (per the site, only
   ~233 of ~500 listed models do; the rest are metadata/reference-only).
   `KNOWN_RELEVANT_MODEL_IDS` is the resulting query list.
4. Fetch per-model group-level Biomass/PB/QB for each disseminable model
   (`fetch_model_inputs()`), explicitly detecting the "data not available
   despite the flag" case (confirmed against a real response for model 418).
   `-9999` (Ecopath's own sentinel for "not provided") is converted to `NA`.
5. Save a full status audit (`ecobase_mediterranean_models_status.csv`) and
   the failure-only subset (`ecobase_models_without_data.csv`).
6. Compile the final summary table: `FG_name, PB, QB, EwE_model, year,
   authors` (model metadata columns matched by best-guess candidate names
   against `model_list`, since EcoBase's exact field names aren't confirmed).

## Outputs

* `ecobase_model_list_full.csv`, `ecobase_mediterranean_models.csv`,
  `ecobase_mediterranean_models_status.csv`, `ecobase_models_without_data.csv`,
  `ecobase_literature_pb_qb_raw.csv`,
  **`ecobase_literature_pb_qb_simple.csv`** (the file `04_pbqb_calc.R` reads).
* No workbook writes from this script directly — `04_pbqb_calc.R` is what
  merges this CSV into the workbook's `Ecobase` sheet.

### Column reference — `ecobase_literature_pb_qb_simple.csv`

| Column | Meaning |
|---|---|
| `FG_name` | EcoBase group name (matched to your own `FG_name` by text in `04_pbqb_calc.R` — group naming won't line up automatically across different published models) |
| `PB` | Literature Production/Biomass (`-9999`, EcoBase's own "not provided" sentinel, converted to `NA`) |
| `QB` | Literature Consumption/Biomass (same `-9999`→`NA` handling) |
| `EwE_model` | Source model name |
| `year` | Source model's reference year |
| `authors` | Source model's authors |

---

# Script 4 — `04_pbqb_calc.R`

**Role:** Pipeline Step 4 of 4 — run **last**. Requires Step 1's
`species_density_regional_combined.csv`. Optionally uses Step 2's
`fg_catch_timeseries_<DATASET_VERSION>.csv` (`FG_YIELD_SOURCE` toggle) and
Step 3's `ecobase_literature_pb_qb_simple.csv` — both degrade gracefully with
a clear message if missing, they don't hard-fail this script.

Estimates Production/Biomass (PB) and Consumption/Biomass (QB) for every EwE
Functional Group, following the taxon-specific methods in *"Quick guide on
how to calculate P/B and Q/B for EwE models"* (Vilas, Coll, Piroddi,
Steenbeek). Growth parameters, length-weight a/b, maturity, body weight,
longevity, trophic level, temperature, depth and aspect ratio are all fetched
automatically from FishBase + SeaLifeBase (`rfishbase` ≥ 4.0.0 required — the
script hard-stops with install instructions if an old, broken pre-4.0 CRAN
build, or a `duckdbfs` install missing `duckdb_config`, is detected).

## Inputs

| File | Format | Role |
|---|---|---|
| `species_density_regional_combined.csv` | csv | **From Step 1.** Reshaped via `lib_build_species_df_from_survey.R::build_species_df_from_survey()` into `species_df` (Species, FG, Biomass density, Yield=NA) |
| `fg_catch_timeseries_GFCM_2025.csv` | csv (optional) | **From Step 2.** FG-level catch, used to compute a per-FG fishing-mortality rate `Fmort_FG` (`FG_YIELD_SOURCE = "fg_catch_csv"`) |
| `strata_area_by_area.csv` | csv | **From Step 1.** Study area, needed to convert FG catch totals into a density comparable to Biomass |
| `ecobase_literature_pb_qb_simple.csv` | csv (optional) | **From Step 3.** Literature PB/QB for gap-filling/comparison |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet `fg_wmed_95` | xlsx | FG reference (loaded by explicit sheet **name**, since the workbook also has an unrelated `fg_ebro delta_21` sheet) |
| `landings_by_species_gsa_year.csv` | csv (optional, placeholder path — not yet a real data source) | Species-level landings, for a species-level Fmort path (`YIELD_SOURCE = "landings_csv"`, currently `"none"`) |
| FishBase / SeaLifeBase (`rfishbase`) | web API | Growth, length-weight, maturity, ecology/trophic level, swimming/aspect-ratio traits |
| WoRMS (`worrms`, via `lib_worms_taxonomy_lookup.R`) | web API | Taxonomic classification (Class → dispatch group: fish/mammal/seabird/phytoplankton/invertebrate) |
| FishLife R package | local DB (`FishBase_and_RAM$ParentChild_gz`) | Phylogenetically-imputed Loo/K/M/Winf, walking up the taxonomic tree when a direct match is missing |
| TropFishR R package | local | `M_empirical()` — Hoenig (1983), Then et al. (2015), Alverson & Carney (1975) natural-mortality estimators |

### Column reference — key input files

**`species_density_regional_combined.csv`** (from Step 1 — required columns,
enforced by `build_species_df_from_survey()`):

| Column | Meaning |
|---|---|
| `Year` | Survey year |
| `FG_num`, `FG_name` | Functional Group id/name |
| `ScientificName` | Species scientific name |
| `mean_density` | Species density, t/km² |

**`fg_catch_timeseries_GFCM_2025.csv`** (from Step 2, optional):

| Column | Meaning |
|---|---|
| `Year` | Catch year |
| `FG_num`, `FG_name` | Functional Group id/name |
| `Catch_t` | Total landed catch, tonnes (region-wide, landings-only) |

**`strata_area_by_area.csv`** (from Step 1):

| Column | Meaning |
|---|---|
| `AreaID` | GSA number (or `1` for a custom-region run) |
| `Stratum` | Depth-stratum number |
| `area_km2` | Stratum area within that GSA/region, km² |
| `prop` | That stratum's proportion of the area's total |

**`FG_WMed.xlsx`, sheet `fg_wmed_95`** — loaded by explicit sheet **name**
(not position), since the same workbook also has an unrelated
`fg_ebro delta_21` sheet. `readxl` disambiguates duplicate-looking headers
as `FG_name...2`/`FG_name...7` — handled with a flagged best-guess in the
code rather than assumed.

**`landings_by_species_gsa_year.csv`** (placeholder path — not yet a real
data source; schema below is what `attach_yield_from_landings()` expects if
this is ever wired up):

| Column | Meaning |
|---|---|
| `ScientificName` | Species scientific name |
| `Year` | Landings year |
| `AreaID` | GSA number |
| `catch_t` | Landed catch, tonnes |

```{r pbqb-comparison-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="fish_PB_methods_comparison.png — all 6 fish PB/M methods per species, chosen method highlighted."}
safe_include_graphics(
  "figures/fish_PB_methods_comparison.png",
  script = "04_pbqb_calc.R",
  out_dir_note = "<out_dir>/plots/fish_PB_methods_comparison.png"
)
```

## Processing steps

1. **Setup** — package loading, `rfishbase`/`duckdbfs` version guard, GUI or
   console progress tracker across 13 named pipeline stages.
2. **Load `species_df`** (`SPECIES_DF_SOURCE = "survey"`, the real path) via
   `lib_build_species_df_from_survey.R`.
3. **Fishing mortality attachment** — two independent attachment points:
   * (A) species-level, via `attach_yield_from_landings()` — currently
     inactive (`YIELD_SOURCE = "none"`; no species×GSA×year landings source
     exists yet).
   * (B) FG-level, via `fg_catch_timeseries_*.csv` — currently inactive
     (`FG_YIELD_SOURCE = "none"` by default, but wired and working when
     enabled) — computed as `Yield_FG / Biomass_FG` per FG, applied to fish
     PB inside `calc_fish()` as a **fallback** when species-level Yield is
     absent (species-level always takes priority when both exist).
     **Known, unresolved caveat (same root cause as `Catches_Ecopath`/
     `Catches_Ecosim` below):** `Yield_FG`'s denominator (`area_total_km2`,
     summed from `strata_area_by_area.csv`) is the *survey's* GSA-filtered
     area, while its numerator (`fg_catch_timeseries_*.csv`'s `Catch_t`) is
     filtered by GFCM Division, which for the Western Med example covers
     GSA 1–11 **plus** GSA 12 — so `Yield_FG`, and therefore `Fmort_FG` and
     the fish PB it feeds, is a **slight overestimate** wherever this path
     is enabled. Not fixable from this side (GFCM's catch data can't be
     split below Division resolution); see the same caveat under
     [Script 2 — `02_fao_catches.R`] for the full explanation.
     **Also not implemented — no fleet or gear disaggregation:** both
     `Fmort_species` (path A, inactive) and `Fmort_FG` (path B) are a
     single Yield/Biomass ratio lumped across the whole region and every
     gear/fleet that landed that species/FG — neither the
     `landings_by_species_gsa_year.csv` placeholder schema nor
     `fg_catch_timeseries_*.csv` carries a fleet or gear column, so
     there's nothing here that could produce a fleet- or gear-specific F
     even once a real landings file is plugged in. See the matching note
     under [Script 2 — `02_fao_catches.R`] for what a real fix would
     require (GFCM's Operational-Unit data or STECF's FDI database).
     **`Fmort` itself stays single-fleet even where `Catches_` sheets are
     split:** unlike `Catches_Ecopath`/`Catches_Ecosim` (which can now be
     split by fleet via `fleet_structure` — see the "Single fleet by
     default; multi-fleet via an optional `fleet_structure` input" note
     under [Script 2 — `02_fao_catches.R`]), this script never sees
     `fleet_structure` and computes `Fmort_species`/`Fmort_FG` from the
     same un-split `fg_catch_timeseries_*.csv` regardless. A genuinely
     fleet-specific F would need Yield split by fleet *before* this ratio,
     not just the `Catches_` sheets split after the fact.
4. **Taxonomic classification** — `worms_taxonomy_lookup()` for every species
   in `species_df`, dispatch-classified into `fish` / `mammal` / `seabird` /
   `phytoplankton` / `invertebrate` (default) by WoRMS Class, with a
   synonym→accepted-name resolution fallback for species WoRMS returns no
   Class for.
5. **Trait fetching (Step 2a–2f)** — `select_best_rows()` picks the best
   population-specific study per species from FishBase/SeaLifeBase
   multi-record tables, tiered: Western-Med locality keywords → broader
   Mediterranean keywords → most recent year → first available.
   `extract_provenance()` captures Locality/Year/Reference citation for each
   selected row (used later for the `References` sheet/CSV).
6. **Gap-filling** — `fill_by_taxonomic_proximity()`: local within-dataset
   donors first (Genus→Family→Order→Class, restricted to directly-computed
   values), then, for methods in `EXTERNAL_FORMULA_REGISTRY` (fish/mammal/
   general Gascuel formulas plus raw-trait identity mappings), an **external**
   FishBase/SeaLifeBase donor lookup — querying other species sharing the
   same taxonomic rank, not limited to species already in `species_df`.
7. **Taxon-specific PB/QB calculation** (dispatch by group) — the diagram
   below shows the actual dispatch and method-cascade logic:

   ```{=html}
   <div class="mermaid">
   %%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
   flowchart LR
       T["worms_taxonomy_lookup()&lt;br/&gt;WoRMS Class"] --> F["calc_fish()&lt;br/&gt;PB: 6 M methods + Fmort&lt;br/&gt;QB: 4 methods"]
       T --> M["calc_mammal()&lt;br/&gt;PB: Siler surv. to Gascuel&lt;br/&gt;QB: Innes&amp;Trites to Q/P=3"]
       T --> S["calc_seabird()&lt;br/&gt;PB: Gascuel only&lt;br/&gt;QB: Nilsson-Nilsson (no Q/P=3)"]
       T --> I["calc_invertebrate() (default)&lt;br/&gt;PB: Tumbiolo/Brey to Gascuel&lt;br/&gt;QB: Q/P=3 only"]
       F --> D["fill_by_taxonomic_proximity()&lt;br/&gt;local then external donors"]
       M --> D
       S --> D
       I --> D
       D --> AGG["FG-level aggregation&lt;br/&gt;biomass-weighted mean"]
       AGG --> OUT["Ecobase merge -&gt; PB_QB,&lt;br/&gt;PB_QB_spp, Ecobase, References"]

       classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
       classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
       classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
       classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
       class T gray
       class F,M,S,I purple
       class D teal
       class AGG,OUT coral
   </div>
   ```

   *Phytoplankton is flagged by `worms_taxonomy_lookup()` but not computed
   (needs separate production-sampling data, out of scope for this
   pipeline) — not shown above.*

   * `calc_fish()` — PB from 6 independent M methods (Pauly 1980, FishLife,
     Gascuel 2008, Hoenig 1983, Then et al. 2015, Alverson & Carney 1975) +
     Fmort; QB from 4 methods (Palomares & Pauly 1998 Eq.27/Eq.26, Christensen
     & Pauly 1992, Q/P=3 fallback). Chosen value via priority order or mean
     (`PB_QB_SELECTION_MODE`).
   * `calc_mammal()` — PB via Siler-survivorship (Barlow & Boveng 1991),
     fallback Gascuel (2008); QB via Innes & Trites (1997), fallback Q/P=3.
   * `calc_seabird()` — PB via Gascuel (2008) only; QB via Nilsson & Nilsson
     (1976), **no** Q/P=3 fallback (that heuristic is fish/ectotherm-derived
     and would badly underestimate endotherm QB).
   * `calc_invertebrate()` — PB via Tumbiolo & Downing (1994), Brey (1999),
     fallback Gascuel (2008); QB via Q/P=3 only (Brey's 2012 ANN model
     flagged as non-automatable — top-10 highest-biomass invertebrates
     exported instead for manual cross-check).
   * Phytoplankton — flagged, not computed (needs separate production-
     sampling data, out of scope for this pipeline).
8. **FG-level aggregation** — biomass-weighted mean PB/QB per FG, with
   coverage/count audit columns (`n_species_with_PB`, `biomass_coverage_PB`,
   etc.).
9. **EcoBase literature merge** (if `ecobase_literature_pb_qb_simple.csv`
   exists) — averages multiple matching literature models per `FG_name`,
   flags >2×/<0.5× divergence from the empirical estimate, gap-fills only
   where the empirical value is `NA`. Writes sheet **`Ecobase`**.
10. **FG-level F top-up** (if `FG_YIELD_SOURCE = "fg_catch_csv"`) — applied
    only to FGs with no species-level F already baked in (mainly mammal/
    seabird/invertebrate FGs), to avoid double-counting against fish's own
    species-level Fmort.
11. **Workbook write** — `add_pbqb_to_ecopath_workbook()` writes sheet
    **`PB_QB`** (one row per FG — the biomass-weighted rollup), and, since
    `results` (the species-level table) is passed in as its
    `species_pb_qb` argument, also writes sheet **`PB_QB_spp`** (one row
    per species, mirroring `FG_spp_Ecopath`'s shape — every species that
    went into computing `PB_FG`/`QB_FG`, with three proportion columns:
    `prop_biomass_FG` — share of the FG's total biomass, regardless of
    whether that species got a PB/QB; `prop_biomass_PB`/`prop_biomass_QB`
    — that species' actual weight inside `PB_FG`'s/`QB_FG`'s own
    biomass-weighted mean, `NA` if that species' own PB/QB didn't
    compute. This is the sheet to check *which* species and *how much*
    of each FG's biomass a given `PB_FG`/`QB_FG` number was built from —
    without it, that detail only existed outside the workbook, in
    `species_pb_qb_by_taxon_group.csv`). A separate `upsert_workbook_sheets()`
    call writes sheet **`References`** (later renamed `PB_QB_References_`
    by `finalize_workbook_sheet_order()`, run from Script 1).
12. **Diagnostic plots** — per-method PB/QB comparison plots for fish, plus
    paginated species-level and FG-level PB/QB comparison plots across every
    dispatch group (auto-paginated when a plot would exceed `ggsave`'s
    50-inch height limit).
13. **Ecopath-ready CSV export** — `ecopath_ready_PB_QB.csv`, formatted to
    match a real Ecopath Basic Input CSV's exact column layout and
    European comma-decimal number formatting (verified against a real
    exported file, `westernmed90s-Basic_input.csv`).

## Outputs

* **CSVs:** `invertebrates_for_brey_manual_check.csv`,
  `species_pb_qb_by_taxon_group.csv`, `fg_pb_qb_weighted.csv`,
  `phytoplankton_needs_separate_method.csv`,
  `fg_pb_qb_weighted_with_ecobase.csv`, `fg_pb_qb_weighted_with_F.csv`,
  `species_parameter_references.csv`, `ecopath_ready_PB_QB.csv`.
* **Plots:** `fish_PB_methods_comparison.png`,
  `fish_QB_methods_comparison.png`,
  `species_PB_QB_comparison[_pageN].png`, `FG_PB_QB_comparison[_pageN].png`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written: `PB_QB`,
  `PB_QB_spp`, `Ecobase`, `References` (renamed `PB_QB_References_`).

---

# Shared library — `lib_survey_fg_density_functions.R`

Sourced by all four numbered scripts; has **no top-level driver code** of its
own — every symbol below is a function definition only.

## FG-matching functions

Every observed species runs through this cascade, in order, until it
resolves to a Functional Group or falls through to manual review — the
diagram shows the actual fallback order implemented in the code below
(same style as `medbs_pipeline_methodology.qmd`'s `@sec-fg-matching`
diagram; each stage only attempts species the previous one left
unmatched).

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["match_species_to_fg()&lt;br/&gt;direct ScientificName match"] --> A2["match_nominate_subspecies_fg()&lt;br/&gt;strip repeated trinomial word, retry"]
    A2 --> A3["apply_seed_fg_rules()&lt;br/&gt;manual rules, zero-reference FGs"]
    A3 --> A4["fallback_match_fg_by_taxonomy()&lt;br/&gt;Genus to Phylum, exactly-one-FG rule"]
    A4 --> A5["summarize_unresolved_species()&lt;br/&gt;manual review"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    class A1,A2,A3,A4 purple
    class A5 gray
</div>
```

| Function | Purpose | Returns |
|---|---|---|
| `validate_survey_data(dataframe1, dataframe2, strata = TRUE)` | Fail-fast schema/uniqueness check on the two standardized input tables (required columns present, `SampleID` unique per survey) | `invisible(TRUE)` |
| `prepare_fg_lookup(dataframe2)` | Deduplicates species→FG reference; keeps genuine stanza splits (a species exclusively mapped to ≥2 FGs), excludes/flags ambiguous multi-FG species | deduped `data.table(ScientificName, FG_num, FG_name)` |
| `match_species_to_fg(dataframe1, fg_lookup_safe)` | Direct left-merge by `ScientificName` | merged `dt` |
| `match_nominate_subspecies_fg(dt, fg_lookup_safe)` | Strips a repeated trinomial word (e.g. "Diplodus sargus sargus" → "Diplodus sargus") and retries the match | `dt` with newly resolved rows |
| `fetch_taxonomy(species_names, taxonomy_source = "worms")` | Dispatches to `fetch_taxonomy_worms()`, `fetch_taxonomy_fishbase()`, or both | `data.table(ScientificName, Genus, Family, Order, Class, Phylum)` |
| `fallback_match_fg_by_taxonomy(dt, fg_lookup_safe, taxonomy_source = "worms", rank_levels = c("Genus","Family","Order","Class","Phylum"))` | Data-driven rank fallback — a rank value is used only if it maps to exactly one FG among reference species | `dt`, with `attr(dt,"fetched_taxonomy")` and `attr(dt,"still_unresolved_taxonomy")` attached |
| `apply_seed_fg_rules(dt, dataframe2, seed_rules, species_exceptions = NULL)` | Explicit small rule table for FGs with **zero** reference species (nothing for the automatic fallback to learn from); species-level exceptions applied first | `dt` |
| `summarize_unresolved_species(dt, taxonomy = NULL)` | Mean biomass + appearance count per still-unmatched species, sorted by appearance count | summary `data.table` |

## Area / stratum / density functions

| Function | Purpose | Returns |
|---|---|---|
| `match_samples_to_area(dt, area_shp, area_id_col)` | Spatial join (`sf::st_join`) of sample points to polygons; no-op if `AreaID` already fully populated | `dt` |
| `assign_depth_stratum(dt, strata_def)` | Bins `Depth` into strata via `findInterval()`; out-of-range depths → `Stratum = NA` | `dt` |
| `filter_samples_by_area(dt, area_filter, filter_type = c("shapefile","bbox"), points_crs = 4326)` | Custom sub-region filter, independent of GSA/`AreaID` | `dt` filtered to in-area `SampleID`s |
| `compute_sample_densities(dt)` | `Density := Biomass / Effort` | `dt` |
| `apply_catchability_correction(dt, catchability_table, default_q = 1, species_taxonomy = NULL, exempt_fg_names = NULL, exempt_species = NULL)` | 5-tier resolution: direct species → genus "spp" entries → explicit broad-rank entries → taxonomic-proximity fallback → `default_q`; exemptions override last. Divides `Density`/`Biomass` by resolved `q` | `dt` |
| `remove_sample_outliers(dt, threshold = 70, min_samples = 5, drop_outliers = TRUE, log_scale = TRUE)` | Per-species/per-area robust z-score (median/MAD) on `log1p(Density)`; sets flagged `Density`/`Biomass` to `NA` if `drop_outliers = TRUE` | `dt`, with `attr(dt,"flagged_outliers")` |
| `compute_fg_densities_by_stratum(dt, strata = TRUE)` | Sums species density within FG per sample, then across samples by area/year/[stratum]/FG | `per_group_fg` (`density_sum`) |
| `compute_strata_area_by_area(area_ids, area_shp, area_id_col, strata_def, cache_path = NULL, resolution = 1)` | Downloads NOAA bathymetry per area, bins cell area into depth strata (latitude-corrected); optional CSV cache | `strata_area_by_area` table |
| `weight_by_strata(per_group_fg, n_samples_by_stratum, strata_area_by_area)` | Core strata-weighted estimator: `D_area = Σ_h (density_sum_h/n_samples_h) · area_prop_h` | `fg_index`, with `attr(fg_index,"per_stratum")` |
| `simple_area_density(per_group_fg, n_samples_by_area)` | Unstratified path (`strata = FALSE`) | `data.table` |
| `weight_by_area(fg_index_stratified)` | Region-wide weighting from absolute stratum areas (`per_stratum` attribute) — requires the stratified path upstream | `fg_index_regional` |
| `compute_species_densities_by_stratum(dt, strata = TRUE)` / `weight_species_by_area(per_group_sp, n_samples_by_stratum, strata_area_by_area)` | Species-level analogs of the two functions above, feeding the `FG_spp` sheets | `data.table(Year, FG_num, FG_name, ScientificName, mean_density)` |
| `compute_cv_log_by_fg(value_dt, value_col, min_replicates = 2)` | `cv_log = sd(log1p(x))/mean(log1p(x))` per FG/Year, averaged across years — feeds the Ecosim "Weight" row | `data.table(FG_num, FG_name, cv_log, n_years)` |

## Plotting functions

| Function | Produces |
|---|---|
| `plot_sample_map(dt, area_shp, area_id_col, ..., by_year = FALSE, selected_areas = NULL, land_on_top = FALSE)` | Sample coverage map, points colored inside/outside the study area, polygons shaded by sample intensity |
| `plot_fg_timeseries_by_area(fg_index, ...)` / `plot_fg_timeseries_regional(fg_index_regional, ...)` | Faceted FG density time series, per-area vs. region-wide |
| `plot_strata_profile(fg_index_stratified, strata_def, ...)` | Depth-strata validation bar chart (pre-area-weighting, per-sample density by stratum) |

## Excel workbook functions

| Function | Purpose |
|---|---|
| `upsert_workbook_sheets(sheets, out_path)` | Adds/replaces the named sheets in `out_path`, **leaves every other sheet untouched**. Creates the workbook if it doesn't exist yet. Safe to call repeatedly, from any script, in any order. |
| `read_existing_ts_years(out_path, sheet_name)` | Reads year values from column 1 of an existing sheet, e.g. to keep `Catches_Ecosim`'s years aligned with an already-written `Ecosim` sheet |
| `export_ecopath_ecosim_excel(fg_index_regional, species_density_regional, n_samples_by_area_year, dataframe2, year_ecopath, ts_years, out_path, species_taxonomy = NULL, fg_cv_log = NULL, normalize_ts = TRUE)` | Writes `FG_spp_Ecopath`, `Ecopath`, `Ecosim`, `FG_spp_Ecosim` |
| `add_catches_to_ecopath_workbook(fg_catch, fg_lookup, out_path, year_ecopath, area_km2, ts_years = NULL, fleet_structure = NULL)` | Converts raw catch tonnes to t/km² density; writes `Catches_Ecopath`, `Catches_Ecosim`. `fleet_structure` (optional `FG_num`/`Fleet`/`prop_catch` table) splits catch by fleet and also writes `Fleet_Structure`; left `NULL`, output is unchanged (single implicit fleet, no `Fleet_Structure` sheet) |
| `add_pbqb_to_ecopath_workbook(fg_weighted, out_path, species_pb_qb = NULL)` | Writes `PB_QB` (FG-level). If `species_pb_qb` (04_pbqb_calc.R's species-level `results` table) is also passed, writes `PB_QB_spp` too — one row per species, with `prop_biomass_FG`/`prop_biomass_PB`/`prop_biomass_QB` |
| `finalize_workbook_sheet_order(out_path, rename_map, target_order)` | **Run last.** Renames `FG_lookup`→`FG` and `References`→`PB_QB_References_`, then reorders every sheet to `target_order`. Sheets not yet present are skipped with a message (not an error); sheets present but not in `target_order` are kept, appended at the end, with a warning. *(As of the latest patch, if a rename target is already occupied by a stale sheet from an earlier run, that stale sheet is replaced rather than raising an error — see the "Known issue" note below.)* |

The diagram below shows which script/function writes which sheet into the
one shared workbook, and how `finalize_workbook_sheet_order()` renames and
reorders it at the end.

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    S1["Step 1a/1b&lt;br/&gt;export_ecopath_ecosim_excel()"] --> W["ecopath_ecosim_inputs.xlsx&lt;br/&gt;(shared workbook)"]
    S2["Step 2&lt;br/&gt;add_catches_to_ecopath_workbook()"] --> W
    S4a["Step 4&lt;br/&gt;add_pbqb_to_ecopath_workbook()"] --> W
    S4b["Step 4&lt;br/&gt;upsert_workbook_sheets()&lt;br/&gt;(Ecobase, References)"] --> W
    W --> F["finalize_workbook_sheet_order()&lt;br/&gt;rename + reorder, run last"]
    F --> O["output/ecopath_ecosim_inputs.xlsx&lt;br/&gt;(final, ordered)"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class S1,S2,S4a,S4b purple
    class W gray
    class F,O coral
</div>
```

**Final target sheet order** (`target_order` used in every numbered
script's own final call):

1. `FG` (renamed from `FG_lookup`)
2. `Ecopath`
3. `Catches_Ecopath`
4. `FG_spp_Ecopath`
5. `PB_QB`
6. `PB_QB_spp`
7. `Ecobase`
8. `PB_QB_References_` (renamed from `References`)
9. `traits_ewe`
10. `Ecosim`
11. `FG_spp_Ecosim`
12. `Catches_Ecosim`
13. `Fleet_Structure` — only present when `add_catches_to_ecopath_workbook()` was called with a `fleet_structure` that resolved to more than one fleet; skipped (not an error) otherwise, same as any other not-yet-present target-order sheet

### Known issue with `finalize_workbook_sheet_order()` reruns

`upsert_workbook_sheets()` always writes fresh data back under the
**pre-rename** literal sheet name (e.g. `FG_lookup`), since it has no way to
know a previous call to `finalize_workbook_sheet_order()` already renamed
that sheet to `FG`. On a second run of any Step-1 script against an existing
workbook, this recreates the exact rename collision
(`finalize_workbook_sheet_order()` trying to rename `FG_lookup`→`FG` while a
stale `FG` sheet from the previous run is still there) every time. The fix
applied to this library treats that as expected re-run behaviour: the stale
sheet at the target name is replaced with the current run's data rather than
raising an error.

---

# Other library files

## `lib_worms_taxonomy_lookup.R`

Library file, no top-level driver code — `source()`'d automatically by the
numbered scripts. Defines a single function:

**`worms_taxonomy_lookup(names_vector)`** — the single taxonomy source used
everywhere in the pipeline (so genus/family comparisons are always
apples-to-apples, never a FishBase-derived family compared against a
WoRMS-derived one for the same actual family). Strips `"Genus spp."`
suffixes so genus-level placeholders still resolve, chunks the query list
(50 names/chunk) against `worrms::wm_records_names()`, picks the best
(`status == "accepted"` preferred) match per name, and returns
`data.table(original_name, search_term, AphiaID, matched_name, rank, status,
phylum, class, order, family, genus)`. Explicitly namespaces `purrr::map()`
rather than calling `map()` bare — the `maps` package (loaded elsewhere in
this pipeline for basemap plotting) also exports an unrelated `map()`, and
whichever package attached most recently silently wins an unqualified call.

## `lib_build_species_df_from_survey.R`

Library file — bridges `01_survey_density_westmed.R`'s survey output and
`04_pbqb_calc.R`'s PB/QB estimation. Defines:

**`build_species_df_from_survey(survey_csv_path, year_ecopath_range,
out_rds_path = NULL)`** — reads `species_density_regional_combined.csv`,
restricts to `YEAR_ECOPATH`'s snapshot years (the same years used for the
`Ecopath` sheet's Biomass column, so PB/QB weighting and the reported Biomass
describe the same time snapshot), collapses to one mean-density `Biomass`
value per species (stops with an error if any species maps to more than one
FG within that year range — shouldn't happen given upstream FG-matching
logic), sets `Yield := NA` (this survey pipeline has no catch/landings data
of its own — F = Yield/Biomass isn't computable from it directly). Optionally
saves the result as an `.rds`. Returns
`species_df(Species, FG, FG_name, Biomass, Yield)`.

---

# Standalone cartography utilities (not part of the numbered pipeline)

## `export_gsa_boundaries.R`

Standalone — does not need `01_survey_density_westmed.R` or the shared
library sourced first. Downloads the same GFCM GSA shapefile via the same URL
and GSA-numbering fix (`SMU_CODE` 111/112 → 11). **Western-Med-specific as
delivered:** restricts to GSA 1–11 — change the `gsa_num %in% 1:11` filter to
any other GSA range to export a different region's boundaries. Simplifies
the result for web embedding via
`rmapshaper::ms_simplify(keep = 0.05, keep_shapes = TRUE)` (keeps small
polygons like Alboran Island, GSA 2, from being deleted outright), and writes
`gsa_boundaries.geojson` — the file embedded in
`medbs_pipeline_methodology.qmd`'s interactive Leaflet map.

## `fig_WMed_basemap.R`

Standalone map-generation script (not sourced by anything else), purely
cartographic — not required by the numbered pipeline at all.
**Western-Med-specific as delivered:** downloads the GFCM GSA shapefile,
splits polygons into Western Med (GSA 1–11, the target study area) vs.
context GSAs (12, 16 — Northern Tunisia, South of Sicily) and draws the
Sicily Channel divide (Cape Bon–Cape Lilibeo) as the Western/Central
Mediterranean boundary; all three of those choices would need editing to
produce an equivalent basemap for a different region. The general
technique — label placement via the "pole of inaccessibility" algorithm
(`polylabelr`, more robust than `st_point_on_surface()` for narrow,
coastline-hugging GSAs) plus a high-resolution (`rnaturalearthhires`)
coastline — is reusable as-is for any region. Output:
`output/plots/westmed_gsa_map.png`.

---

# Adapting this pipeline to a different region

Everything below applies whether the target region is a single GSA, an
arbitrary GSA subset, the whole Mediterranean GSA-by-GSA, or a fully
custom (non-GSA) boundary.

**To model a different GSA or set of GSAs** (Adriatic, Ionian, Aegean,
Levant, or the whole basin), starting from `01_survey_density_westmed.R` as
a template:

* `FILTER_AREAS` — set to the target GSA number(s). No other GSA-based code
  in `lib_survey_fg_density_functions.R` hardcodes a Western-Med-only
  assumption — `compute_strata_area_by_area()`, `weight_by_strata()`,
  `weight_by_area()` etc. are all region-agnostic and operate on whatever
  `area_ids`/`area_shp` they're given.
* `ALBORAN_GSAS`/`fix_alboran_longitude_sign()` — a Western-Med-only
  correction (GSA 1–3); harmless no-op elsewhere, delete if not needed.
* `MEDIAS_HANDBOOK_AREAS`/`GSA_AREA_CROSSWALK` — only populated for Western
  Med GSAs; add the target region's own rows from the MEDIAS Handbook
  (or rely entirely on the bathymetry-derived 10–200 m fallback if MEDIAS
  data or handbook-reported areas aren't available for that region).
* `out_dir`/`pcloud_dir`/`git_dir` paths, and the raw MEDITS/MEDIAS/FG-
  reference file paths under them — point at the target region's own data.
* `export_gsa_boundaries.R`/`fig_WMed_basemap.R` — optional cartography
  only; edit the GSA-range filters if you want an equivalent map for the
  new region, otherwise skip both entirely (nothing downstream needs them).

**To model a fully custom (non-GSA) region anywhere in the Mediterranean**,
use `01_survey_density_custom.R` directly: set `AREA_NAME`, `CUSTOM_AREA_TYPE`
(`"bbox"` or `"shapefile"`), and `CUSTOM_BBOX`/`CUSTOM_SHAPEFILE_PATH` to the
target boundary — no GSA-specific code path is involved at all in this
script (the Alboran-sign fix is present but is a documented no-op outside
GSA 1–3).

Everything downstream of Step 1 — `02_fao_catches.R`, `03_ecobase_query.R`,
`04_pbqb_calc.R`, and the shared library's Excel-export/PB-QB functions — is
already region-agnostic: they operate on whatever `strata_area_by_area.csv`
and `species_density_regional_combined.csv` Step 1 produced, with no
Western-Med-specific logic beyond `02_fao_catches.R`'s `area_filter_values`
(the FAO/GFCM division codes 37.1.1–37.1.3, which likewise need changing to
match the target region's own FAO division codes) and `03_ecobase_query.R`'s
`MED_LON_MIN/MAX`/`MED_LAT_MIN/MAX` bounding box (already covers the **full**
Mediterranean by default — Gibraltar to the Levant — so it does not need
editing for a sub-region; narrow it only if you specifically want to exclude
literature models from elsewhere in the basin).

---

# Methodology document — `medbs_pipeline_methodology.qmd`

A Quarto methodology write-up (`format: html`, all R chunks `eval: false` —
illustrative, not executed at render time) documenting
`lib_survey_fg_density_functions.R` and the two Step-1 driver scripts. Key
features:

* **Embedded interactive map** — the actual `gsa_boundaries.geojson` (from
  `export_gsa_boundaries.R`) is inlined as a JSON blob and rendered via
  Leaflet.js. Lets a reader toggle between the GSA-based and custom-region
  workflows, live-preview `FILTER_AREAS`/`CUSTOM_BBOX`, and generate a
  ready-to-paste R config block — all client-side JavaScript, no R executes
  behind it.
* **Section structure:** Overview (with mermaid diagrams for the pipeline,
  FG-matching sub-flow, and density/weighting sub-flow) → Input files →
  Interactive configuration → Functional Group matching (worked examples of
  each matching stage) → Study-area definition (GSA vs. custom) → Per-haul
  density and corrections (swept-area formula, depth strata, catchability,
  outliers) → Strata-weighted density → Region-wide aggregation → MEDIAS
  acoustic survey → Validation plots → Sample coverage map → Outputs → a
  full `ecopath_ecosim_inputs.xlsx` deep-dive, including the same final
  sheet-order table reproduced above.
* Real output files are copied into `figures/`/`downloads/` folders next to
  the rendered document (via `safe_include_graphics()`/`safe_download_link()`
  helpers) so the rendered HTML is portable off the author's own machine,
  since `out_dir` is otherwise an absolute local path.
