---
title: "Biomass, Fisheries, PB/QB-Traits & Diet Pipeline"
subtitle: "Code documentation for the four EwE input-building scripts — biomass, fisheries, PB/QB & traits, diet — for any GFCM GSA subset or user-defined region in the Mediterranean"
author: "Daniel Vilas"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true
execute:
  eval: false
  echo: true
editor: visual
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(tidy = "styler")  # eval/echo now set via the YAML execute: block above
# figure chunks below override eval/echo individually so this document can
# show real pipeline output when it's available; everything else in the
# document stays non-executed (this is documentation, not a live report).

# safe_include_graphics(): shows the real figure if it's been copied into
# figures/ next to this .Rmd (see "Populating the real figures" below);
# otherwise prints a plain placeholder message instead of erroring, so this
# document always knits cleanly whether or not the pipeline has been run
# yet against the real WestMed EwE pCloud drive data.
safe_include_graphics <- function(path, script, out_dir_note) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(
      "*Figure not yet available.* Run **", script, "**",
      " against the real WestMed EwE pCloud drive data, then copy\n`",
      out_dir_note, "`\ninto a `figures/` folder next to this `.Rmd` ",
      "(as `", basename(path), "`) and re-knit.",
      sep = ""
    )
  }
}
```

```{css, echo=FALSE}
body {
  max-width: 900px;
}
/* keep the whole page from ever scrolling sideways */
html, body {
  overflow-x: hidden;
}
/* wide tables: scroll horizontally inside their own box instead of
   pushing the page wider than the viewport */
.table-responsive,
table {
  display: block;
  max-width: 100%;
  overflow-x: auto;
}
table {
  width: auto !important;
}
table td, table th {
  white-space: normal;
  word-break: break-word;
}
/* code blocks (the ASCII pipeline diagram, chunk output, etc.):
   wrap long lines instead of overflowing, but still allow a
   horizontal scrollbar as a fallback for anything that truly can't wrap */
pre, code {
  white-space: pre-wrap;
  word-break: break-word;
  overflow-x: auto;
  max-width: 100%;
}
/* mermaid diagrams (see medbs_pipeline_methodology.qmd for the same look):
   don't let the pre-wrap rule above touch them - mermaid.js replaces the
   div's contents with an SVG at render time */
div.mermaid {
  overflow-x: auto;
  max-width: 100%;
  text-align: center;
}
```

<script src="https://cdn.jsdelivr.net/npm/mermaid@10.9.8/dist/mermaid.min.js"></script>
<script>
  mermaid.initialize({ startOnLoad: true, theme: "neutral",
                        flowchart: { padding: 20, nodeSpacing: 60, rankSpacing: 90 } });
</script>

# Purpose of this document

**What this produces and why it matters:** running this pipeline end to end
produces one Excel workbook, `ecopath_ecosim_inputs.xlsx`, with the eleven
sheets an Ecopath-with-Ecosim (EwE) model needs to get off the ground for a
region — baseline biomass, catches/landings/discards by fleet, natural and
fishing mortality, production and consumption ratios, life-history traits,
and the predator-prey diet matrix. Having these sheets, already computed
from real survey and catch data rather than left as blanks, is what lets a
modeller open Ecopath and actually build and balance a working model for
this region instead of starting from a spreadsheet of guesses.

This document describes a general-purpose R pipeline for turning MEDITS/
MEDIAS trawl-survey and catch data into input sheets for an Ecopath-with-
Ecosim (EwE) model, for **any region in the Mediterranean Sea** — a single
GFCM Geographical Sub-Area (GSA), any subset or the full set of GSAs, or a
completely user-defined region (a bounding box or an arbitrary shapefile
that doesn't line up with GSA lines at all, e.g. a bay, an MPA, or a
footprint spanning parts of several GSAs). It covers every script provided,
what each one reads and writes, and every function in the shared library
those scripts call.

The scripts as delivered are configured for the **Western Mediterranean**
(GSA 1–11) as a **worked example** — this is what `01_biomass.R`,
and `export_gsa_boundaries.R` currently point at, and
what a few reference tables (e.g. the MEDIAS Handbook area crosswalk) are
currently populated for. The underlying mechanism — GSA-based selection via
`FILTER_AREAS`, or an arbitrary custom region via `01_biomass.R`'s
`AREA_MODE <- "custom"` switch — is not Western-Med-specific; see
[Adapting this pipeline to a different region]
below for exactly what to change to point the same pipeline at another part
of the Mediterranean.

All scripts read from and write to **one shared workbook**,
`output/ecopath_ecosim_inputs.xlsx`, via two library functions designed for
exactly this purpose:

* `upsert_workbook_sheets()` — adds/replaces named sheets, leaves every other
  sheet in the workbook untouched. Safe to call from any script, in any order,
  any number of times.
* `finalize_workbook_sheet_order()` — called at the end of every numbered
  script. Drops intermediate sheets (they live as CSV in each block's folder)
  and orders the 11 final sheets (see [Final target sheet order (11 sheets)]).

Because of this design, the numbered pipeline scripts are
**order-independent with respect to the shared workbook** — except where a
script's own calculations literally need a file another script produced
first (documented per-script below).

# Region, time window & configuration

Before anything else, this is what a viewer needs to see to know **what
was actually modeled**: which region, which years, and which top-level
settings.

* **Region — `AREA_MODE`** (set at the top of `01_biomass.R`):
  - `"westmed"` (the default, and what this worked example uses) —
    GSA-based selection via `FILTER_AREAS <- 1:11`, i.e. **GSA 1–11, the
    Western Mediterranean**. Any other GSA or set of GSAs anywhere in the
    GFCM area can be modeled by changing this one line.
  - `"custom"` — a fully custom, non-GSA study area: a bounding box
    (`CUSTOM_BBOX`, e.g. `WESTMED_BBOX`), an arbitrary shapefile
    (`CUSTOM_SHAPEFILE_PATH`), or a custom grouping of GSAs
    (`CUSTOM_GSA_IDS`), set via `CUSTOM_AREA_TYPE`. Useful for a bay, an
    MPA, or a footprint spanning parts of several GSAs that don't line up
    with GSA boundaries. See [`AREA_MODE == "custom"`] below for the full
    mechanism.
* **Time window:**
  - `YEAR_ECOPATH <- 1994:1996` — the Ecopath baseline snapshot years (the
    single 3-year average that `Ecopath_B`, PB, QB, Landings and Discards
    are all computed against).
  - `TS_YEARS`/`START_YEAR`/`END_YEAR` <- `1995:2023` — the multi-year
    Ecosim time series window.
  - Both are ordinary variables at the top of each script, guarded so they
    can be overridden from a calling/driver script without editing the
    script itself — see [Adapting this pipeline to a different region]
    below.
* **Run configuration — where code, data and outputs live:**
  - `git_dir` — the local clone of the WestMed EwE GitHub repository
    (code only, no raw data).
  - `pcloud_dir` — the local mount/sync of the WestMed EwE pCloud drive
    (all raw survey, catch and reference data; never code).
  - `out_dir` — a local folder every script writes its CSVs, plots and the
    shared workbook to; not synced back to pCloud or GitHub by the scripts
    themselves.
  - All three are plain local paths resolved at the top of each script —
    hardcoded for one user, or picked interactively via
    `rstudioapi::selectDirectory()` for anyone else — never fixed
    locations baked into the pipeline's logic.

See [Data and code locations] just below for more on how code/data/output
separation works, and [Adapting this pipeline to a different region] for
every other knob (`TARGET_COUNTRIES`, `DATASET_VERSION`, etc.) a different
region or time window would need.

# Data and code locations

* **Code** — every script and library file documented here lives in the
  **WestMed EwE GitHub repository** (`WMed_EwE`, cloned locally to what the
  scripts call `git_dir`). All shared-library `source()` calls
  (`lib_survey_fg_density_functions.R`, `lib_worms_taxonomy_lookup.R`,
  `lib_fishbase_diet_matrix.R`, `lib_aquamaps_depth_extension.R`,
  `lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R`)
  resolve against this repository's
  `scripts/` folder.
* **Input data** — every raw survey, catch and reference file each script
  reads (`FG_WMed_2026.csv`, `TM_list_(April_2019).xlsx`, MEDITS `TA.csv`/
  `TB.csv`, MEDIAS `abundance.csv`/`biomass.csv`, `ASFIS_sp_2026.1.csv`,
  the catchability CSV, GFCM capture/reference CSVs) is read from the
  **WestMed EwE pCloud drive** (`EwE Western Med 2026`, mounted/synced
  locally to what the scripts call `pcloud_dir`), **not** from the GitHub
  repository — the repository holds code only, no raw data.
* **Outputs** — every CSV, plot, and the shared `ecopath_ecosim_inputs.xlsx`
  workbook are written to a local `out_dir` (not synced back to pCloud or
  GitHub by the scripts themselves).

Throughout this document, `<pcloud_dir>` means *the WestMed EwE pCloud
drive* and `<git_dir>` means *the WestMed EwE GitHub repository*. Retargeting
the pipeline to a different region (see [Adapting this pipeline to a
different region]) does not require moving off pCloud/GitHub — only
pointing `pcloud_dir`/`git_dir` at that region's own drive/repository, or
adding that region's data alongside the Western Med data in the same
drive/repository.

**A note on the plots and column tables in this document:** this
documentation was written without direct access to the WestMed EwE pCloud
drive's actual data files. The **column-reference tables** throughout are
not guesses — they list exactly the column names each script's own code
reads (`select()`/`merge()`/`setnames()` calls, etc.), so they're accurate
regardless of the actual data values. The **figures**, however, are the
pipeline's own real output and only appear once you've generated them — see
[Populating the real figures] below.

## Populating the real figures

Several sections below embed a figure via a small `safe_include_graphics()`
helper (defined in this document's setup chunk). Each one checks for a file
under `figures/` (a folder that must sit next to this `.Rmd`) and shows the
real image if present, or a plain "not yet available" message with the
exact script/output path to copy from if not — so this document always
knits cleanly either way. To populate them for real:

1. Run the relevant script (named in each placeholder message) against the
   real WestMed EwE pCloud drive data.
2. Copy the named `.png` from that script's `<out_dir>/plots/` into a
   `figures/` folder next to this `.Rmd`, keeping the same filename.
3. Re-knit.

Four **other** diagrams in this document (the pipeline flow, the
FG-matching cascade, the workbook sheet assembly, and the PB/QB taxon
dispatch diagrams) are a different kind of figure: **Mermaid flowcharts**,
in the same style as `medbs_pipeline_methodology.qmd`'s own diagrams
(same `flowchart LR` layout, the same gray/purple/teal/coral `classDef`
palette). They show the code's own logic — which function calls which,
which sheet gets written by what, which fallback branch runs when — built
directly from the script and library contents described here, not from
any survey or model output. Being plain text (Mermaid syntax rendered via
`mermaid.js` from a CDN), they always render, with or without a connection
to the pCloud drive — unlike the `safe_include_graphics()` placeholders
above. They do need a connection *to the Mermaid CDN* at view time to draw
the SVG in the browser; `rmarkdown::html_document(self_contained = TRUE)`
inlines the `<script>` tags instead if that CDN is unreachable from
wherever the HTML is opened.

# Pipeline overview

The diagram shows which script produces which intermediate file and which workbook sheets each one writes. It is a schematic of the code, not a plot of data.

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["Step 1&lt;br/&gt;01_biomass.R&lt;br/&gt;(sources 01b_biomass_unsurveyed.R, 03b_ecobase.R)"] --> B1["output/biomass/&lt;br/&gt;species_density_regional_combined.csv,&lt;br/&gt;strata_area_by_area.csv, biomass_proportion_by_species_fg.csv"]
    A1 --> C1["xlsx: Ecopath_B, FG_spp, Ecosim_ts (B)"]
    B1 --> A2["Step 2&lt;br/&gt;02_fisheries.R&lt;br/&gt;(sources 02b_sau_extraction.R)"]
    C1 --> A2
    A2 --> C2["xlsx: Ecopath_L, Ecopath_Di, Ecosim_ts (C, effort)&lt;br/&gt;output/fisheries/: F_by_fg.csv, F_by_species_fg.csv"]
    B1 --> A3["Step 3&lt;br/&gt;03_pbqb-traits.R&lt;br/&gt;(sources 03b_ecobase.R)"]
    C2 --> A3
    A3 --> C3["xlsx: Ecopath_PBQB, Ecopath_traits, Ecobase"]
    B1 --> A4["Step 4&lt;br/&gt;04_diets.R&lt;br/&gt;(sources lib_fishbase_diet_matrix.R)"]
    A4 --> C4["xlsx: Ecopath_diet, FG_References&lt;br/&gt;output/diet/"]
    C1 --> A5["Step 5&lt;br/&gt;05_validation.R"]
    C2 --> A5
    C3 --> A5
    C4 --> A5
    A5 --> O["output/validation/, plots/validation/validation_plots_ALL.pdf"]

    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class A1,A2,A3,A4,A5 purple
    class B1 gray
    class C1,C2,C3,C4 teal
    class O coral
</div>
```

| Step | Script | What it writes | Needs first |
|---|---|---|---|
| 1 | `01_biomass.R` (+ `01b_biomass_unsurveyed.R`, `lib_aquamaps_depth_extension.R`, `lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R`) | `Ecopath_B`, `FG_spp`, B part of `Ecosim_ts`; `output/biomass/` | — |
| 2 | `02_fisheries.R` (+ `02b_sau_extraction.R`) | `Ecopath_L`, `Ecopath_Di`, catch and effort part of `Ecosim_ts`; `F_by_fg.csv`; `output/fisheries/` | 1 (`strata_area_by_area.csv`, `species_density_regional_combined.csv`, `Ecopath_B`) |
| 3 | `03_pbqb-traits.R` (+ `03b_ecobase.R`) | `Ecopath_PBQB`, `Ecopath_traits`, `Ecobase`; `output/pbqb-traits/` | 1, 2 (`F_by_species_fg.csv`, `F_by_fg.csv`) |
| 4 | `04_diets.R` (+ `lib_fishbase_diet_matrix.R`, `lib_worms_taxonomy_lookup.R`) | `Ecopath_diet`, `FG_References`; `output/diet/` | 1 (`biomass_proportion_by_species_fg.csv`, `Ecopath_B`) |
| 5 | `05_validation.R` | `output/validation/`, `plots/validation/` | 1–4 |

Every t/km² in every sheet uses the same model area, $A_{model}$ = sum of `strata_area_by_area.csv` = 175,660 km² (MEDITS 10–800 m strata of GSAs 1, 2, 5–11). Each script writes a time-stamped log, `output/YYYYMMDD_HHMMSS_<script>_log.txt`.

Library files (no driver code): `lib_survey_fg_density_functions.R` (survey density, workbook writing, `safe_fwrite()`/`csv_has_nul()` for the pCloud drive, `compute_stanza_catch_split_from_length()`), `lib_worms_taxonomy_lookup.R`, `lib_fishbase_diet_matrix.R`, `lib_aquamaps_depth_extension.R`, `lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R`. Helper scripts: `combine_STAR_RAMlegacy.R` (builds `combined_medbs_star_ramlegacy.csv`), `export_gsa_boundaries.R`; standalone exploration scripts not called by the pipeline: `posidonia.R`, `coral_gorgonian.R`, `phyto.R`, `emodnet_benthic_habitat_biomass_extraction.R`.

---

# Detailed methodology per output block — B, C, L, Di, M, TRAITS, F, PB, QB, DIET

Full equations with references are in `medbs_pipeline_methodology.qmd` (sections Biomass, Fisheries, PB/QB, Diet, Validation) and in the `NOTATION, EQUATIONS AND DATA SOURCES` header at the top of each script. This section summarises each block and where its code lives.

| Block | What it is | Primary source(s) | Key caveat |
|---|---|---|---|
| **B** — Biomass | `Ecopath_B` (1994–1996 mean, t/km²) and the B series in `Ecosim_ts` | MEDITS / q, MEDIAS (+ per-GSA small-pelagic series), stock assessments (hake stanzas, BFT, SWO), literature × habitat maps, ISIMIP3a COBALT plankton | groups the surveys cannot sample come from cited literature and models, each with its own assumptions (see `01b`) |
| **C** — Catches | FG × year catch of the target countries | GFCM capture production, STECF FDI, Sea Around Us, FishMIP discard ratios, ICCAT Task I | only the coastal states of the model GSAs (Spain, France, Italy); ICCAT scaled to the model GSAs by FDI landings share |
| **L** — Landings | `Ecopath_L`: FG × fleet | C split by FDI / SAU fleet shares, ICCAT gear field for BFT/SWO/ALB | unresolved fleet → "Target countries - fleet not resolved" column |
| **Di** — Discards | `Ecopath_Di`: FG × fleet | FDI per-fleet discard ratio, else FishMIP | no source → discards 0 |
| **M** — Natural mortality | fish $M$ inside $P/B = M + F$ | FishBase/SeaLifeBase, TropFishR, FishLife | priority Then > Pauly > Hoenig > Alverson & Carney > FishLife > Gascuel |
| **F** — Fishing mortality | $F_f = (\bar C_f/A_{model})/B_f$, 1994–1996 | C (02) and the final `Ecopath_B` (01) | one region-wide, all-gear rate per FG |
| **PB** | fish $M + F$; mammals Barlow & Boveng / Gascuel; seabirds Gascuel; invertebrates Tumbiolo & Downing / Brey / Gascuel, $+F_f$ for fished FGs | 03 | invertebrate empirical P/B treated as $M$ (ASSUMPTION) |
| **QB** | Palomares & Pauly / Christensen & Pauly (fish); Innes/Trites (mammals); Nilsson & Nilsson (seabirds); Q/P = 3 (invertebrates) | 03 | fish diet dummies fixed |
| **TRAITS** | `Ecopath_traits` | FishBase/SeaLifeBase | unresolved taxonomy → generic bucket, flagged |
| **DIET** | `Ecopath_diet` | metaweb (empty template) → MAISHA → FishBase/SeaLifeBase → EcoBase | 3 FGs still have no diet |

## B — Biomass

### 1. What it is
The 1994–1996 (`YEAR_ECOPATH`) mean biomass density per FG in t/km² of $A_{model}$ (`Ecopath_B`), and the yearly B series in `Ecosim_ts` (relative to 1995).

### 2. Inputs / data sources
- MEDITS bottom-trawl survey (TA/TB/TC) and MEDIAS acoustic survey, processed with `lib_survey_fg_density_functions.R`.
- Catchability table `catchability_factors_ecotrans_medits_2021_spp.csv` (ECOTRANS).
- `combined_medbs_star_ramlegacy.csv` (GFCM STAR + RAM Legacy v4.44; built by `combine_STAR_RAMlegacy.R`): small-pelagic trends, hake SSB.
- `01b_biomass_unsurveyed.R` reference tables on the pCloud drive: `marine_megafauna_biomass.csv`, `literature_density_biomass.csv`, `benthic_habitat_reference_tables/habitat_shapefile_biomass.csv`, `medits_reference_tables/density_unit_to_wet_g_m2.csv`; EUSeaMap 2025 geodatabase (cached area: `euseamap2025_seagrass_area_by_gsa_ge10m.csv`); ISIMIP3a GFDL-MOM6-COBALT2 obsclim (cached `isimip3a_cobalt2_plankton_detritus_timeseries.csv`).
- `FG_WMed_2026.csv` (species → FG and taxonomy).

### 3. Computation method
1. Haul density → catchability → outlier screen → stratified mean per GSA → model-area mean → FG sum (equations 1–7 in the 01 header).
2. MEDIAS shelf density rescaled to $A_{model}$; sardine and anchovy from the per-GSA acoustic series with stock-assessment-trend gap filling (`small_pelagics_per_gsa_series_REVIEW.csv`).
3. Priority per FG × year: single-species stock assessment (FGs the surveys don't measure) > MEDIAS > MEDITS; survey-exempt FGs take only the `01b` value (never survey bycatch).
4. Hake stanzas from the assessment (`HAKE_STOCK_GSAS`): $B_{adult} \approx SSB$, $B_{juv} = B_{adult}(1-r)/r$, $r = 0.42$, back-cast with the MEDITS index (adult 0.0746, juvenile 0.103 t/km²).
5. `01b_biomass_unsurveyed.R` (see its own section below) adds megafauna, BFT/SWO, benthic habitats, plankton, suprabenthos and detritus.
6. `Ecopath_B` = mean over `YEAR_ECOPATH`; `FG_spp` = species share of FG biomass (`prop_sp_fg`); EcoBase biomass is evaluated only (`ecobase_biomass_evaluation_NOT_INCORPORATED.csv`).

### 4. Known caveats / limitations
- 5 FGs have no Ecopath_B (3 Expanding FGs, Salps — left for Ecopath — and Discards).
- 123 species in 7 FGs use an even within-FG split.
- `FG_WMed_2026.csv` has no `survey_exempt` column; the hard-coded `EXEMPT_FG_NAMES` list is used (log message).
- Assumptions in the unsurveyed groups are listed in the `01b` section and in each row's `Source_citation`.

### 5. Code location
`01_biomass.R` (survey cascade, small pelagics, hake stanza block, priority rule, `Ecopath_B`/`FG_spp`/`Ecosim_ts`); `01b_biomass_unsurveyed.R` (`load_manual_cited_biomass_group()`, `extrapolate_density_row()`, `estimate_habitat_area_km2()`, `compute_euseamap_habitat_area()`, ISIMIP/CMEMS plankton block); `lib_survey_fg_density_functions.R`.

### 6. References
See the 01 and 01b headers; main ones: MEDITS Working Group 2017; Leonori et al. 2021; Ricard et al. 2012; Cochran 1977; Pante & Simon-Bouhet 2013; Stock et al. 2020; Bernardeau-Esteller et al. 2023; Coppari et al. 2019; Ballesteros 2006; Colléter et al. 2015.

## DIET — Ecopath_diet sheet / diet matrix

### 1. What it is
Predator FG × prey FG diet composition $DC_{f,j}$ (each predator column sums to 1), written to `Ecopath_diet` and to `diet_composition_ewe.csv` (EwE import format).

### 2. Inputs / data sources
- Mediterranean trophic metaweb `data_entry_metaweb_empty.xlsx` (still an empty template).
- MAISHA v1.2 (Vascotto et al. 2026, doi:10.13120/hwpn1e, CC-BY 4.0), Darwin Core Archive at `<pcloud_dir>/data/diet/MAISHA/MAISHA_DwC-A` (or `.zip`, or any `maisha*.xlsx/csv` under `data/diet`; `MAISHA_PATH`).
- FishBase and SeaLifeBase diet tables via `rfishbase` (cached in `output/diet`).
- `diet_reference_tables/fishbase_food_category_to_fg.csv` (food category → FG rules), `diet_metric_priority.csv`.
- `biomass_proportion_by_species_fg.csv` and `Ecopath_B` from 01; `FG_WMed_2026.csv`.

### 3. Computation method
1. Tier per predator FG: metaweb → MAISHA (per species with records) → FishBase/SeaLifeBase → EcoBase (`DIET_FALLBACK_ENABLE_MAISHA`, `DIET_FALLBACK_ENABLE_FISHBASE`, `DIET_FALLBACK_ENABLE_ECOBASE`; `DIET_FISHBASE_METHOD = "fg_matrix"`).
2. Study proportions $p_{s,d,k} = v_k/\sum_k v_k$ with $v$ = first available of IRI, %W, %N, %FO (`DIET_METRIC_PRIORITY`); Mediterranean studies only when a species has any.
3. Prey → FG: species → genus/family/order/class/phylum (equal split, `DIET_TAXON_SPLIT`) → food-category regex (split by Ecopath B, `DIET_CATEGORY_SPLIT`) → WoRMS second pass; eggs/larvae → zooplankton; detritus dropped for predators with TL ≥ 3 (`DIET_DETRITUS_MAX_TL`); non-food items dropped and the study renormalised.
4. $DC_{s,j}$ = mean over studies; $DC_{f,j} = \sum_s b_{s|f} DC_{s,j} / \sum_s b_{s|f}$; columns renormalised to 1.
5. Checks: column sums, cannibalism > 0.1.

### 4. Known caveats / limitations
- The metaweb has no rows yet, so every diet comes from MAISHA, FishBase/SeaLifeBase or nothing.
- Category shares split by prey biomass is an assumption (prey eaten in proportion to availability).
- EcoBase's web service exposes no diet matrices, so tier 4 finds nothing.
- No diet for corals and gorgonians, meso and micro zooplankton, suprabenthos (2026-10-07 run).

### 5. Code location
`04_diets.R` (config, tiers, `build_fg_diet()`, export, checks); `lib_fishbase_diet_matrix.R` (`maisha_diet_records()`, `fb_fetch_diet_tables()`, `fishbase_diet_records()`, `diet_resolve_prey()`, `diet_records_to_fg_matrix()`, `write_diet_source_outputs()`); `lib_worms_taxonomy_lookup.R`.

### 6. References
Vascotto et al. 2026 (MAISHA); Froese & Pauly (FishBase); Palomares & Pauly (SeaLifeBase); Boettiger, Lang & Wainwright 2012 (rfishbase); Hyslop 1980; Pinkas, Oliphant & Iverson 1971 (IRI); Christensen, Walters & Pauly 2005; Colléter et al. 2015.


---


## References

Alverson, D.L., & Carney, M.J. (1975). A graphic review of the growth and decay of population cohorts. *Journal of the Conseil International pour l'Exploration de la Mer*, 36(2), 133–143.

Barlow, J., & Boveng, P. (1991). Modeling age-specific mortality for marine mammal populations. *Marine Mammal Science*, 7(1), 50–65.

Brey, T. (1999). Growth performance and mortality in aquatic macrobenthic invertebrates. *Advances in Marine Biology*, 35, 153–223.

Brey, T. (2012). *Population dynamics in benthic invertebrates. A virtual handbook.* Alfred Wegener Institute for Polar and Marine Research.

Christensen, V., & Pauly, D. (1992). ECOPATH II — a software for balancing steady-state ecosystem models. *Ecological Modelling*, 61(3–4), 169–185.

Christensen, V., Walters, C.J., & Pauly, D. (2008). *Ecopath with Ecosim: a User's Guide*. Fisheries Centre, University of British Columbia.

Colléter, M., Palomares, M.L.D., Christensen, V., Coll, M., Steenbeek, J., Falautano, M., Sagarese, S., & Pauly, D. (2015). EcoBase: a repository solution to gather and communicate information from EwE models. *Ecological Modelling*, 302, 42–48.

Froese, R., & Pauly, D. (Eds.). FishBase. World Wide Web electronic publication. www.fishbase.org.

Gascuel, D., Morissette, L., Palomares, M.L.D., & Christensen, V. (2008). Trophic flow kinetics in marine ecosystems: toward a theoretical approach to ecosystem functioning. *Ecological Modelling*, 217(1–2), 33–47.

GFCM (General Fisheries Commission for the Mediterranean). MEDBS (Mediterranean and Black Sea) Working Group stock-assessment methodology; Data Collection Reference Framework (DCRF).

Hoenig, J.M. (1983). Empirical use of longevity data to estimate mortality rates. *Fishery Bulletin*, 82(1), 898–903.

Innes, S., Lavigne, D.M., Earle, W.M., & Kovacs, K.M. (1987). Feeding rates of seals and whales. *Journal of Animal Ecology*, 56(1), 115–130.

Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S.O., Rees, T., & Froese, R. AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org.

Mildenberger, T.K., Taylor, M.H., & Wolff, M. (2017). TropFishR: an R package for fisheries analysis with length-frequency data. *Methods in Ecology and Evolution*, 8(11), 1520–1527.

Nilsson, S.G., & Nilsson, I.N. (1976). Number, food consumption, and fish predation by birds in Lake Möckeln, Sweden. *Ornis Scandinavica*, 7(1), 61–70.

Palomares, M.L.D., & Bailly, N. (Eds.). SeaLifeBase. World Wide Web electronic publication. www.sealifebase.org.

Palomares, M.L.D., & Pauly, D. (1998). Predicting food consumption of fish populations as functions of mortality, food type, morphometrics, temperature and salinity. *Marine and Freshwater Research*, 49(5), 447–453.

Pauly, D. (1980). On the interrelationships between natural mortality, growth parameters, and mean environmental temperature in 175 fish stocks. *ICES Journal of Marine Science*, 39(2), 175–192.

Pauly, D., & Zeller, D. (Eds.) (2016). *Global Atlas of Marine Fisheries: A Critical Appraisal of Catches and Ecosystem Impacts.* Island Press. [Sea Around Us]

Pinkas, L., Oliphant, M.S., & Iverson, I.L.K. (1971). Food habits of albacore, bluefin tuna, and bonito in California waters. *California Department of Fish and Game, Fish Bulletin*, 152.

STECF (Scientific, Technical and Economic Committee for Fisheries). Fisheries Dependent Information (FDI) database.

Then, A.Y., Hoenig, J.M., Hall, N.G., & Hewitt, D.A. (2015). Evaluating the predictive performance of empirical estimators of natural mortality rate using information on over 200 fish species. *ICES Journal of Marine Science*, 72(1), 82–92.

Thorson, J.T., Munch, S.B., Cope, J.M., & Gao, J. (2017). Predicting life history parameters for all fishes worldwide. *Ecological Applications*, 27(8), 2262–2276.

Tumbiolo, M.L., & Downing, J.A. (1994). An empirical model for the prediction of secondary production in marine benthic invertebrate populations. *Marine Ecology Progress Series*, 114, 165–174.

WoRMS Editorial Board. World Register of Marine Species. www.marinespecies.org.

ICCAT (International Commission for the Conservation of Atlantic Tunas). GBYP (Grand Bluefin Tuna Year Program) aerial survey, CREEM Phase 11 report (2021).

---

# Run order and dependencies

| Step | Script | Depends on | Must run before |
|---|---|---|---|
| 1 | `01_biomass.R` (`AREA_MODE` picks westmed vs. custom; sources `01b_biomass_unsurveyed.R`) | nothing | 2, 3, 4 |
| 2 | `02_fisheries.R` | 1: `strata_area_by_area.csv`, `species_density_regional_combined.csv`, `Ecopath_B` sheet | 3 |
| 3 | `03_pbqb-traits.R` (sources `03b_ecobase.R`) | 1: `FG_spp`, biomass proportions; 2: `F_by_species_fg.csv`, `F_by_fg.csv` | 5 |
| 4 | `04_diets.R` | 1: `biomass_proportion_by_species_fg.csv`, `Ecopath_B` | 5 |
| 5 | `05_validation.R` | 1–4 | — |

After a change in 01, rerun 02 → 03 → 04 → 05, because $F$ (02) uses `Ecopath_B` and P/B (03) uses $F$. Each script calls `finalize_workbook_sheet_order()` at the end, so the workbook is always in its final layout.

---

# Script 1a — `01_biomass.R`

**Role:** Pipeline Step 1 of 5 — a single script for **any region in the
Mediterranean**, GSA-based or fully custom, picked by the `AREA_MODE`
config variable (`"westmed"`, the default, or `"custom"` — see
[`AREA_MODE == "custom"`] below). No dependency on the other numbered
scripts; `02_fisheries.R`, `03_pbqb-traits.R` and `04_diets.R` all require this step's
output to exist first, regardless of which `AREA_MODE` produced it.

Runs two independent surveys over the same GSAs/FG scheme (MEDITS bottom
trawl + MEDIAS acoustic) and combines both into one set of Ecopath/Ecosim
sheets, rather than producing separate MEDIAS sheets. As delivered it's
configured for the Western Med (GSA 1–11, via `FILTER_AREAS <- 1:11`) as a
worked example; retargeting it to a different GSA subset elsewhere in the
Mediterranean — the Adriatic, the Aegean, the whole basin — is a matter of
changing that one line and the region-specific reference tables listed in
[Adapting this pipeline to a different region].

## Inputs

| File | Format | Role |
|---|---|---|
| `<pcloud_dir>/data/FG_WMed_2026.csv` | csv (`;`) | Species → Functional Group reference with taxonomy (`dataframe2`: `species`→`ScientificName`, `FG_number`→`FG_num`, `FG_name`; `Genus`…`Phylum`, `source`, `status`). The old `FG_WMed.xlsx` is retired. |
| `<pcloud_dir>/.../TM_list_(April_2019).xlsx` | xlsx | MEDITS species-code → scientific-name lookup |
| `<pcloud_dir>/.../Demersal/TA.csv` | csv | MEDITS haul/sample metadata (position, depth, gear, swept-area inputs) |
| `<pcloud_dir>/.../Demersal/TB.csv` | csv | MEDITS catch-per-haul data |
| `<pcloud_dir>/.../Acoustic/abundance.csv`, `biomass.csv` | csv | MEDIAS acoustic-survey raw records |
| `<pcloud_dir>/data/ASFIS_sp_2026.1.csv` | csv | FAO ASFIS species code list (MEDIAS species-code → scientific name) |
| `<pcloud_dir>/data/catchability_factors_ecotrans_medits_2021_spp.csv` | csv (optional) | Per-species catchability correction factors (q); skipped with a message if absent |
| GFCM GSA shapefile | auto-downloaded zip from `gfcmsitestorage.blob.core.windows.net/website/5.Data/ArcGIS/GFCM_GSA.zip` | Study-area polygons (`area_shp`); `SMU_CODE` 111/112 (W/E Sardinia) merged into GSA 11 |
| NOAA bathymetry | auto-downloaded via `marmap::getNOAA.bathy()` inside `compute_strata_area_by_area()` | Depth-stratum area calculation |

`FILTER_AREAS <- 1:11` (this example's Western Med GSAs — set to any GSA
number or set of GSA numbers, anywhere in the GFCM area, to retarget the
whole script to a different region), `YEAR_ECOPATH <- 1994:1996`,
`TS_YEARS <- 1995:2023`, `STRATA <- TRUE`, `DROP_OUTLIERS <- FALSE`,
`NORMALIZE_TS <- TRUE`, `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- FALSE` (opt-in —
see Step 7b below).

### Step 7b — AquaMaps shallow/deep depth adjustment (opt-in)

MEDITS' own 5 standard strata (`MEDITS_STRATA`) only cover 10–800 m.
`compute_strata_area_by_area()`'s area-proportion weighting normalizes
across just those 5 strata, so as delivered, this pipeline's Biomass/
density figures implicitly treat the whole modeled area as if it were
entirely 10–800 m seafloor — a species with real biomass shallower than
10 m or deeper than 800 m is invisible to these numbers, not just
imprecisely measured. Setting `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- TRUE`
(default `FALSE`) adds genuine **derived** pseudo-strata: one shallow
(0–10 m) plus, by default, the deep zone split into **two** sub-bands —
800–1000 m and 1000–2850 m — rather than one lumped 800–6000 m stratum
(2850 m is a practical ceiling for the western Mediterranean's own real
depths specifically, not the Mediterranean-wide ~5267 m Calypso Deep,
which sits in the eastern/Ionian basin outside these GSAs). Each
pseudo-stratum gets its own real bathymetric area (the same
`compute_strata_area_by_area()` machinery already used for the 5 real
strata) and a density **derived, not measured**, anchored to the ONE
real MEDITS stratum immediately adjacent to it — the shallowest real
stratum (10–50 m) for the shallow pseudo-band, the deepest real stratum
(500–800 m) for every deep pseudo-band — rather than an area-weighted
average across all 5 real strata. Each pseudo-stratum's density equals
that adjacent stratum's own **measured** density (`compute_species_
anchor_stratum_density()`), scaled by the ratio of AquaMaps mean
suitability between the pseudo-band and that same adjacent stratum's own
depth range. A species at equal AquaMaps suitability in both bands gets
a multiplier of exactly 1 — its pseudo-stratum density literally equals
what MEDITS measured next door, with no distortion from averaging across
strata the species may barely occupy. (An earlier version of this
adjustment compared each pseudo-band's suitability against the whole
10–800 m composite range instead — that systematically understated
narrow pseudo-bands relative to the ~790 m-wide reference purely from
the width mismatch, independent of actual suitability level; the
anchor-stratum design replaces it.) The trade-off: a species with real
catch elsewhere in 10–800 m but none specifically in the one adjacent
anchor stratum gets no pseudo-stratum density at all for that area
(flagged in the audit sheet), rather than being extrapolated from
unrelated, non-adjacent strata. These pseudo-strata are folded into the
exact same `weight_by_strata()`/`weight_species_by_area()` calls used
for the 5 real strata — not a separate formula — so `fg_index`/
`fg_index_regional`/`species_density_regional` end up as one consistent
area-weighted sum across all 8 strata (5 real + 3 derived), real and
derived alike. The number and boundaries of deep sub-bands are
controlled by `apply_aquamaps_depth_adjustment()`'s `deep_ranges`
argument — a list of `c(from, to)` pairs — so this can be collapsed back
to one lump (`list(c(800, 6000))`) or split further without changing
anything else; all deep sub-bands anchor to the same deepest real
stratum, they don't chain off each other.

**Core assumption, stated plainly rather than absorbed silently**: a
species' relative density across depth is assumed proportional to
AquaMaps' relative occurrence-suitability across depth (its trapezoidal
`DepthMin`/`DepthPrefMin`/`DepthPrefMax`/`DepthMax` envelope, per
AquaMaps' own "Algorithm and Data Sources" documentation). Occurrence
suitability is not the same thing as density — this is a standard
extrapolation approach when direct sampling doesn't cover a species' full
depth range, but an approximation nonetheless. Every resulting multiplier
is written to the `AquaMaps_Depth_Adjustment` sheet (species, its
`DepthMin`/`DepthPrefMin`/`DepthPrefMax`/`DepthMax`, probability masses,
resulting `multiplier_shallow`/`multiplier_deep_800_1000`/
`multiplier_deep_1000_2850` (one column per configured deep sub-band,
named by its own depth bounds), and a `note` explaining species that got
NO adjustment for a given band — unmatched to AquaMaps, no depth
envelope in `am_hspen()`, or essentially zero AquaMaps suitability in
that band's own adjacent real anchor stratum, making a ratio against
that unstable) — reviewable per species, not a black box.

Requires the `aquamapsdata` R package (github.com/raquamaps/aquamapsdata),
the only verified way to get AquaMaps' per-species HSPEN depth-envelope
table without scraping — its `download_db()` pulls a real, deliberate
**~2GB (~10GB unpacked) local SQLite database once** (`ensure_aquamaps_db()`
uses `force = FALSE`, so this never re-downloads on subsequent runs).
Implemented in `lib_aquamaps_depth_extension.R` (species → AquaMaps
`SpeciesID` resolution, envelope fetch, envelope math, pseudo-strata
construction) — see that file's own header for the full mechanism.
Requires `STRATA <- TRUE` (errors clearly if not — there's no unstratified
equivalent). Available under **either** `AREA_MODE` — the adjustment
operates on species depth envelopes and strata generically, with no
GSA-specific assumptions, so it applies the same way to a custom region.

Two outputs make the strata (real + derived) inspectable directly,
independent of `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT` — they cover whichever
strata are actually in play, 5 or 8:

- **`survey_fg_depth_strata_profile.png`** (Step 8, `plot_strata_profile()`)
  — a bar chart of raw per-sample density by stratum, faceted per FG, for
  sanity-checking that known deep-water FGs actually peak in the deep
  strata and vice versa. Reads `weight_by_strata()`'s `per_stratum_raw`
  attribute specifically (not `per_stratum`), so a stratum with real
  sample data but no computed area-proportion for a given AreaID still
  shows up here rather than silently vanishing.
- **`FG_Density_by_Stratum`** sheet (Step 12c, `build_fg_density_by_stratum_sheet()`)
  — the same information as an Excel table: one row per FG, one column
  per stratum (region-wide, area-weighted, `YEAR_ECOPATH`-averaged), plus
  a `Total_Density` column that reproduces `fg_index_regional`'s own
  `mean_density` exactly, since each stratum column is that stratum's own
  additive contribution to the total, not an independently-normalized
  figure.

### Column reference — key input files

Columns as actually referenced by the code (from the `<pcloud_dir>` files
above). Not every column in the real file is necessarily listed — only the
ones the script reads.

**`FG_WMed_2026.csv`** (→ `dataframe2`, the species→FG master; semicolon-separated):

| Column | Renamed to | Meaning |
|---|---|---|
| `species` | `ScientificName` | Species scientific name |
| `FG_number` | `FG_num` | Functional Group number |
| `FG_name` | `FG_name` | Functional Group name |
| `Genus`, `Family`, `Order`, `Class`, `Phylum` | same | taxonomy, used by the taxonomy fallback and the diet prey matching |
| `source`, `status` | same | where the row came from (MEDITS, MEDIAS, others) and whether it was reviewed |

Life-history traits (`traits_ewe`) are no longer read here; `03_pbqb-traits.R` builds `Ecopath_traits` from FishBase/SeaLifeBase.

**`TM_list_(April_2019).xlsx`, sheet 1 (skip first 2 rows)** — the code
locates its two needed columns by regex rather than a fixed name, since the
exact header text has drifted between file versions: a column matching
`/MEDITS/i` (the MEDITS species code) and a column matching `/Scientific/i`
(the scientific name).

**MEDITS `TA.csv`** (haul/sample metadata):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID` (concatenated, `_`-separated) |
| `distance` | Trawl distance towed → swept area |
| `wing_opening` | Trawl wing opening (decimetres) → swept area |
| `shooting_depth`, `hauling_depth` | Averaged into `mean_depth` |
| `shooting_latitude`, `shooting_longitude` | DDMM.mmm format — converted to decimal degrees |
| `shooting_quadrant` | Present but **not used** — flagged in the code as an unresolved caveat around hemisphere sign |

**MEDITS `TB.csv`** (catch-per-haul data):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID`, joined against `TA.csv` |
| `genus`, `species` | Concatenated into `species_code`, joined against `TM_list` |
| `ptot` | Total catch weight (grams; negative values treated as `NA`, converted to tonnes) |

**MEDIAS `abundance.csv` / `biomass.csv`** (acoustic survey raw records):

| Column | Used for |
|---|---|
| `country`, `year`, `species` | Grouping key for aggregation |
| `area` | Text field containing a GSA number, extracted via regex (`\d+`) |
| `sex` | Filtered to `"C"` (combined sexes) when present |
| `lengthclass*` (many columns) | Summed row-wise into a single `total_value` (negative values treated as missing) |

**`ASFIS_sp_2026.1.csv`** (FAO species reference list):

| Column | Used for |
|---|---|
| `Alpha3_Code` | MEDIAS species code → joined against `abundance`/`biomass` |
| `Scientific_Name` | Resolved scientific name |
| *(also present: `ISSCAAP_Group`, `Taxonomic_Code`, `English_name`, …)* | Not read by this pipeline |

**Catchability CSV** (`catchability_factors_ecotrans_medits_2021_spp.csv`,
optional):

| Column | Renamed to | Meaning |
|---|---|---|
| `species` | `ScientificName` | Species scientific name |
| `q_FACTOR` | `q` | Catchability correction factor |

```{r sample-coverage-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_sample_map() output — MEDITS sample coverage by area."}
safe_include_graphics(
  "figures/survey_sample_coverage_map.png",
  script = "01_biomass.R",
  out_dir_note = "<out_dir>/plots/survey_sample_coverage_map.png"
)
```

## Processing steps

1. **Config** — resolve `out_dir`/`pcloud_dir`/`git_dir`, download/cache the GSA
   shapefile, source `lib_survey_fg_density_functions.R` and
   `lib_worms_taxonomy_lookup.R`.
2. **Load MEDITS data** — read `TA.csv`/`TB.csv`, compute swept area
   (`distance × wing_opening_m / 1e6`, wing opening in decimetres), convert
   DDMM.mmm lat/lon to decimal degrees, fix the Alboran Sea (GSA 1–3)
   longitude sign — a Western-Med-specific correction, harmless (a no-op) if
   your `FILTER_AREAS` doesn't include GSA 1–3 — reshape into the
   standardized `dataframe1`/`dataframe2` format. Runs `validate_survey_data()`
   and `plot_sample_map()` (→ `survey_sample_coverage_map.png`,
   `survey_sample_coverage_map_by_year.png`).
3. **FG matching (Steps 3–4)** — `prepare_fg_lookup()` →
   `match_species_to_fg()` → `match_nominate_subspecies_fg()` → manual
   overrides (species codes absent from the MEDITS taxonomic list, e.g.
   `ARGRACU`→*Argyropelecus aculeatus*) → `fallback_match_fg_by_taxonomy()`
   (Genus→Family→Order→Class→Phylum) → `apply_seed_fg_rules()` (FGs with zero
   reference species, e.g. Holothuroidea→"Sea cucumbers") →
   `summarize_unresolved_species()` → `survey_unmatched_for_manual_review.csv`.
4. **Densities (Step 5)** — `match_samples_to_area()`,
   `assign_depth_stratum()`, `compute_sample_densities()`, restrict to
   `FILTER_AREAS`, `apply_catchability_correction()` (if the catchability CSV
   exists), `remove_sample_outliers()` (→ `survey_outliers_flagged.csv` if
   any flagged), `compute_fg_densities_by_stratum()`.
5. **Strata/area weighting (Steps 6–7)** — `compute_strata_area_by_area()`
   (cached to `strata_area_by_area.csv`), `weight_by_strata()`,
   `weight_by_area()`, `compute_species_densities_by_stratum()`,
   `weight_species_by_area()`.
6. **MEDITS plots (Step 8)** — `plot_fg_timeseries_by_area()`,
   `plot_fg_timeseries_regional()`, `plot_strata_profile()` →
   `survey_fg_density_timeseries.png`,
   `survey_fg_density_timeseries_regional.png`,
   `survey_fg_depth_strata_profile.png`.

   ```{r fg-timeseries-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_fg_timeseries_by_area() output — MEDITS trawl survey density by FG and area (strata-weighted)."}
   safe_include_graphics(
     "figures/survey_fg_density_timeseries.png",
     script = "01_biomass.R",
     out_dir_note = "<out_dir>/plots/survey_fg_density_timeseries.png"
   )
   ```

   ```{r strata-profile-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_strata_profile() output — raw per-sample density across depth strata, by FG."}
   safe_include_graphics(
     "figures/survey_fg_depth_strata_profile.png",
     script = "01_biomass.R",
     out_dir_note = "<out_dir>/plots/survey_fg_depth_strata_profile.png"
   )
   ```
7. **MEDIAS acoustic survey (Step 9, sub-steps 9a–9g)** — load and collapse
   raw abundance/biomass records, map species codes via ASFIS, match to FG
   the same way as MEDITS, attach area (MEDIAS Handbook Table 1, hand-keyed
   `MEDIAS_HANDBOOK_AREAS` — as delivered this table only has Western Med
   entries, since that's this example's scope; extending to another region
   means adding that region's own Handbook rows, with a bathymetry-derived
   10–200 m fallback via `compute_strata_area_by_area()` covering any
   country/GSA combination the table doesn't have a row for), compute
   region-wide FG and species density, then
   **combine MEDITS + MEDIAS** into one FG-level and one species-level table
   (averaged on Year/FG or Year/species overlap, overlaps flagged/printed).
   Outputs: `medias_fg_annual_density_by_area.csv`,
   `medias_fg_annual_density_regional.csv`,
   `medias_fg_biomass_density_timeseries_regional.png`,
   `medias_fg_abundance_density_timeseries_regional.png`,
   `species_density_regional_combined.csv` (**contract file** — read by
   `02_fisheries.R` for F and the NEI cap, and by `03_pbqb-traits.R`).
8. **Excel export (Step 10)** — `survey_fg_annual_index.csv`,
   `survey_fg_annual_index_regional_combined.csv`, then
   `export_ecopath_ecosim_excel()` writes sheets `FG_spp_Ecopath`, `Ecopath`,
   `Ecosim`, `FG_spp_Ecosim` into `ecopath_ecosim_inputs.xlsx`. **Also writes
   a `PB_QB_spp` scaffold** — reusing `FG_spp_Ecopath`'s own `Species`/
   `FG_num`/`Density`/`prop_sp_fg` (renamed to `PB_QB_spp`'s convention:
   `Biomass`, `prop_biomass_FG`), with `PB`/`prop_biomass_PB`/`QB`/
   `prop_biomass_QB` all blank and a `Note` column — so `PB_QB_spp` exists in
   the workbook after this script alone, without needing `03_pbqb-traits.R`
   (Step 4) to have run first. This function has no PB/QB computation of its
   own (that's entirely Step 4's, from life-history parameters, not survey
   density), so the scaffold's numbers are genuinely blank, not estimated.
   Running Step 4 afterward replaces the scaffold with the real sheet
   automatically (`upsert_workbook_sheets()` always fully replaces a
   same-named sheet). Guarded against clobbering the other way: if
   `PB_QB_spp` already has real (non-`NA`) `PB` values — i.e. Step 4 already
   ran against this exact workbook — re-running this script leaves it
   untouched rather than overwriting it with a blank scaffold again.
9. **Complete `FG_spp_Ecopath`/`FG_lookup` (Step 11)** — union of observed +
   reference-catalog species, zero-fill species with no base-year density,
   attach taxonomy, defensive check for FGs with no species at all. Written
   via `upsert_workbook_sheets()`.
10. **`traits_ewe` (Step 12)** — moved to `03_pbqb-traits.R` (`Ecopath_traits`).
11. **Finalize sheet order (Step 13)** — `finalize_workbook_sheet_order()`.

## Outputs

* **CSVs** (in `out_dir`): `survey_sample_coverage_map.png` and other plot
  PNGs (in `out_dir/plots`), `survey_unmatched_for_manual_review.csv`,
  `survey_outliers_flagged.csv`, `strata_area_by_area.csv`,
  `medias_fg_annual_density_by_area.csv`,
  `medias_fg_annual_density_regional.csv`,
  `species_density_regional_combined.csv`, `survey_fg_annual_index.csv`,
  `survey_fg_annual_index_regional_combined.csv`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written/updated:
  `FG_spp_Ecopath`, `Ecopath`, `Ecosim`, `FG_spp_Ecosim`, `FG_lookup`
  (renamed to `FG`), `traits_ewe`, `PB_QB_spp` (scaffold, unless it
  already has real data from `03_pbqb-traits.R`), `AquaMaps_Depth_Adjustment`
  (only when `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- TRUE` — see Step 7b), and
  `FG_Density_by_Stratum` (only when `STRATA <- TRUE` — Step 12c; density
  by FG x depth stratum, region-wide, covering whichever strata are
  actually in play).

---

# Script 1b — `01b_biomass_unsurveyed.R`

**Role:** sourced by `01_biomass.R` (`source(..., local = TRUE)`, inside the `AREA_MODE == "westmed"` block, after MEDIAS). It builds every biomass that does not come from MEDITS/MEDIAS and returns `stock_assessment_fg_year` (stock assessments + literature + models), `stock_assessment_fg_year_fisheries_only` and `species_lit_biomass` to 01. All values are converted to t WW/km² of $A_{model}$.

## Sources (the "SOURCE A–D" roadmap in the script)

| Source | FGs | Method | Reference table / data |
|---|---|---|---|
| A. Stock assessment | Bluefin tuna, swordfish; marine megafauna | ICCAT density (GBYP aerial survey for BFT; $0.72\,B_{MSY}$ for SWO); megafauna density or count × body mass | `marine_megafauna_biomass.csv`; ICCAT reports |
| B. Literature density × depth band | Benthic mollusc, other macro-benthos (literature first, MEDITS fallback), sea pens, suprabenthos | $B = D \times A(z_{min}, z_{max}) / A_{model}$, $A$ from the strata/bathymetry | `literature_density_biomass.csv` |
| C. Habitat map × density × occupancy | Posidonia, Cymodocea, macroalgae, coralligenous corals and gorgonians | $B = D \times c_{WW} \times A_{map} \times occ / A_{model}$ | `habitat_shapefile_biomass.csv`, `density_unit_to_wet_g_m2.csv`, EUSeaMap 2025, EMODnet macroalgae |
| D. Plankton models | Large and small phytoplankton, meso + micro zooplankton, detritus | ISIMIP3a GFDL-MOM6-COBALT2 obsclim (CMEMS MedBFM/PISCES/satellite as cross-check); macrozooplankton = krill field minimum (`MACROZOO_SOURCE`) | ISIMIP cache, `lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R` |

B and C are added together per FG × year.

## Habitat areas (EUSeaMap 2025)

`compute_euseamap_habitat_area()` reads living Posidonia (EUNIS2019C MB252, MB2522; dead matte MB2523 excluded), coralligenous (MC151, MC251) and tries two Cymodocea codes (MB5521, MB5534 — not present in the 2025 map) from the geodatabase, clips them to the model GSAs, dissolves per GSA and measures them in an equal-area Albers projection. Only habitat deeper than the shallowest MEDITS stratum (`EUSEAMAP_MIN_DEPTH_M` = 10 m) is counted: `.frac_deeper_than()` samples up to 5,000 points per GSA polygon and reads NOAA ETOPO depth via `marmap` (15 arc-sec, 1 arc-min fallback; coarse for the 0–10 m band). Result cached in `euseamap2025_seagrass_area_by_gsa_ge10m.csv` (delete to recompute). 2026-10-07 run: Posidonia 3,111.9 km² (4,351.2 km² at all depths), coralligenous 1,375.3 km².

## Values and assumptions (2026-10-07 run)

| FG | Density / method | Ecopath_B (t/km²) | Assumptions |
|---|---|---|---|
| Posidonia | 168 g DW/m² leaf (280 at 2–4 m × depth factor 0.6) × 4.17 × 3,111.9 km² × 0.7 | 8.68 | depth factor, no seasonal correction, occupancy |
| Cymodocea | 410.4 g DW/m² × 4.17 × (10–15 m band) × 0.6 | 14.70 | no map; may include below-ground biomass |
| Macroalgae | 1,650 g DW/m² × 7 × 325.05 km² × 0.4 | 8.55 | DW→WW × 7, occupancy, ICES-ecoregion area |
| Corals and gorgonians | 37.3 g DW/m² × 2.5 × 1,375.3 km² | 0.729 | DW→WW × 2.5 |
| Large / small phytoplankton | COBALT, C:WW 0.16 | 5.70 / 6.30 | model value |
| Meso and micro zooplankton | COBALT zmicro + zmeso, C:WW 0.07 / 0.09 | 29.6 | above the field range (05, block 11) |
| Macro zooplankton | krill field minimum | 0.124 | lower bound |
| Suprabenthos | literature density, 10–388 m + 389–1,255 m | 0.022 | hand-read figure |
| Detritus | Pauly et al. 1993 equation | 288 | — |
| Salps | none (Ecopath EE) | — | — |

Full citations are in the script header (`REFERENCES`) and in each row's `Source_citation`, which is copied to `FG_References`.

## Outputs (in `output/biomass/`)
`habitat_shapefile_biomass_by_fg.csv`, `literature_density_biomass_by_fg.csv`, `marine_megafauna_biomass_by_fg.csv`, `uncertain_fg_biomass_status_REVIEW.csv`, `temporal_mismatch_REVIEW_*.csv`, `ecobase_biomass_evaluation_NOT_INCORPORATED.csv`, `fg_missing_ecopath_B_REVIEW.csv`.

---

# `AREA_MODE == "custom"` — a fully custom study area, in `01_biomass.R`

**Role:** Not a separate script — a config switch inside `01_biomass.R`
(`AREA_MODE <- "custom"`, instead of the `"westmed"` default) for **any
study area anywhere in the Mediterranean that doesn't line up with GSA
boundaries** — a single custom bounding box or shapefile (e.g. one EwE
model's own footprint, a bay, an MPA, or a region spanning parts of several
GSAs, in any part of the basin), or a custom subset/grouping of real GSAs.

**Merged into `01_biomass.R`.** This used to be a second script,
`01_survey_density_custom.R`, run *instead of* `01_biomass.R`. It is now
merged into `01_biomass.R` itself, so there's one script and one set of
fixes/improvements both regions benefit from. `01_survey_density_custom.R`
is retired and no longer in the repository.

Structurally identical to `AREA_MODE == "westmed"` for the MEDITS loading
and FG-matching stages (Steps 1–4 use the exact same shared-library
functions, unconditionally). What's genuinely different:

* **MEDIAS acoustic survey and GFCM stock-assessment biomass are skipped**
  (Step 9, and the stock-assessment/priority blocks right after it) —
  both are West-Med-subregion-scoped data sources (MEDIAS Handbook Table 1
  areas, STAR/RAM Legacy species-by-GSA assessments) with no defined
  meaning outside the named GSAs they're built around. `fg_index_regional_combined`/
  `species_density_regional_combined`/`fg_cv_log_combined` — the same
  variable names Step 10 onward reads either way — are built as plain
  MEDITS-only aliases instead.
* **`AREA_NAME`** — a user-supplied label for this custom study area (typed
  interactively via `rstudioapi::showPrompt()`, or hardcoded for the
  `daniel` user). Used to build a **dedicated output subfolder**
  (`out_dir/<AREA_NAME>/`), so a custom-region run never collides with the
  default Western-Med `out_dir` or with another custom run under a
  different `AREA_NAME`.
* **Study-area definition** — `CUSTOM_AREA_TYPE <- "bbox"` (default; a
  rectangle defined by `CUSTOM_BBOX <- c(xmin, xmax, ymin, ymax)`),
  `"shapefile"` (an arbitrary `sf` polygon read via `st_read()`), or **`"gsa"`**
  (one or more GFCM GSAs, picked out of the same GSA shapefile downloaded
  for `AREA_MODE == "westmed"`, via `CUSTOM_GSA_IDS <- c(...)` — e.g. a
  single GSA, or a different regional grouping than the fixed `1:11`
  Western-Med range). For `"bbox"`/`"shapefile"` the result is a
  single-feature `area_shp` with `AREA_ID_COL = "area_id"`, always
  `area_id = 1` — there is no GSA sub-division inside a custom region unless
  the user builds one. For `"gsa"`, `area_shp` can have **one row per
  selected GSA**, each keeping its own `gsa_num` as `area_id` (same
  convention as `AREA_MODE == "westmed"`), so a multi-GSA custom region is
  NOT forced down to a single `area_id = 1`.
* **`AreaID` is derived spatially** for `"bbox"`/`"shapefile"`, not read off
  MEDITS' own `area` (GSA) column — every sample gets `AreaID = 1`, then
  `filter_samples_by_area()` drops everything outside the custom boundary.
  For `"gsa"`, `AreaID` follows the selected GSA(s) directly, same as
  `AREA_MODE == "westmed"`.
* **Strata-area computation (Step 6)** — `compute_strata_area_by_area()`
  called with whichever `AreaID`s are actually in scope (a single custom
  polygon for `"bbox"`/`"shapefile"`, or the selected `CUSTOM_GSA_IDS` for
  `"gsa"`) — the function itself is agnostic to this distinction either way,
  no branch needed.

Three files stay at plain, unsuffixed names regardless of `AREA_MODE`,
because `02_fisheries.R`, `03_pbqb-traits.R` and `04_diets.R` read them by that exact
fixed name: `species_density_regional_combined.csv`,
`ecopath_ecosim_inputs.xlsx`, `strata_area_by_area.csv`. The dedicated
`AREA_NAME` output subfolder (not a filename suffix) is what keeps a custom
run's copies of these three from colliding with the default run's.

## Inputs

Same as `AREA_MODE == "westmed"` minus the MEDIAS/ASFIS inputs (the GSA
shapefile is still downloaded when `CUSTOM_AREA_TYPE == "gsa"`), plus (for
`CUSTOM_AREA_TYPE == "shapefile"`) a user-supplied boundary shapefile.

## Processing steps

Steps 1–8 and 11–13 are identical code to `AREA_MODE == "westmed"` (see
above), running against the custom `area_shp`/`AreaID`. Step 9 (MEDIAS +
stock-assessment + priority combine) is skipped entirely — its `else{}`
branch aliases `fg_index_regional`/`species_density_regional`/
`fg_cv_log_medits` (already computed in Steps 5–7) as the `*_combined`
tables instead, so the `species_density_regional` table feeding Step 10 is
MEDITS-only — no MEDITS+MEDIAS averaging step.

## Outputs

Same file set and workbook sheets as `AREA_MODE == "westmed"` (see above),
written to `out_dir/<AREA_NAME>/`, with the three contract files above
staying unsuffixed.

---

# Script 2 — `02_fisheries.R`

**Role:** Step 2. Builds catches (C), landings (L) and discards (Di) by FG and fleet, the effort series, and fishing mortality (F). Needs Step 1's `strata_area_by_area.csv`, `species_density_regional_combined.csv` and the `Ecopath_B` sheet. Sources `02b_sau_extraction.R` (Sea Around Us extract) and `lib_survey_fg_density_functions.R`. The former scripts `02_fao_catches.R`, `02a_fisheries_multisource.R` and `02b_fisheries_priority_merge.R` are retired; nothing in the current run reads them.

## Configuration and domain

- `FILTER_AREAS` = the GSAs of `strata_area_by_area.csv` (1, 2, 5–11); `Total_Area_km2` = its sum (175,660 km²), the same $A_{model}$ as the biomass.
- `GSA_COASTAL_STATE` (GFCM Res. GFCM/33/2009/2) → `TARGET_COUNTRIES` = coastal states of the model GSAs = Spain, France, Italy (`TARGET_COUNTRIES_OVERRIDE` to change). Catch of other countries in GFCM divisions 37.1.1–37.1.3 is excluded and reported (2026-10-07 run: Algeria 3.25 Mt, Morocco 0.86 Mt, Tunisia 0.64 Mt over all years).
- `FLEET_REGISTER`, `GEAR_TO_FLEETTYPE` and FishMIP effort are filtered to `TARGET_COUNTRIES`; STECF FDI is filtered to the model GSAs.
- `STAR_STOCK_GSA_OVERRIDE` corrects stock-unit labels (e.g. HAKEMEDGSA1-7 = GSAs 1, 5, 6, 7).
- `YEAR_ECOPATH` = 1994–1996; `START_YEAR`–`END_YEAR` = 1994–2023; `DATASET_VERSION` = GFCM_2025.

## Processing steps

1. **GFCM catch → FG.** Species matched to FGs by scientific name, common name, the FAO `Name_En` → `Scientific_Name` bridge (ASFIS), genus, taxonomic rank (`match_bare_rank_to_fg()`) and NEI keywords (`nei_keyword_to_rank.csv`). Every match goes through `resolve_matches_safely()` (non-commercial rule below).
2. **Magnitude calibration** to STECF FDI for EU countries: $k_{c,f} = \sum C^{FDI}/\sum C^{GFCM}$ (2014–2023), not applied when > 10 (`catch_calibration_factor_capped_REVIEW.csv`).
3. **Excluded gears** (`FleetType = EXCLUDED` in `gear_to_fleettype.csv`) removed by their FDI (2014+) or SAU share.
4. **Discards** $Di = C \times r$ (FDI per-fleet ratio, else FishMIP ratio); $L = C - Di$.
5. **ICCAT** (BFT, SWO, ALB): Task I nominal catch of `TARGET_COUNTRIES`, Mediterranean rows only, × `fdi_model_gsa_share` (each country's FDI landings share inside the model GSAs, 2014+); replaces the GFCM value for these cells and gives a real gear split (`resolve_iccat_fleet_type()`, checked against ICCAT ST01 vessel length).
6. **Hake stanzas.** `compute_stanza_catch_split_from_length()` (library) uses MEDITS TC length frequencies of 1994–1996 in the model GSAs, VBGF ($L_\infty$ 110 cm, $K$ 0.178, $t_0$ 0; Mellon-Duval et al. 2010), $W = aL^{3.035}$, juvenile = age < 2 (L < 32.9 cm), MCRS 20 cm: 55.9 % of landings and 100 % of discards → FG 26, the rest → FG 27; applied at FG and fleet level (`multistanza_juv_adult_age_split.csv`).
7. **Fleet split** $C_l = C_{c,f,y}\,p_{l,c,f,y}$ (FDI → SAU hindcast → FDI multi-year mean → country mix → equal share, each flagged). Residual target-country catch with no fleet → column "Target countries - fleet not resolved".
8. **Effort.** $E_{l,y} = NE_{c,y}\,\kappa_c\,s_{l,y}$ (Rousseau nominal effort calibrated to FDI); fleet shares from FDI since 2014 and, before, the catch × days-per-tonne share rescaled per fleet to the 2014–2016 FDI mean (no jump at 2014; boundary check printed in the log). Technology creep below.
9. **Ecopath sheets.** `Ecopath_L`, `Ecopath_Di`: 1994–1996 mean / $A_{model}$ by FG × fleet.
10. **Fishing mortality.** `F_by_fg.csv`: $F_f = (\bar C_f/A_{model})/B_f$ with $B_f$ = the final `Ecopath_B` read from the workbook (helper `.use_ecopath_b()`; survey density only for FGs with no Ecopath_B; column `B_source`). `F_by_species_fg.csv` repeats the FG value per species. Changed on 2026-10-07 (before: survey density for every FG).
11. **Ecosim.** Catch $C_{f,y}/A_{model}$ and effort relative to 1995 in `Ecosim_ts`.

## Technology-creep correction

Raw nominal effort — FDI's own `Effort_days`/`Effort_kWdays_total` and the
Rousseau-calibrated series for the target countries — understates
real fishing power over time: gear, electronics and engine efficiency
improve ("technology creep") even when the number of days fished stays
flat. `tech_creep_multiplier()` converts nominal to an *effective* effort
figure to correct for this.

For a calendar year $y$ and gear class $g \in \{\text{bottom\_trawl},
\text{other}\}$, the annual creep rate $r(y,g)$ (`TECH_CREEP_PERIOD_RATES`):

$$
r(y, g) =
\begin{cases}
0.79\%/\text{yr} & 1994 \le y \le 2013,\ g = \text{bottom\_trawl} \\
2.00\%/\text{yr} & 1994 \le y \le 2013,\ g = \text{other} \\
4.50\%/\text{yr} & 2014 \le y \le 2023,\ \text{any } g
\end{cases}
$$

- **0.79%/year, 1994–2013 bottom-trawl (FDI's "DTS" code)** — Damalas, D.,
  Maravelias, C.D., Osio, G.C., Maynou, F., Sbrana, M., Sartor, P. (2015).
  "Once upon a Time in the Mediterranean" — long term trends of
  Mediterranean fisheries resources based on fishers' Traditional
  Ecological Knowledge. *PLoS ONE*, 10(3), e0119330, as directly quoted by
  Tsagarakis et al. (2022, *Frontiers in Marine Science* 9:919793).
- **2.0%/year, 1994–2013 every other gear** — Tsagarakis et al. (2022)'s
  own summary of the wider Mediterranean creep literature (best-fitting of
  0%/1%/2%/year tested for a Mediterranean Ecopath hindcast, 1993–2020).
- **4.5%/year, 2014–2023, flat across gears** — Palomares & Pauly (2019)'s
  $C\% = 13.8 \times Y^{-0.511}$, evaluated **once** at $Y=9$ (the window's
  full length) = 4.49%/year — not re-evaluated at each year's own elapsed
  distance, which would spike to 13.8%/year at the window's start (correct
  algebraically, not a realistic annual rate, since $Y$ is meant as a
  whole-window average).

A year outside every listed period uses whichever listed period's midpoint
is closest. The rate is further scaled by a country multiplier
$c(\text{Country})$ (`TECH_CREEP_COUNTRY_MULTIPLIER`: EU-3 = 1.0; the
0.7 for Morocco/Algeria/Tunisia is unused while they are outside `TARGET_COUNTRIES`) and a sector multiplier $s(\text{Sector})$
(`TECH_CREEP_SECTOR_MULTIPLIER`: Industrial = 1.0; Artisanal/Recreational =
0.6) — both flagged in the code as judgment calls, not literature-derived.
Combining $cm = c \times s$, the cumulative multiplier compounds year by
year from `TECH_CREEP_BASE_YEAR`:

$$
M(\text{Country}, Y, \text{Sector}, g) =
\begin{cases}
1 & Y = \text{base\_year} \\[4pt]
\displaystyle\prod_{y=\text{base\_year}+1}^{Y} \left(1 + \frac{r(y,g)\cdot cm}{100}\right) & Y > \text{base\_year} \\[10pt]
\displaystyle\prod_{y=Y+1}^{\text{base\_year}} \left(1 + \frac{r(y,g)\cdot cm}{100}\right)^{-1} & Y < \text{base\_year}
\end{cases}
$$

$$
\text{Effort}_{\text{effective}}(\text{Country}, Y, \text{Sector}, g) = \text{Effort}_{\text{raw}}(\text{Country}, Y, \text{Sector}, g) \times M(\text{Country}, Y, \text{Sector}, g)
$$

Implemented as `tech_creep_rate_for_year()` / `tech_creep_multiplier()`,
applied identically to FDI's `Effort_kWdays_total` and FishMIP's
`nom_active_kWdays`, just with a different Country/Sector/gear_class set
per source.

## NEI catch fallback with an F≤1 cap

Raw catch records reported under a broad FAO "NEI" (not elsewhere
included) common name — "Marine fishes NEI", "Cuttlefish, bobtail squids
NEI", "Marine crustaceans NEI", "Various squids NEI", etc. — previously
fell through every matching step unresolved (~97% of unresolved catch
tonnage in a real review). `nei_keyword_to_rank.csv` maps each such
keyword to a target taxonomic Class/Phylum; the matched catch is then
biomass-weight-split across the commercial FGs under that rank (reusing
`match_bare_rank_to_fg()`, the same tiered Genus→Family→Order→Class→Phylum
splitter the bare-rank taxonomic fallback uses elsewhere), explicitly
excluding non-commercial FGs from the split (see the next section).

Before this split is applied to `catches_discards_fg`, each candidate FG's
resulting fishing mortality — the same $F = (\text{Catch}_{t,\text{avg}} /
\text{Total\_Area\_km}^2) / B_f$ used for the F block, with $B_f$ the final
`Ecopath_B` (survey density only where an FG has no Ecopath_B; since
2026-10-07), averaged over `YEAR_ECOPATH` — is checked. Any FG that would exceed
$F=1$ because of its NEI-attributed catch has that catch trimmed back down
to exactly its $F=1$ cap; the removed tonnage is redistributed (a single
proportional pass across the other candidate FGs under the same rank that
still have headroom — not a fully iterative water-filling algorithm) to
FGs with room; whatever can't be placed anywhere without pushing another
FG over $F=1$ is left capped out and reported via a console message, never
silently dropped or forced through. Writes `nei_keyword_fcap_adjustments.csv`
(Species × FG × delta, `YEAR_ECOPATH` only).

## Non-commercial-FG exclusion rule

A raw catch record can never resolve onto a non-commercial FG — nobody
reports landed catch of a group nobody is allowed/able to land. This rule
is enforced in two places with different scopes; the second one is the
real backstop, applying to **every** catch-to-FG match the cascade
produces, not only the taxonomic-rank fallback.

**`match_bare_rank_to_fg()`** (the shared taxonomic-rank splitter used by
both the ordinary bare-rank fallback and the NEI keyword fallback above)
drops any FG whose name matches `non-commercial|noncommercial`
(case-insensitive) from a rank's candidate set *before* computing biomass
weights — and if a rank's candidates are now empty because they were ALL
non-commercial, falls through to try the next broader rank instead of
returning an empty/non-commercial result.

**`resolve_matches_safely()`** — the one function nearly every merge step in
the cascade funnels its result through (direct name match, FAO
`Scientific_Name` bridge, genus fallback, word-containment fallback, manual
overrides, the rerun cascade for late-arriving species, and both SAU match
tiers) — originally only checked for non-commercial candidates in its
*ambiguous* branch (`n_distinct_fg > 1`): when an ambiguous species' FGs
span a commercial-vs-non-commercial split of the same taxon, dropping the
non-commercial candidate(s) could leave exactly one FG, rescuing the match
onto the commercial one. That left a real gap, flagged in review: a
species that already resolved to exactly **one** FG via any of those
merge steps — a direct scientific-name hit, a common-name match, the FAO
bridge, a genus match, SAU, and so on — had no non-commercial check at
all, because the check only ever ran on the `n_distinct_fg > 1` branch. If
that one matching FG happened to be non-commercial, it was accepted as a
valid match with nothing to catch it.

The fix broadens the check to run before the safe/ambiguous split, not
just inside it: for every query, if **all** of its candidate FG(s) — one
candidate or several — are non-commercial, that query is moved out of
`safe_queries` into `ambiguous_queries`, regardless of how many candidates
it had. This isn't a "drop the species" change: every caller of
`resolve_matches_safely()` only removes `$safe`'s species from its own
`remaining` table before moving to the next cascade step, so a query that
lands in `$ambiguous` instead simply stays in `remaining` and gets another
attempt at a later, broader step (taxonomy fallback, SAU, STAR/RAM, ICCAT,
...) exactly as an ambiguous match always has — it is never force-matched
to a non-commercial FG, and never silently lost. A species with no
remaining commercial candidate anywhere in the cascade ends up correctly
unresolved (flagged for manual review), rather than wrongly assigned.

## Effort of non-target countries

The FishMIP forward-fill for Morocco/Algeria/Tunisia (effort held flat after FishMIP's last year, 2017) only matters when those countries are in `TARGET_COUNTRIES`. With the current domain rule they are excluded, so their effort no longer appears in the outputs.

## Outputs (in `output/fisheries/` unless stated)

| File | Content |
|---|---|
| `Ecopath_L`, `Ecopath_Di` (workbook) | FG × fleet landings and discards, t/km²/yr, 1994–1996 |
| `F_by_fg.csv`, `F_by_species_fg.csv` | F per FG (with `B_source`), repeated per species |
| `Catches_Discards_FG_ts.csv`, `Catches_Ecopath`, `Catches_Ecosim` | FG × year series |
| `nei_keyword_fcap_adjustments.csv` | NEI catch moved by the F ≤ 1 cap |
| `catch_calibration_factor_capped_REVIEW.csv`, `gfcm_stecf_catch_magnitude_calibration_factors.csv` | GFCM → FDI calibration |
| `ecopath_L_fg_coverage_check.csv`, `fg_missing_ecopath_L_Di_REVIEW.csv` | fleet coverage, zero-catch FGs |
| `multistanza_juv_adult_age_split.csv` | hake juvenile/adult catch shares |
| `plots/fisheries/fisheries_validation_plots.pdf` and PNGs `01`–`08` | catch, discards, fleets, effort, data-source coverage |

## Log values of the 2026-10-07 run

- ICCAT model-GSA shares: France 1 (all); Italy ALB 0.23, BFT 0.57, SWO 0.59; Spain ALB 1, BFT 0.95, SWO 0.89.
- Excluded gears, 1994–1996 mean: France 1,158 t/yr, Italy 15,188 t/yr, Spain 9,855 t/yr.
- Effort 2013 → 2014 ratio: 0.53–3.26 (5 cells above ±50 %).
- Rousseau vs FDI effort correlation, 2014–2017: France 0.96, Spain 0.95, Italy −0.24.
- Highest F (before the Ecopath_B change): deep-water rose shrimp 5.86; suprabenthos and macro zooplankton 1.0 (capped against survey biomass — fixed by the change); coastal benthic cephalopods 0.85.

---

# Script 3 — `03b_ecobase.R`

**Role:** Pipeline Step 3 of 4. Independent of Steps 1/2 — can run anytime
before Step 4. **Optional** input to Step 4 for literature-based PB/QB
gap-filling and comparison against the empirical formula estimates.

Queries **EcoBase** (ecobase.ecopath.org), an open-access repository of
published Ecopath models, via its SOAP/XML web service — the official
EcoBase R example ported from RCurl/XML to httr/xml2. The script author
notes this was **not tested against the live service** from within the
authoring sandbox (no network access to `sirs.agrocampus-ouest.fr`) — the
diagnostics throughout are there specifically to surface an endpoint change,
not just fail silently.

**Fixed: `merge(mediterranean_models, all_status, ...)` — "x has duplicated
column name [comments_objectives]".** Root cause traced to
`xml_node_to_dt_row()` (formerly an inline anonymous function, at two call
sites — the model-list parse and, in `fetch_model_inputs()`, the per-model
group parse): when a single `<model>` (or `<group>`) XML node repeats a
child tag — confirmed EcoBase's own XML does this, e.g. a duplicated
`<comments_objectives>` — `setNames(as.list(xml_text(fields)),
xml_name(fields))` produced a list with two entries sharing one name, and
`as.data.table()` on that silently creates a data.table with two
same-named columns. data.table allows that at creation time; it only
rejects it the moment something needs unique names — which is why the
error surfaced three functions and one `rbindlist()` away from where the
duplicate actually originated, at the `merge()` call. Fix: both parse
sites now go through one shared `xml_node_to_dt_row()` that detects a
repeated tag name and collapses its values into ONE column, joined with
`"; "` — nothing is dropped, there's just no longer two columns fighting
over one name.

## Inputs (network)

| Endpoint | Purpose |
|---|---|
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client_3.php` | Full model list |
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client.php?no_model=<id>` | Per-model input values (Biomass/PB/QB per group) |

`pcloud_dir`/`git_dir` are resolved for consistency with the other scripts
but not actually used — this script is self-contained.

## Processing steps

1. Fetch the full EcoBase model list; save `ecobase_model_list_full.csv`.
2. Filter geographically to the Mediterranean, using each model's
   `geographic_extent` bounding box (not text/keyword matching on
   country/ecosystem names — more robust). Save
   `ecobase_mediterranean_models.csv`.
3. Filter further to `dissemination_allow == "true"` — EcoBase's own flag for
   which models actually have downloadable input data (per the site, only
   ~233 of ~500 listed models do; the rest are metadata/reference-only).
   `KNOWN_RELEVANT_MODEL_IDS` is the resulting query list.
4. Fetch per-model group-level Biomass/PB/QB for each disseminable model
   (`fetch_model_inputs()`), explicitly detecting the "data not available
   despite the flag" case (confirmed against a real response for model 418).
   `-9999` (Ecopath's own sentinel for "not provided") is converted to `NA`.
5. Save a full status audit (`ecobase_mediterranean_models_status.csv`) and
   the failure-only subset (`ecobase_models_without_data.csv`).
6. Compile the final summary table: `FG_name, PB, QB, EwE_model, year,
   authors` (model metadata columns matched by best-guess candidate names
   against `model_list`, since EcoBase's exact field names aren't confirmed).

## Outputs

* `ecobase_model_list_full.csv`, `ecobase_mediterranean_models.csv`,
  `ecobase_mediterranean_models_status.csv`, `ecobase_models_without_data.csv`,
  `ecobase_literature_pb_qb_raw.csv`,
  **`ecobase_literature_pb_qb_simple.csv`** (the file `03_pbqb-traits.R` reads).
* No workbook writes from this script directly — `03_pbqb-traits.R` is what
  merges this CSV into the workbook's `Ecobase` sheet.

### Column reference — `ecobase_literature_pb_qb_simple.csv`

| Column | Meaning |
|---|---|
| `FG_name` | EcoBase group name (matched to your own `FG_name` by text in `03_pbqb-traits.R` — group naming won't line up automatically across different published models) |
| `PB` | Literature Production/Biomass (`-9999`, EcoBase's own "not provided" sentinel, converted to `NA`) |
| `QB` | Literature Consumption/Biomass (same `-9999`→`NA` handling) |
| `EwE_model` | Source model name |
| `year` | Source model's reference year |
| `authors` | Source model's authors |

---

# Script 4 — `03_pbqb-traits.R`

**Role:** Pipeline Step 3. Requires Step 1 (`FG_spp`, `biomass_proportion_by_species_fg.csv`) and Step 2 (`F_by_species_fg.csv`, `F_by_fg.csv` in `output/fisheries/`); sources `03b_ecobase.R` for the EcoBase cross-check. Missing optional inputs produce a message, not an error.

Estimates Production/Biomass (PB) and Consumption/Biomass (QB) for every EwE
Functional Group, following the taxon-specific methods in *"Quick guide on
how to calculate P/B and Q/B for EwE models"* (Vilas, Coll, Piroddi,
Steenbeek). Growth parameters, length-weight a/b, maturity, body weight,
longevity, trophic level, temperature, depth and aspect ratio are all fetched
automatically from FishBase + SeaLifeBase (`rfishbase` ≥ 4.0.0 required — the
script hard-stops with install instructions if an old, broken pre-4.0 CRAN
build, or a `duckdbfs` install missing `duckdb_config`, is detected).

## Inputs

| File | Format | Role |
|---|---|---|
| `species_density_regional_combined.csv`, `FG_spp` | csv / sheet | **From Step 1.** `species_df` is built from `FG_WMed_2026.csv` (`FG_REFERENCE_CSV_PATH`) with each species' biomass from `FG_spp` (Species, FG, Biomass density, Yield=NA) |
| `output/fisheries/F_by_species_fg.csv`, `F_by_fg.csv` | csv | **From Step 2.** $F$ per FG ($C/B$ with the final Ecopath_B); fish get $P/B = M + F$ per species, non-fish FGs get $F$ added at FG level |
| `strata_area_by_area.csv` | csv | **From Step 1.** Study area, needed to convert FG catch totals into a density comparable to Biomass |
| `ecobase_literature_pb_qb_simple.csv` | csv (optional) | **From Step 3.** Literature PB/QB for gap-filling/comparison |
| `<pcloud_dir>/data/FG_WMed_2026.csv` | csv | FG reference and species list (`FG_REFERENCE_CSV_PATH`) |
| `landings_by_species_gsa_year_<FISHERIES_DATA_SOURCE>.csv` | csv (legacy, normally absent) | Species-level landings from the retired `02a` script; when missing (the normal case) the log says so and F comes from Step 2 |
| FishBase / SeaLifeBase (`rfishbase`) | web API | Growth, length-weight, maturity, ecology/trophic level, swimming/aspect-ratio traits |
| WoRMS (`worrms`, via `lib_worms_taxonomy_lookup.R`) | web API | Taxonomic classification (Class → dispatch group: fish/mammal/seabird/phytoplankton/invertebrate) |
| FishLife R package | local DB (`FishBase_and_RAM$ParentChild_gz`) | Phylogenetically-imputed Loo/K/M/Winf, walking up the taxonomic tree when a direct match is missing |
| TropFishR R package | local | `M_empirical()` — Hoenig (1983), Then et al. (2015), Alverson & Carney (1975) natural-mortality estimators |

### Column reference — key input files

**`species_density_regional_combined.csv`** (from Step 1 — required columns,
enforced by `build_species_df_from_survey()`):

| Column | Meaning |
|---|---|
| `Year` | Survey year |
| `FG_num`, `FG_name` | Functional Group id/name |
| `ScientificName` | Species scientific name |
| `mean_density` | Species density, t/km² |

**`F_by_fg.csv`** (from Step 2):

| Column | Meaning |
|---|---|
| `FG_num`, `FG_name` | Functional Group id/name |
| `Biomass_density_avg` | $B_f$, t/km² (final Ecopath_B; `B_source` says which) |
| `Catch_t_avg`, `Catch_density_avg` | 1994–1996 mean catch, t/yr and t/km²/yr |
| `F` | $C/B$, yr⁻¹ |

**`strata_area_by_area.csv`** (from Step 1):

| Column | Meaning |
|---|---|
| `AreaID` | GSA number (or `1` for a custom-region run) |
| `Stratum` | Depth-stratum number |
| `area_km2` | Stratum area within that GSA/region, km² |
| `prop` | That stratum's proportion of the area's total |

**`landings_by_species_gsa_year.csv`** (placeholder path — not yet a real
data source; schema below is what `attach_yield_from_landings()` expects if
this is ever wired up):

| Column | Meaning |
|---|---|
| `ScientificName` | Species scientific name |
| `Year` | Landings year |
| `AreaID` | GSA number |
| `catch_t` | Landed catch, tonnes |

```{r pbqb-comparison-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="fish_PB_methods_comparison.png — all 6 fish PB/M methods per species, chosen method highlighted."}
safe_include_graphics(
  "figures/fish_PB_methods_comparison.png",
  script = "03_pbqb-traits.R",
  out_dir_note = "<out_dir>/plots/fish_PB_methods_comparison.png"
)
```

## Processing steps

1. **Setup** — package loading, `rfishbase`/`duckdbfs` version guard, GUI or
   console progress tracker across 13 named pipeline stages.
2. **Load `species_df`** (`SPECIES_DF_SOURCE = "survey"`): species and FGs from `FG_WMed_2026.csv`, biomass from `FG_spp` (log: "Built species_df: … rows").
3. **Fishing mortality attachment** — `F_by_species_fg.csv` from `02_fisheries.R` gives every species of a fished FG the FG value `Fmort_FG` (log: "[F] Read F_by_species_fg.csv …"). Fish use it inside `calc_fish()` ($P/B = M + F$). The legacy paths (`FISHERIES_DATA_SOURCE`, `YIELD_SOURCE`, `FG_YIELD_SOURCE`, which read files of the retired `02_fao_catches.R`/`02a_fisheries_multisource.R`) stay in the code as fallbacks; with the current run they find no file and only print a message. $F$ is one region-wide, all-gear rate per FG.
4. **Taxonomic classification** — `worms_taxonomy_lookup()` for every species
   in `species_df`, dispatch-classified into `fish` / `mammal` / `seabird` /
   `phytoplankton` / `invertebrate` (default) by WoRMS Class, with a
   synonym→accepted-name resolution fallback for species WoRMS returns no
   Class for.
5. **Trait fetching (Step 2a–2f)** — `select_best_rows()` picks the best
   population-specific study per species from FishBase/SeaLifeBase
   multi-record tables, tiered: Western-Med locality keywords → broader
   Mediterranean keywords → most recent year → first available.
   `extract_provenance()` captures Locality/Year/Reference citation for each
   selected row (used later for the `References` sheet/CSV).
6. **Gap-filling** — `fill_by_taxonomic_proximity()`: local within-dataset
   donors first (Genus→Family→Order→Class, restricted to directly-computed
   values), then, for methods in `EXTERNAL_FORMULA_REGISTRY` (fish/mammal/
   general Gascuel formulas plus raw-trait identity mappings), an **external**
   FishBase/SeaLifeBase donor lookup — querying other species sharing the
   same taxonomic rank, not limited to species already in `species_df`.
7. **Taxon-specific PB/QB calculation** (dispatch by group) — the diagram
   below shows the actual dispatch and method-cascade logic:

   ```{=html}
   <div class="mermaid">
   %%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
   flowchart LR
       T["worms_taxonomy_lookup()&lt;br/&gt;WoRMS Class"] --> F["calc_fish()&lt;br/&gt;PB: 6 M methods + Fmort&lt;br/&gt;QB: 4 methods"]
       T --> M["calc_mammal()&lt;br/&gt;PB: Siler surv. to Gascuel&lt;br/&gt;QB: Innes&amp;Trites to Q/P=3"]
       T --> S["calc_seabird()&lt;br/&gt;PB: Gascuel only&lt;br/&gt;QB: Nilsson-Nilsson (no Q/P=3)"]
       T --> I["calc_invertebrate() (default)&lt;br/&gt;PB: Tumbiolo/Brey to Gascuel&lt;br/&gt;QB: Q/P=3 only"]
       F --> D["fill_by_taxonomic_proximity()&lt;br/&gt;local then external donors"]
       M --> D
       S --> D
       I --> D
       D --> AGG["FG-level aggregation&lt;br/&gt;biomass-weighted mean"]
       AGG --> OUT["Ecobase gap-fill -&gt; Ecopath_PBQB&lt;br/&gt;(PB_QB/PB_QB_spp/Ecobase CSVs)"]

       classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
       classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
       classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
       classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
       class T gray
       class F,M,S,I purple
       class D teal
       class AGG,OUT coral
   </div>
   ```

   *Phytoplankton is flagged by `worms_taxonomy_lookup()` but not computed
   (needs separate production-sampling data, out of scope for this
   pipeline) — not shown above.*

   * `calc_fish()` — PB from 6 independent M methods (Pauly 1980, FishLife,
     Gascuel 2008, Hoenig 1983, Then et al. 2015, Alverson & Carney 1975) +
     Fmort; QB from 4 methods (Palomares & Pauly 1998 Eq.27/Eq.26, Christensen
     & Pauly 1992, Q/P=3 fallback). Chosen value via priority order or mean
     (`PB_QB_SELECTION_MODE`).
   * `calc_mammal()` — PB via Siler-survivorship (Barlow & Boveng 1991),
     fallback Gascuel (2008); QB via Innes & Trites (1997), fallback Q/P=3.
   * `calc_seabird()` — PB via Gascuel (2008) only; QB via Nilsson & Nilsson
     (1976), **no** Q/P=3 fallback (that heuristic is fish/ectotherm-derived
     and would badly underestimate endotherm QB).
   * `calc_invertebrate()` — PB via Tumbiolo & Downing (1994), Brey (1999),
     fallback Gascuel (2008); QB via Q/P=3 only (Brey's 2012 ANN model
     flagged as non-automatable — top-10 highest-biomass invertebrates
     exported instead for manual cross-check).
   * Phytoplankton — flagged, not computed (needs separate production-
     sampling data, out of scope for this pipeline).
8. **FG-level aggregation** — biomass-weighted mean PB/QB per FG, with
   coverage/count audit columns (`n_species_with_PB`, `biomass_coverage_PB`,
   etc.).
9. **EcoBase literature merge** (if `ecobase_literature_pb_qb_simple.csv`
   exists) — averages multiple matching literature models per `FG_name`,
   flags >2×/<0.5× divergence from the empirical estimate, gap-fills only
   where the empirical value is `NA`. Writes sheet **`Ecobase`** (one row per
   FG, averaged PB/QB across every matching model, plus a `References`
   column listing each contributing model) **and `Ecobase_by_Model`** (one
   row per FG × individual published model — `FG_num`, `FG_name`,
   `EwE_model`, `year`, `authors`, `PB`, `QB` — the un-averaged detail behind
   `Ecobase`'s own per-FG mean, since a single FG typically has several
   different published models backing it and the averaged sheet alone loses
   which model said what).
10. **FG-level F top-up** — FGs with no species-level F (invertebrates, mammals, seabirds) get $P/B_f = P/B^{emp}_f + F_f$ from `F_by_fg.csv`; for all-invertebrate FGs $Q/B_f$ is recomputed as $3\,P/B_f$. Written to `fg_pb_qb_weighted_with_F.csv` (`PB_FG_before_F`, `QB_FG_before_F`, `F_FG`). Added on 2026-10-07; before, this step only ran with the retired `fg_catch_csv` input, so fished invertebrate FGs had P/B without $F$. ASSUMPTION: the empirical invertebrate P/B is treated as natural mortality.
11. **Workbook write** — `add_pbqb_to_ecopath_workbook()` writes sheet
    **`PB_QB`** (one row per FG — the biomass-weighted rollup), and, since
    `results` (the species-level table) is passed in as its
    `species_pb_qb` argument, also writes sheet **`PB_QB_spp`** (one row
    per species, mirroring `FG_spp_Ecopath`'s shape — every species that
    went into computing `PB_FG`/`QB_FG`, with three proportion columns:
    `prop_biomass_FG` — share of the FG's total biomass, regardless of
    whether that species got a PB/QB; `prop_biomass_PB`/`prop_biomass_QB`
    — that species' actual weight inside `PB_FG`'s/`QB_FG`'s own
    biomass-weighted mean, `NA` if that species' own PB/QB didn't
    compute. This is the sheet to check *which* species and *how much*
    of each FG's biomass a given `PB_FG`/`QB_FG` number was built from —
    without it, that detail only existed outside the workbook, in
    `species_pb_qb_by_taxon_group.csv`). If `01_biomass.R` already
    wrote a `PB_QB_spp` **scaffold** into this workbook (Species/FG_num/
    Biomass/`prop_biomass_FG` only, no real PB/QB yet), this call replaces
    it wholesale with the full sheet — `upsert_workbook_sheets()` always
    fully overwrites a same-named sheet, so no separate merge step is
    needed. A separate `upsert_workbook_sheets()`
    call writes sheet **`References`** (later renamed `PB_QB_References_`
    by `finalize_workbook_sheet_order()`, run from Script 1).
12. **Diagnostic plots** — per-method PB/QB comparison plots for fish, plus
    paginated species-level and FG-level PB/QB comparison plots across every
    dispatch group (auto-paginated when a plot would exceed `ggsave`'s
    50-inch height limit).
13. **Ecopath-ready CSV export** — `ecopath_ready_PB_QB.csv`, formatted to
    match a real Ecopath Basic Input CSV's exact column layout and
    European comma-decimal number formatting (verified against a real
    exported file, `westernmed90s-Basic_input.csv`).

## Outputs

* **CSVs (all under `output/pbqb-traits/`, not the top-level
  `output/`):** `invertebrates_for_brey_manual_check.csv`,
  `species_pb_qb_by_taxon_group.csv`, `fg_pb_qb_weighted.csv`,
  `phytoplankton_needs_separate_method.csv`,
  `fg_pb_qb_weighted_with_ecobase.csv`, `fg_pb_qb_weighted_with_F.csv`,
  `species_parameter_references.csv` (species-level parameter provenance —
  not the same table as the workbook's `FG_References` sheet, see below),
  `ecopath_ready_PB_QB.csv`, native `PB_QB.csv`/`PB_QB_spp.csv`.
* **Plots:** `fish_PB_methods_comparison.png`,
  `fish_QB_methods_comparison.png`,
  `species_PB_QB_comparison[_pageN].png`, `FG_PB_QB_comparison[_pageN].png`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, final sheets written:
  `Ecopath_PBQB`, `Ecopath_traits`. (`PB_QB`/`PB_QB_spp`/`Ecobase` are
  native/intermediate CSVs only now, not workbook sheets — see "Excel
  workbook functions" in the shared-library section below.)

---

# Shared library — `lib_survey_fg_density_functions.R`

Sourced by `01_biomass.R`, `02_fisheries.R`, `03_pbqb-traits.R` and `04_diets.R`; has **no top-level driver code** of its
own — every symbol below is a function definition only.

## FG-matching functions

Every observed species runs through this cascade, in order, until it
resolves to a Functional Group or falls through to manual review — the
diagram shows the actual fallback order implemented in the code below
(same style as `medbs_pipeline_methodology.qmd`'s `@sec-fg-matching`
diagram; each stage only attempts species the previous one left
unmatched).

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["match_species_to_fg()&lt;br/&gt;direct ScientificName match"] --> A2["match_nominate_subspecies_fg()&lt;br/&gt;strip repeated trinomial word, retry"]
    A2 --> A3["apply_seed_fg_rules()&lt;br/&gt;manual rules, zero-reference FGs"]
    A3 --> A4["fallback_match_fg_by_taxonomy()&lt;br/&gt;Genus to Phylum, exactly-one-FG rule"]
    A4 --> A5["summarize_unresolved_species()&lt;br/&gt;manual review"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    class A1,A2,A3,A4 purple
    class A5 gray
</div>
```

| Function | Purpose | Returns |
|---|---|---|
| `validate_survey_data(dataframe1, dataframe2, strata = TRUE)` | Fail-fast schema/uniqueness check on the two standardized input tables (required columns present, `SampleID` unique per survey) | `invisible(TRUE)` |
| `prepare_fg_lookup(dataframe2)` | Deduplicates species→FG reference; keeps genuine stanza splits (a species exclusively mapped to ≥2 FGs), excludes/flags ambiguous multi-FG species | deduped `data.table(ScientificName, FG_num, FG_name)` |
| `match_species_to_fg(dataframe1, fg_lookup_safe)` | Direct left-merge by `ScientificName` | merged `dt` |
| `match_nominate_subspecies_fg(dt, fg_lookup_safe)` | Strips a repeated trinomial word (e.g. "Diplodus sargus sargus" → "Diplodus sargus") and retries the match | `dt` with newly resolved rows |
| `fetch_taxonomy(species_names, taxonomy_source = "worms")` | Dispatches to `fetch_taxonomy_worms()`, `fetch_taxonomy_fishbase()`, or both | `data.table(ScientificName, Genus, Family, Order, Class, Phylum)` |
| `fallback_match_fg_by_taxonomy(dt, fg_lookup_safe, taxonomy_source = "worms", rank_levels = c("Genus","Family","Order","Class","Phylum"))` | Data-driven rank fallback — a rank value is used only if it maps to exactly one FG among reference species | `dt`, with `attr(dt,"fetched_taxonomy")` and `attr(dt,"still_unresolved_taxonomy")` attached |
| `apply_seed_fg_rules(dt, dataframe2, seed_rules, species_exceptions = NULL)` | Explicit small rule table for FGs with **zero** reference species (nothing for the automatic fallback to learn from); species-level exceptions applied first | `dt` |
| `summarize_unresolved_species(dt, taxonomy = NULL)` | Mean biomass + appearance count per still-unmatched species, sorted by appearance count | summary `data.table` |
| `build_fg_manual_review_template(csv_out_dir, out_path = file.path(csv_out_dir, "FG_manual_review_template.xlsx"))` | Combines `survey_unmatched_for_manual_review.csv` + `medias_unmatched_for_manual_review.csv` into one de-duplicated review workbook, with blank `GF`/`FG_name` columns ready to hand-fill | invisibly returns the combined `data.table`; writes an `.xlsx` |

`build_fg_manual_review_template()` is a one-off, run-it-yourself helper, not
part of the automated pipeline — nothing calls it from `01_biomass.R` itself.
After a run leaves species unmatched, call it against that run's
`output/biomass/` folder to get one workbook (a `README` sheet plus a
`For_review` sheet, sorted by how often each species actually showed up in
the data — review the highest-impact species first). Its `ESPECIE`/`GF`/
`FG_name` columns deliberately use the exact same names as `FG_WMed_2026.csv`'s
own sheet 4 (`fg_wmed_95`) — fill in `GF`/`FG_name` for whichever species you
want to resolve, then copy columns A:C straight into a **new** version of
that sheet (never overwrite the original in place), and point `fg_file` at
the new file for the next run. There's no code path that writes the filled-in
review back into `FG_WMed_2026.csv` automatically — that hand-off is deliberately
manual, since it's a genuine judgment call about which FG a given species
belongs to.

| `snapshot_full_extent_species(matched_dt, species_taxonomy, source_label, csv_out_dir, sci_name_col = "ScientificName")` | One row per distinct species a source's own FG-matching step saw, taken BEFORE that source's `FILTER_AREAS`/year restriction — i.e. every species anywhere in the raw file(s) that source loaded, independent of this run's config. Called once each for MEDITS (Steps 3-4), MEDIAS (Step 9), and stock assessment | writes `all_species_<source>_full_extent.csv` |
| `build_full_fg_species_catalog(csv_out_dir, out_path = file.path(csv_out_dir, "FG_WMed_full_species_catalog.csv"))` | Combines the three `all_species_*_full_extent.csv` snapshots above into ONE species catalog — matched species (a real `FG_number`/`FG_name`) and still-unmatched species (`FG_number`/`FG_name` blank) together, with a `sources` column (e.g. `"MEDITS+MEDIAS"`) showing which source(s) actually observed each one. Note the column is named `FG_number` here (not `GF`, as in `FG_WMed.xlsx`/`build_fg_manual_review_template()` above) since this catalog spans all three sources rather than being a direct fill-in template for the Excel sheet | invisibly returns the combined `data.table`; writes a CSV |

`01_biomass.R` calls `build_full_fg_species_catalog(csv_out_dir)` itself,
near the very end (Step 14, right after `trim_workbook_to_final_sheets()`)
— **`FG_WMed_full_species_catalog.csv`** lands in `output/biomass/` on every
run without any extra step. It's a candidate "sheet 4" replacement covering
every species — matched (`status = "matched"`, already correctly resolved)
and unmatched (`status = "needs review"`, blank) — in one file, sorted with
matched rows first. Same read-only relationship to `FG_WMed_2026.csv` as
`build_fg_manual_review_template()` above — nothing writes back into it
automatically.

Unlike the matched/unmatched CSVs everything else in this script reads and
writes, the three `all_species_*_full_extent.csv` snapshots — and so this
catalog — are **NOT scoped by `FILTER_AREAS`/`YEAR_ECOPATH`/`TS_YEARS`**, by
construction: `snapshot_full_extent_species()` is called right after each
source's own FG-matching step and *before* that source's own area
restriction (MEDITS: before `dt <- dt[AreaID %in% FILTER_AREAS]`; MEDIAS:
`acoustic_matched` itself is never restricted in place, only derived summary
tables are; stock assessment: `star_wm` is already the whole West Med
subregion, not this run's `FILTER_AREAS`). So this catalog always covers the
**whole available survey extent** — every GSA and year the raw MEDITS/
MEDIAS/stock-assessment files actually contain — on every run, regardless
of what region/years that particular run happens to be configured for. A
source's snapshot (and its contribution to the catalog) simply won't exist
for a run where that source didn't execute at all — `AREA_MODE == "custom"`
skips MEDIAS and stock assessment entirely, so a custom run's catalog is
MEDITS-only there, which is a real limit of that mode (no MEDIAS/stock-
assessment data exists to catalog), not something a wider `FILTER_AREAS`
would fix.

## Area / stratum / density functions

| Function | Purpose | Returns |
|---|---|---|
| `match_samples_to_area(dt, area_shp, area_id_col)` | Spatial join (`sf::st_join`) of sample points to polygons; no-op if `AreaID` already fully populated | `dt` |
| `assign_depth_stratum(dt, strata_def)` | Bins `Depth` into strata via `findInterval()`; out-of-range depths → `Stratum = NA` | `dt` |
| `filter_samples_by_area(dt, area_filter, filter_type = c("shapefile","bbox"), points_crs = 4326)` | Custom sub-region filter, independent of GSA/`AreaID` | `dt` filtered to in-area `SampleID`s |
| `compute_sample_densities(dt)` | `Density := Biomass / Effort` | `dt` |
| `apply_catchability_correction(dt, catchability_table, default_q = 1, species_taxonomy = NULL, exempt_fg_names = NULL, exempt_species = NULL)` | 5-tier resolution: direct species → genus "spp" entries → explicit broad-rank entries → taxonomic-proximity fallback → `default_q`; exemptions override last. Divides `Density`/`Biomass` by resolved `q` | `dt` |
| `remove_sample_outliers(dt, threshold = 70, min_samples = 5, drop_outliers = TRUE, log_scale = TRUE)` | Per-species/per-area robust z-score (median/MAD) on `log1p(Density)`; sets flagged `Density`/`Biomass` to `NA` if `drop_outliers = TRUE` | `dt`, with `attr(dt,"flagged_outliers")` |
| `compute_fg_densities_by_stratum(dt, strata = TRUE)` | Sums species density within FG per sample, then across samples by area/year/[stratum]/FG | `per_group_fg` (`density_sum`) |
| `compute_strata_area_by_area(area_ids, area_shp, area_id_col, strata_def, cache_path = NULL, resolution = 1)` | Downloads NOAA bathymetry per area, bins cell area into depth strata (latitude-corrected); optional CSV cache | `strata_area_by_area` table |
| `weight_by_strata(per_group_fg, n_samples_by_stratum, strata_area_by_area)` | Core strata-weighted estimator: `D_area = Σ_h (density_sum_h/n_samples_h) · area_prop_h` | `fg_index`, with `attr(fg_index,"per_stratum")` |
| `simple_area_density(per_group_fg, n_samples_by_area)` | Unstratified path (`strata = FALSE`) | `data.table` |
| `weight_by_area(fg_index_stratified)` | Region-wide weighting from absolute stratum areas (`per_stratum` attribute); denominator = area of all cells sampled that year (`sampled_area_by_year` attribute), so zero catches count (fixed 2026-10-07) — requires the stratified path upstream | `fg_index_regional` |
| `sampled_area_by_year(n_samples_by_stratum, strata_area_by_area)` | Area of the (area, stratum) cells with ≥ 1 haul per year — denominator of the region-wide means | `data.table(Year, sampled_area_km2)` |
| `flag_ts_upward_spikes(year, x, half = 2, k = 3, min_fold = 3)` | One-sided Hampel filter on log values; used by `export_ecopath_ecosim_excel(spike_filter = TRUE)` to blank isolated spikes in Ecosim B series | logical vector |
| `compute_species_densities_by_stratum(dt, strata = TRUE)` / `weight_species_by_area(per_group_sp, n_samples_by_stratum, strata_area_by_area)` | Species-level analogs of the two functions above, feeding the `FG_spp` sheets | `data.table(Year, FG_num, FG_name, ScientificName, mean_density)` |
| `compute_cv_log_by_fg(value_dt, value_col, min_replicates = 2)` | `cv_log = sd(log1p(x))/mean(log1p(x))` per FG/Year, averaged across years — feeds the Ecosim "Weight" row | `data.table(FG_num, FG_name, cv_log, n_years)` |

## Plotting functions

| Function | Produces |
|---|---|
| `plot_sample_map(dt, area_shp, area_id_col, ..., by_year = FALSE, selected_areas = NULL, land_on_top = FALSE)` | Sample coverage map, points colored inside/outside the study area, polygons shaded by sample intensity |
| `plot_fg_timeseries_by_area(fg_index, ...)` / `plot_fg_timeseries_regional(fg_index_regional, ...)` | Faceted FG density time series, per-area vs. region-wide |
| `plot_strata_profile(fg_index_stratified, strata_def, ...)` | Depth-strata validation bar chart (pre-area-weighting, per-sample density by stratum) |

## Excel workbook functions

`ecopath_ecosim_inputs.xlsx`
carries **only the 10 final target sheets** — every native/intermediate
table each block used to also keep as a workbook sheet is now CSV-only
(see "Per-block output CSV subfolders" and "Final target sheet order"
below). The table below reflects the current function signatures, not the
older ones this document used to describe.

| Function | Purpose |
|---|---|
| `upsert_workbook_sheets(sheets, out_path)` | Adds/replaces the named sheets in `out_path`, **leaves every other sheet untouched**. Creates the workbook if it doesn't exist yet. Safe to call repeatedly, from any script, in any order. |
| `write_native_sheet_csv(dt, name, out_dir)` / `write_native_sheets_csv(sheets, out_dir)` / `read_native_sheet_csv(name, out_dir)` | Generic CSV read/write for a block's own native/intermediate tables — `out_dir` is always passed explicitly by the caller (the block's own subfolder), never assumed |
| `read_existing_ts_years(out_path, sheet_name, csv_dir = NULL)` | Reads year values from column 1 of an existing native CSV sheet (e.g. `Ecosim`), so `Catches_Ecosim`'s years stay aligned with it. `csv_dir` defaults to `dirname(out_path)` for back-compat; the numbered scripts pass the correct block's subfolder explicitly |
| `read_full_fg_reference(out_path, csv_dir = NULL)` | Reads `FG_lookup.csv` (a biomass-block output) — the full FG_num/FG_name reference, used to expand PB_QB/FG_References/etc. to every model FG, not just the ones with data this run |
| `export_ecopath_ecosim_excel(fg_index_regional, species_density_regional, n_samples_by_area_year, dataframe2, year_ecopath, ts_years, out_path, species_taxonomy = NULL, fg_cv_log = NULL, normalize_ts = TRUE, csv_out_dir = NULL)` | Writes the final `Ecopath_B` sheet directly to the workbook, plus native/intermediate CSVs (`FG_spp_Ecopath`, `FG_lookup`, `Ecosim`, `FG_spp_Ecosim`, a `PB_QB_spp` scaffold) into `csv_out_dir` (01_biomass.R's own `output/biomass/` subfolder) |
| `add_catches_to_ecopath_workbook(fg_catch, fg_lookup, out_path, year_ecopath, area_km2, ts_years = NULL, fleet_structure = NULL, csv_out_dir = NULL, biomass_csv_dir = NULL)` | Converts raw catch tonnes to t/km² density; writes native CSVs `Catches_Ecopath`/`Catches_Ecosim`/`Fleet_Structure` into `csv_out_dir` (02_fisheries.R's `output/fisheries/`). Reads `Ecosim.csv` from `biomass_csv_dir` when `ts_years` isn't passed explicitly |
| `add_pbqb_to_ecopath_workbook(fg_weighted, out_path, species_pb_qb = NULL, csv_out_dir = NULL, biomass_csv_dir = NULL)` | Writes native CSVs `PB_QB`/`PB_QB_spp` into `csv_out_dir` (03_pbqb-traits.R's `output/pbqb-traits/`). Reads `FG_lookup.csv` from `biomass_csv_dir` to expand to every model FG |
| `finalize_ecopath_ecosim_summary_sheets(out_path, year_ecopath, biomass_csv_dir = NULL, fisheries_csv_dir = NULL)` | Builds the final `Ecosim_ts` sheet by combining native CSVs from **two** blocks — `Ecosim` (biomass) plus `Catches_Discards_FG_ts`/`Catches_Ecosim`/`Fishing_Effort_by_Fleet` (fisheries) — hence the two separate directory arguments. Guards against a blank/non-numeric Year cell in `Ecosim.csv` (which would otherwise crash the relative-scaling-reference comparison) by falling back to raw, un-rescaled kW-days for that fleet with an explicit message. |
| `build_fg_references_sheet(out_path, biomass_csv_dir = NULL, fisheries_csv_dir = NULL, pbqb_csv_dir = NULL, diet_csv_dir = NULL)` | Builds the final `FG_References` sheet — one row per FG, consolidating which data source fed each block (see "The FG_References sheet" below) |
| `trim_workbook_to_final_sheets(out_path)` | **Run last**, by every numbered script. Reduces the workbook to exactly the 10 final target sheets (see below), dropping any native/intermediate sheet a much older run of this workbook might still be carrying |

The diagram below shows which script/function writes which sheet into the
one shared workbook, and the final trim that runs after each script.

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    S1["01_biomass.R&lt;br/&gt;export_ecopath_ecosim_excel()"] --> W["ecopath_ecosim_inputs.xlsx&lt;br/&gt;(shared workbook, top-level out_dir)"]
    S2["02_fisheries.R&lt;br/&gt;add_catches_to_ecopath_workbook()"] --> W
    S3["03_pbqb-traits.R&lt;br/&gt;add_pbqb_to_ecopath_workbook()"] --> W
    S4["04_diets.R&lt;br/&gt;upsert_workbook_sheets()&lt;br/&gt;(Ecopath_diet)"] --> W
    S5["build_fg_references_sheet()&lt;br/&gt;(run by 04_diets.R, last)"] --> W
    W --> F["trim_workbook_to_final_sheets()&lt;br/&gt;run after every script"]
    F --> O["output/ecopath_ecosim_inputs.xlsx&lt;br/&gt;(10 final sheets only)"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class S1,S2,S3,S4,S5 purple
    class W gray
    class F,O coral
</div>
```

## Per-block output CSV subfolders

Every native/intermediate CSV a block writes now lands in that block's own
subfolder under the shared `out_dir`, instead of a single flat directory
shared by all scripts:

* `01_biomass.R` (+ `01b`) → `output/biomass/` (plots `output/plots/biomass/`)
* `02_fisheries.R` → `output/fisheries/` (plots `output/plots/fisheries/`)
* `03_pbqb-traits.R` → `output/pbqb-traits/`
* `04_diets.R` → `output/diet/` (plots `output/plots/diets/`)
* `05_validation.R` → `output/validation/` (plots `output/plots/validation/`, including the P/B–Q/B, F and P/Q plots built from 03's CSVs)
* a custom-region run (`AREA_MODE = "custom"`) writes to `output/<AREA_NAME>/`

`ecopath_ecosim_inputs.xlsx` itself stays at the top-level `output/` — all
four scripts read and write it, so it has no single owning block. Each
script defines its own `csv_out_dir` near the top and passes it (or a
`*_csv_dir` pointing at a *different* block's subfolder, for the handful of
cross-block reads — species density, strata area, `FG_lookup.csv`,
`Ecosim.csv`, `biomass_proportion_by_species_fg.csv`, catch time series)
into the library functions above. Every one of those new directory
arguments defaults to `dirname(out_path)` (the workbook's own top-level
folder) when omitted, so nothing breaks if some other caller doesn't pass
them.

## Final target sheet order (11 sheets)

`target_order`, used by `trim_workbook_to_final_sheets()` / `finalize_workbook_sheet_order()` at the end of every numbered script (order confirmed by the 2026-10-07 logs):

1. `info` — run metadata (areas, countries, YEAR_ECOPATH, TS_YEARS, …)
2. `FG_spp` — FG × species × taxonomy and biomass share, written by `01_biomass.R`
3. `Ecopath_B` — FG biomass (t/km²), 1994–1996 mean, with a `Reference` column
4. `Ecopath_L` — FG × fleet landings, `02_fisheries.R`
5. `Ecopath_Di` — FG × fleet discards, `02_fisheries.R`
6. `Ecopath_PBQB` — FG P/B, Q/B and their sources, `03_pbqb-traits.R`
7. `Ecopath_traits` — species traits under FG header rows, `03_pbqb-traits.R`
8. `Ecopath_diet` — predator FG × prey FG, `04_diets.R`
9. `Ecosim_ts` — B, catch and effort series
10. `FG_References` — per-FG sources (`B_ref`, `L_ref`, `Di_ref`, `PBQB_ref`, `PBQB_method_ref`, `traits_ref`, `diet_ref`), built last by `04_diets.R`
11. `Ecobase` — EcoBase values used for comparison (reference only)

Every native/intermediate table (`Ecopath`, `Catches_Ecopath`, `PB_QB`,
`Ecosim`, `Catches_Ecosim`, `Fishing_Effort_by_Fleet`, `FG_spp_Ecopath`,
`FG_spp_Ecosim`, `FG_lookup`, `traits_ewe`, `fg_traits_weighted`,
`PB_QB_spp`, `References` [species-parameter-level, `03_pbqb-traits.R`'s
own audit CSV — a different table from the `FG_References` workbook sheet
above, despite the similar name], `Ecobase`, `Catches_Discards_FG_ts`,
`diet_references_by_fg`, ...) lives only as CSV, in its own block's
subfolder, never as a workbook sheet — `trim_workbook_to_final_sheets()`
drops anything else it finds regardless.

## The FG_References sheet

Built by `build_fg_references_sheet()`
(`lib_survey_fg_density_functions.R`), run by `04_diets.R` after every other
script has had a chance to run. One row per FG, with columns:

| Column | What it summarizes | Read from |
|---|---|---|
| `FG_num`, `FG_name` | The full FG reference | `FG_lookup.csv` (`output/biomass/`) |
| `B_ref` | Which biomass source (MEDITS/MEDIAS survey vs. stock assessment) fed `Ecopath_B`/`Ecosim_ts`'s B block for this FG | `biomass_source` in `survey_fg_annual_index_regional_combined.csv` (`output/biomass/`) |
| `L_ref`, `Di_ref` | Which catch/discard source (GFCM/FDI/SAU-derived vs. stock assessment) fed `Ecopath_L`/`Ecopath_Di`/`Ecosim_ts`'s catch time series for this FG | `catch_source` in `Catches_Discards_FG_ts.csv` (`output/fisheries/`) |
| `PBQB_ref`, `traits_ref` | Whether `PB_FG`/`QB_FG` came from the empirical species-level calculation or an EcoBase literature gap-fill | `PB_source`/`QB_source` in `PB_QB.csv` (`output/pbqb-traits/`), falling back to a generic "empirical" note when EcoBase gap-filling didn't run |
| `diet_ref` | Which diet studies (metaweb references, MAISHA study citations, FishBase/SeaLifeBase references) fed this FG's column in `Ecopath_diet` | `diet_references_by_fg.csv` (`output/diet/`), rolled up from `build_species_diet()`'s per-predator citations via the same species→FG mapping the diet matrix itself uses |
| `PBQB_method_ref` | The calculation METHOD/literature behind `PB_FG`/`QB_FG` (as opposed to `PBQB_ref`, which is about the *data*, not the method) | `dispatch_group` in `PB_QB_spp.csv` (`output/pbqb-traits/`), mapped to a literature citation |

Each column is a best-effort summary (distinct values `" | "`-joined when a
FG drew on more than one source across years/species/studies), not a full
audit trail — the individual block's own CSV named in the table above has
the complete detail. A column is left blank for a FG when the block behind
it hasn't run yet against this workbook, with a message saying so rather
than an error; running `build_fg_references_sheet()` again later (as
`04_diets.R` does every run) fills in whatever is available by then.

---

> **Note:** the detailed guide for adapting this pipeline to a different region, and the standalone library/cartography files, are moved to [Appendix: Adapting this pipeline & other library files] at the end of this document, out of the way of the main demo walk-through.

---


# Script 5 — `04_diets.R`

**Role:** Step 4. Builds the predator FG × prey FG diet matrix and the `FG_References` sheet. Auto-runs on `source()`. Sources `lib_survey_fg_density_functions.R`, `lib_worms_taxonomy_lookup.R`, `lib_fishbase_diet_matrix.R` and `03b_ecobase.R`. Must run after `01_biomass.R` (species biomass shares and `Ecopath_B`).

## Configuration

| Variable | Default | Meaning |
|---|---|---|
| `METAWEB_XLSX_PATH` | `data/Complementary data/data_entry_metaweb_empty.xlsx` | metaweb (empty template) |
| `DIET_FALLBACK_ENABLE_MAISHA` | `TRUE` | use MAISHA |
| `MAISHA_PATH` | first existing of `data/diet/MAISHA/MAISHA_DwC-A`, `…/MAISHA_DwC-A.zip`, any `maisha*.xlsx/csv` under `data/diet` (case-insensitive) | MAISHA location |
| `DIET_FALLBACK_ENABLE_FISHBASE` | `TRUE` | use FishBase/SeaLifeBase |
| `DIET_FISHBASE_METHOD` | `"fg_matrix"` | build the FG matrix with `lib_fishbase_diet_matrix.R` |
| `DIET_FALLBACK_ENABLE_ECOBASE` | `TRUE` | try EcoBase diet matrices |
| `DIET_TAXON_SPLIT` / `DIET_CATEGORY_SPLIT` | `"equal"` / `"biomass"` | split of a prey taxon / food category over its FGs |
| `DIET_DETRITUS_MAX_TL` | 3 | detritus dropped for predators with TL ≥ 3 |
| `DIET_METRIC_PRIORITY` | `diet_metric_priority.csv` | IRI > %W > %N > %FO > presence |

## Inputs

| File | Role |
|---|---|
| `<pcloud_dir>/data/FG_WMed_2026.csv` | species → FG and taxonomy |
| `<pcloud_dir>/data/Complementary data/data_entry_metaweb_empty.xlsx` | metaweb (`DATA_ENTRY`, `Taxonomic_codes`, `Non_taxonomic_groups`) |
| `<pcloud_dir>/data/diet/MAISHA/MAISHA_DwC-A` | MAISHA v1.2 Darwin Core Archive: `occurrence.txt` (one "Predator" record per study × predator, `occurrenceRemarks == "Predator"`, with citation, date and coordinates, listing its prey records in `associatedOccurrences`; prey records carry the WoRMS `scientificName`) and `extendedmeasurementorfact.txt` (per prey: gravimetric %, numerosity %, frequency of occurrence; per predator: the unknown/debris/detritus fractions — detritus kept as a prey item, unknown and debris dropped) |
| FishBase/SeaLifeBase (`rfishbase`) | DIET, DIETITEMS, FOODITEMS; cached as `output/diet/fishbase_diet_raw_*.csv` |
| `<pcloud_dir>/data/Complementary data/diet_reference_tables/fishbase_food_category_to_fg.csv` | food-category regex → FG rules (about 44), including non-food exclusions |
| `output/biomass/biomass_proportion_by_species_fg.csv`, workbook `Ecopath_B` | $b_{s|f}$ and $B_j$ |
| `<pcloud_dir>/data/ewe_group_table.csv` (optional) | predator flag override |

## Method

See [DIET — Ecopath_diet sheet / diet matrix] above and `medbs_pipeline_methodology.qmd` (Diet Matrix) for the equations. In short: tier per predator FG (metaweb → MAISHA → FishBase/SeaLifeBase → EcoBase); study proportions by the first available metric; prey → FG by species, then taxonomic rank (equal split), then food category (biomass split), then a WoRMS second pass (`worms_diet_prey_cache.rds`); eggs and larvae → zooplankton; detritus dropped for TL ≥ 3; studies averaged; species weighted by biomass share; columns renormalised to 1.

## Outputs (in `output/diet/`)

| File | Content |
|---|---|
| `MAISHA/`, `FishBase/` | each source alone: species-level diets, FG matrix (long/wide), item mapping and unassigned items (REVIEW) |
| `external_diet_matrix_fg_long.csv`, `external_diet_matrix_fg_wide.csv` | MAISHA + FishBase combined (used in `Ecopath_diet`) |
| `external_diet_item_mapping_REVIEW.csv`, `external_diet_unassigned_items_REVIEW.csv` | how each prey item was assigned / not assigned |
| `diet_provenance_by_predator_fg.csv`, `diet_references_by_fg.csv` | tier and studies per FG |
| `diet_still_missing_REVIEW.csv`, `diet_cannibalism_REVIEW.csv`, `diet_matrix_column_sum_REVIEW.csv` (only written when a column ≠ 1) | checks |
| `diet_composition_ewe.csv` | EwE import format |
| workbook `Ecopath_diet`, `FG_References` | final sheets |
| `plots/diets/diet_matrix_heatmap.png` | heatmap |

## Key functions

| Function | File | Role |
|---|---|---|
| `maisha_diet_records()` | `lib_fishbase_diet_matrix.R` | reads MAISHA (DwC-A folder/zip, or a flat table), tags Mediterranean studies by coordinates (lon −6 to 36.5, lat 30 to 46; non-Mediterranean studies are dropped when the predator has Mediterranean ones), returns study × predator × prey rows with W/N/FO |
| `fb_fetch_diet_tables()`, `fishbase_diet_records()` | `lib_fishbase_diet_matrix.R` | FishBase/SeaLifeBase tables with per-server DietCode joins; Mediterranean studies first |
| `diet_resolve_prey()` | `lib_fishbase_diet_matrix.R` | prey item → FG weights $w_{k \to j}$ |
| `diet_records_to_fg_matrix()` | `lib_fishbase_diet_matrix.R` | species diets → FG matrix, with the WoRMS second pass |
| `write_diet_source_outputs()` | `lib_fishbase_diet_matrix.R` | writes each source's files |
| `build_species_diet()`, `build_fg_diet()` | `04_diets.R` | metaweb tier |
| `export_ewe_matrix_csv()`, `build_ecopath_diet_sheet()` | `04_diets.R` | exports |
| `build_fg_references_sheet()` | `lib_survey_fg_density_functions.R` | `FG_References` |

---

# Script 6 — `05_validation.R`

**Role:** Step 5. Reads the workbook and the REVIEW files of 01–04; changes no input sheet.

- **STEP A** — compares `Ecopath_B`, `Ecopath_PBQB` (and L/Di when the reference has them) with the "estimates" sheet of the 1995 Western Mediterranean model (`WMed_EwE.xlsx`, $A_{old}$ = 846,002 km²). Old → new groups via `validation_reference_tables/old_to_new_fg_crosswalk.csv` (many → one summed, one → many compared with the sum). Old densities are also expressed on the new area, $X_{old}\,A_{old}/A_{new}$ (same total tonnes). Flags ratios outside 0.5–2. Plots: `biomass_vs_old_model_by_fg.png` (both values side by side, same total tonnes) and `…_as_published.png`.
- **STEP B** — `validation_summary_ALL.csv`: index of every check (P/B, Q/B, GE; P/Q; diet sums; cannibalism; missing diets; EcoBase evaluations; FGs without B; even species splits; temporal mismatches; fleet coverage; zero-catch FGs; unmatched names).
- **Block 11** — total zooplankton field check against Fernández de Puelles et al. 2023 (*Water* 15:4267): field 2–17 mg DM/m³ over the mean sampled depth $z_{eff}$ = 86.7 m, DW:WW 0.1–0.2 (ASSUMPTION), compared with FGs 68 + 69 + 71 + 72; salp upper bound with salp DW:WW 0.0804. Output `zooplankton_total_field_check_REVIEW.csv`.
- **STEP C** — one PNG per check and `plots/validation/validation_plots_ALL.pdf` (19 pages), also embedding the P/B–Q/B, F, P/Q and diet plots.

Outputs in `output/validation/`: `comparison_with_old_westmed_estimates_REVIEW.csv`, `biomass_vs_old_model_REVIEW.csv`, `old_workbook_unmatched_fg_names_REVIEW.csv`, `validation_summary_ALL.csv`, `zooplankton_total_field_check_REVIEW.csv`.

---

# Status of the 2026-10-07 run and open issues

All five scripts completed without errors (logs `20261007_094022_01_biomass_log.txt` to `20261007_100250_05_validation_log.txt`). Changes made after this run — **rerun 02 → 03 → 05**:

1. `02_fisheries.R` — F and the NEI F ≤ 1 cap use the final `Ecopath_B` (adult hake FG 27 now gets an F; macrozooplankton and suprabenthos are no longer capped against survey bycatch). Residual fleet column renamed "Target countries - fleet not resolved". "6 target countries" wording removed from messages.
2. `03_pbqb-traits.R` — FG-level F from `F_by_fg.csv` added to P/B of non-fish FGs with catch (cephalopods, shrimps, lobsters, molluscs, macrobenthos); invertebrate Q/B recomputed as 3 × P/B. Previously this step only ran with a retired catch file, so these FGs had P/B without fishing mortality.
3. `05_validation.R`, `01b_biomass_unsurveyed.R` — comments and the residual column name brought up to date.
4. `lib_survey_fg_density_functions.R`, `01_biomass.R` (evening of 2026-10-07) — **rerun 01 → 05**. Region-wide MEDITS and MEDIAS densities now count zero catches: the denominator is the area of all cells sampled that year, not only the cells where the FG was caught. This lowers `Ecopath_B` of patchy or depth-restricted FGs: in an offline test the 1994–1996 means were ×1.7 too high at the median and ×3 or more for 17 FGs. F, P/B and the balance change with them. New Ecosim trend rules (haul cap at q95, years with < 80 % area coverage blank, upward-spike filter) change only the B series, not `Ecopath_B`; see `ecosim_haul_cap_effect_REVIEW.csv`, `survey_area_coverage_by_year.csv` and `ecosim_ts_spikes_removed_REVIEW.csv`.

Open issues, by block:

| Block | Issue | Where to look |
|---|---|---|
| F | Deep-water rose shrimp F = 5.86 (catch 3,713 t/yr, B 0.0036 t/km²) | `F_by_fg.csv`; MEDITS q; NEI decapod split |
| F | Benthic mollusc F ≈ 5 once Ecopath_B (0.022 t/km²) is used — catch 20,414 t/yr, probably largely from < 10 m and lagoons | `F_by_fg.csv`; decide whether this catch belongs to the model area |
| C/L | 9 FGs with < 50 % of landings in a named fleet; 24 FGs with species but zero catch; 564 zero-tonne rows with no fleet share | `ecopath_L_fg_coverage_check.csv`, `fg_missing_ecopath_L_Di_REVIEW.csv` |
| Effort | Italy Rousseau vs FDI r = −0.24 | 02 log |
| B | Cymodocea 14.7 > Posidonia 8.7 t/km² (no map, density may include rhizomes) | `habitat_shapefile_biomass.csv` |
| B | Meso + micro zooplankton 29.6 t/km², 2× the field maximum | `zooplankton_total_field_check_REVIEW.csv` |
| B | Macrozooplankton 0.124 (krill minimum) vs 30 in the old model; salps without B | Ecopath_B; `MACROZOO_SOURCE` |
| B | Hake assessments in RAM end in 2014; old adult hake 1.81 vs new 0.075 t/km² (same area) | `hake_stanza_stock_assessment_REVIEW.csv` |
| B | 123 species on an even within-FG split; pre-MEDIAS acoustic values to check against the PDFs | `fg_species_biomass_needs_review.csv`, `small_pelagics_per_gsa_series_REVIEW.csv` |
| PB/QB | 8 FGs without P/B or Q/B; 28 FGs outside P/Q 0.05–0.3; 23 FGs > 2× from EcoBase | `plausibility_flag_REVIEW.csv`, 03 log |
| DIET | No diet: corals and gorgonians, meso and micro zooplankton, suprabenthos | `diet_still_missing_REVIEW.csv` |
| DIET | Cannibalism: other macro-benthos 0.286, non-commercial decapods 0.101 | `diet_cannibalism_REVIEW.csv` |
| Data | `FG_WMed_2026.csv` lacks a `survey_exempt` column | 01 log |
| Output | `plots/pbqb-traits/` still holds copies of validation plots from an older layout (05 now writes to `plots/validation/`) | delete the old copies by hand |

---

# Methodology document — `medbs_pipeline_methodology.qmd`

A Quarto write-up of the method (R chunks `eval: false`, except the setup chunk that copies figures and outputs). It documents `lib_survey_fg_density_functions.R` and `01_biomass.R` in both `AREA_MODE`s (FG matching, study area, haul density, catchability, outliers, stratified and regional estimators, MEDIAS), then one section per script with its equations and references (Biomass with `01b`, Fisheries, PB/QB, Diet, Validation) and the status and open issues of the latest run. It embeds the GFCM GSA GeoJSON (from `export_gsa_boundaries.R`) in a Leaflet map with a client-side configuration generator. The setup chunk copies figures from `out_dir/plots/<block>/` (and `out_dir/<AREA_NAME>/plots/` for the custom example) into `figures/`, and key outputs into `downloads/`, so the rendered HTML is portable.

---

# Appendix: Adapting this pipeline & other library files

*(Moved here from the main body — useful for a future maintainer retargeting this pipeline to a different region, not needed to follow the demo.)*

## Other library files

### `lib_worms_taxonomy_lookup.R`

Library file, no top-level driver code — `source()`'d automatically by the
numbered scripts. Defines a single function:

**`worms_taxonomy_lookup(names_vector)`** — the single taxonomy source used
everywhere in the pipeline (so genus/family comparisons are always
apples-to-apples, never a FishBase-derived family compared against a
WoRMS-derived one for the same actual family). Strips `"Genus spp."`
suffixes so genus-level placeholders still resolve, chunks the query list
(50 names/chunk) against `worrms::wm_records_names()`, picks the best
(`status == "accepted"` preferred) match per name, and returns
`data.table(original_name, search_term, AphiaID, matched_name, rank, status,
phylum, class, order, family, genus)`. Explicitly namespaces `purrr::map()`
rather than calling `map()` bare — the `maps` package (loaded elsewhere in
this pipeline for basemap plotting) also exports an unrelated `map()`, and
whichever package attached most recently silently wins an unqualified call.

### `lib_build_species_df_from_survey.R` (retired — not in the repository)

No longer used: `03_pbqb-traits.R` builds `species_df` itself (see Script 4, step 2). Former description, kept for reference: library file — bridges `01_biomass.R`'s survey output and
`03_pbqb-traits.R`'s PB/QB estimation. Defines:

**`build_species_df_from_survey(survey_csv_path, year_ecopath_range,
out_rds_path = NULL)`** — reads `species_density_regional_combined.csv`,
restricts to `YEAR_ECOPATH`'s snapshot years (the same years used for the
`Ecopath` sheet's Biomass column, so PB/QB weighting and the reported Biomass
describe the same time snapshot), collapses to one mean-density `Biomass`
value per species (stops with an error if any species maps to more than one
FG within that year range — shouldn't happen given upstream FG-matching
logic), sets `Yield := NA` (this survey pipeline has no catch/landings data
of its own — F = Yield/Biomass isn't computable from it directly). Optionally
saves the result as an `.rds`. Returns
`species_df(Species, FG, FG_name, Biomass, Yield)`.

---

## Standalone cartography utilities (not part of the numbered pipeline)

### `export_gsa_boundaries.R`

Standalone — does not need `01_biomass.R` or the shared
library sourced first. Downloads the same GFCM GSA shapefile via the same URL
and GSA-numbering fix (`SMU_CODE` 111/112 → 11). **Western-Med-specific as
delivered:** restricts to GSA 1–11 — change the `gsa_num %in% 1:11` filter to
any other GSA range to export a different region's boundaries. Simplifies
the result for web embedding via
`rmapshaper::ms_simplify(keep = 0.05, keep_shapes = TRUE)` (keeps small
polygons like Alboran Island, GSA 2, from being deleted outright), and writes
`gsa_boundaries.geojson` — the file embedded in
`medbs_pipeline_methodology.qmd`'s interactive Leaflet map.

### `fig_WMed_basemap.R` (not in the repository)

Standalone map-generation script (not sourced by anything else), purely
cartographic — not required by the numbered pipeline at all.
**Western-Med-specific as delivered:** downloads the GFCM GSA shapefile,
splits polygons into Western Med (GSA 1–11, the target study area) vs.
context GSAs (12, 16 — Northern Tunisia, South of Sicily) and draws the
Sicily Channel divide (Cape Bon–Cape Lilibeo) as the Western/Central
Mediterranean boundary; all three of those choices would need editing to
produce an equivalent basemap for a different region. The general
technique — label placement via the "pole of inaccessibility" algorithm
(`polylabelr`, more robust than `st_point_on_surface()` for narrow,
coastline-hugging GSAs) plus a high-resolution (`rnaturalearthhires`)
coastline — is reusable as-is for any region. Output:
`output/plots/westmed_gsa_map.png`.

---

## Adapting this pipeline to a different region

Everything below applies whether the target region is a single GSA, an
arbitrary GSA subset, the whole Mediterranean GSA-by-GSA, or a fully
custom (non-GSA) boundary.

### Preferred method: source the scripts with pre-set options

`01_biomass.R`, `02_fisheries.R`,
`03_pbqb-traits.R`, `04_diets.R` and `05_validation.R` are **source-able**: every region/year/method knob each
one uses (`FILTER_AREAS`, `TARGET_COUNTRIES`, `YEAR_ECOPATH`,
`START_YEAR`/`END_YEAR`, `TS_YEARS`, `DATASET_VERSION`,
`TARGET_COUNTRIES_OVERRIDE`, `OUTLIER_METHOD`, `PB_QB_SELECTION_MODE`, `MACROZOO_SOURCE`, `MAISHA_PATH`,
`out_dir`/`pcloud_dir`/`git_dir`, and a handful more — each script's own
"Configuration" section, under a comment block headed `SOURCE-ABLE
SCRIPT`, lists exactly which variables it recognizes) is wrapped in
`if (!exists("VAR", envir = .GlobalEnv, inherits = FALSE)) VAR <- <default>`.
That means:

* If a variable is **already set** in the calling environment before the
  script is `source()`-d, the script uses that value and skips its own
  default entirely.
* If it is **not set**, the script falls back to its original Western Med
  default (GSA 1–11, 1994–2023) — running any of the three scripts on its
  own, exactly as before, is completely unchanged.

`run_pipeline_demo.R` is a driver script demonstrating this: it sets
`out_dir`/`pcloud_dir`/`git_dir` once, then `source()`s
`01_biomass.R` → `02_fisheries.R` →
`03_pbqb-traits.R` → `04_diets.R` in that dependency order (01 must run
first — see [Adapting this pipeline to a different region] below and each
script's own header comment for why; 04 specifically needs 01's
`FG_spp_Ecopath` sheet in the shared workbook for its species'
share-of-FG-biomass values, and is otherwise optional — diet composition
isn't needed for every run). It includes two ready-to-run examples,
switched via a `RUN_MODE` flag at the top:

* `RUN_MODE <- "westmed_default"` — sets nothing beyond the three paths;
  every script uses its built-in West Med default.
* `RUN_MODE <- "custom_example"` — sets `FILTER_AREAS <- 17:18`,
  `TARGET_COUNTRIES <- c("Italy", "Croatia", "Slovenia")`,
  `YEAR_ECOPATH <- 2005:2007`, `TS_YEARS <- 2000:2020`, and
  `START_YEAR`/`END_YEAR <- 2000`/`2020` as a worked Adriatic-GSA example
  — copy and adapt this block for any other region/year range, setting
  only the variables that need to differ from the West Med default.

`YEAR_ECOPATH` in particular is read by all three scripts and must stay
consistent across them (it defines the single Biomass snapshot the PB/QB
weights are computed against) — setting it once in the driver script
before any of the three are sourced keeps them in sync automatically,
which is easy to forget when editing each script's hardcoded copy by hand.

This sourcing approach and the manual per-script editing approach
described in the rest of this section are not mutually exclusive — the
`if (!exists(...))` guards mean a variable can still be hardcoded by
editing the script directly (equivalent to always setting it), or left
alone to pick up a caller's override, on a variable-by-variable basis.

**To model a different GSA or set of GSAs** (Adriatic, Ionian, Aegean,
Levant, or the whole basin), starting from `01_biomass.R` as
a template:

* `FILTER_AREAS` — set to the target GSA number(s). No other GSA-based code
  in `lib_survey_fg_density_functions.R` hardcodes a Western-Med-only
  assumption — `compute_strata_area_by_area()`, `weight_by_strata()`,
  `weight_by_area()` etc. are all region-agnostic and operate on whatever
  `area_ids`/`area_shp` they're given.
* `ALBORAN_GSAS`/`fix_alboran_longitude_sign()` — a Western-Med-only
  correction (GSA 1–3); harmless no-op elsewhere, delete if not needed.
* `MEDIAS_HANDBOOK_AREAS`/`GSA_AREA_CROSSWALK` — only populated for Western
  Med GSAs; add the target region's own rows from the MEDIAS Handbook
  (or rely entirely on the bathymetry-derived 10–200 m fallback if MEDIAS
  data or handbook-reported areas aren't available for that region).
* `out_dir`/`pcloud_dir`/`git_dir` paths, and the raw MEDITS/MEDIAS/FG-
  reference file paths under them — point at the target region's own data.
* `export_gsa_boundaries.R` — optional cartography
  only; edit the GSA-range filters if you want an equivalent map for the
  new region, otherwise skip it (nothing downstream needs it).

**To model a fully custom (non-GSA) region anywhere in the Mediterranean**,
set `AREA_MODE <- "custom"` before sourcing `01_biomass.R`, plus `AREA_NAME`,
`CUSTOM_AREA_TYPE` (`"bbox"` (default), `"shapefile"`, or `"gsa"`), and
`CUSTOM_BBOX`/`CUSTOM_SHAPEFILE_PATH`/`CUSTOM_GSA_IDS` set to the target
boundary — see [`AREA_MODE == "custom"`] above. No GSA-specific code path
is involved for `"bbox"`/`"shapefile"` (the Alboran-sign fix is present but
is a documented no-op outside GSA 1–3); MEDIAS and stock-assessment biomass
are skipped entirely under `AREA_MODE == "custom"`, regardless of
`CUSTOM_AREA_TYPE`.

Everything downstream of Step 1 — `02_fisheries.R`, `03b_ecobase.R`,
`03_pbqb-traits.R`, and the shared library's Excel-export/PB-QB functions — is
already region-agnostic: they operate on whatever `strata_area_by_area.csv`
and `species_density_regional_combined.csv` Step 1 produced, with no
Western-Med-specific logic beyond `02_fisheries.R`'s GFCM division codes
(37.1.1–37.1.3) and `GSA_COASTAL_STATE` table (which country is the coastal
state of which GSA; `TARGET_COUNTRIES` is derived from it and from
`FILTER_AREAS`) — both need checking for another region — and `03b_ecobase.R`'s
`MED_LON_MIN/MAX`/`MED_LAT_MIN/MAX` bounding box (already covers the **full**
Mediterranean by default — Gibraltar to the Levant — so it does not need
editing for a sub-region; narrow it only if you specifically want to exclude
literature models from elsewhere in the basin).

---

