---
title: "Biomass, Fisheries, PB/QB-Traits & Diet Pipeline"
subtitle: "Code documentation for the four EwE input-building scripts — biomass, fisheries, PB/QB & traits, diet — for any GFCM GSA subset or user-defined region in the Mediterranean"
author: "Daniel Vilas"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true
execute:
  eval: false
  echo: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = "styler")  # eval/echo now set via the YAML execute: block above
# figure chunks below override eval/echo individually so this document can
# show real pipeline output when it's available; everything else in the
# document stays non-executed (this is documentation, not a live report).

# safe_include_graphics(): shows the real figure if it's been copied into
# figures/ next to this .Rmd (see "Populating the real figures" below);
# otherwise prints a plain placeholder message instead of erroring, so this
# document always knits cleanly whether or not the pipeline has been run
# yet against the real WestMed EwE pCloud drive data.
safe_include_graphics <- function(path, script, out_dir_note) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(
      "*Figure not yet available.* Run **", script, "**",
      " against the real WestMed EwE pCloud drive data, then copy\n`",
      out_dir_note, "`\ninto a `figures/` folder next to this `.Rmd` ",
      "(as `", basename(path), "`) and re-knit.",
      sep = ""
    )
  }
}
```

```{css, echo=FALSE}
body {
  max-width: 900px;
}
/* keep the whole page from ever scrolling sideways */
html, body {
  overflow-x: hidden;
}
/* wide tables: scroll horizontally inside their own box instead of
   pushing the page wider than the viewport */
.table-responsive,
table {
  display: block;
  max-width: 100%;
  overflow-x: auto;
}
table {
  width: auto !important;
}
table td, table th {
  white-space: normal;
  word-break: break-word;
}
/* code blocks (the ASCII pipeline diagram, chunk output, etc.):
   wrap long lines instead of overflowing, but still allow a
   horizontal scrollbar as a fallback for anything that truly can't wrap */
pre, code {
  white-space: pre-wrap;
  word-break: break-word;
  overflow-x: auto;
  max-width: 100%;
}
/* mermaid diagrams (see medbs_pipeline_methodology.qmd for the same look):
   don't let the pre-wrap rule above touch them - mermaid.js replaces the
   div's contents with an SVG at render time */
div.mermaid {
  overflow-x: auto;
  max-width: 100%;
  text-align: center;
}
```

<script src="https://cdn.jsdelivr.net/npm/mermaid@10.9.8/dist/mermaid.min.js"></script>
<script>
  mermaid.initialize({ startOnLoad: true, theme: "neutral",
                        flowchart: { padding: 20, nodeSpacing: 60, rankSpacing: 90 } });
</script>

# Purpose of this document

This document describes a general-purpose R pipeline for turning MEDITS/
MEDIAS trawl-survey and catch data into input sheets for an Ecopath-with-
Ecosim (EwE) model, for **any region in the Mediterranean Sea** — a single
GFCM Geographical Sub-Area (GSA), any subset or the full set of GSAs, or a
completely user-defined region (a bounding box or an arbitrary shapefile
that doesn't line up with GSA lines at all, e.g. a bay, an MPA, or a
footprint spanning parts of several GSAs). It covers every script provided,
what each one reads and writes, and every function in the shared library
those scripts call.

The scripts as delivered are configured for the **Western Mediterranean**
(GSA 1–11) as a **worked example** — this is what `01_biomass.R`,
`export_gsa_boundaries.R` and `fig_WMed_basemap.R` currently point at, and
what a few reference tables (e.g. the MEDIAS Handbook area crosswalk) are
currently populated for. The underlying mechanism — GSA-based selection via
`FILTER_AREAS`, or an arbitrary custom region via `01_biomass.R`'s
`AREA_MODE <- "custom"` switch — is not Western-Med-specific; see
[Adapting this pipeline to a different region]
below for exactly what to change to point the same pipeline at another part
of the Mediterranean.

All scripts read from and write to **one shared workbook**,
`output/ecopath_ecosim_inputs.xlsx`, via two library functions designed for
exactly this purpose:

* `upsert_workbook_sheets()` — adds/replaces named sheets, leaves every other
  sheet in the workbook untouched. Safe to call from any script, in any order,
  any number of times.
* `finalize_workbook_sheet_order()` — run **last**, after every upstream
  script has run at least once. Renames a couple of sheets (`FG_lookup` →
  `FG`, `References` → `PB_QB_References_`) and reorders every sheet into the
  final layout Ecopath/Ecosim expects.

Because of this design, the four numbered pipeline scripts are
**order-independent with respect to the shared workbook** — except where a
script's own calculations literally need a file another script produced
first (documented per-script below).

# Data and code locations

* **Code** — every script and library file documented here lives in the
  **WestMed EwE GitHub repository** (`WMed_EwE`, cloned locally to what the
  scripts call `git_dir`). All shared-library `source()` calls
  (`lib_survey_fg_density_functions.R`, `lib_worms_taxonomy_lookup.R`,
  `lib_build_species_df_from_survey.R`) resolve against this repository's
  `scripts/` folder.
* **Input data** — every raw survey, catch and reference file each script
  reads (`FG_WMed.xlsx`, `TM_list_(April_2019).xlsx`, MEDITS `TA.csv`/
  `TB.csv`, MEDIAS `abundance.csv`/`biomass.csv`, `ASFIS_sp_2026.1.csv`,
  the catchability CSV, GFCM capture/reference CSVs) is read from the
  **WestMed EwE pCloud drive** (`EwE Western Med 2026`, mounted/synced
  locally to what the scripts call `pcloud_dir`), **not** from the GitHub
  repository — the repository holds code only, no raw data.
* **Outputs** — every CSV, plot, and the shared `ecopath_ecosim_inputs.xlsx`
  workbook are written to a local `out_dir` (not synced back to pCloud or
  GitHub by the scripts themselves).

Throughout this document, `<pcloud_dir>` means *the WestMed EwE pCloud
drive* and `<git_dir>` means *the WestMed EwE GitHub repository* — both are
plain local paths the top of each script resolves (hardcoded for one user,
or picked interactively via `rstudioapi::selectDirectory()` for anyone
else), not fixed locations baked into the code. Retargeting the pipeline to
a different region (see [Adapting this pipeline to a different region])
does not require moving off pCloud/GitHub — only pointing `pcloud_dir`/
`git_dir` at that region's own drive/repository, or adding that region's
data alongside the Western Med data in the same drive/repository.

**A note on the plots and column tables in this document:** this
documentation was written without direct access to the WestMed EwE pCloud
drive's actual data files. The **column-reference tables** throughout are
not guesses — they list exactly the column names each script's own code
reads (`select()`/`merge()`/`setnames()` calls, etc.), so they're accurate
regardless of the actual data values. The **figures**, however, are the
pipeline's own real output and only appear once you've generated them — see
[Populating the real figures] below.

## Populating the real figures

Several sections below embed a figure via a small `safe_include_graphics()`
helper (defined in this document's setup chunk). Each one checks for a file
under `figures/` (a folder that must sit next to this `.Rmd`) and shows the
real image if present, or a plain "not yet available" message with the
exact script/output path to copy from if not — so this document always
knits cleanly either way. To populate them for real:

1. Run the relevant script (named in each placeholder message) against the
   real WestMed EwE pCloud drive data.
2. Copy the named `.png` from that script's `<out_dir>/plots/` into a
   `figures/` folder next to this `.Rmd`, keeping the same filename.
3. Re-knit.

Four **other** diagrams in this document (the pipeline flow, the
FG-matching cascade, the workbook sheet assembly, and the PB/QB taxon
dispatch diagrams) are a different kind of figure: **Mermaid flowcharts**,
in the same style as `medbs_pipeline_methodology.qmd`'s own diagrams
(same `flowchart LR` layout, the same gray/purple/teal/coral
`classDef` palette). They show the code's own logic — which function
calls which, which sheet gets written by what, which fallback branch runs
when — built directly from the script and library contents described
here, not from any survey or model output. Because they're plain text
(Mermaid syntax rendered in-browser via `mermaid.js`, loaded from a CDN
in this document's setup), they contain no data values and no image
files to go missing — unlike the four `safe_include_graphics()`
placeholders above, they always render, with or without an internet
connection to the pCloud drive. They do need an internet connection *to
the Mermaid CDN* at view time (not render time) to draw the SVG in the
browser; if you'd rather they render standalone with no external
dependency, `rmarkdown::html_document(self_contained = TRUE)` still
inlines the two `<script>` tags below the CSS chunk, so this only
matters if that CDN is unreachable from wherever the HTML is opened.

# Pipeline overview

The diagram below is a schematic of the actual script/function call
structure documented in this document — which script produces which
intermediate file, and which workbook sheets each one writes. It is not a
plot of any survey data or model output (same diagram style as
`medbs_pipeline_methodology.qmd`'s "Pipeline flow" section).

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["Step 1&lt;br/&gt;01_biomass.R"] --> B1["output/biomass/&lt;br/&gt;species_density_regional_combined.csv,&lt;br/&gt;strata_area_by_area.csv, FG_lookup.csv, Ecosim.csv"]
    B1 --> C1["xlsx: Ecopath_B, FG_spp"]
    B1 --> A2["Step 2&lt;br/&gt;02_fisheries.R"]
    A2 --> C2["xlsx: Ecopath_L, Ecopath_Di&lt;br/&gt;output/fisheries/: Catches_Ecopath,&lt;br/&gt;Catches_Ecosim, Catches_Discards_FG_ts"]
    B1 --> A3["Step 3&lt;br/&gt;03_pbqb-traits.R"]
    A3 --> C3["xlsx: Ecopath_PBQB, Ecopath_traits&lt;br/&gt;output/pbqb-traits/: PB_QB, PB_QB_spp, Ecobase"]
    B1 --> A4["Step 4 (run last)&lt;br/&gt;04_diets.R"]
    A4 --> C4["xlsx: Ecopath_diet, FG_References&lt;br/&gt;output/diet/: diet_references_by_fg.csv"]
    C1 --> F["trim_workbook_to_final_sheets()&lt;br/&gt;run after every script"]
    C2 --> F
    C3 --> F
    C4 --> F
    F --> O["output/ecopath_ecosim_inputs.xlsx&lt;br/&gt;(10 final sheets)"]

    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class A1,A2,A3,A4 purple
    class B1 gray
    class C1,C2,C3,C4 teal
    class F,O coral
</div>
```

**Step 1 — one script, `AREA_MODE` picks the region:**

* `01_biomass.R`, `AREA_MODE <- "westmed"` (default)
  — MEDITS + MEDIAS, GFCM GSA areas, **or**
* `01_biomass.R`, `AREA_MODE <- "custom"`
  — MEDITS only, custom bbox/shapefile/custom-GSA grouping

(`01_survey_density_custom.R`, the old separate script for the second case,
is deprecated — its logic now lives in `01_biomass.R` behind `AREA_MODE`.)

&#8595; produces `species_density_regional_combined.csv`,
`strata_area_by_area.csv`, and sheets `FG_spp_Ecopath`, `Ecopath`, `Ecosim`,
`FG_spp_Ecosim`, `FG` (renamed from `FG_lookup`), `traits_ewe`.

**Step 2:**

* `02_fao_catches.R`
  (reads Step 1's `strata_area_by_area.csv`)

&#8595; produces sheets `Catches_Ecopath`, `Catches_Ecosim`.

**Step 3 — independent, can run anytime before Step 4:**

* `03b_ecobase.R`

&#8595; produces `ecobase_literature_pb_qb_simple.csv` (optional Step 4 input).

**Step 4 — run last:**

* `03_pbqb-traits.R`
  (needs Step 1's `species_density_regional_combined.csv`; optionally reads
  Step 2's `fg_catch_timeseries_*.csv` and Step 3's
  `ecobase_literature_pb_qb_simple.csv`)

&#8595; produces sheets `PB_QB`, `PB_QB_spp`, `Ecobase`, `References` (renamed
to `PB_QB_References_`).

**Finally**, `finalize_workbook_sheet_order()` — called from whichever
script runs last — reorders every sheet into the final layout:

&#8595; `output/ecopath_ecosim_inputs.xlsx`, final, ordered, ready for
Ecopath import.

Two more files are **not** part of the numbered pipeline, but support it:

* `lib_survey_fg_density_functions.R` and `lib_worms_taxonomy_lookup.R` and
  `lib_build_species_df_from_survey.R` — shared library files, `source()`'d
  by the numbered scripts, no top-level driver code of their own.
* `export_gsa_boundaries.R` and `fig_WMed_basemap.R` — standalone cartography
  utilities, not required by the numbered pipeline. `export_gsa_boundaries.R`
  in particular produces the GeoJSON embedded in `medbs_pipeline_methodology.qmd`'s
  interactive map.

---

# Detailed methodology per output block — B, C, L, Di, M, TRAITS, F, PB, QB, DIET

*(Added 2026-09-24, generated against the current `01_biomass.R` / `02_fisheries.R` (+ `02b_fisheries_priority_merge.R`) / `03_pbqb-traits.R` + `03b_ecobase.R` / `04_diets.R` (+ `04b_diet_matrix_crosswalk.R`) scripts. This section is the authoritative, per-block methodology reference — including the exact priority cascades, formulas, and literature references behind every number in the final workbook. It supersedes any conflicting description elsewhere in this document that still refers to the older `01_survey_density_westmed.R`/`02_fao_catches.R`-as-primary/`03_ecobase_query.R`/`04_pbqb_calc.R` naming — those scripts are either renamed or now secondary/legacy fallbacks, as noted per-section below.)*

This document describes the **current, running** WMed EwE input-data pipeline as implemented in:

- `01_biomass.R` — Biomass (B)
- `02_fisheries.R` (plus helper `02b_fisheries_priority_merge.R`) — Catches (C), Landings (L), Discards (Di), Fishing mortality (F)
- `03_pbqb-traits.R` + sourced helper `03b_ecobase.R` — Natural mortality (M), Production/Biomass (PB), Consumption/Biomass (QB), life-history Traits (TRAITS)
- `04_diets.R` (plus helper `04b_diet_matrix_crosswalk.R`) — Diet matrix (DIET)
- Shared library `lib_survey_fg_density_functions.R`, plus `lib_worms_taxonomy_lookup.R`, `lib_aquamaps_depth_extension.R`, `lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R`, `lib_sau_extraction.R`, `iccat_bft_swo_automation.R`

**This supersedes any earlier documentation that refers to `01_survey_density_westmed.R`, `02_fao_catches.R` (as the primary catch script), `03_ecobase_query.R`, or `04_pbqb_calc.R`.** Those names describe an older iteration of this pipeline. `02_fao_catches.R` in particular still exists on disk and is read by `03_pbqb-traits.R` only as a **legacy fallback** path for the FG-level fishing-mortality rate (`Fmort_FG`) when `02_fisheries.R`'s own `F_by_fg.csv` output is not available — the current, preferred F source for every section below is `02_fisheries.R`.

All ten input blocks (B, C, L, Di, M, TRAITS, F, PB, QB, DIET) ultimately write into one shared Excel workbook, `ecopath_ecosim_inputs.xlsx` (path held in `ECOPATH_WORKBOOK_PATH`), via `upsert_workbook_sheets()` / `finalize_ecopath_ecosim_summary_sheets()` in `lib_survey_fg_density_functions.R`, and each block also writes native, audit-friendly CSVs into its own output subfolder (`biomass/`, `fisheries/`, `pbqb/`, `diet/`).

---

## B — Biomass

### 1. What it is
The Ecopath base-year (baseline snapshot, `year_ecopath`, default 1994–1996) biomass density per functional group (FG), in t·km⁻², written to the `Ecopath_B` sheet; and the multi-year `Ecosim_ts` biomass time series used to drive/validate Ecosim.

### 2. Inputs / data sources
- **MEDITS** bottom-trawl survey (dataframe1/dataframe2 standardized survey tables) and **MEDIAS** pelagic acoustic survey — both processed through the shared density-computation library.
- **Catchability correction** table (`CATCHABILITY_CSV_PATH`, columns `ScientificName`, `q`) — MEDITS/MEDIAS gear catchability is <1 for many taxa; density/biomass is divided by the resolved `q`.
- **GFCM STAR / RAM Legacy** combined stock-assessment output (`combined_medbs_star_ramlegacy.csv`, produced by an external `combine_STAR_RAMlegacy.R`), read from `STAR_RAMLEGACY_DIR`.
- **ICCAT** stock-assessment SSB for Atlantic bluefin tuna (BFT), swordfish (SWO), albacore (ALB) — manual/cited per-stock SSB tables, converted to density via `ICCAT_STOCK_AREA_KM2`.
- **Manual/cited biomass CSVs**: `MEGAFAUNA_BIOMASS_PATH` (marine megafauna: cetaceans, pinnipeds, sea turtles, elasmobranch megafauna) and `PRIMARY_PRODUCER_BIOMASS_PATH` (zooplankton, phytoplankton, macroalgae, seagrass/Posidonia, gorgonians+corals) — both auto-drafted from EcoBase/CMEMS/satellite when missing, always tagged as a draft.
- **EcoBase** (existing published Ecopath models covering the Mediterranean) — via `fetch_ecobase_literature_biomass()` (`03b_ecobase.R`), keyword-matched per target group, closest-year-to-1995 model picked.
- **CMEMS/satellite ocean-colour chlorophyll-a** (`lib_cmems_phytoplankton_biomass.R`, `lib_satellite_phytoplankton_biomass.R`) as a phytoplankton-biomass fallback when EcoBase has no match.
- FG taxonomy/species-list reference: `FG_WMed_2026.csv`.

### 3. Computation method
The core is a **biomass-source PRIORITY CASCADE** built via `fcase()` per FG × Year (in `01_biomass.R`, ~lines 2310–2483):

1. **`FG_ECOLOGY_TYPE` classification** (checked first, ~line 2325):
   - `"survey_exempt"` — FG name is in `EXEMPT_FG_NAMES` (`Cymodocea`, `Posidonia`, `Macroalgae`, `Corals and gorgonians`, `Macro zooplankton`, `Meso and micro zooplankton`, `Suprabenthos`) — groups a bottom-trawl/acoustic survey is not designed to sample representatively at all.
   - `"single_species_assessed"` — exactly one species in the FG **and** that FG has a GFCM STAR/RAM or ICCAT stock assessment.
   - `"mixed"` — everything else (default).
2. **`mean_density` fcase priority**, evaluated top-down (first TRUE wins):
   - `survey_exempt` **and** a stock-assessment/EcoBase/manual density exists → use it (tagged `"<source> - survey-exempt FG..."`).
   - `survey_exempt` and **no** such density → left explicitly `NA` (never falls through to a survey figure — a deliberate 2026-09-24 fix; the survey noise for these groups is worse than a genuine gap).
   - `single_species_assessed` and a stock-assessment density exists → use it.
   - MEDIAS density exists → MEDIAS (preferred over MEDITS: acoustic survey covers pelagic groups MEDITS' bottom trawl misses).
   - MEDITS density exists → MEDITS.
   - Otherwise, any remaining stock-assessment density.
   - A final backfill (`avg_density`, the MEDITS+MEDIAS average) fills any FG still `NA` **except** `survey_exempt` rows, which are guarded from ever being backfilled with survey noise.
3. **Stock-assessment table build** (`stock_assessment_fg_year`): STAR/RAM species matched to FG by exact scientific name, then by genus fallback; biomass summed per FG × Year, divided by `total_area_km2_biomass` to get density. ICCAT SSB is folded in on top, **winning any overlap** with STAR/RAM for the same FG × Year cell (an anti-join removes the STAR/RAM row first). `ICCAT_STOCK_AREA_KM2` (BFT=1 — i.e. biomass reported directly as density already; SWO/ALB=2,510,000 km²) converts SSB tonnage to density; rows where the density conversion isn't valid are left with density `NA` and excluded from the priority rule rather than guessed.
4. **Manual/cited groups** (megafauna, primary producers/plankton) are merged into the same `stock_assessment_fg_year` table via `merge_manual_cited_biomass()`, each tagged `"literature/survey estimate (<label> - manual, cited entry)"`, winning any overlap with an existing STAR/RAM/ICCAT row for that cell.
5. **Catchability correction** (`apply_catchability_correction()`, `lib_survey_fg_density_functions.R` ~line 1625): resolves a catchability coefficient `q` per species using a taxonomic cascade — species-level exact match → genus (`"Genus spp"` entries) → genus-level taxonomic proximity (average `q` of related species already in the table) → family proximity → order (explicit entry, then proximity) → class (explicit only) → phylum (explicit only) → `EXEMPT_FG_NAMES`/`exempt_species` forced to `q=1` (overrides everything) → `default_q=1` for anything left unmatched. `Density`/`Biomass` are divided by the resolved `q`.
6. **Ecopath_B sheet build** (`export_ecopath_ecosim_excel()`, ~line 3661): one row per FG (from the full FG reference list, not just observed FGs), `Biomass` = the mean density across every year in `year_ecopath` (not the single first year), matching the averaging window used by PB/QB/L/Di. A parallel `Biomass_<baseyear>` single-year figure is retained in an audit-only CSV (`biomass_by_fg_ecopath_detail.csv`).
7. **Nearest-year baseline fallback** (`resolve_baseline_with_nearest_year_fallback()`, ~line 3504): if an FG's `year_ecopath` baseline density is zero/NA but it has a real (>0) density in some *other* survey year, that nearest-year value is borrowed instead of reporting a false zero — **except** for any FG whose name contains `"Expanding"` (a genuinely range-expanding/colonizing group, where a real zero baseline is the correct signal).
8. **`Ecosim_ts`**: built the same way per year across `ts_years`, optionally rescaled to a reference index (`NORMALIZE_TS`/`normalize_ts = TRUE`: first non-NA value in the series set to 1, everything else relative to it) or left as raw density.
9. **EcoBase last-resort fallback (STEP 15b)**: after all of the above, whatever FG still has no biomass at all is matched against EcoBase's other published Mediterranean models by keyword on the FG's own name/scientific name; a hit patches only the `Ecopath_B` baseline-year snapshot (`fg_biomass_base_year`), **not** the full `Ecosim_ts` series.

### 4. Known caveats / limitations
- `survey_exempt` FGs with no stock-assessment/EcoBase/manual source are deliberately left `NA` in `Ecopath_B` (flagged `"MISSING - survey-exempt FG..."`), not filled with survey noise — the modeller must supply a real figure.
- The `EXEMPT_FG_NAMES` list is hand-typed against `FG_WMed_2026.csv`'s exact FG-name spelling; a mismatch is a silent no-op, not an error (flagged explicitly in code comments — "Suprabenthos" specifically called out as unconfirmed spelling).
- `ICCAT_STOCK_AREA_KM2`'s BFT row (area = 1) is a placeholder to avoid dividing a real tonnage — flagged as "not a real biomass number" until fixed; rows with `uniform_density_valid == FALSE` are excluded rather than guessed.
- The catchability table's proximity fallback can silently fail to type-check (a documented `data.table` gotcha with `sapply()` over zero-row subsets) if not guarded — the code explicitly pre-initializes `match_value` as character to avoid this.
- Marine-megafauna keyword coverage (`MEGAFAUNA_TAXON_KEYWORDS`) was previously missing a working monk-seal match; comments flag this as fixed but call out that "Suprabenthos" and other entries were never independently re-verified against the real FG-name spelling.
- EcoBase biomass figures are explicitly "closest available, not a real 1994–1996 measurement" — every filled value is tagged with source model/authors/year and how far that year is from the target, never silently blended in as if it were a direct observation.
- The Ecosim sheet's exact export shape was matched against a real prior exported example, "not independently re-verified" against an actual Ecosim-imported file (per the function's own header comment).

### 5. Code location
- `01_biomass.R`: `EXEMPT_FG_NAMES` (~line 942), `ICCAT_STOCK_AREA_KM2` (~line 1807), `MEGAFAUNA_TAXON_KEYWORDS` (~line 2074), `FG_ECOLOGY_TYPE`/priority `fcase()` (~lines 2310–2483), STEP 15b EcoBase fallback (~line 2952).
- `lib_survey_fg_density_functions.R`: `apply_catchability_correction()` (~line 1625), `resolve_baseline_with_nearest_year_fallback()` (~line 3504), `export_ecopath_ecosim_excel()` (~line 3560, `Ecopath_B`/`Ecosim` construction ~3661–3740).
- `03b_ecobase.R`: `fetch_ecobase_literature_biomass()` (~line 493).

### 6. References
Colléter et al. (2015) — EcoBase; MEDITS Working Group (survey handbook); ICCAT GBYP aerial survey program (CREEM Phase 11, 2021); Kaschner et al. — AquaMaps.

---

## C — Catches

### 1. What it is
The FG × Year (and Country × FG × Year) **gross catch** (`Catch_t`) time series that feeds both Landings (L) and Discards (Di) — the reported/landed tonnage grossed up by an estimated discard ratio.

### 2. Inputs / data sources
- **GFCM Capture Production** (Task 1-style) statistics — `gfcm_country_fg`/`gfcm_species_division_fg` — the primary bulk catch source, restricted to `START_YEAR`–`END_YEAR` and `TARGET_COUNTRIES`.
- **STECF FDI** (Fisheries Dependent Information) — EU fleet-level landings/discards, trusted only from `STECF_FDI_START_YEAR = 2014` onward (2013 excluded for data quality); provides the discard ratio (`Discarded_t/(Landed_t+Discarded_t)`) and fleet composition for the EU-flagged portion of the fleet.
- **Sea Around Us (SAU)** reconstructed catch — `sau_download_raw_extract()`/`sau_get_raw_extract()` (`lib_sau_extraction.R`), EEZ-level bulk download, restricted to `TARGET_COUNTRIES` and the study period; supplies its own `catch_type`-derived discard ratio and pre-2014 fleet-share hindcast.
- **FishMIP** fisheries-effort/catch product, matched via a FishMIP-FG crosswalk (`FISHMIP_FG_CROSSWALK_PATH`) — supplies a FG × Year (and flat, all-years-pooled) discard ratio from "reported vs total" tonnage.
- **GFCM STAR/RAM Legacy** stock-assessment catch figures — cross-checked against (not blindly substituted for) the SAU-built series (`star_catch_pct_diff`), with an explicit override step further down for exact FG × Year matches.
- **Belhabib et al.** total-catch reconstruction (used for Morocco/Algeria/Tunisia where GFCM/FDI coverage is thin).
- Technology-creep correction table (`tech_creep_rate_for_year()`/`tech_creep_multiplier()`), anchored to `TECH_CREEP_BASE_YEAR = STECF_FDI_START_YEAR`.

### 3. Computation method
1. `catch_with_discards <- merge(gfcm_country_fg, discard_by_fg, ...)`: gross catch is computed as `Catch_t = Landings_t / (1 - discard_ratio)` wherever a FishMIP discard ratio exists for that FG × Year; `Discard_t = Catch_t - Landings_t`. Where no ratio exists at all, `Catch_t = Landings_t` (landings-only, no discard assumed).
2. **Discard-ratio priority override**, applied in sequence (each step recomputes `Catch_t`/`Discard_t` if it fires): FishMIP (base) → **SAU's own discard ratio** (hindcast/bias-corrected against FDI for the pre-2014 years, Country × FG × Year, preferred over FishMIP's Med-wide FG-level ratio) → **STECF FDI's own discard ratio** (Country × FG × Year, preferred over SAU when both exist, since it is fleet-reported rather than reconstructed).
3. **Hindcast calibration**: for years before `STECF_FDI_START_YEAR`, SAU's discard ratio and fleet-composition shares are calibrated against the FDI/SAU overlap years (`discard_calibration`, a per-Country×FG multiplicative factor), clamped to a plausible `[0, 0.95]` range, rather than trusting SAU's flat all-years figure uncorrected.
4. **Belhabib override** for Morocco/Algeria/Tunisia: `Catch_t_belhabib` is split into `Discard_t`/`Landings_t` using whichever `discard_ratio` tier already resolved for that cell, if any; otherwise the whole Belhabib figure is treated as gross catch with no discard split (`"Belhabib et al. total catch (gross) - no discard_ratio tier available"`).
5. **STAR/RAM cross-check** (not an automatic override): `star_ram_combined`, filtered to `Western Mediterranean` subregion, matched to FG by exact species name then genus fallback; compared against `catches_discards_fg` and written as `star_catch_pct_diff`/`star_discard_ratio_diff_pp` for manual review.
6. The full FG × Year grid is completed via `CJ()` cross-join to `full_fg_list`, so every FG has an explicit (possibly zero) row rather than being silently absent.

### 4. Known caveats / limitations
- GFCM's own catch source is restricted to a defined coverage window and country set; species reported only at NEI/genus-level aggregates are a known source of FG mis-assignment risk (flagged where relevant to F, see below).
- The FishMIP discard ratio is Mediterranean-wide and FG-level, coarser than the SAU/FDI country-specific ratios that are preferred over it wherever available.
- The pre-2014 SAU hindcast calibration factor defaults to 1 (no correction) whenever there are no FDI/SAU overlap years to calibrate against for that Country × FG — a judgement call, explicitly given the same "status" in comments as the technology-creep multiplier (i.e., an assumption, not a measurement).
- `catch_area_mismatch_flags` — a GSA/Division area-scope mismatch between GFCM's own Catches_Ecopath/Ecosim area and the survey's area can lead F (see below) to be a slight overestimate relative to Biomass's area; called out explicitly as a "known, unresolved mismatch" in `03_pbqb-traits.R`'s comments.
- Belhabib-sourced catch for Morocco/Algeria/Tunisia frequently has no discard-ratio tier to split against, so it is carried as gross, undifferentiated catch.

### 5. Code location
- `02_fisheries.R`: discard-ratio tiers and `Catch_t`/`Discard_t` computation (~lines 3390–3437, ~3536–3584), hindcast calibration (~lines 2892–2926), STAR/RAM cross-check (~lines 3636–3691), tech-creep functions (~lines 1841–1896).
- `02b_fisheries_priority_merge.R` — merges the multi-source catch/discard/fleet tiers by priority (see L/Di section).
- `lib_sau_extraction.R` — SAU download functions.

### 6. References
Pauly & Zeller (2016) — Sea Around Us; GFCM MEDBS working group; STECF (FDI database); ICCAT/GFCM Task-1/Task-2 catch reporting conventions.

---

## L — Landings

### 1. What it is
The `Ecopath_L` workbook sheet: FG × Fleet landings density (t·km⁻²·year⁻¹), averaged over `YEAR_ECOPATH`, one column per named fleet.

### 2. Inputs / data sources
Same underlying sources as Catches (C) above; the fleet dimension additionally draws on:
- `FLEET_REGISTER` / `parse_gfcm_fleet_register()` — the GFCM Regional Fleet Register, giving each of six named countries (Spain, France, Italy, Morocco, Algeria, Tunisia) its Sector (Artisanal/Industrial) × Gear (`FleetType`) taxonomy.
- STECF FDI's own fleet-composition shares (per Country × FG, averaged across covered years) as a pre-2014 prior.
- SAU's own year-by-year fleet-composition shares (calibrated against FDI where overlap years exist).

### 3. Computation method
1. `Landings_t = Catch_t − Discard_t` per fleet — the fleet split is applied to the *fleet-resolved* long table `ecopath_fleet_long` (built from `fleet_split_out`), not derived independently.
2. Fleet-composition proportions (`fleet_prop_final`) are resolved in **four tiers**, combined via `rbindlist()`: (a) STECF FDI's real Country × FG × Year composition where FDI covers that cell; (b) SAU's hindcast-calibrated composition for pre-2014 years; (c) FDI's FG-specific multi-year average as a prior for cells FDI covers in *some* year but not this one; (d) a crude flat/equal split as the last resort where none of the above cover the cell.
3. `landings_ecopath_by_fleet_wide <- dcast(ecopath_fleet_long, FG_num + FG_name ~ Fleet, value.var = "Landings_t_km2_avg", fill = 0)`, then completed to every FG in `full_fg_list` (0-filled where a FG has no fleet catch at all).
4. **"Other GFCM countries — Unclassified" residual column**: `FLEET_REGISTER` only names fleets for the 6 target countries, so catch attributed to any other GFCM-reporting Mediterranean country (Greece, Libya, Malta, Cyprus, Turkey, Egypt, etc.) would otherwise silently vanish from `Ecopath_L`. This residual column is computed as `pmax(0, broad_FG_landings_density - rowSums(named_fleet_columns))`, keeping each FG's row total consistent with `Catches_Ecopath`'s own FG total.
5. A `Fleet_Structure` diagnostic table (`fleet_structure_out`) separately reports each fleet's real share of an FG's total catch across all 6 countries combined, `prop_catch` summing to 1 within each `FG_num`.

### 4. Known caveats / limitations
- A `ecopath_L_fg_coverage_check.csv` diagnostic flags FGs that get **less than half** their total landings from a named 6-country fleet — for those, the fleet/gear breakdown is materially coarser (dominated by the "Other GFCM countries" residual), even though the FG-level total still reconciles with `Catches_Ecopath`.
- 48 of 63 all-zero FG rows in a real run were found to have real matched crosswalk catch that was silently dropped before the "Other GFCM countries" fix — flagged explicitly as "not a small edge case."
- No fleet/gear-specific fishing mortality exists anywhere downstream (see F, below) — the fleet split here is landings/discards-only bookkeeping, not an independently fleet-resolved F.

### 5. Code location
- `02_fisheries.R`: fleet-split tier merge (~lines 4685–4710), `Ecopath_L`/`Ecopath_Di` fleet pivot (~lines 5186–5250), coverage-check diagnostic (~lines 5261–5306).

### 6. References
GFCM Regional Fleet Register (data collection reference framework, DCRF); STECF FDI database documentation.

---

## Di — Discards

### 1. What it is
The `Ecopath_Di` workbook sheet: FG × Fleet discards density (t·km⁻²·year⁻¹), the complement of Landings within the same gross-catch total, built identically to `Ecopath_L` but from `Discard_t_km2_avg`.

### 2. Inputs / data sources
Same discard-ratio sources as Catches (C): FishMIP reported-vs-total tonnage ratio, SAU's own `catch_type`-derived discard ratio, STECF FDI's own landed/discarded tonnage.

### 3. Computation method
- Built via the identical `dcast(ecopath_fleet_long, FG_num + FG_name ~ Fleet, value.var = "Discard_t_km2_avg", fill = 0)` pattern as `Ecopath_L`, completed to every FG, plus the same "Other GFCM countries — Unclassified" residual column computed from `broad_discards_density`.
- The FG-level discard ratio itself (before the fleet pivot) is resolved via the same three-tier priority as under Catches (C): FishMIP (Med-wide, FG × Year time-varying, falling back to an all-years-pooled flat ratio per FG when the per-year cell is too thin/implausible) → SAU (Country × FG × Year, hindcast-calibrated pre-2014) → STECF FDI (Country × FG × Year, wins wherever it exists).
- Implausible ratios are filtered at each tier (`is.finite(discard_ratio) & discard_ratio >= 0 & discard_ratio <= 0.95`) rather than accepted uncritically.
- A documented `data.table` bug fix (2026-09) ensures multiple FishMIP `f_group`s mapping to the same FG with *differing* discard ratios are averaged into exactly one ratio per FG × Year cell, rather than silently producing duplicate rows.

### 4. Known caveats / limitations
- The 0.95 cap on any discard ratio is a plausibility guard, not a measured ceiling — a genuinely near-total-discard fishery would be truncated.
- Discard ratios ultimately trace to at most three independent, imperfectly-overlapping data products (FishMIP, SAU, FDI); where none of the three cover a cell, `Discard_t` stays `NA` and `Catch_t = Landings_t` (i.e., zero assumed discards, not a genuine "unknown").
- Same residual-column caveat as Landings (L): the "Other GFCM countries" column is not broken down by fleet/gear.

### 5. Code location
- `02_fisheries.R`: discard-ratio tiers (~lines 2477–2496, 3063–3114, 3213–3238), fleet pivot (~lines 5200–5249).

### 6. References
Same as Catches (C): FishMIP; Sea Around Us (Pauly & Zeller, 2016); STECF FDI.

---

## M — Natural mortality

### 1. What it is
The natural-mortality-rate component used **inside** the PB calculation (`PB = M + F`, the standard Ecopath equilibrium-mortality decomposition for fish; other taxa use a direct PB formula without an explicit M+F split — see PB below).

### 2. Inputs / data sources
- **FishBase/SeaLifeBase** growth parameters (`Loo`, `K`, `tmax`) via `rfishbase::popgrowth()`/`species()`, fetched by `fetch_intrinsic_traits()`, with a Western-Mediterranean-first, then-broader-Mediterranean, then-most-recent-year row-selection rule (`select_best_rows()`, using `WMED_KEYWORDS`/`MED_KEYWORDS`).
- **FishLife** (Thorson et al.) — `fetch_fishlife()`, a phylogenetic-imputation database (`FishLife::FishBase_and_RAM`) giving a direct `M` prediction (`M_FishLife_2023`) plus imputed `Loo`/`K`/`Winfinity`, without needing this species' own growth data.
- **TropFishR::M_empirical()** — `fetch_tropfishr_M()`, bundling three independent empirical estimators from a single call (parsed from the function's row-named output matrix, not column names — confirmed against TropFishR's actual source).
- Ambient temperature (`Temp`), defaulting to `DEFAULT_TEMP = 16` (°C) where no species/area-specific value is supplied.

### 3. Computation method (fish; `calc_fish()`)
Six independent M methods are computed for **every** fish species (not gated behind missing-data fallback logic, so all methods are genuinely comparable):
1. **Pauly (1980)**, Eq. 9: `M = 10^(-0.0066 - 0.279·log10(Loo) + 0.6543·log10(K) + 0.4634·log10(Temp))`.
2. **FishLife**'s own phylogenetically-imputed `M` (used directly, not re-derived from FishLife's `Loo`/`K` via Pauly, to preserve FishLife's joint-uncertainty modelling).
3. **Gascuel et al. (2008)**, Eq. 15: `M = 2.31 · TrophicLevel^(-1.72) · exp(0.053·Temp)` — needs only trophic level + temperature, so it is nearly always computable.
4. **Hoenig (1983)** (via TropFishR; two internal rows — "Joint Equation" and "Fish Equation" — averaged into one value).
5. **Then et al. (2015)** (via TropFishR) — identified in its own validation paper as the best-performing empirical M estimator across 200+ fish species; given first priority when selecting the "chosen" PB method.
6. **Alverson & Carney (1975)** (via TropFishR).

Every M-method column is then gap-filled by `fill_by_taxonomic_proximity()`: **Stage 1**, borrow the mean of local donors (species already in this model's own `species_df`) at increasingly broad ranks — Genus → Family → Order → Class — using only species with a **direct** (non-borrowed) value as donors. **Stage 2**, if no local donor exists at any rank and the method is in `EXTERNAL_FORMULA_REGISTRY` (or is a TropFishR method), query FishBase/SeaLifeBase directly via `rfishbase::species_list()` for *external* species sharing that rank, fetch their traits, and compute the same formula for them (capped at 20 candidate species per lookup). Every filled value is tagged `"direct"` / `"borrowed (<rank>)"` / `"external FishBase/SeaLifeBase donor (<rank>)"`.

For non-fish taxa (mammals, seabirds), M is not computed as a separate column — the taxon-specific PB formulas below (Barlow & Boveng Siler survivorship; Gascuel general) directly estimate total mortality/production without an explicit M+F split, since no species-level catch data exists for most of those groups anyway.

### 4. Known caveats / limitations
- `Fmort`/`Fmort_species`/`Fmort_FG` (the F side of `PB=M+F`) is a **single lumped Yield/Biomass ratio across the whole region and all gears/fleets** — nothing in this pipeline currently produces a fleet- or gear-specific F even though `Ecopath_L`/`Ecopath_Di` do carry a fleet dimension; explicitly flagged as "NOT IMPLEMENTED" in code comments, and noted that GFCM's Operational-Unit-level DCRF data (fleet segment × ISSCFG gear × GSA) would be the real source for that, not derivable from `GFCM_Capture_Quantity.csv`.
- A known, unresolved GSA/Division area-scope mismatch between the catch data's area and the survey's biomass area means `Fmort_FG`/PB is a slight *over*-estimate relative to Biomass's own area.
- `M_FishLife_2023` depends on FishLife's internal `ParentChild_gz` structure matching what the code was verified against (version 3.1.0); a mismatch causes FishLife to be silently skipped (with a message) rather than erroring, leaving the other five M methods to carry the estimate.

### 5. Code location
- `03_pbqb-traits.R`: `calc_fish()` (~line 2148, M methods ~2163–2196), `fetch_fishlife()` (~line 1966), `fetch_tropfishr_M()` (~line 2079), `fill_by_taxonomic_proximity()` (~line 1452), `fetch_intrinsic_traits()` (~line 1361).

### 6. References
Pauly, D. (1980). On the interrelationships between natural mortality, growth parameters, and mean environmental temperature in 175 fish stocks. *ICES Journal of Marine Science*, 39(2), 175–192.
Hoenig, J.M. (1983). Empirical use of longevity data to estimate mortality rates. *Fishery Bulletin*, 82(1), 898–903.
Then, A.Y., Hoenig, J.M., Hall, N.G., & Hewitt, D.A. (2015). Evaluating the predictive performance of empirical estimators of natural mortality rate using information on over 200 fish species. *ICES Journal of Marine Science*, 72(1), 82–92.
Alverson, D.L., & Carney, M.J. (1975). A graphic review of the growth and decay of population cohorts. *Journal of the Conseil International pour l'Exploration de la Mer*, 36(2), 133–143.
Gascuel, D., Morissette, L., Palomares, M.L.D., & Christensen, V. (2008) [PB/QB "quick guide" protocol equations, as used by Vilas/Coll/Piroddi/Steenbeek-style EwE workflows].
Thorson, J.T., Munch, S.B., Cope, J.M., & Gao, J. (2017). Predicting life history parameters for all fishes worldwide. *Ecological Applications*, 27(8), 2262–2276. [FishLife]
Froese, R., & Pauly, D. (Eds.). FishBase. www.fishbase.org.
Palomares, M.L.D., & Bailly, N. (Eds.). SeaLifeBase. www.sealifebase.org.
Mildenberger, T.K., Taylor, M.H., & Wolff, M. (2017). TropFishR: an R package for fisheries analysis with length-frequency data. *Methods in Ecology and Evolution*, 8(11), 1520–1527.

---

## TRAITS — Ecopath_traits sheet

### 1. What it is
A per-species life-history/ecology trait table (`traits_ewe.csv`, exported to the workbook as `Ecopath_traits`, grouped by FG with header rows) — Loo, K, Winf, TrophicLevel, MaxWeight, Longevity, AspectRatio, temperature, plus a coarse `Organism` classification (Fishes/Mammals/Birds/Reptiles/Algae/Plants/Bacteria/Fungi/Other).

### 2. Inputs / data sources
- **FishBase/SeaLifeBase** taxonomy (`fetch_taxonomy_fishbase()`, wrapping `rfishbase::load_taxa()`), Class/Order/Family/Genus/Phylum/Kingdom.
- Same growth/ecology/morphometrics tables as M/PB/QB (`popgrowth`, `ecology`, `species`, `swimming`, `maturity`).
- **rfishbase::synonyms()** for species whose name is a synonym not directly recognized by `load_taxa()` (e.g. genus-level placeholders like `"Sepiola spp."`).

### 3. Computation method
- **`dispatch_group`** (the coarse bucket used to pick a PB/QB formula set): `classify_dispatch()` maps `Class` to `"fish"` / `"mammal"` / `"seabird"` / `"phytoplankton"`, defaulting anything else (including reptiles) to `"invertebrate"`.
- **`Organism`** (a separate, finer trait-sheet column, intentionally independent of `dispatch_group`): `classify_organism()` maps Class → Fishes/Mammals/Birds/Reptiles/Algae; Phylum → Plants (`Tracheophyta`, `Magnoliophyta`, `Streptophyta` — vascular seagrasses); Kingdom → Bacteria/Fungi; everything else defaults to `"Other"`.
- **Unresolved-taxonomy fallback**: species with no `Class` from the initial lookup are re-queried via a single **batched** `rfishbase::synonyms()` call (mapping synonym → currently-valid name) followed by a single batched taxonomy re-fetch for all resolved valid names — explicitly re-engineered (2026-09-23) from an earlier per-species loop that reloaded FishBase's/SeaLifeBase's **entire** table on every call, which was the actual cause of a "taking forever" performance problem, not the synonym lookup itself. Whatever remains unresolved after this defaults `dispatch_group` to `"invertebrate"` and `Organism` to `"Other"` — a taxonomically-neutral default, not a guess borrowed from another source.
- **Best-row selection** for multi-record FishBase/SeaLifeBase tables (`select_best_rows()`): Western-Mediterranean locality keywords first, then broader-Mediterranean, then most-recent year, then first available.
- **`build_grouped_traits_sheet()`** lays the flat trait table out with a `"<FG_num>: <FG_name>"` header row and a blank separator above each FG's species block, matching the legacy `FG_WMed.xlsx` sheet's visual convention — the workbook's `Ecopath_traits` sheet is this human-readable layout; `traits_ewe.csv` (flat, one row per species) is the version anything downstream should read programmatically.

### 4. Known caveats / limitations
- The taxonomic-proximity gap-filling used across M/PB/QB also silently defines what counts as "close" for trait imputation — Genus → Family → Order → Class, with an external-donor fallback capped at 20 species per lookup.
- `Organism == "Other"` covers any species whose Class/Phylum/Kingdom never resolved to any category — a genuine unknown, flagged for audit via a printed list of "Unclassified/unusual Class values."
- Bacteria/Fungi classification only works if the FishBase/SeaLifeBase version installed actually carries a `Kingdom` column; otherwise these fall through to `"Other"` rather than erroring.

### 5. Code location
- `03_pbqb-traits.R`: `classify_dispatch()` (~line 949), `classify_organism()` (~line 985), synonym-resolution fallback (~lines 1008–1166), `select_best_rows()` (~line 1200), `build_grouped_traits_sheet()` (~line 3383).

### 6. References
Froese, R., & Pauly, D. (Eds.). FishBase; Palomares, M.L.D., & Bailly, N. (Eds.). SeaLifeBase; WoRMS Editorial Board. World Register of Marine Species (www.marinespecies.org).

---

## F — Fishing mortality

### 1. What it is
The instantaneous fishing-mortality rate per FG, `F = Yield/Biomass` (catch density ÷ biomass density), computed both as a standalone `F_by_fg.csv`/`F_by_species_fg.csv` output in `02_fisheries.R` and reused inside `03_pbqb-traits.R`'s `PB = M + F` chain (fish) via `Fmort`/`Fmort_FG`.

### 2. Inputs / data sources
- `catches_discards_fg` (this pipeline's own combined GFCM/STECF/SAU/FishMIP/Belhabib catch table — see Catches, C).
- `species_density_regional_combined.csv` (01_biomass.R's Step-1 output) for the biomass-density side.
- `Total_Area_km2` — the same West-Med study area used consistently across Biomass, Catches and F, so all three sheets are mutually area-consistent.

### 3. Computation method
- `f_by_fg`: species density summed to FG level per year, averaged over `YEAR_ECOPATH` → `Biomass_density_avg`; catch averaged over the same years → `Catch_t_avg`; `F = (Catch_t_avg / Total_Area_km2) / Biomass_density_avg`.
- `f_by_species_fg`: the **same FG-level F value is repeated for every species in that FG** — F is *not* independently estimated per species, since GFCM catch data exists only at FG resolution. This is the standard EwE simplifying assumption of uniform fishing pressure within an FG. What genuinely differs per species is its own biomass share (`prop_sp_fg`) and the resulting implied catch density (`Species_catch_density_avg = F × Species_density_avg`).
- Inside `03_pbqb-traits.R`, `FG_YIELD_SOURCE` first tries to read `F_by_fg.csv` directly from `02_fisheries.R`'s output (preferred, current path); only if that file is missing does it fall back to reconstructing an FG-level F from the **legacy** `02_fao_catches.R` output (`fg_catch_csv`). Species-level `Fmort_species = Yield/Biomass` is used when a species-level yield exists (currently never, since no species-level landings source is implemented anywhere in the pipeline); otherwise `Fmort := Fmort_FG`.
- `PB_<method> := M_<method> + Fmort` for every one of the six fish M methods, so F is applied identically regardless of which M source was used.

### 4. Known caveats / limitations
- No fleet-, gear-, or GSA-specific F exists anywhere in the pipeline — F is a single, region-wide, all-gears rate per FG (explicitly flagged; see M section above for the same caveat).
- F undefined (`NA`/`Inf`) wherever `Catch_t_avg` is 0 (no matched catch) or `Biomass_density_avg` is 0/NA (FG not observed in the survey) — not imputed.
- A known area-scope mismatch between the catch data's coverage area and the survey's biomass area biases F slightly high relative to Biomass's own area (same root cause noted under Catches/M above).
- F is landings-only in `02_fao_catches.R`'s legacy fallback path (true F including discards is understated there) — flagged explicitly in code comments as a "REMINDER."

### 5. Code location
- `02_fisheries.R`: `f_by_fg`/`f_by_species_fg` computation (~lines 5337–5399).
- `03_pbqb-traits.R`: `FG_YIELD_SOURCE` resolution and `F_by_fg.csv` read (~lines 764–907), application inside `calc_fish()` (~lines 2197–2247).

### 6. References
Christensen, V., & Pauly, D. (1992). ECOPATH II — a software for balancing steady-state ecosystem models. *Ecological Modelling*, 61(3–4), 169–185. [standard Ecopath PB=M+F formulation]

---

## PB — Production/Biomass ratio

### 1. What it is
The Ecopath production/biomass ratio (year⁻¹) per species, aggregated to FG level for the `Ecopath_PBQB` workbook sheet — computed by a **taxon-dispatched** set of independent methods, each kept as its own comparable column.

### 2. Inputs / data sources
Same FishBase/SeaLifeBase/FishLife/TropFishR sources as M (fish); plus, for non-fish taxa:
- **Marine mammals**: `Longevity` (max age) and taxonomic family (mapped to a Siler-model "surrogate type" via `MAMMAL_SURROGATE_DEFAULT`).
- **Seabirds/invertebrates**: `TrophicLevel`, `Temp`, and (invertebrates only) `Biomass`, `MaxWeight`, `Depth`, and max age/max body-mass-in-kJ.
- **EcoBase literature PB/QB** (`fetch_ecobase_literature_pb_qb()`, `03b_ecobase.R`) as an FG-level gap-filler/cross-check, matched on `FG_name` (EcoBase carries no `FG_num`), averaged across all matching published models per FG.

### 3. Computation method, by taxon (`dispatch_group`)
- **Fish (`calc_fish()`)**: PB = chosen M (see M section) + Fmort. The "chosen" PB method is selected by `select_chosen()` in priority order: **Then et al. (2015)** first (validated best-in-class), then **Pauly (1980)**, then Hoenig (1983), Alverson & Carney (1975), FishLife, Gascuel (2008) — but every method's PB column remains in the output for direct comparison regardless of which was "chosen."
- **Marine mammals (`calc_mammal()`)**: **Method 1 — Barlow & Boveng (1991)** Siler three-parameter survivorship model (`siler_pb()`): juvenile/constant/senescent hazard components combined from family-specific surrogate parameter sets (`surrogate_type` 2=phocid-like, 3=delphinid/ziphiid/physeterid-like, 4=a third set), integrated over `Longevity` to a mean instantaneous mortality (= PB at equilibrium). **Method 2 — Gascuel (2008) general**, Eq. 23: `PB = 20.19 · TrophicLevel^(-3.26) · exp(0.041·Temp)`, computable independent of longevity data. Chosen PB prefers Barlow & Boveng when available, else Gascuel.
- **Seabirds (`calc_seabird()`)**: **only** Gascuel (2008) general (Eq. 23) — "no dedicated seabird PB method exists in the protocol," explicitly flagged; gap-filled by taxonomic proximity since previously ungap-filled seabirds with missing `TrophicLevel` got no PB at all.
- **Invertebrates (`calc_invertebrate()`)**: three independent methods — **Tumbiolo & Downing (1994)**, Eq. 17 (`logP = 0.24 + 0.96·log10(Biomass) - 0.21·log10(MaxWeight) + 0.03·Temp - 0.16·log10(Depth+1)`, then `PB = 10^logP / Biomass`; needs this model's own `Biomass`, so it is local-donor-only, never extended to external species); **Brey (1999)**, Eq. 19 (`log(P/B) = 1.672 + 0.993·log10(1/Amax) - 0.035·log10(Mmax) - 300.447·(1/(T+273))`, using max age as `Amax` and max body mass converted to kJ via a rough wet-weight energy-density constant of **4.5 kJ/g**, flagged explicitly as an approximation); **Gascuel (2008) general**. Chosen PB prefers Tumbiolo & Downing, then Brey, then Gascuel.
- **EcoBase gap-fill/cross-check** (in the main script, after taxon dispatch): FG-level empirical PB/QB (`PB_FG`, biomass-weighted across the FG's species) is compared to EcoBase's literature figure (`PB_ecobase`, averaged across matching models); FGs where the ratio `PB_FG/PB_ecobase` exceeds 2× or falls below 0.5× are flagged (`PB_diverges`) for manual review, but **not automatically overridden**. Where the empirical `PB_FG` is `NA` (no species-level data at all for that FG), the EcoBase literature value fills the gap (`PB_source := "EcoBase (literature)"`).

### 4. Known caveats / limitations
- Seabirds have no automatable dedicated PB method at all — every seabird's PB is the generic Gascuel formula, a materially coarser approximation than the fish/mammal-specific methods.
- The invertebrate energy-density constant (4.5 kJ/g) used for Brey (1999) is described in comments as "commonly cited order of magnitude," not a precise per-taxon value — adds real uncertainty beyond the formula itself.
- `PB_TumbioloDowning_1994` can only ever be direct for species with their own `MaxWeight` **and** `Depth` — SeaLifeBase's `MaxWeight` coverage for invertebrates is noted as "patchy," so the local-donor pool for this method can be permanently empty for some taxa (external-donor filling is deliberately disabled for this method since it needs this model's own Biomass).
- EcoBase's `-9999` sentinel (Ecopath's own "not provided" code) is converted to `NA` before use — a documented, verified fix, not an assumption.
- The `> 2x`/`< 0.5x` divergence threshold against EcoBase is a review trigger, not evidence either figure is wrong.

### 5. Code location
- `03_pbqb-traits.R`: `calc_fish()` (~line 2148), `siler_pb()`/`calc_mammal()` (~lines 2330–2392), `calc_seabird()` (~line 2400), `calc_invertebrate()` (~line 2434), EcoBase PB merge/divergence flagging (~lines 2682–2784).
- `03b_ecobase.R`: `fetch_ecobase_literature_pb_qb()` (~line 391).

### 6. References
Pauly (1980); Hoenig (1983); Then et al. (2015); Alverson & Carney (1975) [all as under M].
Gascuel, D., Morissette, L., Palomares, M.L.D., & Christensen, V. (2008) [PB/QB quick-guide protocol].
Barlow, J., & Boveng, P. (1991). Modeling age-specific mortality for marine mammal populations. *Marine Mammal Science*, 7(1), 50–65.
Tumbiolo, M.L., & Downing, J.A. (1994). An empirical model for the prediction of secondary production in marine benthic invertebrate populations. *Marine Ecology Progress Series*, 114, 165–174.
Brey, T. (1999). Growth performance and mortality in aquatic macrobenthic invertebrates. *Advances in Marine Biology*, 35, 153–223. (See also Brey, T. (2012) — Population dynamics in benthic invertebrates. A virtual handbook.)
Colléter, M., Palomares, M.L.D., Christensen, V., et al. (2015). EcoBase: a repository solution to gather and communicate information from EwE models. *Ecological Modelling*, 302, 42–48.

---

## QB — Consumption/Biomass ratio

### 1. What it is
The Ecopath consumption/biomass ratio (year⁻¹) per species/FG — the food-intake rate, computed by taxon-specific empirical formulas paralleling the PB section.

### 2. Inputs / data sources
Fish: `Winf` (asymptotic weight), `AspectRatio` (caudal-fin aspect ratio, `rfishbase::swimming()`), `Temp`, chosen `PB`. Marine mammals: `MaxWeight`. Seabirds: `MaxWeight`. Invertebrates: chosen `PB` only (via the Q/P=3 heuristic). EcoBase literature QB as a gap-filler/cross-check, same mechanism as PB.

### 3. Computation method, by taxon
- **Fish (`calc_fish()`)**: `Tprime = 1000/(Temp+273.15)`. **Method 1 — Palomares & Pauly (1998), Eq. 27 (Z-based)**: `logQB = 5.847 + 0.280·log10(PB) - 0.152·log10(Winf) - 1.360·T' + 0.062·AspectRatio + 0.510·h + 0.390·d` (herbivore/detritivore dummies `h`,`d` default to 0 — "no reliable automated diet-type classification available yet"), most information-rich since it uses `PB` (=Z) directly. **Method 2 — Palomares & Pauly (1998), Eq. 26 (non-Z)**: same form without the `log10(PB)` term — an independent cross-check that doesn't inherit PB's own error. **Method 3 — Christensen & Pauly (1992), Eq. 24**: `logQB = 6.37 - 1.5045·T' - 0.168·log10(Winf) + 0.1399·Pf + 0.2765·h` (predator dummy `Pf` defaults to 1). **Verified OCR-artifact fix**: the source protocol PDF's extracted text showed "log10 T′" for this equation, but using T′ *without* the log transform is used instead — with a log transform, QB values reached physically implausible tens-of-thousands, whereas T′ used directly (matching Eqs. 26/27's own convention) brings the result into the realistic ~1–20/year range. **Method 4 — Q/P=3** (`QB = 3 × chosen PB`; Christensen et al. 2008 convention). Chosen QB priority: Palomares & Pauly Z-based > Palomares & Pauly non-Z > Christensen & Pauly (1992) > Q/P=3.
- **Marine mammals (`calc_mammal()`)**: **Method 1 — Innes & Trites**, Eq. 31: `W_kg = MaxWeight/1000`; `QB = (0.1·W_kg^0.8 / W_kg) × 365`. **Method 2 — Q/P=3** fallback. Chosen QB prefers Innes/Trites.
- **Seabirds (`calc_seabird()`)**: **Nilsson & Nilsson (1976)**, Eq. 30: `logDR = -0.293 + 0.85·log10(MaxWeight)`; `QB = (10^logDR / MaxWeight) × 365`. **Deliberately no Q/P=3 fallback** — flagged explicitly, since that heuristic derives from ectotherm/fish biology and would badly underestimate an endotherm's much higher metabolic QB; if Nilsson & Nilsson fails even after gap-filling, QB is left `NA` rather than guessed.
- **Invertebrates (`calc_invertebrate()`)**: **Q/P=3 only** (`QB = 3 × chosen PB`) — the only automatable method; a more accurate Brey-based invertebrate QB method exists in the literature but is flagged as "non-automatable," left for manual review rather than implemented.
- **EcoBase gap-fill/cross-check**: identical mechanism to PB — `QB_ratio = QB_FG/QB_ecobase`, flagged if outside [0.5×, 2×]; EcoBase literature value fills any FG with no empirical QB at all.

### 4. Known caveats / limitations
- The herbivore/detritivore/predator dummy variables (`h`, `d`, `Pf`) in the fish QB equations are hardcoded defaults (carnivore/predator), not derived from any actual diet-type classification — flagged explicitly as a current limitation ("no reliable automated diet-type classification available yet").
- The Christensen & Pauly (1992) equation's T′ treatment is a documented correction of a likely OCR error in the source protocol extraction, not independently re-derived from the original 1992 paper — flagged for anyone cross-checking against the primary source.
- Seabird QB has no fallback at all if Nilsson & Nilsson's inputs are missing.
- Q/P=3 is applied to invertebrates and mammals as the only or secondary QB method — a coarse heuristic, not taxon-specific.

### 5. Code location
- `03_pbqb-traits.R`: fish QB (~lines 2261–2317), mammal QB (~lines 2377–2387), seabird QB (~lines 2409–2421), invertebrate QB (~lines 2475–2477), EcoBase QB merge (~lines 2682–2784).

### 6. References
Palomares, M.L.D., & Pauly, D. (1998). Predicting food consumption of fish populations as functions of mortality, food type, morphometrics, temperature and salinity. *Marine and Freshwater Research*, 49(5), 447–453.
Christensen, V., & Pauly, D. (1992). ECOPATH II — a software for balancing steady-state ecosystem models. *Ecological Modelling*, 61(3–4), 169–185.
Christensen, V., Walters, C.J., & Pauly, D. (2008). *Ecopath with Ecosim: a User's Guide*. Fisheries Centre, University of British Columbia. [Q/P=3 convention]
Innes, S., Lavigne, D.M., Earle, W.M., & Kovacs, K.M. (1987). Feeding rates of seals and whales. *Journal of Animal Ecology*, 56(1), 115–130. [Innes & Trites QB basis]
Nilsson, S.G., & Nilsson, I.N. (1976). Number, food consumption, and fish predation by birds in Lake Möckeln, Sweden. *Ornis Scandinavica*, 7(1), 61–70.

---

## DIET — Ecopath_diet sheet / diet matrix

### 1. What it is
The predator × prey diet-composition matrix at FG resolution (`Ecopath_diet` workbook sheet and `diet_composition_ewe.csv`, EwE's plain-text import format) — each predator FG's diet proportions summing to 1 across all prey FGs.

### 2. Inputs / data sources
- **The Mediterranean trophic metaweb database** (`data_entry_metaweb_empty.xlsx`, sheets `DATA_ENTRY`, `Taxonomic_codes`, `Non_taxonomic_groups`) — a compiled stomach-content/diet-study literature database with predator/prey codes and diet metrics (IRI, %Weight, %Number, Frequency, Presence). **Note:** the currently-shipped file is an empty template — no `DATA_ENTRY` rows exist yet; the earlier `diet_to_ewe.R` tool (documented separately in the project's `diet_to_ewe_tool_notes.md`) instead supported a stomach-content Access export or a direct FishBase/SeaLifeBase diet pull, but that path has been **superseded** — the current `04_diets.R` reads *only* the metaweb file, plus two small fallback CSVs (see below); it does not itself query FishBase/SeaLifeBase for diet data.
- `FG_WMed_2026.csv` — species→FG membership (same reference file used by 01/02/03).
- `01_biomass.R`'s own output `biomass_proportion_by_species_fg.csv` (`prop_sp_fg` column) — each species' share of its FG's biomass, used to weight multi-species FGs' blended diet.
- Two small **fallback-only** manual CSVs: `species_to_fg_missing.csv` (species not in `FG_WMed_2026.csv`) and `species_biomass_in_fg_missing.csv` (species × FG pairs missing a biomass share).
- `ewe_group_table.csv` — the final group numbering/order and predator/non-predator flag used for matrix layout.
- `04b_diet_matrix_crosswalk.R` (standalone helper, not run in the main 01→04 pipeline order) — re-expresses an *existing* diet matrix (e.g. from a published EcoBase model, or an older version of this model's own FG structure) against a different target FG list.

### 3. Computation method
1. **Code resolution** (`build_code_lookup()`/`resolve_codes()`): `Code_predator`/`Code_prey` resolved to scientific names/generic-group labels via `Taxonomic_codes` first (specific taxonomic identifications), then `Non_taxonomic_groups` (catch-all bins, e.g. "Pelagic fish"); an unresolved code is left as its raw code and reported, never silently dropped.
2. **Metric selection per DATA_ENTRY row** (`pick_diet_metric()`), priority order: **IRI** (Index of Relative Importance — folds in Frequency and either Weight or Number) → **%Weight** (biomass-based, closest to what Ecopath actually wants) → **%Number** → **Frequency** alone (weakest, presence/occurrence only) → **Presence** (1/0, converted to an equal split among a study's present prey when nothing else is available for that predator×reference).
3. **Species-level diet build** (`build_species_diet()`): each DATA_ENTRY row's diet value is renormalized within its own study (`Reference` × `Code_predator`) so that study's prey sum to 1; then averaged **across studies**, weighted by `predator_number` (sample size) where reported, else an unweighted mean; finally renormalized again per predator so its overall diet sums to 1. Distinct references per predator are retained separately (`predator_references`) to populate the final workbook's `FG_References` sheet's `diet_ref` column.
4. **Species → FG diet aggregation** (`build_fg_diet()`): each predator-FG's diet is built by weighting its constituent species by `pred_weight = pred_fg_share × pred_biomass_share` — i.e., a species' contribution to its FG's blended diet is weighted by **both** its taxonomic-assignment share (`species_to_fg`'s `proportion`, for species split across more than one FG) **and** its share of the FG's total biomass (from `biomass_in_fg`, sourced from `01_biomass.R`'s `prop_sp_fg`). Prey is similarly expanded across FGs via `prey_fg_share`; a prey with no FG mapping is kept as its own raw name with `prey_fg_share = 1`. The result is renormalized so each predator FG's diet sums to 1.
5. Where a species×FG pair has **no** biomass share from either `01_biomass.R`'s output or the fallback CSV, `build_biomass_in_fg()` defaults to an **equal split** among that FG's other mapped species — flagged with a message naming the affected species.
6. **Export**: `export_ewe_matrix_csv()` writes EwE's own plain-text diet-composition layout (predator columns by group number, restricted to `is_predator==TRUE` groups; prey rows by group number+name for the *full* group list; decimal-comma-quoted values; blank for zero; `Import`/`Sum`/`1-Sum` footer rows). `build_ecopath_diet_sheet()` writes the equivalent numeric (non-quoted) matrix directly into the shared workbook's `Ecopath_diet` sheet.
7. **`04b_diet_matrix_crosswalk.R`** (used only when re-expressing an existing external/older diet matrix against a new FG structure): handles the prey axis as a mass/contribution axis (splitting a source prey FG's contribution across several target preys via `Split_Weight`, or additively merging several source preys into one target) and the predator axis as a composition axis (a split source predator's diet vector is **copied** unchanged to each target predator; several source predators merging into one target predator are combined via a `Merge_Weight`-weighted average of their diet vectors). Any source FG in the matrix with no crosswalk row is excluded and flagged (`diet_crosswalk_UNMAPPED_REVIEW.csv`), never silently dropped or silently kept as its own category; any target predator whose post-conversion diet no longer sums to 1 because some of its prey were excluded is flagged with the % of diet mass lost (`diet_mass_conservation_REVIEW.csv`) before being renormalized back to 1.

### 4. Known caveats / limitations
- **The metaweb `DATA_ENTRY` sheet currently ships empty** — the entire species-level diet pipeline (`build_species_diet()` onward) has no real rows to process yet; `run_pipeline()` will stop with an explicit error ("No usable diet records...") until the database is populated.
- Contrary to what the task's own framing might suggest, the *current* `04_diets.R` does **not** call FishBase/SeaLifeBase or EcoBase for diet data at all — those are mentioned only as the superseded `diet_to_ewe.R` tool's alternate source paths; the only diet source in the current pipeline is the metaweb file.
- Both `Split_Weight` and `Merge_Weight` in the crosswalk helper default to an **equal** split/average when not explicitly supplied — a genuine biomass-weighted aggregation requires the modeller to supply real shares (e.g. from `mean_density` or `biomass_proportion_by_species_fg.csv`).
- A species×FG pair with no biomass-share source anywhere defaults to an equal split among its FG's other species — a simplifying assumption, not a measured share.
- Presence-only records are converted to a rough equal-split proxy, the weakest of the five diet metrics in the priority order.

### 5. Code location
- `04_diets.R`: `build_code_lookup()`/`resolve_codes()` (~line 138), `pick_diet_metric()` (~line 173), `build_species_diet()` (~line 196), `build_biomass_in_fg()` (~line 426), `build_fg_diet()` (~line 458), `export_ewe_matrix_csv()` (~line 502), `build_ecopath_diet_sheet()` (~line 541).
- `04b_diet_matrix_crosswalk.R`: `apply_prey_crosswalk()` (~line 248), `apply_predator_crosswalk()` (~line 266), `renormalize_and_flag()` (~line 284).

### 6. References
Christensen, V., & Pauly, D. (1992). ECOPATH II — a software for balancing steady-state ecosystem models. *Ecological Modelling*, 61(3–4), 169–185. [diet-matrix/mass-balance conventions]
Colléter, M., Palomares, M.L.D., Christensen, V., et al. (2015). EcoBase: a repository solution to gather and communicate information from EwE models. *Ecological Modelling*, 302, 42–48. [diet-matrix crosswalk use case]
Pinkas, L., Oliphant, M.S., & Iverson, I.L.K. (1971). Food habits of albacore, bluefin tuna, and bonito in California waters. *California Department of Fish and Game, Fish Bulletin*, 152. [Index of Relative Importance, IRI, methodological basis]

---

## References

Alverson, D.L., & Carney, M.J. (1975). A graphic review of the growth and decay of population cohorts. *Journal of the Conseil International pour l'Exploration de la Mer*, 36(2), 133–143.

Barlow, J., & Boveng, P. (1991). Modeling age-specific mortality for marine mammal populations. *Marine Mammal Science*, 7(1), 50–65.

Brey, T. (1999). Growth performance and mortality in aquatic macrobenthic invertebrates. *Advances in Marine Biology*, 35, 153–223.

Brey, T. (2012). *Population dynamics in benthic invertebrates. A virtual handbook.* Alfred Wegener Institute for Polar and Marine Research.

Christensen, V., & Pauly, D. (1992). ECOPATH II — a software for balancing steady-state ecosystem models. *Ecological Modelling*, 61(3–4), 169–185.

Christensen, V., Walters, C.J., & Pauly, D. (2008). *Ecopath with Ecosim: a User's Guide*. Fisheries Centre, University of British Columbia.

Colléter, M., Palomares, M.L.D., Christensen, V., Coll, M., Steenbeek, J., Falautano, M., Sagarese, S., & Pauly, D. (2015). EcoBase: a repository solution to gather and communicate information from EwE models. *Ecological Modelling*, 302, 42–48.

Froese, R., & Pauly, D. (Eds.). FishBase. World Wide Web electronic publication. www.fishbase.org.

Gascuel, D., Morissette, L., Palomares, M.L.D., & Christensen, V. (2008). Trophic flow kinetics in marine ecosystems: toward a theoretical approach to ecosystem functioning. *Ecological Modelling*, 217(1–2), 33–47.

GFCM (General Fisheries Commission for the Mediterranean). MEDBS (Mediterranean and Black Sea) Working Group stock-assessment methodology; Data Collection Reference Framework (DCRF).

Hoenig, J.M. (1983). Empirical use of longevity data to estimate mortality rates. *Fishery Bulletin*, 82(1), 898–903.

Innes, S., Lavigne, D.M., Earle, W.M., & Kovacs, K.M. (1987). Feeding rates of seals and whales. *Journal of Animal Ecology*, 56(1), 115–130.

Kaschner, K., Rius-Barile, J., Kesner-Reyes, K., Garilao, C., Kullander, S.O., Rees, T., & Froese, R. AquaMaps: Predicted range maps for aquatic species. www.aquamaps.org.

Mildenberger, T.K., Taylor, M.H., & Wolff, M. (2017). TropFishR: an R package for fisheries analysis with length-frequency data. *Methods in Ecology and Evolution*, 8(11), 1520–1527.

Nilsson, S.G., & Nilsson, I.N. (1976). Number, food consumption, and fish predation by birds in Lake Möckeln, Sweden. *Ornis Scandinavica*, 7(1), 61–70.

Palomares, M.L.D., & Bailly, N. (Eds.). SeaLifeBase. World Wide Web electronic publication. www.sealifebase.org.

Palomares, M.L.D., & Pauly, D. (1998). Predicting food consumption of fish populations as functions of mortality, food type, morphometrics, temperature and salinity. *Marine and Freshwater Research*, 49(5), 447–453.

Pauly, D. (1980). On the interrelationships between natural mortality, growth parameters, and mean environmental temperature in 175 fish stocks. *ICES Journal of Marine Science*, 39(2), 175–192.

Pauly, D., & Zeller, D. (Eds.) (2016). *Global Atlas of Marine Fisheries: A Critical Appraisal of Catches and Ecosystem Impacts.* Island Press. [Sea Around Us]

Pinkas, L., Oliphant, M.S., & Iverson, I.L.K. (1971). Food habits of albacore, bluefin tuna, and bonito in California waters. *California Department of Fish and Game, Fish Bulletin*, 152.

STECF (Scientific, Technical and Economic Committee for Fisheries). Fisheries Dependent Information (FDI) database.

Then, A.Y., Hoenig, J.M., Hall, N.G., & Hewitt, D.A. (2015). Evaluating the predictive performance of empirical estimators of natural mortality rate using information on over 200 fish species. *ICES Journal of Marine Science*, 72(1), 82–92.

Thorson, J.T., Munch, S.B., Cope, J.M., & Gao, J. (2017). Predicting life history parameters for all fishes worldwide. *Ecological Applications*, 27(8), 2262–2276.

Tumbiolo, M.L., & Downing, J.A. (1994). An empirical model for the prediction of secondary production in marine benthic invertebrate populations. *Marine Ecology Progress Series*, 114, 165–174.

WoRMS Editorial Board. World Register of Marine Species. www.marinespecies.org.

ICCAT (International Commission for the Conservation of Atlantic Tunas). GBYP (Grand Bluefin Tuna Year Program) aerial survey, CREEM Phase 11 report (2021).

---

# Run order and dependencies

| Step | Script | Depends on | Must run before |
|---|---|---|---|
| 1 | `01_biomass.R` (`AREA_MODE` picks westmed vs. custom) | nothing | 2, 4 |
| 2 | `02_fao_catches.R` | Step 1's `strata_area_by_area.csv` | — |
| 2a | `02a_fisheries_multisource.R` (optional, alongside Step 2) | Step 1's `strata_area_by_area.csv` and FG reference; optionally Step 2's `fg_catch_timeseries_*.csv` for the FAO-GFCM source | 4 (optional input, via `FISHERIES_DATA_SOURCE`) |
| 3 | `03b_ecobase.R` | nothing (independent web query) | 4 (optional input) |
| 4 | `03_pbqb-traits.R` | Step 1's `species_density_regional_combined.csv`; optionally Step 2's/Step 2a's `fg_catch_timeseries_*.csv`, Step 2a's `landings_by_species_gsa_year_*.csv`, and Step 3's `ecobase_literature_pb_qb_simple.csv` | — |

`finalize_workbook_sheet_order()` should be called at the end of whichever
script ran *last* in a given session (each of the numbered scripts already
does this itself at its own final step).

---

# Script 1a — `01_biomass.R`

**Role:** Pipeline Step 1 of 4 — a single script for **any region in the
Mediterranean**, GSA-based or fully custom, picked by the `AREA_MODE`
config variable (`"westmed"`, the default, or `"custom"` — see
[`AREA_MODE == "custom"`] below). No dependency on the other numbered
scripts; `02_fao_catches.R` and `03_pbqb-traits.R` both require this step's
output to exist first, regardless of which `AREA_MODE` produced it.

Runs two independent surveys over the same GSAs/FG scheme (MEDITS bottom
trawl + MEDIAS acoustic) and combines both into one set of Ecopath/Ecosim
sheets, rather than producing separate MEDIAS sheets. As delivered it's
configured for the Western Med (GSA 1–11, via `FILTER_AREAS <- 1:11`) as a
worked example; retargeting it to a different GSA subset elsewhere in the
Mediterranean — the Adriatic, the Aegean, the whole basin — is a matter of
changing that one line and the region-specific reference tables listed in
[Adapting this pipeline to a different region].

## Inputs

| File | Format | Role |
|---|---|---|
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → Functional Group reference (`dataframe2`: `ESPECIE`→`ScientificName`, `GF`→`FG_num`, `FG_name`) |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet `traits_ewe` | xlsx | Per-species life-history/ecology traits |
| `<pcloud_dir>/.../TM_list_(April_2019).xlsx` | xlsx | MEDITS species-code → scientific-name lookup |
| `<pcloud_dir>/.../Demersal/TA.csv` | csv | MEDITS haul/sample metadata (position, depth, gear, swept-area inputs) |
| `<pcloud_dir>/.../Demersal/TB.csv` | csv | MEDITS catch-per-haul data |
| `<pcloud_dir>/.../Acoustic/abundance.csv`, `biomass.csv` | csv | MEDIAS acoustic-survey raw records |
| `<pcloud_dir>/data/ASFIS_sp_2026.1.csv` | csv | FAO ASFIS species code list (MEDIAS species-code → scientific name) |
| `<pcloud_dir>/data/catchability_factors_ecotrans_medits_2021_spp.csv` | csv (optional) | Per-species catchability correction factors (q); skipped with a message if absent |
| GFCM GSA shapefile | auto-downloaded zip from `gfcmsitestorage.blob.core.windows.net/website/5.Data/ArcGIS/GFCM_GSA.zip` | Study-area polygons (`area_shp`); `SMU_CODE` 111/112 (W/E Sardinia) merged into GSA 11 |
| NOAA bathymetry | auto-downloaded via `marmap::getNOAA.bathy()` inside `compute_strata_area_by_area()` | Depth-stratum area calculation |

`FILTER_AREAS <- 1:11` (this example's Western Med GSAs — set to any GSA
number or set of GSA numbers, anywhere in the GFCM area, to retarget the
whole script to a different region), `YEAR_ECOPATH <- 1994:1996`,
`TS_YEARS <- 1995:2023`, `STRATA <- TRUE`, `DROP_OUTLIERS <- FALSE`,
`NORMALIZE_TS <- TRUE`, `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- FALSE` (opt-in —
see Step 7b below).

### Step 7b — AquaMaps shallow/deep depth adjustment (opt-in)

MEDITS' own 5 standard strata (`MEDITS_STRATA`) only cover 10–800 m.
`compute_strata_area_by_area()`'s area-proportion weighting normalizes
across just those 5 strata, so as delivered, this pipeline's Biomass/
density figures implicitly treat the whole modeled area as if it were
entirely 10–800 m seafloor — a species with real biomass shallower than
10 m or deeper than 800 m is invisible to these numbers, not just
imprecisely measured. Setting `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- TRUE`
(default `FALSE`) adds genuine **derived** pseudo-strata: one shallow
(0–10 m) plus, by default, the deep zone split into **two** sub-bands —
800–1000 m and 1000–2850 m — rather than one lumped 800–6000 m stratum
(2850 m is a practical ceiling for the western Mediterranean's own real
depths specifically, not the Mediterranean-wide ~5267 m Calypso Deep,
which sits in the eastern/Ionian basin outside these GSAs). Each
pseudo-stratum gets its own real bathymetric area (the same
`compute_strata_area_by_area()` machinery already used for the 5 real
strata) and a density **derived, not measured**, anchored to the ONE
real MEDITS stratum immediately adjacent to it — the shallowest real
stratum (10–50 m) for the shallow pseudo-band, the deepest real stratum
(500–800 m) for every deep pseudo-band — rather than an area-weighted
average across all 5 real strata. Each pseudo-stratum's density equals
that adjacent stratum's own **measured** density (`compute_species_
anchor_stratum_density()`), scaled by the ratio of AquaMaps mean
suitability between the pseudo-band and that same adjacent stratum's own
depth range. A species at equal AquaMaps suitability in both bands gets
a multiplier of exactly 1 — its pseudo-stratum density literally equals
what MEDITS measured next door, with no distortion from averaging across
strata the species may barely occupy. (An earlier version of this
adjustment compared each pseudo-band's suitability against the whole
10–800 m composite range instead — that systematically understated
narrow pseudo-bands relative to the ~790 m-wide reference purely from
the width mismatch, independent of actual suitability level; the
anchor-stratum design replaces it.) The trade-off: a species with real
catch elsewhere in 10–800 m but none specifically in the one adjacent
anchor stratum gets no pseudo-stratum density at all for that area
(flagged in the audit sheet), rather than being extrapolated from
unrelated, non-adjacent strata. These pseudo-strata are folded into the
exact same `weight_by_strata()`/`weight_species_by_area()` calls used
for the 5 real strata — not a separate formula — so `fg_index`/
`fg_index_regional`/`species_density_regional` end up as one consistent
area-weighted sum across all 8 strata (5 real + 3 derived), real and
derived alike. The number and boundaries of deep sub-bands are
controlled by `apply_aquamaps_depth_adjustment()`'s `deep_ranges`
argument — a list of `c(from, to)` pairs — so this can be collapsed back
to one lump (`list(c(800, 6000))`) or split further without changing
anything else; all deep sub-bands anchor to the same deepest real
stratum, they don't chain off each other.

**Core assumption, stated plainly rather than absorbed silently**: a
species' relative density across depth is assumed proportional to
AquaMaps' relative occurrence-suitability across depth (its trapezoidal
`DepthMin`/`DepthPrefMin`/`DepthPrefMax`/`DepthMax` envelope, per
AquaMaps' own "Algorithm and Data Sources" documentation). Occurrence
suitability is not the same thing as density — this is a standard
extrapolation approach when direct sampling doesn't cover a species' full
depth range, but an approximation nonetheless. Every resulting multiplier
is written to the `AquaMaps_Depth_Adjustment` sheet (species, its
`DepthMin`/`DepthPrefMin`/`DepthPrefMax`/`DepthMax`, probability masses,
resulting `multiplier_shallow`/`multiplier_deep_800_1000`/
`multiplier_deep_1000_2850` (one column per configured deep sub-band,
named by its own depth bounds), and a `note` explaining species that got
NO adjustment for a given band — unmatched to AquaMaps, no depth
envelope in `am_hspen()`, or essentially zero AquaMaps suitability in
that band's own adjacent real anchor stratum, making a ratio against
that unstable) — reviewable per species, not a black box.

Requires the `aquamapsdata` R package (github.com/raquamaps/aquamapsdata),
the only verified way to get AquaMaps' per-species HSPEN depth-envelope
table without scraping — its `download_db()` pulls a real, deliberate
**~2GB (~10GB unpacked) local SQLite database once** (`ensure_aquamaps_db()`
uses `force = FALSE`, so this never re-downloads on subsequent runs).
Implemented in `lib_aquamaps_depth_extension.R` (species → AquaMaps
`SpeciesID` resolution, envelope fetch, envelope math, pseudo-strata
construction) — see that file's own header for the full mechanism.
Requires `STRATA <- TRUE` (errors clearly if not — there's no unstratified
equivalent). Available under **either** `AREA_MODE` — the adjustment
operates on species depth envelopes and strata generically, with no
GSA-specific assumptions, so it applies the same way to a custom region.

Two outputs make the strata (real + derived) inspectable directly,
independent of `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT` — they cover whichever
strata are actually in play, 5 or 8:

- **`survey_fg_depth_strata_profile.png`** (Step 8, `plot_strata_profile()`)
  — a bar chart of raw per-sample density by stratum, faceted per FG, for
  sanity-checking that known deep-water FGs actually peak in the deep
  strata and vice versa. Reads `weight_by_strata()`'s `per_stratum_raw`
  attribute specifically (not `per_stratum`), so a stratum with real
  sample data but no computed area-proportion for a given AreaID still
  shows up here rather than silently vanishing.
- **`FG_Density_by_Stratum`** sheet (Step 12c, `build_fg_density_by_stratum_sheet()`)
  — the same information as an Excel table: one row per FG, one column
  per stratum (region-wide, area-weighted, `YEAR_ECOPATH`-averaged), plus
  a `Total_Density` column that reproduces `fg_index_regional`'s own
  `mean_density` exactly, since each stratum column is that stratum's own
  additive contribution to the total, not an independently-normalized
  figure.

### Column reference — key input files

Columns as actually referenced by the code (from the `<pcloud_dir>` files
above). Not every column in the real file is necessarily listed — only the
ones the script reads.

**`FG_WMed.xlsx`, sheet 4** (→ `dataframe2`, the species→FG master):

| Column | Renamed to | Meaning |
|---|---|---|
| `ESPECIE` | `ScientificName` | Species scientific name |
| `GF` | `FG_num` | Functional Group number |
| `FG_name` | `FG_name` | Functional Group name |

**`FG_WMed.xlsx`, sheet `traits_ewe`** (alternating "N: Group Name" header
rows / species data rows — see Script 1a Step 10 below):

| Column (original header) | Renamed to |
|---|---|
| *(unnamed index column)* | `row_index` |
| `Species` | `Species_col` |
| `Organism` | `Organism` |
| `Ecology` | `Ecology` |
| `Occurrence status` | `Occurrence_status` |
| `Biomass contribution` | `Biomass_contribution` |
| `Catch contribution` | `Catch_contribution` |
| `IUCN conservation status` | `IUCN_conservation_status` |
| `Exploitation status` | `Exploitation_status` |
| `Vulnerability index (?)` | `Vulnerability_index` |
| `Mean length (?)` | `Mean_length` |
| `Max length (?)` | `Max_length` |
| `Mean weight (?)` | `Mean_weight` |
| `Mean life span (year)` | `Mean_lifespan_years` |

**`TM_list_(April_2019).xlsx`, sheet 1 (skip first 2 rows)** — the code
locates its two needed columns by regex rather than a fixed name, since the
exact header text has drifted between file versions: a column matching
`/MEDITS/i` (the MEDITS species code) and a column matching `/Scientific/i`
(the scientific name).

**MEDITS `TA.csv`** (haul/sample metadata):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID` (concatenated, `_`-separated) |
| `distance` | Trawl distance towed → swept area |
| `wing_opening` | Trawl wing opening (decimetres) → swept area |
| `shooting_depth`, `hauling_depth` | Averaged into `mean_depth` |
| `shooting_latitude`, `shooting_longitude` | DDMM.mmm format — converted to decimal degrees |
| `shooting_quadrant` | Present but **not used** — flagged in the code as an unresolved caveat around hemisphere sign |

**MEDITS `TB.csv`** (catch-per-haul data):

| Column | Used for |
|---|---|
| `country`, `area`, `vessel`, `year`, `haul_number`, `month`, `day` | Built into `SampleID`, joined against `TA.csv` |
| `genus`, `species` | Concatenated into `species_code`, joined against `TM_list` |
| `ptot` | Total catch weight (grams; negative values treated as `NA`, converted to tonnes) |

**MEDIAS `abundance.csv` / `biomass.csv`** (acoustic survey raw records):

| Column | Used for |
|---|---|
| `country`, `year`, `species` | Grouping key for aggregation |
| `area` | Text field containing a GSA number, extracted via regex (`\d+`) |
| `sex` | Filtered to `"C"` (combined sexes) when present |
| `lengthclass*` (many columns) | Summed row-wise into a single `total_value` (negative values treated as missing) |

**`ASFIS_sp_2026.1.csv`** (FAO species reference list):

| Column | Used for |
|---|---|
| `Alpha3_Code` | MEDIAS species code → joined against `abundance`/`biomass` |
| `Scientific_Name` | Resolved scientific name |
| *(also present: `ISSCAAP_Group`, `Taxonomic_Code`, `English_name`, …)* | Not read by this pipeline |

**Catchability CSV** (`catchability_factors_ecotrans_medits_2021_spp.csv`,
optional):

| Column | Renamed to | Meaning |
|---|---|---|
| `species` | `ScientificName` | Species scientific name |
| `q_FACTOR` | `q` | Catchability correction factor |

```{r sample-coverage-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_sample_map() output — MEDITS sample coverage by area."}
safe_include_graphics(
  "figures/survey_sample_coverage_map.png",
  script = "01_biomass.R",
  out_dir_note = "<out_dir>/plots/survey_sample_coverage_map.png"
)
```

## Processing steps

1. **Config** — resolve `out_dir`/`pcloud_dir`/`git_dir`, download/cache the GSA
   shapefile, source `lib_survey_fg_density_functions.R` and
   `lib_worms_taxonomy_lookup.R`.
2. **Load MEDITS data** — read `TA.csv`/`TB.csv`, compute swept area
   (`distance × wing_opening_m / 1e6`, wing opening in decimetres), convert
   DDMM.mmm lat/lon to decimal degrees, fix the Alboran Sea (GSA 1–3)
   longitude sign — a Western-Med-specific correction, harmless (a no-op) if
   your `FILTER_AREAS` doesn't include GSA 1–3 — reshape into the
   standardized `dataframe1`/`dataframe2` format. Runs `validate_survey_data()`
   and `plot_sample_map()` (→ `survey_sample_coverage_map.png`,
   `survey_sample_coverage_map_by_year.png`).
3. **FG matching (Steps 3–4)** — `prepare_fg_lookup()` →
   `match_species_to_fg()` → `match_nominate_subspecies_fg()` → manual
   overrides (species codes absent from the MEDITS taxonomic list, e.g.
   `ARGRACU`→*Argyropelecus aculeatus*) → `fallback_match_fg_by_taxonomy()`
   (Genus→Family→Order→Class→Phylum) → `apply_seed_fg_rules()` (FGs with zero
   reference species, e.g. Holothuroidea→"Sea cucumbers") →
   `summarize_unresolved_species()` → `survey_unmatched_for_manual_review.csv`.
4. **Densities (Step 5)** — `match_samples_to_area()`,
   `assign_depth_stratum()`, `compute_sample_densities()`, restrict to
   `FILTER_AREAS`, `apply_catchability_correction()` (if the catchability CSV
   exists), `remove_sample_outliers()` (→ `survey_outliers_flagged.csv` if
   any flagged), `compute_fg_densities_by_stratum()`.
5. **Strata/area weighting (Steps 6–7)** — `compute_strata_area_by_area()`
   (cached to `strata_area_by_area.csv`), `weight_by_strata()`,
   `weight_by_area()`, `compute_species_densities_by_stratum()`,
   `weight_species_by_area()`.
6. **MEDITS plots (Step 8)** — `plot_fg_timeseries_by_area()`,
   `plot_fg_timeseries_regional()`, `plot_strata_profile()` →
   `survey_fg_density_timeseries.png`,
   `survey_fg_density_timeseries_regional.png`,
   `survey_fg_depth_strata_profile.png`.

   ```{r fg-timeseries-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_fg_timeseries_by_area() output — MEDITS trawl survey density by FG and area (strata-weighted)."}
   safe_include_graphics(
     "figures/survey_fg_density_timeseries.png",
     script = "01_biomass.R",
     out_dir_note = "<out_dir>/plots/survey_fg_density_timeseries.png"
   )
   ```

   ```{r strata-profile-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="plot_strata_profile() output — raw per-sample density across depth strata, by FG."}
   safe_include_graphics(
     "figures/survey_fg_depth_strata_profile.png",
     script = "01_biomass.R",
     out_dir_note = "<out_dir>/plots/survey_fg_depth_strata_profile.png"
   )
   ```
7. **MEDIAS acoustic survey (Step 9, sub-steps 9a–9g)** — load and collapse
   raw abundance/biomass records, map species codes via ASFIS, match to FG
   the same way as MEDITS, attach area (MEDIAS Handbook Table 1, hand-keyed
   `MEDIAS_HANDBOOK_AREAS` — as delivered this table only has Western Med
   entries, since that's this example's scope; extending to another region
   means adding that region's own Handbook rows, with a bathymetry-derived
   10–200 m fallback via `compute_strata_area_by_area()` covering any
   country/GSA combination the table doesn't have a row for), compute
   region-wide FG and species density, then
   **combine MEDITS + MEDIAS** into one FG-level and one species-level table
   (averaged on Year/FG or Year/species overlap, overlaps flagged/printed).
   Outputs: `medias_fg_annual_density_by_area.csv`,
   `medias_fg_annual_density_regional.csv`,
   `medias_fg_biomass_density_timeseries_regional.png`,
   `medias_fg_abundance_density_timeseries_regional.png`,
   `species_density_regional_combined.csv` (**contract file** — read by
   `lib_build_species_df_from_survey.R` for `03_pbqb-traits.R`).
8. **Excel export (Step 10)** — `survey_fg_annual_index.csv`,
   `survey_fg_annual_index_regional_combined.csv`, then
   `export_ecopath_ecosim_excel()` writes sheets `FG_spp_Ecopath`, `Ecopath`,
   `Ecosim`, `FG_spp_Ecosim` into `ecopath_ecosim_inputs.xlsx`. **Also writes
   a `PB_QB_spp` scaffold** — reusing `FG_spp_Ecopath`'s own `Species`/
   `FG_num`/`Density`/`prop_sp_fg` (renamed to `PB_QB_spp`'s convention:
   `Biomass`, `prop_biomass_FG`), with `PB`/`prop_biomass_PB`/`QB`/
   `prop_biomass_QB` all blank and a `Note` column — so `PB_QB_spp` exists in
   the workbook after this script alone, without needing `03_pbqb-traits.R`
   (Step 4) to have run first. This function has no PB/QB computation of its
   own (that's entirely Step 4's, from life-history parameters, not survey
   density), so the scaffold's numbers are genuinely blank, not estimated.
   Running Step 4 afterward replaces the scaffold with the real sheet
   automatically (`upsert_workbook_sheets()` always fully replaces a
   same-named sheet). Guarded against clobbering the other way: if
   `PB_QB_spp` already has real (non-`NA`) `PB` values — i.e. Step 4 already
   ran against this exact workbook — re-running this script leaves it
   untouched rather than overwriting it with a blank scaffold again.
9. **Complete `FG_spp_Ecopath`/`FG_lookup` (Step 11)** — union of observed +
   reference-catalog species, zero-fill species with no base-year density,
   attach taxonomy, defensive check for FGs with no species at all. Written
   via `upsert_workbook_sheets()`.
10. **`traits_ewe` sheet (Step 12)** — parse `FG_WMed.xlsx`'s `traits_ewe`
    sheet (alternating "N: Group Name" header rows / species rows,
    fill-down logic), reconcile against `dataframe2`, flag naming-variant and
    missing-traits species. Written via `upsert_workbook_sheets()`.
11. **Finalize sheet order (Step 13)** — `finalize_workbook_sheet_order()`.

## Outputs

* **CSVs** (in `out_dir`): `survey_sample_coverage_map.png` and other plot
  PNGs (in `out_dir/plots`), `survey_unmatched_for_manual_review.csv`,
  `survey_outliers_flagged.csv`, `strata_area_by_area.csv`,
  `medias_fg_annual_density_by_area.csv`,
  `medias_fg_annual_density_regional.csv`,
  `species_density_regional_combined.csv`, `survey_fg_annual_index.csv`,
  `survey_fg_annual_index_regional_combined.csv`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written/updated:
  `FG_spp_Ecopath`, `Ecopath`, `Ecosim`, `FG_spp_Ecosim`, `FG_lookup`
  (renamed to `FG`), `traits_ewe`, `PB_QB_spp` (scaffold, unless it
  already has real data from `03_pbqb-traits.R`), `AquaMaps_Depth_Adjustment`
  (only when `APPLY_AQUAMAPS_DEPTH_ADJUSTMENT <- TRUE` — see Step 7b), and
  `FG_Density_by_Stratum` (only when `STRATA <- TRUE` — Step 12c; density
  by FG x depth stratum, region-wide, covering whichever strata are
  actually in play).

---

# `AREA_MODE == "custom"` — a fully custom study area, in `01_biomass.R`

**Role:** Not a separate script — a config switch inside `01_biomass.R`
(`AREA_MODE <- "custom"`, instead of the `"westmed"` default) for **any
study area anywhere in the Mediterranean that doesn't line up with GSA
boundaries** — a single custom bounding box or shapefile (e.g. one EwE
model's own footprint, a bay, an MPA, or a region spanning parts of several
GSAs, in any part of the basin), or a custom subset/grouping of real GSAs.

**2026-09-19: merged from a separate script.** This used to be a second
script, `01_survey_density_custom.R`, run *instead of* `01_biomass.R`. It's
now merged into `01_biomass.R` itself, so there's one script and one set of
fixes/improvements both regions benefit from — see `01_biomass.R`'s own
header comment. `01_survey_density_custom.R` still exists on disk but only
`stop()`s with a pointer back here; it is not maintained.

Structurally identical to `AREA_MODE == "westmed"` for the MEDITS loading
and FG-matching stages (Steps 1–4 use the exact same shared-library
functions, unconditionally). What's genuinely different:

* **MEDIAS acoustic survey and GFCM stock-assessment biomass are skipped**
  (Step 9, and the stock-assessment/priority blocks right after it) —
  both are West-Med-subregion-scoped data sources (MEDIAS Handbook Table 1
  areas, STAR/RAM Legacy species-by-GSA assessments) with no defined
  meaning outside the named GSAs they're built around. `fg_index_regional_combined`/
  `species_density_regional_combined`/`fg_cv_log_combined` — the same
  variable names Step 10 onward reads either way — are built as plain
  MEDITS-only aliases instead.
* **`AREA_NAME`** — a user-supplied label for this custom study area (typed
  interactively via `rstudioapi::showPrompt()`, or hardcoded for the
  `daniel` user). Used to build a **dedicated output subfolder**
  (`out_dir/<AREA_NAME>/`), so a custom-region run never collides with the
  default Western-Med `out_dir` or with another custom run under a
  different `AREA_NAME`.
* **Study-area definition** — `CUSTOM_AREA_TYPE <- "bbox"` (default; a
  rectangle defined by `CUSTOM_BBOX <- c(xmin, xmax, ymin, ymax)`),
  `"shapefile"` (an arbitrary `sf` polygon read via `st_read()`), or **`"gsa"`**
  (one or more GFCM GSAs, picked out of the same GSA shapefile downloaded
  for `AREA_MODE == "westmed"`, via `CUSTOM_GSA_IDS <- c(...)` — e.g. a
  single GSA, or a different regional grouping than the fixed `1:11`
  Western-Med range). For `"bbox"`/`"shapefile"` the result is a
  single-feature `area_shp` with `AREA_ID_COL = "area_id"`, always
  `area_id = 1` — there is no GSA sub-division inside a custom region unless
  the user builds one. For `"gsa"`, `area_shp` can have **one row per
  selected GSA**, each keeping its own `gsa_num` as `area_id` (same
  convention as `AREA_MODE == "westmed"`), so a multi-GSA custom region is
  NOT forced down to a single `area_id = 1`.
* **`AreaID` is derived spatially** for `"bbox"`/`"shapefile"`, not read off
  MEDITS' own `area` (GSA) column — every sample gets `AreaID = 1`, then
  `filter_samples_by_area()` drops everything outside the custom boundary.
  For `"gsa"`, `AreaID` follows the selected GSA(s) directly, same as
  `AREA_MODE == "westmed"`.
* **Strata-area computation (Step 6)** — `compute_strata_area_by_area()`
  called with whichever `AreaID`s are actually in scope (a single custom
  polygon for `"bbox"`/`"shapefile"`, or the selected `CUSTOM_GSA_IDS` for
  `"gsa"`) — the function itself is agnostic to this distinction either way,
  no branch needed.

Three files stay at plain, unsuffixed names regardless of `AREA_MODE`,
because `02_fao_catches.R` and `03_pbqb-traits.R` read them by that exact
fixed name: `species_density_regional_combined.csv`,
`ecopath_ecosim_inputs.xlsx`, `strata_area_by_area.csv`. The dedicated
`AREA_NAME` output subfolder (not a filename suffix) is what keeps a custom
run's copies of these three from colliding with the default run's.

## Inputs

Same as `AREA_MODE == "westmed"` minus the MEDIAS/ASFIS inputs (the GSA
shapefile is still downloaded when `CUSTOM_AREA_TYPE == "gsa"`), plus (for
`CUSTOM_AREA_TYPE == "shapefile"`) a user-supplied boundary shapefile.

## Processing steps

Steps 1–8 and 11–13 are identical code to `AREA_MODE == "westmed"` (see
above), running against the custom `area_shp`/`AreaID`. Step 9 (MEDIAS +
stock-assessment + priority combine) is skipped entirely — its `else{}`
branch aliases `fg_index_regional`/`species_density_regional`/
`fg_cv_log_medits` (already computed in Steps 5–7) as the `*_combined`
tables instead, so the `species_density_regional` table feeding Step 10 is
MEDITS-only — no MEDITS+MEDIAS averaging step.

## Outputs

Same file set and workbook sheets as `AREA_MODE == "westmed"` (see above),
written to `out_dir/<AREA_NAME>/`, with the three contract files above
staying unsuffixed.

---

# Script 2 — `02_fao_catches.R`

**Role:** Pipeline Step 2 of 4. **Requires** Step 1 to have already run — it
reads `strata_area_by_area.csv` (written by Step 1) to convert catch totals
into a t/km² density comparable to Biomass. Running this before Step 1 fails
with a clear "file not found" error pointing back to Step 1.

Config-driven catch-data loader: one `DATASETS` registry entry per source
format, so `GFCM_2025` and (a currently-commented-out) `FAO_2026` share one
processing path (`format = "gfcm_regional"`), while `FAO_2020`
(`format = "legacy_excel"`) is handled separately as a genuinely different
file layout. `DATASET_VERSION <- "GFCM_2025"` is the active, confirmed,
working source; `FAO_2020` is present but its file location is
**unconfirmed** on disk.

## Inputs

| File | Format | Role |
|---|---|---|
| `<git_dir>/data/raw/FI_Regional_2025.1.0/GFCM_Capture_Quantity.csv` | csv | Raw GFCM catch quantities |
| `.../CL_FI_SPECIES_GROUPS.csv` | csv | Species code → name reference (also auto-downloaded from FAO's open-data catalog if not found under `pcloud_dir`) |
| `.../CL_FI_COUNTRY_GROUPS.csv` | csv | Country code → name reference |
| `.../CL_FI_WATERAREA_DIVISION.csv` | csv | FAO/GFCM division code → name reference; catches filtered to divisions `37.1.1`/`37.1.2`/`37.1.3` |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → FG reference, same as Script 1 |
| `strata_area_by_area.csv` | csv | **From Step 1.** Total study-area km² for the density conversion |

### Column reference — key input files

**`GFCM_Capture_Quantity.csv`** (raw catch, `<git_dir>/data/raw/FI_Regional_2025.1.0/`):

| Column | Used for |
|---|---|
| `SPECIES.ALPHA_3_CODE` | Joined against `CL_FI_SPECIES_GROUPS.csv` |
| `COUNTRY.UN_CODE` | Joined against `CL_FI_COUNTRY_GROUPS.csv` |
| `DIVISION.CODE` (`area_code_col`) | Joined against `CL_FI_WATERAREA_DIVISION.csv`; filtered to `37.1.1`/`37.1.2`/`37.1.3` (Western Med divisions — change for a different region) |
| `MEASURE` | Filtered to `"Q_tlw"` (live-weight tonnes) |
| `PERIOD` | Year |
| `VALUE` | Catch quantity |

**Reference CSVs** (same folder):

| File | Columns used |
|---|---|
| `CL_FI_SPECIES_GROUPS.csv` | `3A_Code` (species code) → `Name_En` (species name) |
| `CL_FI_COUNTRY_GROUPS.csv` | `UN_Code` → `Name_En` (country name) |
| `CL_FI_WATERAREA_DIVISION.csv` | `Code` (`area_join_col`) → `Name_En` (division name) |

**`FAO_2020` legacy Excel** (`format = "legacy_excel"`, alternate/unconfirmed
source):

| Column | Renamed to |
|---|---|
| `Species (scientific name)` | `Species` |
| `Area (FAO division)` | `Division` |
| `Area (FAO subarea)` | `Subarea` — filtered to `"Western Med (37.1)"` |
| `Year`, `Quantity`, `Country` | used as-is |

## Processing steps

1. **Config + dataset loading** — resolve directories, `load_gfcm_regional()`
   merges capture/species/country/area reference CSVs, filters to Western
   Med divisions, `MEASURE == "Q_tlw"`, `PERIOD >= 1995`. Builds
   `country_ts`/`species_ts` (Year × Country/Species × Division × Catch).
2. **Plots + audit CSVs** — country- and species-level catch time series
   (`country_catch_timeseries.png`, `species_catch_timeseries.png`),
   `unmatched_species_FG_assignment.csv`, `westmed_species_timeseries.csv`.
3. **Species → FG matching pipeline** (inline, Steps 0–4, same safety
   philosophy as Script 1's taxonomy fallback — a query is only
   auto-resolved if it maps to exactly **one** distinct FG, otherwise flagged
   ambiguous and never guessed):
   - Step 1: direct match (catch-record species text literally equals an
     `FG_name`).
   - Step 2: common name → scientific name via `rfishbase::common_names()`
     (queried **for** the known species list, not searched from all of
     FishBase — avoids cross-matching an unrelated species that happens to
     share a common name).
   - Step 3: FAO genus-level fallback for NEI/aggregate categories (e.g.
     "Sardinellas nei" → genus *Sardinella*).
   - Step 4: word-containment fallback (full token containment only).
   - Manual overrides (e.g. "Turbot" → *Scophthalmus maximus*, correcting a
     wrong FishBase common-name entry).
   - A catch-total sanity check confirms no species record is double-counted
     across matching stages.
   - Output: `species_fg_matched.csv` (one row per species, `status` =
     resolved/unresolved, `match_method`).
4. **FG-level Year timeseries** — joins `species_ts` (has Year, no FG) against
   `species_fg_matched.csv`'s Species→FG mapping, re-aggregates to Year × FG.
   Output: `fg_catch_timeseries_<DATASET_VERSION>.csv` (landings-only, **not**
   confirmed to include discards — read by `03_pbqb-traits.R`'s optional
   `FG_YIELD_SOURCE = "fg_catch_csv"` path).
5. **Add `Catches_Ecopath`/`Catches_Ecosim` to the workbook** — reads
   `strata_area_by_area.csv`, sums to a total study-area km², calls
   `add_catches_to_ecopath_workbook()`. **Known, documented limitation:**
   GFCM Divisions 37.1.1–37.1.3 cover GSA 1–11 **plus** GSA 12 (Northern
   Tunisia), which the survey's `FILTER_AREAS` does not include — the catch
   density is therefore a slight overestimate relative to the Biomass area
   (GFCM's Division-resolution reporting can't isolate GSA 12's specific
   contribution to subtract it out).

**Not implemented — no fleet or gear disaggregation anywhere in this
script.** `GFCM_Capture_Quantity.csv` (the actual, confirmed source above)
resolves only to Country × Species × Division × Year — this is GFCM's
STATLANT-derived aggregate capture-production product, not its
Operational-Unit-level data. GFCM *does* collect catch/effort broken down
by **Operational Unit** (fleet segment × ISSCFG gear code × GSA), via the
DCRF's Task 2 (Catch)/Task 4 (Fleet)/Task 5 (Effort) and the Regional
Fleet Register, submitted by national correspondents — but that's a
genuinely different GFCM product from this file, and isn't confirmed to
have the same kind of open bulk-download `GFCM_Capture_Quantity.csv` does
(it feeds GFCM's own stock-assessment working groups rather than being
published as an open dataset the way STATLANT capture-production is).
Getting real fleet/gear resolution into this pipeline means sourcing that
Operational-Unit data through GFCM's DCRF/SAC channels, or — for the
EU-flagged portion of the fleet specifically — STECF's FDI (Fisheries
Dependent Information) database, which is the same source already
flagged elsewhere in this pipeline as the likely next step for a
discard-inclusive catch figure. Neither is currently plugged in anywhere;
`03_pbqb-traits.R`'s `Fmort_species`/`Fmort_FG` (below) inherit this same
limitation, since they're computed downstream of this file.

**Single fleet by default; multi-fleet via an optional `fleet_structure`
input.** `add_catches_to_ecopath_workbook()` (`lib_survey_fg_density_
functions.R`) now takes an optional `fleet_structure` argument: a table of
`FG_num`, `Fleet`, `prop_catch` (the share, 0–1, of each FG's total catch
attributed to each fleet). Left at its default (`NULL`, and that is still
this script's own default), the workbook is exactly what it always was —
one implicit fleet, `Catches_Ecopath`'s combined `Catch_<year>` columns,
`Catches_Ecosim`'s meta-rows with no `Fleet` field, no `Fleet_Structure`
sheet. Point `FLEET_STRUCTURE_PATH` (near the top of this script) at a CSV
in that `FG_num,Fleet,prop_catch` shape to split catches by fleet instead —
any FG *not* listed there still defaults to a single implicit `"Fleet_1"`,
so a partial table (only the FGs you actually have a fleet split for) is
fine. Once any FG resolves to more than one fleet, **every** FG row in
`Catches_Ecopath` switches to fleet-suffixed columns
(`Catch_<year>_<Fleet>`), `Catches_Ecosim` gets one column-set per FG ×
Fleet, and the resolved fleet table itself is written out as a new
`Fleet_Structure` sheet for traceability.

This only splits catch *after the fact* — it does not source fleet/gear
data itself (see the "not implemented" note above: `GFCM_Capture_
Quantity.csv` still carries no fleet/gear column), so the `fleet_structure`
CSV has to come from somewhere else. **Three ways to get one:** (1) GFCM
Operational Units or STECF FDI, assembled by hand; (2) a hand-built
allocation; or (3) point `FLEET_STRUCTURE_PATH` at
`file.path(out_dir, "fleet_structure_from_sau.csv")` — written automatically
by Script 2a (`02a_fisheries_multisource.R`), country × gear proportions
(e.g. `Spain - bottom trawl`, `France - purse seine`) derived from SAU's own
real fleet-level catch shares per FG. This is an **approximation** — it
borrows SAU's gear mix and applies it to this script's own FAO-GFCM totals,
two different catch reconstructions for the same species/FG, not a real
fleet-resolved FAO-GFCM figure, since FAO-GFCM has no gear data of its own
to resolve directly. If the fleet split should come from the *same*
reconstruction as the catch totals, use SAU's own `Catches_by_Fleet` sheet
(written directly by Script 2a, see below) instead of routing through here.
`03_pbqb-traits.R`'s `Fmort_species`/`Fmort_FG` remain unaffected either way —
they're computed from the same un-split `fg_catch_timeseries_*.csv`
regardless of whether `Catches_Ecopath`/`Catches_Ecosim` end up split by
fleet.

**`CATCHES_DATA_SOURCE`: pick which source feeds `Catches_Ecopath`/
`Catches_Ecosim` — without duplicating Script 2a's own logic here.** This
selector (`NULL | "SAU" | "SAU_no_unreported" | "FishMIP"`, near
`FLEET_STRUCTURE_PATH`) just **selects** which of Script 2a's already-written
contract files to read — the same "point this script at 2a's own output"
pattern `03_pbqb-traits.R`'s `FISHERIES_DATA_SOURCE` already uses — it does
**not** re-implement any of Script 2a's SAU/FishMIP extraction. **Defaults to
`"SAU"`** (not `NULL`) — discards/unreported/IUU catch and a real fleet ×
gear × country split are treated as always-wanted, not an occasional opt-in.
When `"SAU"`/`"SAU_no_unreported"`, this script reads Script 2a's
`fg_catch_timeseries_<SOURCE>.csv` for the catch **total** *and*
`fleet_structure_from_sau.csv` for the fleet **split** — both from the *same*
SAU reconstruction, so this is fully self-consistent, unlike routing SAU's
fleet shares through `FLEET_STRUCTURE_PATH` onto a FAO-GFCM total (that combo
still works and is unaffected, but `CATCHES_DATA_SOURCE`'s own SAU fleet
structure takes priority over `FLEET_STRUCTURE_PATH` whenever both are set).
Set to `"FishMIP"`, only the total is replaced — FishMIP's catch side has no
gear dimension, so there's no fleet split to auto-apply; `FLEET_STRUCTURE_PATH`
still works normally on top of it if you want one anyway. Set to `NULL`
yourself to restore the original FAO-GFCM-only, landings-only, single-fleet
figure for a specific run. `"FAO_GFCM"` is **not** a valid value (it would
just mean reading back this script's own output) and errors clearly if set.
**This script does not auto-run Script 2a for you** — unlike Script 2a's own
`AUTO_RUN_MISSING_SOURCES`, which auto-runs SAU's download and, separately,
auto-runs *this* script for FAO-GFCM — because Script 2 auto-sourcing Script
2a would risk recursing straight back into Script 2. A missing contract file
is therefore a clear `stop()` naming which script to run first, never a
silent fallback to `NULL`'s FAO-GFCM behavior.

**Why the FAO-GFCM auto-run (Script 2a's own STEP 7) is safe from this new
default:** since `CATCHES_DATA_SOURCE` now defaults to `"SAU"`, and Script
2a's SAU step (STEP 5) runs *before* its FAO-GFCM step (STEP 7) in the same
run, `fg_catch_timeseries_SAU.csv` already exists by the time STEP 7
auto-sources this script — so STEP 7 explicitly sets a global
`CATCHES_DATA_SOURCE <- NULL` right before sourcing it, overriding this
script's own default (via the same "reuse if the caller already set it"
pattern `DATASET_VERSION` uses). Without that override, the auto-run would
silently hand back SAU's total mislabeled as the FAO-GFCM figure that
`DataSources_Catch`/`FISHERIES_DATA_SOURCE = "FAO_GFCM"` depend on being
genuinely FAO-GFCM-only.

## Outputs

* **CSVs:** `unmatched_species_FG_assignment.csv`,
  `westmed_species_timeseries.csv`, `species_fg_matched.csv`,
  `fg_catch_timeseries_<DATASET_VERSION>.csv`.
* **Plots:** `country_catch_timeseries.png`, `species_catch_timeseries.png`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written: `Catches_Ecopath`,
  `Catches_Ecosim`, plus `Fleet_Structure` if `FLEET_STRUCTURE_PATH` was set
  and resolved to more than one fleet.

---

# Script 2a — `02a_fisheries_multisource.R` (optional, runs alongside Script 2)

**Role:** Optional additional Step 2 — **requires** Step 1 to have already
run (reads `strata_area_by_area.csv` and the FG reference, same contract as
Script 2/Script 4). Adds **three genuinely different, already-assembled**
catch/effort sources on top of Script 2's own FAO-GFCM source, so all four
can be compared side by side and one of them selected to drive Fishing
mortality (F) and PB = M + F downstream in Script 4 — **without changing
Script 4's own F/PB math at all**, only the input file it reads.

## Why four sources, not one

None of the four is a strict upgrade over the others — each has real gaps of
its own, which is why every one gets written into the workbook
(`DataSources_Catch`) rather than silently picked for the user:

* **SAU (Sea Around Us)** — reconstructed catch (adds back
  unreported/discarded/IUU catch on top of official landings), resolved to
  **country × gear × species × year**. The only source here with a real gear
  dimension, so `Catches_by_Fleet` (below) is built from SAU alone. Yield =
  SAU's own reconstructed `tonnes` (already landings + discards +
  unreported/IUU, by SAU's own definition).
* **SAU w/o unreported** — the same SAU extract, filtered to
  `reporting_status == "Reported"`. This is SAU's own reporting-status
  split, not a clean "landings + discards" figure — kept as its own labeled
  source rather than mislabeled.
* **FishMIP** — a **SAU-derived calibration product** (not independent of
  SAU), resolved to country × sector × functional group × year, with
  reported/IUU/discards already split out. No species dimension and no gear
  dimension on the catch side (gear only exists on FishMIP's separate
  *effort* file — see `Fishing_Effort_by_Fleet` below). Yield = `reported +
  discards` (IUU deliberately excluded, kept as an audit column
  `tonnes_incl_iuu`, per the "landings + discards" definition used
  throughout this script). FishMIP's own functional groups are **not** this
  pipeline's FG scheme — see `FISHMIP_FG_CROSSWALK_PATH` below.
* **FAO-GFCM** — this pipeline's own existing source (Script 2's output),
  passed through unchanged. Landings only (no discards), Division-resolution
  (not GSA), no fleet/gear dimension — same caveats as Script 2 itself.

**Area scope caveat** (same pattern as the GSA-12/Division mismatch already
flagged for Script 2): SAU and FishMIP resolve to **country/EEZ**, not GSA.
Restricting either one to "the same area as the survey" means restricting to
the *countries* whose EEZs overlap that GSA range — an approximation, not an
exact area match. A country with only a small coastal sliver inside the
modeled GSA range still contributes its entire national catch/effort. This
is unavoidable at these sources' native resolution and is reported
explicitly at runtime (`TARGET_COUNTRIES_NOTE`), not silently absorbed.

## Inputs

| File | Format | Notes |
|---|---|---|
| `strata_area_by_area.csv` | csv | From Step 1 |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → FG reference, same as Scripts 1/2 |
| `<pcloud_dir>/data/fisheries/sau_raw_combined_west_med.csv` | csv | SAU raw extract (from `sau_west_med_full.R`/`sau_west_med_analysis.Rmd`) — country × gear × species × year, `reporting_status` column |
| `<pcloud_dir>/data/fisheries/effort_histsoc_*.parquet` | parquet | FishMIP effort — `saup`, `sector`, `gear`, `f_group`, `nom_active` (kW·days) |
| `<pcloud_dir>/data/fisheries/calibration_catch_histsoc_*.parquet` | parquet | FishMIP catch — `saup`, `sector`, `f_group`, `reported`, `iuu`, `discards` |
| `fg_catch_timeseries_<FAO_GFCM_DATASET_VERSION>.csv` | csv | Script 2's own output, read unchanged |
| `FISHMIP_FG_CROSSWALK_PATH` (optional) | csv | User-supplied `fishmip_f_group` → `FG_num` crosswalk — without it, FishMIP is informational-only |

Each source **degrades gracefully** if its input is missing — a clear
console message explains what to run first, the script does not hard-fail.

## Processing steps

1. Configuration — `TARGET_ISO3` (the 6-country Western Med default: Spain,
   France, Italy, Tunisia, Algeria, Morocco), `START_YEAR`/`END_YEAR`
   (1994–2019, matching SAU's public reconstruction cutoff), `N_TOP_GEARS`
   (top 8 gears + "Other gear" bucketing, same convention as the attached
   SAU/FishMIP analysis scripts, to avoid a combinatorial explosion of
   near-empty FG × Fleet × Year rows).
2. `DATA_SOURCE <- "SAU" | "SAU_no_unreported" | "FishMIP" | "FAO_GFCM"` —
   which source drives Script 4's Yield/Fmort/PB = M+F chain. Does **not**
   change this script's own output — every available source is still
   written to `DataSources_Catch`/`Catches_by_Fleet`/`Fishing_Effort_by_Fleet`
   regardless.
2b. `AUTO_RUN_MISSING_SOURCES <- TRUE` (default) — when SAU's raw extract or
   Script 2's `fg_catch_timeseries_<FAO_GFCM_DATASET_VERSION>.csv` is missing,
   this script now **runs the upstream step itself** instead of just telling
   you to go run something else by hand: SAU auto-downloads
   `sau_raw_combined_west_med.csv` straight from SAU's API (per-EEZ, via the
   new `lib_sau_extraction.R` — the same extraction logic as
   `sau_west_med_analysis.Rmd`'s own "Part 0" chunk, lifted out so it doesn't
   require knitting that Rmd by hand every time), and FAO-GFCM auto-sources
   Script 2 (`02_fao_catches.R`) itself, first setting a global
   `DATASET_VERSION` to match this script's own `FAO_GFCM_DATASET_VERSION` so
   the auto-run can't silently regenerate the *wrong* dataset version's file.
   Both attempts are wrapped so a failure (offline, upstream file genuinely
   missing, etc.) degrades to the same "not found, skipping" message as
   before — auto-run is a shortcut, not a new failure mode. **FishMIP is
   deliberately excluded** — its two parquet files are a manual download
   from the FishMIP Input Explorer website, not something any R script here
   can generate, so a missing FishMIP file always just skips with a message
   pointing at that download page. Set `AUTO_RUN_MISSING_SOURCES <- FALSE`
   to go back to the old "message + skip" behavior for every source.
3. Species → FG matching — same `match_species_to_fg()` →
   `match_nominate_subspecies_fg()` → `fallback_match_fg_by_taxonomy()`
   cascade Script 1/Script 2 use, so a species is matched (or left
   unresolved) consistently across the whole pipeline.
4. SAU — year/country filter, species→FG match, builds country-year totals
   (both "SAU total" and "SAU w/o unreported"), FG-year totals, species-year
   totals (for `landings_by_species_gsa_year_<SOURCE>.csv`), and two
   fleet-level (country × top-N-gears, "Other gear" bucket) sheets:
   `Catches_by_Fleet` — **wide**, one row per FG, **one column per Fleet**
   (e.g. `Spain - bottom trawl`, `France - longline`, ...), value = that
   FG × Fleet's catch (t) **averaged over `YEAR_ECOPATH`** (a config
   constant in this script that must match Step 1's own `YEAR_ECOPATH`) —
   same single-snapshot convention `add_catches_to_ecopath_workbook()`
   already uses for `Catches_Ecopath`, but built directly from SAU's real
   per-fleet annual values rather than a static proportion split, so each
   fleet's actual year-to-year share is reflected; and
   `Catches_by_Fleet_AllYears` — the same data in **long** form (one row per
   FG × Fleet × Year), kept for anyone who needs the full annual detail the
   wide sheet collapses out of. Both already include discards (+
   unreported + IUU) — see "Why four sources, not one" above; there is no
   separate landings-only vs. discards-only split at fleet resolution,
   only SAU's already-combined total.
5. FishMIP (if the `arrow` package and both parquet files are available) —
   builds `Fishing_Effort_by_Fleet` (the **only genuine effort data** in this
   whole pipeline — SAU has no effort variable at all) and country-level
   catch totals; FG-level totals only if `FISHMIP_FG_CROSSWALK_PATH` is
   supplied.
6. FAO-GFCM — reads Script 2's own `fg_catch_timeseries_*.csv` unchanged.
7. `DataSources_Catch` — every available source's country-level and FG-level
   Yield, stacked with a `source` column, for side-by-side comparison
   regardless of which one is selected as `DATA_SOURCE`.
8. Writes `Catches_by_Fleet`, `Catches_by_Fleet_AllYears`,
   `Fishing_Effort_by_Fleet`, `DataSources_Catch` into the shared workbook
   via `upsert_workbook_sheets()` (idempotent — re-running replaces these
   sheets, doesn't duplicate them).
9. Writes `fg_catch_timeseries_<DATA_SOURCE>.csv` (and, for
   SAU/SAU_no_unreported only, `landings_by_species_gsa_year_<DATA_SOURCE>.csv`)
   — the contract files Script 4's `FG_YIELD_SOURCE`/`YIELD_SOURCE` mechanism
   reads. **Wiring into Script 4:** Script 4 has its own
   `FISHERIES_DATA_SOURCE` selector (see Script 4 below) which, when set to
   match this script's `DATA_SOURCE`, automatically points
   `FG_CATCH_CSV_PATH`/`LANDINGS_CSV_PATH` at these files — no manual path
   editing needed in Script 4 as long as the two `DATA_SOURCE`/
   `FISHERIES_DATA_SOURCE` values match.

## Outputs

* **CSVs:** `fg_catch_timeseries_<DATA_SOURCE>.csv`, and (SAU/SAU_no_unreported
  only) `landings_by_species_gsa_year_<DATA_SOURCE>.csv`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, sheets written: `Catches_by_Fleet`
  (wide — FG rows × Fleet columns, averaged over `YEAR_ECOPATH`) and
  `Catches_by_Fleet_AllYears` (the same data, long, one row per FG × Fleet ×
  Year) — both SAU only, the only source with real gear × country × species
  resolution; `Fishing_Effort_by_Fleet` (FishMIP only — genuinely different
  fleet resolution than `Catches_by_Fleet`, never to be joined directly to
  it), `DataSources_Catch` (every available source, country- and FG-level,
  side by side).

---

# Script 3 — `03b_ecobase.R`

**Role:** Pipeline Step 3 of 4. Independent of Steps 1/2 — can run anytime
before Step 4. **Optional** input to Step 4 for literature-based PB/QB
gap-filling and comparison against the empirical formula estimates.

Queries **EcoBase** (ecobase.ecopath.org), an open-access repository of
published Ecopath models, via its SOAP/XML web service — the official
EcoBase R example ported from RCurl/XML to httr/xml2. The script author
notes this was **not tested against the live service** from within the
authoring sandbox (no network access to `sirs.agrocampus-ouest.fr`) — the
diagnostics throughout are there specifically to surface an endpoint change,
not just fail silently.

**Fixed: `merge(mediterranean_models, all_status, ...)` — "x has duplicated
column name [comments_objectives]".** Root cause traced to
`xml_node_to_dt_row()` (formerly an inline anonymous function, at two call
sites — the model-list parse and, in `fetch_model_inputs()`, the per-model
group parse): when a single `<model>` (or `<group>`) XML node repeats a
child tag — confirmed EcoBase's own XML does this, e.g. a duplicated
`<comments_objectives>` — `setNames(as.list(xml_text(fields)),
xml_name(fields))` produced a list with two entries sharing one name, and
`as.data.table()` on that silently creates a data.table with two
same-named columns. data.table allows that at creation time; it only
rejects it the moment something needs unique names — which is why the
error surfaced three functions and one `rbindlist()` away from where the
duplicate actually originated, at the `merge()` call. Fix: both parse
sites now go through one shared `xml_node_to_dt_row()` that detects a
repeated tag name and collapses its values into ONE column, joined with
`"; "` — nothing is dropped, there's just no longer two columns fighting
over one name.

## Inputs (network)

| Endpoint | Purpose |
|---|---|
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client_3.php` | Full model list |
| `sirs.agrocampus-ouest.fr/EcoBase/php/webser/soap-client.php?no_model=<id>` | Per-model input values (Biomass/PB/QB per group) |

`pcloud_dir`/`git_dir` are resolved for consistency with the other scripts
but not actually used — this script is self-contained.

## Processing steps

1. Fetch the full EcoBase model list; save `ecobase_model_list_full.csv`.
2. Filter geographically to the Mediterranean, using each model's
   `geographic_extent` bounding box (not text/keyword matching on
   country/ecosystem names — more robust). Save
   `ecobase_mediterranean_models.csv`.
3. Filter further to `dissemination_allow == "true"` — EcoBase's own flag for
   which models actually have downloadable input data (per the site, only
   ~233 of ~500 listed models do; the rest are metadata/reference-only).
   `KNOWN_RELEVANT_MODEL_IDS` is the resulting query list.
4. Fetch per-model group-level Biomass/PB/QB for each disseminable model
   (`fetch_model_inputs()`), explicitly detecting the "data not available
   despite the flag" case (confirmed against a real response for model 418).
   `-9999` (Ecopath's own sentinel for "not provided") is converted to `NA`.
5. Save a full status audit (`ecobase_mediterranean_models_status.csv`) and
   the failure-only subset (`ecobase_models_without_data.csv`).
6. Compile the final summary table: `FG_name, PB, QB, EwE_model, year,
   authors` (model metadata columns matched by best-guess candidate names
   against `model_list`, since EcoBase's exact field names aren't confirmed).

## Outputs

* `ecobase_model_list_full.csv`, `ecobase_mediterranean_models.csv`,
  `ecobase_mediterranean_models_status.csv`, `ecobase_models_without_data.csv`,
  `ecobase_literature_pb_qb_raw.csv`,
  **`ecobase_literature_pb_qb_simple.csv`** (the file `03_pbqb-traits.R` reads).
* No workbook writes from this script directly — `03_pbqb-traits.R` is what
  merges this CSV into the workbook's `Ecobase` sheet.

### Column reference — `ecobase_literature_pb_qb_simple.csv`

| Column | Meaning |
|---|---|
| `FG_name` | EcoBase group name (matched to your own `FG_name` by text in `03_pbqb-traits.R` — group naming won't line up automatically across different published models) |
| `PB` | Literature Production/Biomass (`-9999`, EcoBase's own "not provided" sentinel, converted to `NA`) |
| `QB` | Literature Consumption/Biomass (same `-9999`→`NA` handling) |
| `EwE_model` | Source model name |
| `year` | Source model's reference year |
| `authors` | Source model's authors |

---

# Script 4 — `03_pbqb-traits.R`

**Role:** Pipeline Step 4 of 4 — run **last**. Requires Step 1's
`species_density_regional_combined.csv`. Optionally uses Step 2's
`fg_catch_timeseries_<DATASET_VERSION>.csv` (`FG_YIELD_SOURCE` toggle) and
Step 3's `ecobase_literature_pb_qb_simple.csv` — both degrade gracefully with
a clear message if missing, they don't hard-fail this script.

Estimates Production/Biomass (PB) and Consumption/Biomass (QB) for every EwE
Functional Group, following the taxon-specific methods in *"Quick guide on
how to calculate P/B and Q/B for EwE models"* (Vilas, Coll, Piroddi,
Steenbeek). Growth parameters, length-weight a/b, maturity, body weight,
longevity, trophic level, temperature, depth and aspect ratio are all fetched
automatically from FishBase + SeaLifeBase (`rfishbase` ≥ 4.0.0 required — the
script hard-stops with install instructions if an old, broken pre-4.0 CRAN
build, or a `duckdbfs` install missing `duckdb_config`, is detected).

## Inputs

| File | Format | Role |
|---|---|---|
| `species_density_regional_combined.csv` | csv | **From Step 1.** Reshaped via `lib_build_species_df_from_survey.R::build_species_df_from_survey()` into `species_df` (Species, FG, Biomass density, Yield=NA) |
| `fg_catch_timeseries_GFCM_2025.csv` (or `fg_catch_timeseries_<FISHERIES_DATA_SOURCE>.csv`) | csv (optional) | **From Step 2, or Step 2a if `FISHERIES_DATA_SOURCE` is set.** FG-level catch, used to compute a per-FG fishing-mortality rate `Fmort_FG` (`FG_YIELD_SOURCE = "fg_catch_csv"`) |
| `strata_area_by_area.csv` | csv | **From Step 1.** Study area, needed to convert FG catch totals into a density comparable to Biomass |
| `ecobase_literature_pb_qb_simple.csv` | csv (optional) | **From Step 3.** Literature PB/QB for gap-filling/comparison |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet `fg_wmed_95` | xlsx | FG reference (loaded by explicit sheet **name**, since the workbook also has an unrelated `fg_ebro delta_21` sheet) |
| `landings_by_species_gsa_year.csv` (or `landings_by_species_gsa_year_<FISHERIES_DATA_SOURCE>.csv`) | csv (optional; placeholder path when `FISHERIES_DATA_SOURCE` is `NULL`) | Species-level landings, for a species-level Fmort path (`YIELD_SOURCE = "landings_csv"`) — real data when Step 2a's `FISHERIES_DATA_SOURCE` is set to `"SAU"`/`"SAU_no_unreported"` |
| FishBase / SeaLifeBase (`rfishbase`) | web API | Growth, length-weight, maturity, ecology/trophic level, swimming/aspect-ratio traits |
| WoRMS (`worrms`, via `lib_worms_taxonomy_lookup.R`) | web API | Taxonomic classification (Class → dispatch group: fish/mammal/seabird/phytoplankton/invertebrate) |
| FishLife R package | local DB (`FishBase_and_RAM$ParentChild_gz`) | Phylogenetically-imputed Loo/K/M/Winf, walking up the taxonomic tree when a direct match is missing |
| TropFishR R package | local | `M_empirical()` — Hoenig (1983), Then et al. (2015), Alverson & Carney (1975) natural-mortality estimators |

### Column reference — key input files

**`species_density_regional_combined.csv`** (from Step 1 — required columns,
enforced by `build_species_df_from_survey()`):

| Column | Meaning |
|---|---|
| `Year` | Survey year |
| `FG_num`, `FG_name` | Functional Group id/name |
| `ScientificName` | Species scientific name |
| `mean_density` | Species density, t/km² |

**`fg_catch_timeseries_GFCM_2025.csv`** (from Step 2, optional):

| Column | Meaning |
|---|---|
| `Year` | Catch year |
| `FG_num`, `FG_name` | Functional Group id/name |
| `Catch_t` | Total landed catch, tonnes (region-wide, landings-only) |

**`strata_area_by_area.csv`** (from Step 1):

| Column | Meaning |
|---|---|
| `AreaID` | GSA number (or `1` for a custom-region run) |
| `Stratum` | Depth-stratum number |
| `area_km2` | Stratum area within that GSA/region, km² |
| `prop` | That stratum's proportion of the area's total |

**`FG_WMed.xlsx`, sheet `fg_wmed_95`** — loaded by explicit sheet **name**
(not position), since the same workbook also has an unrelated
`fg_ebro delta_21` sheet. `readxl` disambiguates duplicate-looking headers
as `FG_name...2`/`FG_name...7` — handled with a flagged best-guess in the
code rather than assumed.

**`landings_by_species_gsa_year.csv`** (placeholder path — not yet a real
data source; schema below is what `attach_yield_from_landings()` expects if
this is ever wired up):

| Column | Meaning |
|---|---|
| `ScientificName` | Species scientific name |
| `Year` | Landings year |
| `AreaID` | GSA number |
| `catch_t` | Landed catch, tonnes |

```{r pbqb-comparison-fig, echo=FALSE, eval=TRUE, results="asis", fig.cap="fish_PB_methods_comparison.png — all 6 fish PB/M methods per species, chosen method highlighted."}
safe_include_graphics(
  "figures/fish_PB_methods_comparison.png",
  script = "03_pbqb-traits.R",
  out_dir_note = "<out_dir>/plots/fish_PB_methods_comparison.png"
)
```

## Processing steps

1. **Setup** — package loading, `rfishbase`/`duckdbfs` version guard, GUI or
   console progress tracker across 13 named pipeline stages.
2. **Load `species_df`** (`SPECIES_DF_SOURCE = "survey"`, the real path) via
   `lib_build_species_df_from_survey.R`.
3. **Fishing mortality attachment** — two independent attachment points,
   both driven by a single **`FISHERIES_DATA_SOURCE`** selector (set near
   `out_dir`/`pcloud_dir`/`git_dir` at the top of this script):
   `NULL | "SAU" | "SAU_no_unreported" | "FishMIP" | "FAO_GFCM"`, matching
   Script 2a's own `DATA_SOURCE` values one-to-one. When set, it
   automatically points `LANDINGS_CSV_PATH`/`FG_CATCH_CSV_PATH` at the
   `_<FISHERIES_DATA_SOURCE>.csv` files Script 2a writes, and flips
   `YIELD_SOURCE`/`FG_YIELD_SOURCE` on to match — no manual path editing
   needed as long as this matches Script 2a's `DATA_SOURCE` for the run that
   produced those files. **Defaults to `"SAU"`** (not `NULL`) — species-level
   resolution and discards/unreported/IUU catch are treated as always-wanted
   for this chain, matching `02_fao_catches.R`'s own `CATCHES_DATA_SOURCE`
   default (also `"SAU"`, same reasoning). Set to `NULL` yourself to restore
   the original hardcoded placeholder behavior from before Script 2a
   existed. Every source's own catch/effort still appears in the
   workbook's `DataSources_Catch`/`Catches_by_Fleet`/`Fishing_Effort_by_Fleet`
   sheets (written by Script 2a) regardless of this setting — only *this*
   script's own Fmort/PB = M+F chain follows whichever source is selected
   here.
   * (A) species-level, via `attach_yield_from_landings()` — active only for
     `FISHERIES_DATA_SOURCE %in% c("SAU", "SAU_no_unreported")` (the only two
     sources with species-level resolution); inactive (`YIELD_SOURCE =
     "none"`) for `NULL`/`"FishMIP"`/`"FAO_GFCM"`.
   * (B) FG-level, via `fg_catch_timeseries_*.csv` — active
     (`FG_YIELD_SOURCE = "fg_catch_csv"`) for any non-`NULL`
     `FISHERIES_DATA_SOURCE` — computed as `Yield_FG / Biomass_FG` per FG, applied to fish
     PB inside `calc_fish()` as a **fallback** when species-level Yield is
     absent (species-level always takes priority when both exist).
     **Known, unresolved caveat (same root cause as `Catches_Ecopath`/
     `Catches_Ecosim` below):** `Yield_FG`'s denominator (`area_total_km2`,
     summed from `strata_area_by_area.csv`) is the *survey's* GSA-filtered
     area, while its numerator (`fg_catch_timeseries_*.csv`'s `Catch_t`) is
     filtered by GFCM Division, which for the Western Med example covers
     GSA 1–11 **plus** GSA 12 — so `Yield_FG`, and therefore `Fmort_FG` and
     the fish PB it feeds, is a **slight overestimate** wherever this path
     is enabled. Not fixable from this side (GFCM's catch data can't be
     split below Division resolution); see the same caveat under
     [Script 2 — `02_fao_catches.R`] for the full explanation.
     **Also not implemented — no fleet or gear disaggregation:** both
     `Fmort_species` (path A, inactive) and `Fmort_FG` (path B) are a
     single Yield/Biomass ratio lumped across the whole region and every
     gear/fleet that landed that species/FG — neither the
     `landings_by_species_gsa_year.csv` placeholder schema nor
     `fg_catch_timeseries_*.csv` carries a fleet or gear column, so
     there's nothing here that could produce a fleet- or gear-specific F
     even once a real landings file is plugged in. See the matching note
     under [Script 2 — `02_fao_catches.R`] for what a real fix would
     require (GFCM's Operational-Unit data or STECF's FDI database).
     **`Fmort` itself stays single-fleet even where `Catches_` sheets are
     split:** unlike `Catches_Ecopath`/`Catches_Ecosim` (which can now be
     split by fleet via `fleet_structure` — see the "Single fleet by
     default; multi-fleet via an optional `fleet_structure` input" note
     under [Script 2 — `02_fao_catches.R`]), this script never sees
     `fleet_structure` and computes `Fmort_species`/`Fmort_FG` from the
     same un-split `fg_catch_timeseries_*.csv` regardless. A genuinely
     fleet-specific F would need Yield split by fleet *before* this ratio,
     not just the `Catches_` sheets split after the fact.
4. **Taxonomic classification** — `worms_taxonomy_lookup()` for every species
   in `species_df`, dispatch-classified into `fish` / `mammal` / `seabird` /
   `phytoplankton` / `invertebrate` (default) by WoRMS Class, with a
   synonym→accepted-name resolution fallback for species WoRMS returns no
   Class for.
5. **Trait fetching (Step 2a–2f)** — `select_best_rows()` picks the best
   population-specific study per species from FishBase/SeaLifeBase
   multi-record tables, tiered: Western-Med locality keywords → broader
   Mediterranean keywords → most recent year → first available.
   `extract_provenance()` captures Locality/Year/Reference citation for each
   selected row (used later for the `References` sheet/CSV).
6. **Gap-filling** — `fill_by_taxonomic_proximity()`: local within-dataset
   donors first (Genus→Family→Order→Class, restricted to directly-computed
   values), then, for methods in `EXTERNAL_FORMULA_REGISTRY` (fish/mammal/
   general Gascuel formulas plus raw-trait identity mappings), an **external**
   FishBase/SeaLifeBase donor lookup — querying other species sharing the
   same taxonomic rank, not limited to species already in `species_df`.
7. **Taxon-specific PB/QB calculation** (dispatch by group) — the diagram
   below shows the actual dispatch and method-cascade logic:

   ```{=html}
   <div class="mermaid">
   %%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
   flowchart LR
       T["worms_taxonomy_lookup()&lt;br/&gt;WoRMS Class"] --> F["calc_fish()&lt;br/&gt;PB: 6 M methods + Fmort&lt;br/&gt;QB: 4 methods"]
       T --> M["calc_mammal()&lt;br/&gt;PB: Siler surv. to Gascuel&lt;br/&gt;QB: Innes&amp;Trites to Q/P=3"]
       T --> S["calc_seabird()&lt;br/&gt;PB: Gascuel only&lt;br/&gt;QB: Nilsson-Nilsson (no Q/P=3)"]
       T --> I["calc_invertebrate() (default)&lt;br/&gt;PB: Tumbiolo/Brey to Gascuel&lt;br/&gt;QB: Q/P=3 only"]
       F --> D["fill_by_taxonomic_proximity()&lt;br/&gt;local then external donors"]
       M --> D
       S --> D
       I --> D
       D --> AGG["FG-level aggregation&lt;br/&gt;biomass-weighted mean"]
       AGG --> OUT["Ecobase gap-fill -&gt; Ecopath_PBQB&lt;br/&gt;(PB_QB/PB_QB_spp/Ecobase CSVs)"]

       classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
       classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
       classDef teal fill:#E1F5EE,stroke:#0F6E56,color:#04342C
       classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
       class T gray
       class F,M,S,I purple
       class D teal
       class AGG,OUT coral
   </div>
   ```

   *Phytoplankton is flagged by `worms_taxonomy_lookup()` but not computed
   (needs separate production-sampling data, out of scope for this
   pipeline) — not shown above.*

   * `calc_fish()` — PB from 6 independent M methods (Pauly 1980, FishLife,
     Gascuel 2008, Hoenig 1983, Then et al. 2015, Alverson & Carney 1975) +
     Fmort; QB from 4 methods (Palomares & Pauly 1998 Eq.27/Eq.26, Christensen
     & Pauly 1992, Q/P=3 fallback). Chosen value via priority order or mean
     (`PB_QB_SELECTION_MODE`).
   * `calc_mammal()` — PB via Siler-survivorship (Barlow & Boveng 1991),
     fallback Gascuel (2008); QB via Innes & Trites (1997), fallback Q/P=3.
   * `calc_seabird()` — PB via Gascuel (2008) only; QB via Nilsson & Nilsson
     (1976), **no** Q/P=3 fallback (that heuristic is fish/ectotherm-derived
     and would badly underestimate endotherm QB).
   * `calc_invertebrate()` — PB via Tumbiolo & Downing (1994), Brey (1999),
     fallback Gascuel (2008); QB via Q/P=3 only (Brey's 2012 ANN model
     flagged as non-automatable — top-10 highest-biomass invertebrates
     exported instead for manual cross-check).
   * Phytoplankton — flagged, not computed (needs separate production-
     sampling data, out of scope for this pipeline).
8. **FG-level aggregation** — biomass-weighted mean PB/QB per FG, with
   coverage/count audit columns (`n_species_with_PB`, `biomass_coverage_PB`,
   etc.).
9. **EcoBase literature merge** (if `ecobase_literature_pb_qb_simple.csv`
   exists) — averages multiple matching literature models per `FG_name`,
   flags >2×/<0.5× divergence from the empirical estimate, gap-fills only
   where the empirical value is `NA`. Writes sheet **`Ecobase`** (one row per
   FG, averaged PB/QB across every matching model, plus a `References`
   column listing each contributing model) **and `Ecobase_by_Model`** (one
   row per FG × individual published model — `FG_num`, `FG_name`,
   `EwE_model`, `year`, `authors`, `PB`, `QB` — the un-averaged detail behind
   `Ecobase`'s own per-FG mean, since a single FG typically has several
   different published models backing it and the averaged sheet alone loses
   which model said what).
10. **FG-level F top-up** (if `FG_YIELD_SOURCE = "fg_catch_csv"`) — applied
    only to FGs with no species-level F already baked in (mainly mammal/
    seabird/invertebrate FGs), to avoid double-counting against fish's own
    species-level Fmort.
11. **Workbook write** — `add_pbqb_to_ecopath_workbook()` writes sheet
    **`PB_QB`** (one row per FG — the biomass-weighted rollup), and, since
    `results` (the species-level table) is passed in as its
    `species_pb_qb` argument, also writes sheet **`PB_QB_spp`** (one row
    per species, mirroring `FG_spp_Ecopath`'s shape — every species that
    went into computing `PB_FG`/`QB_FG`, with three proportion columns:
    `prop_biomass_FG` — share of the FG's total biomass, regardless of
    whether that species got a PB/QB; `prop_biomass_PB`/`prop_biomass_QB`
    — that species' actual weight inside `PB_FG`'s/`QB_FG`'s own
    biomass-weighted mean, `NA` if that species' own PB/QB didn't
    compute. This is the sheet to check *which* species and *how much*
    of each FG's biomass a given `PB_FG`/`QB_FG` number was built from —
    without it, that detail only existed outside the workbook, in
    `species_pb_qb_by_taxon_group.csv`). If `01_survey_density_*.R` already
    wrote a `PB_QB_spp` **scaffold** into this workbook (Species/FG_num/
    Biomass/`prop_biomass_FG` only, no real PB/QB yet), this call replaces
    it wholesale with the full sheet — `upsert_workbook_sheets()` always
    fully overwrites a same-named sheet, so no separate merge step is
    needed. A separate `upsert_workbook_sheets()`
    call writes sheet **`References`** (later renamed `PB_QB_References_`
    by `finalize_workbook_sheet_order()`, run from Script 1).
12. **Diagnostic plots** — per-method PB/QB comparison plots for fish, plus
    paginated species-level and FG-level PB/QB comparison plots across every
    dispatch group (auto-paginated when a plot would exceed `ggsave`'s
    50-inch height limit).
13. **Ecopath-ready CSV export** — `ecopath_ready_PB_QB.csv`, formatted to
    match a real Ecopath Basic Input CSV's exact column layout and
    European comma-decimal number formatting (verified against a real
    exported file, `westernmed90s-Basic_input.csv`).

## Outputs

* **CSVs (2026-09-17: all under `output/pbqb-traits/`, not the top-level
  `output/`):** `invertebrates_for_brey_manual_check.csv`,
  `species_pb_qb_by_taxon_group.csv`, `fg_pb_qb_weighted.csv`,
  `phytoplankton_needs_separate_method.csv`,
  `fg_pb_qb_weighted_with_ecobase.csv`, `fg_pb_qb_weighted_with_F.csv`,
  `species_parameter_references.csv` (species-level parameter provenance —
  not the same table as the workbook's `FG_References` sheet, see below),
  `ecopath_ready_PB_QB.csv`, native `PB_QB.csv`/`PB_QB_spp.csv`.
* **Plots:** `fish_PB_methods_comparison.png`,
  `fish_QB_methods_comparison.png`,
  `species_PB_QB_comparison[_pageN].png`, `FG_PB_QB_comparison[_pageN].png`.
* **Workbook** `ecopath_ecosim_inputs.xlsx`, final sheets written:
  `Ecopath_PBQB`, `Ecopath_traits`. (`PB_QB`/`PB_QB_spp`/`Ecobase` are
  native/intermediate CSVs only now, not workbook sheets — see "Excel
  workbook functions" in the shared-library section below.)

---

# Shared library — `lib_survey_fg_density_functions.R`

Sourced by all four numbered scripts; has **no top-level driver code** of its
own — every symbol below is a function definition only.

## FG-matching functions

Every observed species runs through this cascade, in order, until it
resolves to a Functional Group or falls through to manual review — the
diagram shows the actual fallback order implemented in the code below
(same style as `medbs_pipeline_methodology.qmd`'s `@sec-fg-matching`
diagram; each stage only attempts species the previous one left
unmatched).

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    A1["match_species_to_fg()&lt;br/&gt;direct ScientificName match"] --> A2["match_nominate_subspecies_fg()&lt;br/&gt;strip repeated trinomial word, retry"]
    A2 --> A3["apply_seed_fg_rules()&lt;br/&gt;manual rules, zero-reference FGs"]
    A3 --> A4["fallback_match_fg_by_taxonomy()&lt;br/&gt;Genus to Phylum, exactly-one-FG rule"]
    A4 --> A5["summarize_unresolved_species()&lt;br/&gt;manual review"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    class A1,A2,A3,A4 purple
    class A5 gray
</div>
```

| Function | Purpose | Returns |
|---|---|---|
| `validate_survey_data(dataframe1, dataframe2, strata = TRUE)` | Fail-fast schema/uniqueness check on the two standardized input tables (required columns present, `SampleID` unique per survey) | `invisible(TRUE)` |
| `prepare_fg_lookup(dataframe2)` | Deduplicates species→FG reference; keeps genuine stanza splits (a species exclusively mapped to ≥2 FGs), excludes/flags ambiguous multi-FG species | deduped `data.table(ScientificName, FG_num, FG_name)` |
| `match_species_to_fg(dataframe1, fg_lookup_safe)` | Direct left-merge by `ScientificName` | merged `dt` |
| `match_nominate_subspecies_fg(dt, fg_lookup_safe)` | Strips a repeated trinomial word (e.g. "Diplodus sargus sargus" → "Diplodus sargus") and retries the match | `dt` with newly resolved rows |
| `fetch_taxonomy(species_names, taxonomy_source = "worms")` | Dispatches to `fetch_taxonomy_worms()`, `fetch_taxonomy_fishbase()`, or both | `data.table(ScientificName, Genus, Family, Order, Class, Phylum)` |
| `fallback_match_fg_by_taxonomy(dt, fg_lookup_safe, taxonomy_source = "worms", rank_levels = c("Genus","Family","Order","Class","Phylum"))` | Data-driven rank fallback — a rank value is used only if it maps to exactly one FG among reference species | `dt`, with `attr(dt,"fetched_taxonomy")` and `attr(dt,"still_unresolved_taxonomy")` attached |
| `apply_seed_fg_rules(dt, dataframe2, seed_rules, species_exceptions = NULL)` | Explicit small rule table for FGs with **zero** reference species (nothing for the automatic fallback to learn from); species-level exceptions applied first | `dt` |
| `summarize_unresolved_species(dt, taxonomy = NULL)` | Mean biomass + appearance count per still-unmatched species, sorted by appearance count | summary `data.table` |
| `build_fg_manual_review_template(csv_out_dir, out_path = file.path(csv_out_dir, "FG_manual_review_template.xlsx"))` | Combines `survey_unmatched_for_manual_review.csv` + `medias_unmatched_for_manual_review.csv` into one de-duplicated review workbook, with blank `GF`/`FG_name` columns ready to hand-fill | invisibly returns the combined `data.table`; writes an `.xlsx` |

`build_fg_manual_review_template()` is a one-off, run-it-yourself helper, not
part of the automated pipeline — nothing calls it from `01_biomass.R` itself.
After a run leaves species unmatched, call it against that run's
`output/biomass/` folder to get one workbook (a `README` sheet plus a
`For_review` sheet, sorted by how often each species actually showed up in
the data — review the highest-impact species first). Its `ESPECIE`/`GF`/
`FG_name` columns deliberately use the exact same names as `FG_WMed.xlsx`'s
own sheet 4 (`fg_wmed_95`) — fill in `GF`/`FG_name` for whichever species you
want to resolve, then copy columns A:C straight into a **new** version of
that sheet (never overwrite the original in place), and point `fg_file` at
the new file for the next run. There's no code path that writes the filled-in
review back into `FG_WMed.xlsx` automatically — that hand-off is deliberately
manual, since it's a genuine judgment call about which FG a given species
belongs to.

| `snapshot_full_extent_species(matched_dt, species_taxonomy, source_label, csv_out_dir, sci_name_col = "ScientificName")` | One row per distinct species a source's own FG-matching step saw, taken BEFORE that source's `FILTER_AREAS`/year restriction — i.e. every species anywhere in the raw file(s) that source loaded, independent of this run's config. Called once each for MEDITS (Steps 3-4), MEDIAS (Step 9), and stock assessment | writes `all_species_<source>_full_extent.csv` |
| `build_full_fg_species_catalog(csv_out_dir, out_path = file.path(csv_out_dir, "FG_WMed_full_species_catalog.csv"))` | Combines the three `all_species_*_full_extent.csv` snapshots above into ONE species catalog — matched species (a real `FG_number`/`FG_name`) and still-unmatched species (`FG_number`/`FG_name` blank) together, with a `sources` column (e.g. `"MEDITS+MEDIAS"`) showing which source(s) actually observed each one. Note the column is named `FG_number` here (not `GF`, as in `FG_WMed.xlsx`/`build_fg_manual_review_template()` above) since this catalog spans all three sources rather than being a direct fill-in template for the Excel sheet | invisibly returns the combined `data.table`; writes a CSV |

`01_biomass.R` calls `build_full_fg_species_catalog(csv_out_dir)` itself,
near the very end (Step 14, right after `trim_workbook_to_final_sheets()`)
— **`FG_WMed_full_species_catalog.csv`** lands in `output/biomass/` on every
run without any extra step. It's a candidate "sheet 4" replacement covering
every species — matched (`status = "matched"`, already correctly resolved)
and unmatched (`status = "needs review"`, blank) — in one file, sorted with
matched rows first. Same read-only relationship to `FG_WMed.xlsx` as
`build_fg_manual_review_template()` above — nothing writes back into it
automatically.

Unlike the matched/unmatched CSVs everything else in this script reads and
writes, the three `all_species_*_full_extent.csv` snapshots — and so this
catalog — are **NOT scoped by `FILTER_AREAS`/`YEAR_ECOPATH`/`TS_YEARS`**, by
construction: `snapshot_full_extent_species()` is called right after each
source's own FG-matching step and *before* that source's own area
restriction (MEDITS: before `dt <- dt[AreaID %in% FILTER_AREAS]`; MEDIAS:
`acoustic_matched` itself is never restricted in place, only derived summary
tables are; stock assessment: `star_wm` is already the whole West Med
subregion, not this run's `FILTER_AREAS`). So this catalog always covers the
**whole available survey extent** — every GSA and year the raw MEDITS/
MEDIAS/stock-assessment files actually contain — on every run, regardless
of what region/years that particular run happens to be configured for. A
source's snapshot (and its contribution to the catalog) simply won't exist
for a run where that source didn't execute at all — `AREA_MODE == "custom"`
skips MEDIAS and stock assessment entirely, so a custom run's catalog is
MEDITS-only there, which is a real limit of that mode (no MEDIAS/stock-
assessment data exists to catalog), not something a wider `FILTER_AREAS`
would fix.

## Area / stratum / density functions

| Function | Purpose | Returns |
|---|---|---|
| `match_samples_to_area(dt, area_shp, area_id_col)` | Spatial join (`sf::st_join`) of sample points to polygons; no-op if `AreaID` already fully populated | `dt` |
| `assign_depth_stratum(dt, strata_def)` | Bins `Depth` into strata via `findInterval()`; out-of-range depths → `Stratum = NA` | `dt` |
| `filter_samples_by_area(dt, area_filter, filter_type = c("shapefile","bbox"), points_crs = 4326)` | Custom sub-region filter, independent of GSA/`AreaID` | `dt` filtered to in-area `SampleID`s |
| `compute_sample_densities(dt)` | `Density := Biomass / Effort` | `dt` |
| `apply_catchability_correction(dt, catchability_table, default_q = 1, species_taxonomy = NULL, exempt_fg_names = NULL, exempt_species = NULL)` | 5-tier resolution: direct species → genus "spp" entries → explicit broad-rank entries → taxonomic-proximity fallback → `default_q`; exemptions override last. Divides `Density`/`Biomass` by resolved `q` | `dt` |
| `remove_sample_outliers(dt, threshold = 70, min_samples = 5, drop_outliers = TRUE, log_scale = TRUE)` | Per-species/per-area robust z-score (median/MAD) on `log1p(Density)`; sets flagged `Density`/`Biomass` to `NA` if `drop_outliers = TRUE` | `dt`, with `attr(dt,"flagged_outliers")` |
| `compute_fg_densities_by_stratum(dt, strata = TRUE)` | Sums species density within FG per sample, then across samples by area/year/[stratum]/FG | `per_group_fg` (`density_sum`) |
| `compute_strata_area_by_area(area_ids, area_shp, area_id_col, strata_def, cache_path = NULL, resolution = 1)` | Downloads NOAA bathymetry per area, bins cell area into depth strata (latitude-corrected); optional CSV cache | `strata_area_by_area` table |
| `weight_by_strata(per_group_fg, n_samples_by_stratum, strata_area_by_area)` | Core strata-weighted estimator: `D_area = Σ_h (density_sum_h/n_samples_h) · area_prop_h` | `fg_index`, with `attr(fg_index,"per_stratum")` |
| `simple_area_density(per_group_fg, n_samples_by_area)` | Unstratified path (`strata = FALSE`) | `data.table` |
| `weight_by_area(fg_index_stratified)` | Region-wide weighting from absolute stratum areas (`per_stratum` attribute) — requires the stratified path upstream | `fg_index_regional` |
| `compute_species_densities_by_stratum(dt, strata = TRUE)` / `weight_species_by_area(per_group_sp, n_samples_by_stratum, strata_area_by_area)` | Species-level analogs of the two functions above, feeding the `FG_spp` sheets | `data.table(Year, FG_num, FG_name, ScientificName, mean_density)` |
| `compute_cv_log_by_fg(value_dt, value_col, min_replicates = 2)` | `cv_log = sd(log1p(x))/mean(log1p(x))` per FG/Year, averaged across years — feeds the Ecosim "Weight" row | `data.table(FG_num, FG_name, cv_log, n_years)` |

## Plotting functions

| Function | Produces |
|---|---|
| `plot_sample_map(dt, area_shp, area_id_col, ..., by_year = FALSE, selected_areas = NULL, land_on_top = FALSE)` | Sample coverage map, points colored inside/outside the study area, polygons shaded by sample intensity |
| `plot_fg_timeseries_by_area(fg_index, ...)` / `plot_fg_timeseries_regional(fg_index_regional, ...)` | Faceted FG density time series, per-area vs. region-wide |
| `plot_strata_profile(fg_index_stratified, strata_def, ...)` | Depth-strata validation bar chart (pre-area-weighting, per-sample density by stratum) |

## Excel workbook functions (2026-09-17 revision)

As of the 2026-09-17 CSV-native refactor, `ecopath_ecosim_inputs.xlsx`
carries **only the 10 final target sheets** — every native/intermediate
table each block used to also keep as a workbook sheet is now CSV-only
(see "Per-block output CSV subfolders" and "Final target sheet order"
below). The table below reflects the current function signatures, not the
older ones this document used to describe.

| Function | Purpose |
|---|---|
| `upsert_workbook_sheets(sheets, out_path)` | Adds/replaces the named sheets in `out_path`, **leaves every other sheet untouched**. Creates the workbook if it doesn't exist yet. Safe to call repeatedly, from any script, in any order. |
| `write_native_sheet_csv(dt, name, out_dir)` / `write_native_sheets_csv(sheets, out_dir)` / `read_native_sheet_csv(name, out_dir)` | Generic CSV read/write for a block's own native/intermediate tables — `out_dir` is always passed explicitly by the caller (the block's own subfolder), never assumed |
| `read_existing_ts_years(out_path, sheet_name, csv_dir = NULL)` | Reads year values from column 1 of an existing native CSV sheet (e.g. `Ecosim`), so `Catches_Ecosim`'s years stay aligned with it. `csv_dir` defaults to `dirname(out_path)` for back-compat; the numbered scripts pass the correct block's subfolder explicitly |
| `read_full_fg_reference(out_path, csv_dir = NULL)` | Reads `FG_lookup.csv` (a biomass-block output) — the full FG_num/FG_name reference, used to expand PB_QB/FG_References/etc. to every model FG, not just the ones with data this run |
| `export_ecopath_ecosim_excel(fg_index_regional, species_density_regional, n_samples_by_area_year, dataframe2, year_ecopath, ts_years, out_path, species_taxonomy = NULL, fg_cv_log = NULL, normalize_ts = TRUE, csv_out_dir = NULL)` | Writes the final `Ecopath_B` sheet directly to the workbook, plus native/intermediate CSVs (`FG_spp_Ecopath`, `FG_lookup`, `Ecosim`, `FG_spp_Ecosim`, a `PB_QB_spp` scaffold) into `csv_out_dir` (01_biomass.R's own `output/biomass/` subfolder) |
| `add_catches_to_ecopath_workbook(fg_catch, fg_lookup, out_path, year_ecopath, area_km2, ts_years = NULL, fleet_structure = NULL, csv_out_dir = NULL, biomass_csv_dir = NULL)` | Converts raw catch tonnes to t/km² density; writes native CSVs `Catches_Ecopath`/`Catches_Ecosim`/`Fleet_Structure` into `csv_out_dir` (02_fisheries.R's `output/fisheries/`). Reads `Ecosim.csv` from `biomass_csv_dir` when `ts_years` isn't passed explicitly |
| `add_pbqb_to_ecopath_workbook(fg_weighted, out_path, species_pb_qb = NULL, csv_out_dir = NULL, biomass_csv_dir = NULL)` | Writes native CSVs `PB_QB`/`PB_QB_spp` into `csv_out_dir` (03_pbqb-traits.R's `output/pbqb-traits/`). Reads `FG_lookup.csv` from `biomass_csv_dir` to expand to every model FG |
| `finalize_ecopath_ecosim_summary_sheets(out_path, year_ecopath, biomass_csv_dir = NULL, fisheries_csv_dir = NULL)` | Builds the final `Ecosim_ts` sheet by combining native CSVs from **two** blocks — `Ecosim` (biomass) plus `Catches_Discards_FG_ts`/`Catches_Ecosim`/`Fishing_Effort_by_Fleet` (fisheries) — hence the two separate directory arguments |
| `build_fg_references_sheet(out_path, biomass_csv_dir = NULL, fisheries_csv_dir = NULL, pbqb_csv_dir = NULL, diet_csv_dir = NULL)` | Builds the final `FG_References` sheet — one row per FG, consolidating which data source fed each block (see "The FG_References sheet" below) |
| `trim_workbook_to_final_sheets(out_path)` | **Run last**, by every numbered script. Reduces the workbook to exactly the 10 final target sheets (see below), dropping any native/intermediate sheet a much older run of this workbook might still be carrying |

The diagram below shows which script/function writes which sheet into the
one shared workbook, and the final trim that runs after each script.

```{=html}
<div class="mermaid">
%%{init: {'flowchart': {'padding': 20, 'nodeSpacing': 60, 'rankSpacing': 90}, 'themeVariables': {'fontSize': '18px'}}}%%
flowchart LR
    S1["01_biomass.R&lt;br/&gt;export_ecopath_ecosim_excel()"] --> W["ecopath_ecosim_inputs.xlsx&lt;br/&gt;(shared workbook, top-level out_dir)"]
    S2["02_fisheries.R&lt;br/&gt;add_catches_to_ecopath_workbook()"] --> W
    S3["03_pbqb-traits.R&lt;br/&gt;add_pbqb_to_ecopath_workbook()"] --> W
    S4["04_diets.R&lt;br/&gt;upsert_workbook_sheets()&lt;br/&gt;(Ecopath_diet)"] --> W
    S5["build_fg_references_sheet()&lt;br/&gt;(run by 04_diets.R, last)"] --> W
    W --> F["trim_workbook_to_final_sheets()&lt;br/&gt;run after every script"]
    F --> O["output/ecopath_ecosim_inputs.xlsx&lt;br/&gt;(10 final sheets only)"]

    classDef purple fill:#EEEDFE,stroke:#534AB7,color:#26215C
    classDef gray fill:#F1EFE8,stroke:#5F5E5A,color:#2C2C2A
    classDef coral fill:#FAECE7,stroke:#993C1D,color:#4A1B0C
    class S1,S2,S3,S4,S5 purple
    class W gray
    class F,O coral
</div>
```

## Per-block output CSV subfolders (2026-09-17)

Every native/intermediate CSV a block writes now lands in that block's own
subfolder under the shared `out_dir`, instead of a single flat directory
shared by all four scripts:

* `01_biomass.R` → `output/biomass/`
* `02_fisheries.R` → `output/fisheries/`
* `03_pbqb-traits.R` → `output/pbqb-traits/`
* `04_diets.R` → `output/diet/`

`ecopath_ecosim_inputs.xlsx` itself stays at the top-level `output/` — all
four scripts read and write it, so it has no single owning block. Each
script defines its own `csv_out_dir` near the top and passes it (or a
`*_csv_dir` pointing at a *different* block's subfolder, for the handful of
cross-block reads — species density, strata area, `FG_lookup.csv`,
`Ecosim.csv`, `biomass_proportion_by_species_fg.csv`, catch time series)
into the library functions above. Every one of those new directory
arguments defaults to `dirname(out_path)` (the workbook's own top-level
folder) when omitted, so nothing breaks if some other caller doesn't pass
them.

## Final target sheet order (10 sheets)

`target_order`, used by `trim_workbook_to_final_sheets()` and run after
every numbered script:

1. `info` — run metadata (areas, countries, YEAR_ECOPATH, TS_YEARS, ...)
2. `FG_spp` — FG × species × taxonomy list, written by `01_biomass.R`
3. `Ecopath_B` — FG-level biomass (t/km²), `YEAR_ECOPATH`-averaged
4. `Ecopath_L` — FG × Fleet landings, written by `02_fisheries.R`
5. `Ecopath_Di` — FG × Fleet discards, written by `02_fisheries.R`
6. `Ecopath_PBQB` — FG-level PB/QB, written by `03_pbqb-traits.R`
7. `Ecopath_traits` — FG_name/species-level traits, written by `01_biomass.R`
8. `Ecopath_diet` — predator FG × prey FG diet matrix, written by `04_diets.R`
9. `Ecosim_ts` — combined B/L/Di/Effort time series (see `finalize_ecopath_ecosim_summary_sheets()`)
10. `FG_References` — one row per FG, the data-source/citation summary (see below) — **added 2026-09-17**

Every native/intermediate table (`Ecopath`, `Catches_Ecopath`, `PB_QB`,
`Ecosim`, `Catches_Ecosim`, `Fishing_Effort_by_Fleet`, `FG_spp_Ecopath`,
`FG_spp_Ecosim`, `FG_lookup`, `traits_ewe`, `fg_traits_weighted`,
`PB_QB_spp`, `References` [species-parameter-level, `03_pbqb-traits.R`'s
own audit CSV — a different table from the `FG_References` workbook sheet
above, despite the similar name], `Ecobase`, `Catches_Discards_FG_ts`,
`diet_references_by_fg`, ...) lives only as CSV, in its own block's
subfolder, never as a workbook sheet — `trim_workbook_to_final_sheets()`
drops anything else it finds regardless.

## The FG_References sheet

Added 2026-09-17, built by `build_fg_references_sheet()`
(`lib_survey_fg_density_functions.R`), run by `04_diets.R` after every other
script has had a chance to run. One row per FG, with columns:

| Column | What it summarizes | Read from |
|---|---|---|
| `FG_num`, `FG_name` | The full FG reference | `FG_lookup.csv` (`output/biomass/`) |
| `ref_B` | Which biomass source (MEDITS/MEDIAS survey vs. stock assessment) fed `Ecopath_B`/`Ecosim_ts`'s B block for this FG | `biomass_source` in `survey_fg_annual_index_regional_combined.csv` (`output/biomass/`) |
| `ref_fisheries` | Which catch/discard source (GFCM/FDI/SAU-derived vs. stock assessment) fed `Ecopath_L`/`Ecopath_Di`/`Ecosim_ts`'s catch time series for this FG | `catch_source` in `Catches_Discards_FG_ts.csv` (`output/fisheries/`) |
| `ref_pbqb_traits` | Whether `PB_FG`/`QB_FG` came from the empirical species-level calculation or an EcoBase literature gap-fill | `PB_source`/`QB_source` in `PB_QB.csv` (`output/pbqb-traits/`), falling back to a generic "empirical" note when EcoBase gap-filling didn't run |
| `ref_diet` | Which diet-study citations (the metaweb's own `Reference` column) fed this FG's row/columns in `Ecopath_diet` | `diet_references_by_fg.csv` (`output/diet/`), rolled up from `build_species_diet()`'s per-predator citations via the same species→FG mapping the diet matrix itself uses |
| `ref_methods` | The calculation METHOD/literature behind `PB_FG`/`QB_FG` (as opposed to `ref_pbqb_traits`, which is about the *data*, not the method) | `dispatch_group` in `PB_QB_spp.csv` (`output/pbqb-traits/`), mapped to a literature citation |

Each column is a best-effort summary (distinct values `" | "`-joined when a
FG drew on more than one source across years/species/studies), not a full
audit trail — the individual block's own CSV named in the table above has
the complete detail. A column is left blank for a FG when the block behind
it hasn't run yet against this workbook, with a message saying so rather
than an error; running `build_fg_references_sheet()` again later (as
`04_diets.R` does every run) fills in whatever is available by then.

---

# Other library files

## `lib_worms_taxonomy_lookup.R`

Library file, no top-level driver code — `source()`'d automatically by the
numbered scripts. Defines a single function:

**`worms_taxonomy_lookup(names_vector)`** — the single taxonomy source used
everywhere in the pipeline (so genus/family comparisons are always
apples-to-apples, never a FishBase-derived family compared against a
WoRMS-derived one for the same actual family). Strips `"Genus spp."`
suffixes so genus-level placeholders still resolve, chunks the query list
(50 names/chunk) against `worrms::wm_records_names()`, picks the best
(`status == "accepted"` preferred) match per name, and returns
`data.table(original_name, search_term, AphiaID, matched_name, rank, status,
phylum, class, order, family, genus)`. Explicitly namespaces `purrr::map()`
rather than calling `map()` bare — the `maps` package (loaded elsewhere in
this pipeline for basemap plotting) also exports an unrelated `map()`, and
whichever package attached most recently silently wins an unqualified call.

## `lib_build_species_df_from_survey.R`

Library file — bridges `01_biomass.R`'s survey output and
`03_pbqb-traits.R`'s PB/QB estimation. Defines:

**`build_species_df_from_survey(survey_csv_path, year_ecopath_range,
out_rds_path = NULL)`** — reads `species_density_regional_combined.csv`,
restricts to `YEAR_ECOPATH`'s snapshot years (the same years used for the
`Ecopath` sheet's Biomass column, so PB/QB weighting and the reported Biomass
describe the same time snapshot), collapses to one mean-density `Biomass`
value per species (stops with an error if any species maps to more than one
FG within that year range — shouldn't happen given upstream FG-matching
logic), sets `Yield := NA` (this survey pipeline has no catch/landings data
of its own — F = Yield/Biomass isn't computable from it directly). Optionally
saves the result as an `.rds`. Returns
`species_df(Species, FG, FG_name, Biomass, Yield)`.

---

# Standalone cartography utilities (not part of the numbered pipeline)

## `export_gsa_boundaries.R`

Standalone — does not need `01_biomass.R` or the shared
library sourced first. Downloads the same GFCM GSA shapefile via the same URL
and GSA-numbering fix (`SMU_CODE` 111/112 → 11). **Western-Med-specific as
delivered:** restricts to GSA 1–11 — change the `gsa_num %in% 1:11` filter to
any other GSA range to export a different region's boundaries. Simplifies
the result for web embedding via
`rmapshaper::ms_simplify(keep = 0.05, keep_shapes = TRUE)` (keeps small
polygons like Alboran Island, GSA 2, from being deleted outright), and writes
`gsa_boundaries.geojson` — the file embedded in
`medbs_pipeline_methodology.qmd`'s interactive Leaflet map.

## `fig_WMed_basemap.R`

Standalone map-generation script (not sourced by anything else), purely
cartographic — not required by the numbered pipeline at all.
**Western-Med-specific as delivered:** downloads the GFCM GSA shapefile,
splits polygons into Western Med (GSA 1–11, the target study area) vs.
context GSAs (12, 16 — Northern Tunisia, South of Sicily) and draws the
Sicily Channel divide (Cape Bon–Cape Lilibeo) as the Western/Central
Mediterranean boundary; all three of those choices would need editing to
produce an equivalent basemap for a different region. The general
technique — label placement via the "pole of inaccessibility" algorithm
(`polylabelr`, more robust than `st_point_on_surface()` for narrow,
coastline-hugging GSAs) plus a high-resolution (`rnaturalearthhires`)
coastline — is reusable as-is for any region. Output:
`output/plots/westmed_gsa_map.png`.

---

# Adapting this pipeline to a different region

Everything below applies whether the target region is a single GSA, an
arbitrary GSA subset, the whole Mediterranean GSA-by-GSA, or a fully
custom (non-GSA) boundary.

## Preferred method: source the scripts with pre-set options

`01_biomass.R`, `02_fisheries.R`,
`03_pbqb-traits.R`, and `04_diets.R` are **source-able**: every region/year/method knob each
one uses (`FILTER_AREAS`, `TARGET_COUNTRIES`, `YEAR_ECOPATH`,
`START_YEAR`/`END_YEAR`, `TS_YEARS`, `DATASET_VERSION`,
`FISHERIES_DATA_SOURCE`, `OUTLIER_METHOD`, `PB_QB_SELECTION_MODE`,
`out_dir`/`pcloud_dir`/`git_dir`, and a handful more — each script's own
"Configuration" section, under a comment block headed `SOURCE-ABLE
SCRIPT`, lists exactly which variables it recognizes) is wrapped in
`if (!exists("VAR", envir = .GlobalEnv, inherits = FALSE)) VAR <- <default>`.
That means:

* If a variable is **already set** in the calling environment before the
  script is `source()`-d, the script uses that value and skips its own
  default entirely.
* If it is **not set**, the script falls back to its original Western Med
  default (GSA 1–11, 1994–2023) — running any of the three scripts on its
  own, exactly as before, is completely unchanged.

`run_pipeline_demo.R` is a driver script demonstrating this: it sets
`out_dir`/`pcloud_dir`/`git_dir` once, then `source()`s
`01_biomass.R` → `02_fisheries.R` →
`03_pbqb-traits.R` → `04_diets.R` in that dependency order (01 must run
first — see [Adapting this pipeline to a different region] below and each
script's own header comment for why; 04 specifically needs 01's
`FG_spp_Ecopath` sheet in the shared workbook for its species'
share-of-FG-biomass values, and is otherwise optional — diet composition
isn't needed for every run). It includes two ready-to-run examples,
switched via a `RUN_MODE` flag at the top:

* `RUN_MODE <- "westmed_default"` — sets nothing beyond the three paths;
  every script uses its built-in West Med default.
* `RUN_MODE <- "custom_example"` — sets `FILTER_AREAS <- 17:18`,
  `TARGET_COUNTRIES <- c("Italy", "Croatia", "Slovenia")`,
  `YEAR_ECOPATH <- 2005:2007`, `TS_YEARS <- 2000:2020`, and
  `START_YEAR`/`END_YEAR <- 2000`/`2020` as a worked Adriatic-GSA example
  — copy and adapt this block for any other region/year range, setting
  only the variables that need to differ from the West Med default.

`YEAR_ECOPATH` in particular is read by all three scripts and must stay
consistent across them (it defines the single Biomass snapshot the PB/QB
weights are computed against) — setting it once in the driver script
before any of the three are sourced keeps them in sync automatically,
which is easy to forget when editing each script's hardcoded copy by hand.

This sourcing approach and the manual per-script editing approach
described in the rest of this section are not mutually exclusive — the
`if (!exists(...))` guards mean a variable can still be hardcoded by
editing the script directly (equivalent to always setting it), or left
alone to pick up a caller's override, on a variable-by-variable basis.

**To model a different GSA or set of GSAs** (Adriatic, Ionian, Aegean,
Levant, or the whole basin), starting from `01_biomass.R` as
a template:

* `FILTER_AREAS` — set to the target GSA number(s). No other GSA-based code
  in `lib_survey_fg_density_functions.R` hardcodes a Western-Med-only
  assumption — `compute_strata_area_by_area()`, `weight_by_strata()`,
  `weight_by_area()` etc. are all region-agnostic and operate on whatever
  `area_ids`/`area_shp` they're given.
* `ALBORAN_GSAS`/`fix_alboran_longitude_sign()` — a Western-Med-only
  correction (GSA 1–3); harmless no-op elsewhere, delete if not needed.
* `MEDIAS_HANDBOOK_AREAS`/`GSA_AREA_CROSSWALK` — only populated for Western
  Med GSAs; add the target region's own rows from the MEDIAS Handbook
  (or rely entirely on the bathymetry-derived 10–200 m fallback if MEDIAS
  data or handbook-reported areas aren't available for that region).
* `out_dir`/`pcloud_dir`/`git_dir` paths, and the raw MEDITS/MEDIAS/FG-
  reference file paths under them — point at the target region's own data.
* `export_gsa_boundaries.R`/`fig_WMed_basemap.R` — optional cartography
  only; edit the GSA-range filters if you want an equivalent map for the
  new region, otherwise skip both entirely (nothing downstream needs them).

**To model a fully custom (non-GSA) region anywhere in the Mediterranean**,
set `AREA_MODE <- "custom"` before sourcing `01_biomass.R`, plus `AREA_NAME`,
`CUSTOM_AREA_TYPE` (`"bbox"` (default), `"shapefile"`, or `"gsa"`), and
`CUSTOM_BBOX`/`CUSTOM_SHAPEFILE_PATH`/`CUSTOM_GSA_IDS` set to the target
boundary — see [`AREA_MODE == "custom"`] above. No GSA-specific code path
is involved for `"bbox"`/`"shapefile"` (the Alboran-sign fix is present but
is a documented no-op outside GSA 1–3); MEDIAS and stock-assessment biomass
are skipped entirely under `AREA_MODE == "custom"`, regardless of
`CUSTOM_AREA_TYPE`.

Everything downstream of Step 1 — `02_fao_catches.R`, `03b_ecobase.R`,
`03_pbqb-traits.R`, and the shared library's Excel-export/PB-QB functions — is
already region-agnostic: they operate on whatever `strata_area_by_area.csv`
and `species_density_regional_combined.csv` Step 1 produced, with no
Western-Med-specific logic beyond `02_fao_catches.R`'s `area_filter_values`
(the FAO/GFCM division codes 37.1.1–37.1.3, which likewise need changing to
match the target region's own FAO division codes) and `03b_ecobase.R`'s
`MED_LON_MIN/MAX`/`MED_LAT_MIN/MAX` bounding box (already covers the **full**
Mediterranean by default — Gibraltar to the Levant — so it does not need
editing for a sub-region; narrow it only if you specifically want to exclude
literature models from elsewhere in the basin).

---

# Script 5 — `04_diets.R`

**Role:** Pipeline Step 5, optional — builds an EwE diet-composition
matrix (predator FG × prey FG) from the Mediterranean trophic
metaweb database (`data_entry_metaweb_empty.xlsx` - currently an empty template, no DATA_ENTRY rows yet), collapsed from species-level
predator-prey records up to the same Functional Group scheme used
everywhere else in the pipeline. Auto-runs on `source()`, same convention
as `01_biomass.R`/`02_fisheries.R`/`03_pbqb-traits.R` — no separate
function call needed afterward.

**Dependency:** must run **after** `01_biomass.R` has written to the same
`ECOPATH_WORKBOOK_PATH` — it reads `biomass_proportion_by_species_fg.csv`
(01_biomass.R's own native CSV output, `output/biomass/` — **not** a
workbook sheet) for each species' share of its Functional Group's total
biomass ("proportion B"), which weights that species' own diet more heavily
within its FG's combined diet the larger a share of the FG's biomass it
represents. If that CSV isn't there yet (01 hasn't run against this
workbook), it falls back to `SPECIES_BIOMASS_IN_FG_CSV_PATH` (a small manual
CSV, same schema: species/fg_name/biomass_proportion) for whichever
species×FG pairs are still missing, and finally to an equal split among an
FG's remaining mapped species for anything neither source covers.

## Inputs

| File | Format | Role |
|---|---|---|
| `<pcloud_dir>/data/Complementary data/data_entry_metaweb_empty.xlsx` | xlsx | Mediterranean trophic metaweb — predator-prey diet records (`DATA_ENTRY`), taxonomic code lookup (`Taxonomic_codes`), generic/non-taxonomic prey-category lookup (`Non_taxonomic_groups`) |
| `<pcloud_dir>/data/FG_WMed.xlsx`, sheet 4 | xlsx | Species → Functional Group reference — same file `01_biomass.R` uses, so every species already assigned an FG there needs no manual entry here |
| `output/biomass/biomass_proportion_by_species_fg.csv` | csv | Written by `01_biomass.R` — species' share of biomass within their FG ("proportion B"), the preferred source |
| `output/biomass/FG_lookup.csv` | csv | Full FG_num/FG_name reference, used to resolve `FG_num` for the new `diet_references_by_fg.csv` output below |
| `<pcloud_dir>/data/species_to_fg_missing.csv` | csv | Manual species→FG fallback, only for species this run needs that aren't in `FG_WMed.xlsx` at all |
| `<pcloud_dir>/data/species_biomass_in_fg_missing.csv` | csv | Manual biomass-share fallback, only for species×FG pairs missing from `biomass_proportion_by_species_fg.csv` |
| `<pcloud_dir>/data/ewe_group_table.csv` | csv | EwE group numbering (group_number/group_name/is_predator) — defines row/column order and which FGs get a predator column in the exported matrix |

## Outputs

* `output/diet/diet_composition_ewe.csv` (2026-09-17: under `output/diet/`,
  not the top-level `output/`) — EwE's own plain-text diet-composition
  import format: blank + "Prey \\ predator" header, predator columns by
  group number (`is_predator` groups only), prey rows by group number +
  name (full group list), decimal-comma **quoted** values (unquoted would
  silently split into extra columns, since the delimiter is also a comma),
  `Import`/`Sum`/`1-Sum` rows at the bottom.
* `output/diet/diet_references_by_fg.csv` — **added 2026-09-17** — one row
  per FG, the distinct diet-study citations (the metaweb's own `Reference`
  column) behind every species that maps into that FG as a predator. Read
  back by `build_fg_references_sheet()` to populate the final workbook's
  `FG_References` sheet's `ref_diet` column (see the shared-library section
  below).
* `output/ecopath_ecosim_inputs.xlsx`, sheet **`Ecopath_diet`** — the same
  predator-FG × prey-FG matrix, written into the shared workbook as plain
  numeric values (no decimal-comma/quoting — that convention is specific
  to EwE's plain-text import, not needed inside an Excel sheet). Columns:
  `Prey_FG_num`, `Prey_FG_name`, then one `Predator_<num>_<name>` column per
  predator FG.
* `output/ecopath_ecosim_inputs.xlsx`, sheet **`FG_References`** — built
  here too (04_diets.R runs last), after `Ecopath_diet` — see "The
  FG_References sheet" in the shared-library section below.

## Key functions

| Function | Role |
|---|---|
| `read_metaweb(path)` | Reads `DATA_ENTRY`/`Taxonomic_codes`/`Non_taxonomic_groups` from the metaweb workbook |
| `build_code_lookup(tax_codes, nontax_groups)` | Builds one code→name lookup from both tables (Taxonomic_codes wins on a code collision) |
| `resolve_codes(codes, lookup)` | Predator/prey code → scientific name or generic-category label |
| `pick_diet_metric(dt)` | Picks one diet-proportion value per `DATA_ENTRY` row via `DIET_METRIC_PRIORITY` (IRI preferred; Presence-only records get an equal split among a study's present prey) |
| `build_species_diet(data_entry, lookup)` | Species-level predator→prey diet proportions, weighted by study sample size and normalized per predator |
| `read_species_to_fg_from_reference(path, sheet)` | Reads `FG_WMed.xlsx`'s species→FG assignment |
| `build_species_to_fg(needed_species, reference_path, missing_csv_path)` | FG_WMed.xlsx first, manual CSV fallback only for genuinely missing species |
| `read_biomass_proportion_from_workbook(workbook_path, sheet = "FG_spp_Ecopath")` | Reads species' share of FG biomass from `01_biomass.R`'s own output — the preferred "proportion B" source |
| `build_biomass_in_fg(needed_species_fg, workbook_path, fallback_csv_path)` | Combines the workbook source with the manual fallback CSV, reporting coverage counts at each tier |
| `build_fg_diet(species_diet, species_to_fg, biomass_in_fg)` | Expands species-level diet to FG×FG, biomass-weighting each predator species' contribution within its FG, then collapses to one number per (predator FG, prey FG) |
| `export_ewe_matrix_csv(fg_diet, group_table, out_path)` | Writes the EwE plain-text import CSV |
| `build_ecopath_diet_sheet(fg_diet, group_table)` | Builds the same matrix as a plain numeric data.table for the `Ecopath_diet` sheet |
| `run_pipeline(...)` | Runs all of the above end to end; called unconditionally at the bottom of the script |

---

# Methodology document — `medbs_pipeline_methodology.qmd`

A Quarto methodology write-up (`format: html`, all R chunks `eval: false` —
illustrative, not executed at render time) documenting
`lib_survey_fg_density_functions.R` and the two Step-1 driver scripts. Key
features:

* **Embedded interactive map** — the actual `gsa_boundaries.geojson` (from
  `export_gsa_boundaries.R`) is inlined as a JSON blob and rendered via
  Leaflet.js. Lets a reader toggle between the GSA-based and custom-region
  workflows, live-preview `FILTER_AREAS`/`CUSTOM_BBOX`, and generate a
  ready-to-paste R config block — all client-side JavaScript, no R executes
  behind it.
* **Section structure:** Overview (with mermaid diagrams for the pipeline,
  FG-matching sub-flow, and density/weighting sub-flow) → Input files →
  Interactive configuration → Functional Group matching (worked examples of
  each matching stage) → Study-area definition (GSA vs. custom) → Per-haul
  density and corrections (swept-area formula, depth strata, catchability,
  outliers) → Strata-weighted density → Region-wide aggregation → MEDIAS
  acoustic survey → Validation plots → Sample coverage map → Outputs → a
  full `ecopath_ecosim_inputs.xlsx` deep-dive, including the same final
  sheet-order table reproduced above.
* Real output files are copied into `figures/`/`downloads/` folders next to
  the rendered document (via `safe_include_graphics()`/`safe_download_link()`
  helpers) so the rendered HTML is portable off the author's own machine,
  since `out_dir` is otherwise an absolute local path.