---
title: "Biomass Pipeline — Live Explorer"
subtitle: "A Shiny-powered companion to pipeline_documentation.qmd: pick a year range and a set of GSAs, and the figures below actually re-run against real per-sample MEDITS data — not just a generated config block."
author: "Daniel Vilas"
format: html
server: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

suppressWarnings(suppressMessages({
  library(shiny)
  library(data.table)
  library(leaflet)
  library(ggplot2)
  library(sf)
}))

## Resolve every relative path below against THIS DOCUMENT's own directory,
## not whatever the current working directory happens to be when the R
## process starts. quarto/knitr USUALLY sets the working directory to match
## already, but "usually" isn't "always" - it depends on `execute.dir` in a
## _quarto.yml (if this file is part of a project), and on exactly where/how
## `quarto serve`/`quarto preview` was invoked from. A relative path silently
## resolving against the wrong directory is exactly the kind of failure that
## surfaces as a LATER, unrelated-looking error (e.g. deep in a Shiny
## reactive) rather than a clear "file not found" right here - anchoring
## explicitly removes that ambiguity, and every error message below prints
## the full resolved path so a real "it's genuinely not there" case is easy
## to tell apart from a "wrong base directory" case.
DOC_DIR <- tryCatch({
  ci <- knitr::current_input(dir = TRUE)
  if (is.null(ci) || !nzchar(ci)) stop("knitr::current_input() returned nothing")
  dirname(normalizePath(ci, mustWork = TRUE))
}, error = function(e) {
  message("[pipeline_documentation_shiny] Could not resolve this document's own directory via ",
          "knitr::current_input() (", conditionMessage(e), ") - falling back to the current ",
          "working directory instead. If LIB_PATH/CACHE_PATH below can't be found, try running ",
          "quarto from this .qmd's own folder, or set LIB_PATH/CACHE_PATH to absolute paths.")
  getwd()
})
message("[pipeline_documentation_shiny] Resolving paths relative to: ", DOC_DIR)

## This document needs the SAME library file every numbered pipeline script
## sources, since the reactive code below calls its real aggregation/plotting
## functions (compute_fg_densities_by_stratum(), weight_by_strata(),
## weight_by_area(), plot_fg_timeseries_by_area(), plot_fg_timeseries_regional(),
## plot_strata_profile(), plot_sample_map()) - NOT a reimplementation of them.
## Adjust this path if lib_survey_fg_density_functions.R lives somewhere else
## relative to this .qmd (an absolute path works too - DOC_DIR is only
## prepended to a RELATIVE path, via file.path() below).
LIB_PATH <- file.path(DOC_DIR, "lib_survey_fg_density_functions.R")
if (!file.exists(LIB_PATH)) {
  stop("Can't find '", LIB_PATH, "' (resolved from this document's own directory, '", DOC_DIR,
       "') - set LIB_PATH above to wherever lib_survey_fg_density_functions.R actually is, ",
       "relative to this file or as an absolute path.")
}
source(LIB_PATH)

## The data cache 01_biomass.R optionally writes when SAVE_SHINY_CACHE_PATH is
## set (see that variable's own comment in 01_biomass.R's Configuration
## section) - the final per-sample MEDITS table (FG-matched, area/stratum-
## assigned, catchability-corrected, outliers already resolved), plus the GSA
## shapefile and strata definition needed to re-aggregate it. Building this
## cache is a one-time step, done by running 01_biomass.R itself once (ideally
## with FILTER_AREAS/YEAR_ECOPATH/TS_YEARS set to their full available extent,
## so this app has the widest possible range to explore) - see the callout
## below for the exact steps.
##
## Missing cache is NOT a stop() here on purpose - unlike LIB_PATH above
## (genuinely fatal; there's no app at all without the shared functions),
## a missing cache is the normal state the FIRST time anyone opens this
## document, before they've run 01_biomass.R once. CACHE_AVAILABLE gates
## every reactive/render block below via shiny::validate(need(...)), which
## Shiny displays as a clean inline message in place of that output - not
## an uncaught error that kills the whole `quarto serve` process. The app
## still starts and shows the "Before you start" instructions either way.
CACHE_PATH <- file.path(DOC_DIR, "shiny_cache/survey_density_raw.rds")
CACHE_AVAILABLE <- file.exists(CACHE_PATH)
CACHE_MISSING_MSG <- paste0(
  "Data cache not found at '", CACHE_PATH, "' (resolved from this document's own directory, '",
  DOC_DIR, "'). See \"Before you start\" above for how to build it by running 01_biomass.R once ",
  "with SAVE_SHINY_CACHE_PATH set, then reload this page. If you DID build it and it's still not ",
  "found here, check it actually landed in a 'shiny_cache/' folder next to THIS .qmd file, not ",
  "next to 01_biomass.R's own out_dir - those are not the same directory unless you put this ",
  "document there too."
)

## AUTO_BUILD_CACHE: if TRUE (the default) and no cache exists yet, this app
## tries to build one itself on first load, by running 01_biomass.R once
## with SAVE_SHINY_CACHE_PATH/STOP_AFTER_SHINY_CACHE set - the exact manual
## steps in "Before you start" below, done automatically instead of by hand.
## It runs with 01_biomass.R's own built-in defaults (AREA_MODE = "westmed",
## every GSA 1-11, the full 1994-1996 base years and 1995-2023 time series -
## already the widest extent that ships as this pipeline's worked example),
## and stops right after the cache is saved (see STOP_AFTER_SHINY_CACHE's own
## comment in 01_biomass.R) rather than continuing through MEDIAS/stock-
## assessment/Excel-export/plot-file steps this cache doesn't need. It still
## needs the same pCloud data access 01_biomass.R normally needs, and can
## take a few minutes (GSA shapefile + bathymetry downloads, MEDITS loading,
## FG matching) - that wait only happens ONCE, the first time this app is
## opened with no cache yet; every reload after that reads the cache
## instantly, exactly like before. Set AUTO_BUILD_CACHE <- FALSE below to
## skip this and go back to just showing the manual "Before you start"
## instructions.
AUTO_BUILD_CACHE <- TRUE

if (!CACHE_AVAILABLE && AUTO_BUILD_CACHE) {
  BIOMASS_SCRIPT_PATH <- file.path(DOC_DIR, "01_biomass.R")
  if (!file.exists(BIOMASS_SCRIPT_PATH)) {
    message("[pipeline_documentation_shiny] AUTO_BUILD_CACHE is TRUE but '", BIOMASS_SCRIPT_PATH,
            "' isn't next to this document, so there's nothing to auto-run - falling back to the",
            " manual instructions below.")
  } else {
    message("[pipeline_documentation_shiny] No cache found - building one automatically by running",
            " 01_biomass.R once (this can take a few minutes: GSA shapefile + bathymetry downloads,",
            " MEDITS loading, FG matching). This happens only on this first load; every reload after",
            " it succeeds reads the saved cache instantly. Set AUTO_BUILD_CACHE <- FALSE in this",
            " document's setup chunk to skip this and just show the manual instructions instead.")
    auto_build_ok <- tryCatch({
      assign("SAVE_SHINY_CACHE_PATH",   CACHE_PATH, envir = .GlobalEnv)
      assign("STOP_AFTER_SHINY_CACHE",  TRUE,       envir = .GlobalEnv)
      ## local = a fresh env (not .GlobalEnv) - keeps 01_biomass.R's ~2000
      ## lines of intermediate objects (dt, area_shp, out_dir, ...) out of
      ## this document's own environment. The two assign()s just above
      ## still land in the REAL .GlobalEnv, which is where 01_biomass.R's
      ## own `exists(..., envir = .GlobalEnv, inherits = FALSE)` config
      ## checks explicitly look, regardless of where the script itself runs.
      source(BIOMASS_SCRIPT_PATH, local = new.env(parent = globalenv()))
      TRUE
    }, shinyCacheStop = function(e) TRUE, error = function(e) {
      message("[pipeline_documentation_shiny] Automatic cache build failed: ", conditionMessage(e),
              " - falling back to the manual instructions below.")
      FALSE
    })
    CACHE_AVAILABLE <- isTRUE(auto_build_ok) && file.exists(CACHE_PATH)
    if (CACHE_AVAILABLE) {
      message("[pipeline_documentation_shiny] Automatic cache build succeeded - loading it now.")
    }
  }
}

## build_demo_cache(): a small, entirely synthetic stand-in for a real
## SAVE_SHINY_CACHE_PATH cache - same list shape 01_biomass.R saves
## (dt/area_shp/area_id_col/strata_def), but with made-up polygons,
## strata and FGs instead of real MEDITS/GSA data. Its only purpose is
## to give this app SOMETHING to show - populated sliders, a real map,
## real (if meaningless) figures - the moment it's opened, even before
## any real cache exists or an automatic build has had a chance to
## succeed. Every value in it is fabricated; IS_DEMO_DATA (set where
## this is used, below) is what tells the UI to say so.
##
## Deliberately does NOT call compute_strata_area_by_area() - that
## function downloads REAL NOAA bathymetry for whatever polygon it's
## given, which would either fail outright or return nonsense for a
## hand-drawn square that isn't really seafloor. Its output shape
## (AreaID/Stratum/area_km2/prop) is instead hand-built directly here.
build_demo_cache <- function() {
  set.seed(42)

  ## Two small square "GSA-like" polygons, far enough apart to look
  ## like distinct areas on the map, both with an explicit CRS (plot_
  ## sample_map() stop()s without one - see its own comment).
  poly1 <- sf::st_polygon(list(rbind(c(3, 40), c(4, 40), c(4, 41), c(3, 41), c(3, 40))))
  poly2 <- sf::st_polygon(list(rbind(c(5.5, 39.5), c(6.5, 39.5), c(6.5, 40.5), c(5.5, 40.5), c(5.5, 39.5))))
  demo_area_shp <- sf::st_sf(AreaID = c(1, 2), geometry = sf::st_sfc(poly1, poly2, crs = 4326))

  demo_strata_def <- data.table(stratum_num = 1:3, depth_min = c(0, 50, 100), depth_max = c(50, 100, 200))
  demo_fgs <- data.table(FG_num = 1:3, FG_name = c("Demo hake-like FG", "Demo shrimp-like FG", "Demo mullet-like FG"))
  demo_years <- 2015:2023
  n_samples_per_area_year <- 25

  samples <- rbindlist(lapply(list(list(id = 1, poly = poly1), list(id = 2, poly = poly2)), function(a) {
    bb <- sf::st_bbox(a$poly)
    rbindlist(lapply(demo_years, function(yr) {
      n <- n_samples_per_area_year
      data.table(
        SampleID = paste0("DEMO_", a$id, "_", yr, "_", seq_len(n)),
        AreaID   = a$id,
        Year     = yr,
        Stratum  = sample(1:3, n, replace = TRUE, prob = c(0.5, 0.3, 0.2)),
        Lon      = stats::runif(n, bb["xmin"], bb["xmax"]),
        Lat      = stats::runif(n, bb["ymin"], bb["ymax"])
      )
    }))
  }))

  ## Cross each sample with every demo FG (one density observation per
  ## sample/FG, some genuinely zero - a real haul rarely catches every
  ## FG), with a mild upward trend over the demo years for FG 1 only,
  ## so plot_fg_timeseries_by_area()/plot_fg_timeseries_regional() have
  ## something visibly non-flat to draw.
  demo_dt <- samples[rep(seq_len(.N), each = nrow(demo_fgs))]
  demo_dt[, FG_num := rep(demo_fgs$FG_num, times = nrow(samples))]
  demo_dt <- merge(demo_dt, demo_fgs, by = "FG_num")
  demo_dt[, Density := round(stats::rexp(.N, rate = 1 / 4) * (stats::runif(.N) > 0.25), 2)]
  demo_dt[FG_num == 1, Density := round(Density * (1 + 0.06 * (Year - min(demo_years))), 2)]

  ## Hand-built strata areas (km2) - fabricated but internally
  ## consistent (prop sums to 1 within each AreaID), exactly the shape
  ## weight_by_strata()/weight_by_area() expect from a real
  ## compute_strata_area_by_area() call.
  demo_strata_area_by_area <- data.table(
    AreaID    = rep(c(1, 2), each = 3),
    Stratum   = rep(1:3, times = 2),
    area_km2  = c(900, 500, 260, 700, 480, 220)
  )
  demo_strata_area_by_area[, prop := area_km2 / sum(area_km2), by = AreaID]

  list(
    dt                         = demo_dt,
    area_shp                   = demo_area_shp,
    area_id_col                = "AreaID",
    strata_def                 = demo_strata_def,
    species_taxonomy           = NULL,
    cached_at                  = Sys.time(),
    filter_areas_at_cache_time = c(1, 2),
    year_range_at_cache_time   = range(demo_years),
    strata_area_by_area_demo   = demo_strata_area_by_area
  )
}

IS_DEMO_DATA <- FALSE

if (CACHE_AVAILABLE) {
  cache <- readRDS(CACHE_PATH)

  ## Bathymetry-derived strata areas are the slow part (one bathymetry
  ## download per GSA) and depend ONLY on which areas exist in the cache,
  ## never on the year range or GSA selection the user picks interactively -
  ## computed once here at startup, not inside a reactive(), so changing the
  ## year slider or the GSA checkboxes never re-triggers a bathymetry
  ## download.
  ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
  STRATA_AREA_BY_AREA_FULL <- compute_strata_area_by_area(
    area_ids   = ALL_AREA_IDS,
    area_shp   = cache$area_shp,
    area_id_col = cache$area_id_col,
    strata_def = cache$strata_def
  )

  YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
  YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
} else {
  ## Final fallback: no real cache, and either AUTO_BUILD_CACHE is off or
  ## its attempt above failed. Rather than showing an empty "no data yet"
  ## state on first load, fall back to a small synthetic demo dataset
  ## (built purely in-memory, in a few milliseconds - no downloads, no
  ## pCloud access) so every control and figure below has SOMETHING real
  ## to react to immediately. CACHE_AVAILABLE is set TRUE here on purpose:
  ## every validate(need(CACHE_AVAILABLE, ...)) guard downstream means
  ## "is there ANY data to show" (real or demo) - IS_DEMO_DATA is the
  ## separate flag the UI uses to say which one it actually is.
  message("[pipeline_documentation_shiny] ", CACHE_MISSING_MSG,
          " Falling back to a small synthetic demo dataset so this app shows",
          " something immediately - see the banner in the app itself once it",
          " loads. Build a real cache (see \"Before you start\" above) and",
          " reload this page to replace it with your actual MEDITS data.")
  demo_cache <- tryCatch(build_demo_cache(), error = function(e) {
    message("[pipeline_documentation_shiny] Building the synthetic demo dataset also failed (",
            conditionMessage(e), ") - falling back to the plain \"no data yet\" message instead.")
    NULL
  })

  if (!is.null(demo_cache)) {
    cache <- demo_cache
    IS_DEMO_DATA <- TRUE
    CACHE_AVAILABLE <- TRUE

    ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
    STRATA_AREA_BY_AREA_FULL <- cache$strata_area_by_area_demo
    YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
    YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
  } else {
    cache <- NULL
    ALL_AREA_IDS <- integer(0)
    STRATA_AREA_BY_AREA_FULL <- NULL
    YEAR_MIN <- 1990L
    YEAR_MAX <- 2023L
  }
}
```

:::: {.callout-warning}
## Before you start: this needs a live R session, and a one-time data cache

Unlike `pipeline_documentation.qmd`, this file is **not** a static document
you can just open in a browser. `server: shiny` in the YAML above means it
needs an actual R process running behind it - open it with `quarto preview
pipeline_documentation_shiny.qmd` (or `quarto serve`) from a machine with R,
Shiny, and this pipeline's packages installed, or deploy it to
shinyapps.io / Posit Connect / your own Shiny Server. Double-clicking a
rendered `.html` file will not work - there's no R behind a plain HTML file.

It also needs a **one-time data cache** before it can show anything - but
you shouldn't have to build that by hand. **The first time you open this
app with no cache yet, it builds one itself**: it runs `01_biomass.R` once,
automatically, using that script's own built-in defaults (`AREA_MODE =
"westmed"`, every GSA 1-11, its full available year range) and stops right
after the cache is saved - it does *not* run the rest of the pipeline
(MEDIAS, stock assessment, Excel export, plot files) just for this. That
first run still needs the same pCloud data access `01_biomass.R` normally
needs, and can take a few minutes (GSA shapefile + bathymetry downloads,
MEDITS loading, FG matching) - you'll see progress messages in the R
console it's running in. Every reload after that succeeds is instant, since
it just reads the saved cache.

**You don't have to wait for any of that to see how this app works.** If
no real cache exists yet and the automatic build above hasn't finished (or
fails - no pCloud access from wherever this is running, `01_biomass.R` not
sitting next to this `.qmd`, etc.), the app falls back to a small,
entirely synthetic demo dataset - fabricated GSAs, strata, functional
groups and samples, built in-memory in milliseconds - so every control and
figure below is already populated and interactive the first time you open
this page. A grey banner above the controls says plainly when you're
looking at that demo data instead of your real MEDITS results. Once a
real cache exists (auto-built or built by hand), reload the page and the
demo banner disappears in favor of the real thing.

If you'd rather build the real cache yourself instead of waiting on the
automatic build, the manual steps are the same ones it runs under the
hood:

1. In `01_biomass.R`'s Configuration section, temporarily set
   `FILTER_AREAS` to every GSA you have MEDITS data for (not just your usual
   region) and `YEAR_ECOPATH`/`TS_YEARS` to your full available year range -
   the widest extent you want this app to be able to explore later.
2. Set `SAVE_SHINY_CACHE_PATH <- "shiny_cache/survey_density_raw.rds"`
   (relative to wherever `01_biomass.R` writes `out_dir`) and run the script
   once, in full.
3. Copy (or symlink) the resulting `shiny_cache/survey_density_raw.rds` to
   sit next to this `.qmd` file - `CACHE_PATH` in the setup chunk above
   expects it there.
4. Revert `FILTER_AREAS`/`YEAR_ECOPATH`/`TS_YEARS`/`SAVE_SHINY_CACHE_PATH`
   back to their normal values for `01_biomass.R`'s regular use - this cache
   is a frozen snapshot for THIS document only, not something the numbered
   pipeline scripts read back.

Either way, the cache is a snapshot: it does not update itself if the
underlying MEDITS data changes. Delete `shiny_cache/survey_density_raw.rds`
and reload this app (to trigger another automatic build) - or repeat steps
1-3 by hand - whenever you want it reflecting newer data. To turn off the
automatic build entirely and always require the manual steps, set
`AUTO_BUILD_CACHE <- FALSE` in this document's setup chunk.
::::

## What's actually live here, and what isn't

Changing the **year range** or the **GSAs included** below genuinely
re-runs the pipeline's own strata-weighting cascade
(`compute_fg_densities_by_stratum()` → `weight_by_strata()` →
`weight_by_area()`) against the cached per-sample data, and the four figures
re-render from that real result - not a lookup of a pre-rendered image.

Three of the static form's other options don't carry over as live toggles,
for reasons worth being upfront about rather than silently shipping a
control that doesn't really do anything:

* **`DROP_OUTLIERS`** - outlier removal already ran once, at cache-build
  time (`01_biomass.R`'s own Step, before this document's cache is saved).
  Toggling it here would need the raw pre-outlier-removal samples cached
  too, which this cache deliberately doesn't keep (it's already large
  enough as the full per-sample extent).
* **`STRATA`** - `weight_by_area()` (needed for the region-wide figure)
  only implements the strata-weighted path; the unstratified
  (`simple_area_density()`) path has no region-wide aggregation wired up in
  the shared library today. This app always uses `STRATA = TRUE`
  to keep the region-wide figure working; the by-area figure would be the
  only one affected by `STRATA = FALSE` in any case.
* **Custom bounding box** - re-deriving a custom polygon's strata *areas*
  live would mean a fresh bathymetry download per input change, which is
  far too slow for an interactive control. GSA selection is live because
  `STRATA_AREA_BY_AREA_FULL` above is precomputed once for every GSA in the
  cache; a live bbox would need that recomputed on the fly instead.

The **"Generate config block"** panel at the bottom still works exactly like
the static version - it always reflects your current picks above, ready to
paste into `01_biomass.R`, `DROP_OUTLIERS`/`STRATA`/bbox included, even
though only year range and GSA selection also drive the figures live.

```{r ui-inputs}
if (!CACHE_AVAILABLE) {
  div(
    style = paste(
      "padding: 1em 1.25em; margin: 0.5em 0 1.5em; border: 1px solid #f0ad4e;",
      "border-radius: 6px; background: #fcf8e3; color: #664d03;"
    ),
    strong("This app has no data to show yet."), br(),
    CACHE_MISSING_MSG
  )
} else {
  tagList(
    if (isTRUE(IS_DEMO_DATA)) {
      div(
        style = paste(
          "padding: 0.75em 1.25em; margin: 0.5em 0 1em; border: 1px solid #6c757d;",
          "border-radius: 6px; background: #f1f1f1; color: #333;"
        ),
        strong("You're viewing synthetic demo data, not real MEDITS data."), br(),
        "No real cache was found (and either the automatic build above hasn't finished or ",
        "it failed) - so every control and figure below is running against a small, ",
        "made-up dataset (fabricated GSAs, strata, functional groups and samples) just so ",
        "you can see how this app works before your own data is ready. Build a real cache ",
        "(see \"Before you start\" above) and reload this page to replace this with your ",
        "actual MEDITS-derived results."
      )
    },
    fluidRow(
    column(
      width = 4,
      sliderInput("year_range", "Ecosim time-series years (TS_YEARS)",
                  min = YEAR_MIN, max = YEAR_MAX,
                  value = c(YEAR_MIN, YEAR_MAX), step = 1, sep = ""),
      checkboxGroupInput("filter_areas", "GSA numbers to include (FILTER_AREAS)",
                          choices = ALL_AREA_IDS, selected = ALL_AREA_IDS, inline = TRUE),
      hr(),
      checkboxInput("drop_outliers", "DROP_OUTLIERS (config block only - see note above)", value = TRUE),
      checkboxInput("normalize_ts", "NORMALIZE_TS (config block only)", value = TRUE),
      checkboxInput("strata_config", "STRATA (config block only - figures always use TRUE)", value = TRUE),
      hr(),
      numericInput("year_ecopath_start", "Ecopath base year, start", value = 1994),
      numericInput("year_ecopath_end", "Ecopath base year, end", value = 1996),
      hr(),
      downloadButton("download_config", "Download config block (.R)")
    ),
    column(
      width = 8,
      leafletOutput("gsa_map", height = 320)
    )
    )
  )
}
```

```{r reactive-pipeline}
## req(CACHE_AVAILABLE) first in every reactive here - without it, a
## reactive further down the chain (e.g. filtered_dt(), which touches
## cache$dt) would be the first thing to notice the cache is missing,
## with a less clear error than the validate(need(...)) message each
## render*() block below already shows. This just makes the failure
## point explicit and identical everywhere.
selected_areas <- reactive({
  req(CACHE_AVAILABLE, input$filter_areas)
  as.numeric(input$filter_areas)
})

filtered_dt <- reactive({
  cache$dt[Year >= input$year_range[1] & Year <= input$year_range[2] &
             AreaID %in% selected_areas()]
})

per_group_fg <- reactive({
  compute_fg_densities_by_stratum(filtered_dt(), strata = TRUE)
})

n_samples_by_stratum <- reactive({
  filtered_dt()[!is.na(Stratum), .(n_samples = uniqueN(SampleID)), by = .(AreaID, Year, Stratum)]
})

strata_area_selected <- reactive({
  STRATA_AREA_BY_AREA_FULL[AreaID %in% selected_areas()]
})

fg_index <- reactive({
  weight_by_strata(per_group_fg(), n_samples_by_stratum(), strata_area_selected())
})

fg_index_regional <- reactive({
  weight_by_area(fg_index())
})
```

## Sample coverage

```{r plot-sample-map}
renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_sample_map(filtered_dt(), cache$area_shp, cache$area_id_col,
                   selected_areas = selected_areas())
})
```

## FG density by area

```{r plot-by-area}
renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_by_area(fg_index())
})
```

## FG density, region-wide

```{r plot-regional}
renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_regional(fg_index_regional())
})
```

## Density by depth stratum

```{r plot-strata}
renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_strata_profile(fg_index(), cache$strata_def)
})
```

```{r map-render, context="server"}
## context="server" is required here, not optional - this is the actual
## fix for "Error: object 'output' not found" at this chunk. Quarto's
## Shiny engine evaluates every chunk WITHOUT context="server" during an
## initial render pass that only builds the static UI, where `output`
## (and `session`) don't exist yet - that pass is what raised the error.
## reactive()/renderPlot() calls elsewhere in this document survive that
## pass because they just build a promise object to return into the UI;
## an explicit `output$x <- ...` assignment needs `output` to actually
## exist, which only happens inside the live server function - i.e. in a
## context="server" chunk. (config-download below needs the same fix,
## for the same reason - it also assigns into `output`.)
##
## No leafletOutput("gsa_map") even exists in the UI when !CACHE_AVAILABLE
## (see ui-inputs above), so neither block below ever actually runs in
## that case - guarded anyway so this chunk fails the same explicit way
## as everything else if that ever changes, rather than silently.
output$gsa_map <- renderLeaflet({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  shp <- cache$area_shp
  leaflet(shp) |>
    addTiles() |>
    addPolygons(
      fillColor = ~ifelse(get(cache$area_id_col) %in% isolate(selected_areas()), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(cache$area_id_col))
    )
})

observe({
  req(CACHE_AVAILABLE)
  shp <- cache$area_shp
  leafletProxy("gsa_map", data = shp) |>
    clearShapes() |>
    addPolygons(
      fillColor = ~ifelse(get(cache$area_id_col) %in% selected_areas(), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(cache$area_id_col))
    )
})
```

```{r config-download, context="server"}
## context="server" required here too - this chunk assigns into
## output$download_config, same reason as map-render above.
##
## downloadButton("download_config") only exists in the UI when
## CACHE_AVAILABLE (see ui-inputs above), so this is unreachable
## otherwise - req() guard kept for the same reason as map-render.
config_text <- reactive({
  req(CACHE_AVAILABLE)
  paste(
    '## --- generated by "Live Explorer" - 01_biomass.R ---',
    paste0("FILTER_AREAS  <- c(", paste(selected_areas(), collapse = ", "), ")"),
    paste0("STRATA        <- ", if (input$strata_config) "TRUE" else "FALSE"),
    paste0("YEAR_ECOPATH  <- ", input$year_ecopath_start, ":", input$year_ecopath_end),
    paste0("TS_YEARS      <- ", input$year_range[1], ":", input$year_range[2]),
    paste0("DROP_OUTLIERS <- ", if (input$drop_outliers) "TRUE" else "FALSE"),
    paste0("NORMALIZE_TS  <- ", if (input$normalize_ts) "TRUE" else "FALSE"),
    sep = "\n"
  )
})

output$download_config <- downloadHandler(
  filename = function() "biomass_config_block.R",
  content = function(file) writeLines(config_text(), file)
)
```