---
title: "Biomass Pipeline — Live Explorer"
subtitle: "A Shiny-powered companion to pipeline_documentation.qmd: pick a year range and a set of GSAs, and the figures below actually re-run against real per-sample MEDITS data — not just a generated config block."
author: "Daniel Vilas"
format: html
server: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

suppressWarnings(suppressMessages({
  library(shiny)
  library(data.table)
  library(leaflet)
  library(ggplot2)
  library(sf)
}))

## Resolve every relative path below against THIS DOCUMENT's own directory,
## not whatever the current working directory happens to be when the R
## process starts. quarto/knitr USUALLY sets the working directory to match
## already, but "usually" isn't "always" - it depends on `execute.dir` in a
## _quarto.yml (if this file is part of a project), and on exactly where/how
## `quarto serve`/`quarto preview` was invoked from. A relative path silently
## resolving against the wrong directory is exactly the kind of failure that
## surfaces as a LATER, unrelated-looking error (e.g. deep in a Shiny
## reactive) rather than a clear "file not found" right here - anchoring
## explicitly removes that ambiguity, and every error message below prints
## the full resolved path so a real "it's genuinely not there" case is easy
## to tell apart from a "wrong base directory" case.
DOC_DIR <- tryCatch({
  ci <- knitr::current_input(dir = TRUE)
  if (is.null(ci) || !nzchar(ci)) stop("knitr::current_input() returned nothing")
  dirname(normalizePath(ci, mustWork = TRUE))
}, error = function(e) {
  message("[pipeline_documentation_shiny] Could not resolve this document's own directory via ",
          "knitr::current_input() (", conditionMessage(e), ") - falling back to the current ",
          "working directory instead. If LIB_PATH/CACHE_PATH below can't be found, try running ",
          "quarto from this .qmd's own folder, or set LIB_PATH/CACHE_PATH to absolute paths.")
  getwd()
})
message("[pipeline_documentation_shiny] Resolving paths relative to: ", DOC_DIR)

## This document needs the SAME library file every numbered pipeline script
## sources, since the reactive code below calls its real aggregation/plotting
## functions (compute_fg_densities_by_stratum(), weight_by_strata(),
## weight_by_area(), plot_fg_timeseries_by_area(), plot_fg_timeseries_regional(),
## plot_strata_profile(), plot_sample_map()) - NOT a reimplementation of them.
## Adjust this path if lib_survey_fg_density_functions.R lives somewhere else
## relative to this .qmd (an absolute path works too - DOC_DIR is only
## prepended to a RELATIVE path, via file.path() below).
LIB_PATH <- file.path(DOC_DIR, "lib_survey_fg_density_functions.R")
if (!file.exists(LIB_PATH)) {
  stop("Can't find '", LIB_PATH, "' (resolved from this document's own directory, '", DOC_DIR,
       "') - set LIB_PATH above to wherever lib_survey_fg_density_functions.R actually is, ",
       "relative to this file or as an absolute path.")
}
source(LIB_PATH)

## The data cache 01_biomass.R optionally writes when SAVE_SHINY_CACHE_PATH is
## set (see that variable's own comment in 01_biomass.R's Configuration
## section) - the final per-sample MEDITS table (FG-matched, area/stratum-
## assigned, catchability-corrected, outliers already resolved), plus the GSA
## shapefile and strata definition needed to re-aggregate it. Building this
## cache is a one-time step, done by running 01_biomass.R itself once (ideally
## with FILTER_AREAS/YEAR_ECOPATH/TS_YEARS set to their full available extent,
## so this app has the widest possible range to explore) - see the callout
## below for the exact steps.
##
## Missing cache is NOT a stop() here on purpose - unlike LIB_PATH above
## (genuinely fatal; there's no app at all without the shared functions),
## a missing cache is the normal state the FIRST time anyone opens this
## document, before they've run 01_biomass.R once. CACHE_AVAILABLE gates
## every reactive/render block below via shiny::validate(need(...)), which
## Shiny displays as a clean inline message in place of that output - not
## an uncaught error that kills the whole `quarto serve` process. The app
## still starts and shows the "Before you start" instructions either way.
CACHE_PATH <- file.path(DOC_DIR, "shiny_cache/survey_density_raw.rds")
CACHE_AVAILABLE <- file.exists(CACHE_PATH)
CACHE_MISSING_MSG <- paste0(
  "Data cache not found at '", CACHE_PATH, "' (resolved from this document's own directory, '",
  DOC_DIR, "'). See \"Before you start\" above for how to build it by running 01_biomass.R once ",
  "with SAVE_SHINY_CACHE_PATH set, then reload this page. If you DID build it and it's still not ",
  "found here, check it actually landed in a 'shiny_cache/' folder next to THIS .qmd file, not ",
  "next to 01_biomass.R's own out_dir - those are not the same directory unless you put this ",
  "document there too."
)

## AUTO_BUILD_CACHE: if TRUE and no cache exists yet, this app tries to
## build one itself on first load, by running 01_biomass.R once with
## SAVE_SHINY_CACHE_PATH/STOP_AFTER_SHINY_CACHE set - the exact manual steps
## in "Before you start" below, done automatically instead of by hand.
##
## Defaults to FALSE now - NOT because it doesn't work, but because of
## WHERE it runs: the code below is the document's own setup chunk, which
## Quarto/Shiny evaluates synchronously, once, before this page can render
## ANYTHING - no controls, no plots, not even the demo-data banner below.
## A real 01_biomass.R run needs pCloud access and can take several minutes
## (GSA shapefile + bathymetry downloads, MEDITS loading, FG matching) even
## when everything goes right; if pCloud isn't reachable from wherever
## `quarto serve`/`quarto preview` is running, or the run hits any problem
## slow enough to not fail cleanly and quickly, the browser tab just sits
## blank for however long that takes (or indefinitely) - which is exactly
## what "opens to a blank page, nothing shows" looks like from the outside,
## and is indistinguishable from the app being broken. Leaving this FALSE
## means the page always paints immediately: no cache yet -> straight to
## the synthetic demo dataset below (milliseconds, no network), never
## blocked on a live pipeline run. Set AUTO_BUILD_CACHE <- TRUE here only
## if you specifically want that automatic build attempt back (e.g. you
## know pCloud is reachable from this machine and are fine waiting) -
## otherwise build the real cache by running 01_biomass.R yourself (see
## "Before you start" below) and this app will pick it up on the next
## reload, same as always.
AUTO_BUILD_CACHE <- FALSE

if (!CACHE_AVAILABLE && AUTO_BUILD_CACHE) {
  BIOMASS_SCRIPT_PATH <- file.path(DOC_DIR, "01_biomass.R")
  if (!file.exists(BIOMASS_SCRIPT_PATH)) {
    message("[pipeline_documentation_shiny] AUTO_BUILD_CACHE is TRUE but '", BIOMASS_SCRIPT_PATH,
            "' isn't next to this document, so there's nothing to auto-run - falling back to the",
            " manual instructions below.")
  } else {
    message("[pipeline_documentation_shiny] No cache found - building one automatically by running",
            " 01_biomass.R once (this can take a few minutes: GSA shapefile + bathymetry downloads,",
            " MEDITS loading, FG matching). This happens only on this first load; every reload after",
            " it succeeds reads the saved cache instantly. Set AUTO_BUILD_CACHE <- FALSE in this",
            " document's setup chunk to skip this and just show the manual instructions instead.")
    auto_build_ok <- tryCatch({
      assign("SAVE_SHINY_CACHE_PATH",   CACHE_PATH, envir = .GlobalEnv)
      assign("STOP_AFTER_SHINY_CACHE",  TRUE,       envir = .GlobalEnv)
      ## local = a fresh env (not .GlobalEnv) - keeps 01_biomass.R's ~2000
      ## lines of intermediate objects (dt, area_shp, out_dir, ...) out of
      ## this document's own environment. The two assign()s just above
      ## still land in the REAL .GlobalEnv, which is where 01_biomass.R's
      ## own `exists(..., envir = .GlobalEnv, inherits = FALSE)` config
      ## checks explicitly look, regardless of where the script itself runs.
      source(BIOMASS_SCRIPT_PATH, local = new.env(parent = globalenv()))
      TRUE
    }, shinyCacheStop = function(e) TRUE, error = function(e) {
      message("[pipeline_documentation_shiny] Automatic cache build failed: ", conditionMessage(e),
              " - falling back to the manual instructions below.")
      FALSE
    })
    CACHE_AVAILABLE <- isTRUE(auto_build_ok) && file.exists(CACHE_PATH)
    if (CACHE_AVAILABLE) {
      message("[pipeline_documentation_shiny] Automatic cache build succeeded - loading it now.")
    }
  }
}

## build_demo_cache(): a small, entirely synthetic stand-in for a real
## SAVE_SHINY_CACHE_PATH cache - same list shape 01_biomass.R saves
## (dt/area_shp/area_id_col/strata_def), but with made-up polygons,
## strata and FGs instead of real MEDITS/GSA data. Its only purpose is
## to give this app SOMETHING to show - populated sliders, a real map,
## real (if meaningless) figures - the moment it's opened, even before
## any real cache exists or an automatic build has had a chance to
## succeed. Every value in it is fabricated; IS_DEMO_DATA (set where
## this is used, below) is what tells the UI to say so.
##
## Deliberately does NOT call compute_strata_area_by_area() - that
## function downloads REAL NOAA bathymetry for whatever polygon it's
## given, which would either fail outright or return nonsense for a
## hand-drawn square that isn't really seafloor. Its output shape
## (AreaID/Stratum/area_km2/prop) is instead hand-built directly here.
build_demo_cache <- function() {
  set.seed(42)

  ## Two small square "GSA-like" polygons, far enough apart to look
  ## like distinct areas on the map, both with an explicit CRS (plot_
  ## sample_map() stop()s without one - see its own comment).
  poly1 <- sf::st_polygon(list(rbind(c(3, 40), c(4, 40), c(4, 41), c(3, 41), c(3, 40))))
  poly2 <- sf::st_polygon(list(rbind(c(5.5, 39.5), c(6.5, 39.5), c(6.5, 40.5), c(5.5, 40.5), c(5.5, 39.5))))
  demo_area_shp <- sf::st_sf(AreaID = c(1, 2), geometry = sf::st_sfc(poly1, poly2, crs = 4326))

  demo_strata_def <- data.table(stratum_num = 1:3, depth_min = c(0, 50, 100), depth_max = c(50, 100, 200))
  demo_fgs <- data.table(FG_num = 1:3, FG_name = c("Demo hake-like FG", "Demo shrimp-like FG", "Demo mullet-like FG"))
  demo_years <- 2015:2023
  n_samples_per_area_year <- 25

  samples <- rbindlist(lapply(list(list(id = 1, poly = poly1), list(id = 2, poly = poly2)), function(a) {
    bb <- sf::st_bbox(a$poly)
    rbindlist(lapply(demo_years, function(yr) {
      n <- n_samples_per_area_year
      data.table(
        SampleID = paste0("DEMO_", a$id, "_", yr, "_", seq_len(n)),
        AreaID   = a$id,
        Year     = yr,
        Stratum  = sample(1:3, n, replace = TRUE, prob = c(0.5, 0.3, 0.2)),
        Lon      = stats::runif(n, bb["xmin"], bb["xmax"]),
        Lat      = stats::runif(n, bb["ymin"], bb["ymax"])
      )
    }))
  }))

  ## Cross each sample with every demo FG (one density observation per
  ## sample/FG, some genuinely zero - a real haul rarely catches every
  ## FG), with a mild upward trend over the demo years for FG 1 only,
  ## so plot_fg_timeseries_by_area()/plot_fg_timeseries_regional() have
  ## something visibly non-flat to draw.
  demo_dt <- samples[rep(seq_len(.N), each = nrow(demo_fgs))]
  demo_dt[, FG_num := rep(demo_fgs$FG_num, times = nrow(samples))]
  demo_dt <- merge(demo_dt, demo_fgs, by = "FG_num")
  demo_dt[, Density := round(stats::rexp(.N, rate = 1 / 4) * (stats::runif(.N) > 0.25), 2)]
  demo_dt[FG_num == 1, Density := round(Density * (1 + 0.06 * (Year - min(demo_years))), 2)]

  ## Hand-built strata areas (km2) - fabricated but internally
  ## consistent (prop sums to 1 within each AreaID), exactly the shape
  ## weight_by_strata()/weight_by_area() expect from a real
  ## compute_strata_area_by_area() call.
  demo_strata_area_by_area <- data.table(
    AreaID    = rep(c(1, 2), each = 3),
    Stratum   = rep(1:3, times = 2),
    area_km2  = c(900, 500, 260, 700, 480, 220)
  )
  demo_strata_area_by_area[, prop := area_km2 / sum(area_km2), by = AreaID]

  list(
    dt                         = demo_dt,
    area_shp                   = demo_area_shp,
    area_id_col                = "AreaID",
    strata_def                 = demo_strata_def,
    species_taxonomy           = NULL,
    cached_at                  = Sys.time(),
    filter_areas_at_cache_time = c(1, 2),
    year_range_at_cache_time   = range(demo_years),
    strata_area_by_area_demo   = demo_strata_area_by_area
  )
}

IS_DEMO_DATA <- FALSE

if (CACHE_AVAILABLE) {
  cache <- readRDS(CACHE_PATH)

  ## Bathymetry-derived strata areas are the slow part (one bathymetry
  ## download per GSA) and depend ONLY on which areas exist in the cache,
  ## never on the year range or GSA selection the user picks interactively -
  ## computed once here at startup, not inside a reactive(), so changing the
  ## year slider or the GSA checkboxes never re-triggers a bathymetry
  ## download.
  ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
  STRATA_AREA_BY_AREA_FULL <- compute_strata_area_by_area(
    area_ids   = ALL_AREA_IDS,
    area_shp   = cache$area_shp,
    area_id_col = cache$area_id_col,
    strata_def = cache$strata_def
  )

  YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
  YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
} else {
  ## Final fallback: no real cache, and either AUTO_BUILD_CACHE is off or
  ## its attempt above failed. Rather than showing an empty "no data yet"
  ## state on first load, fall back to a small synthetic demo dataset
  ## (built purely in-memory, in a few milliseconds - no downloads, no
  ## pCloud access) so every control and figure below has SOMETHING real
  ## to react to immediately. CACHE_AVAILABLE is set TRUE here on purpose:
  ## every validate(need(CACHE_AVAILABLE, ...)) guard downstream means
  ## "is there ANY data to show" (real or demo) - IS_DEMO_DATA is the
  ## separate flag the UI uses to say which one it actually is.
  message("[pipeline_documentation_shiny] ", CACHE_MISSING_MSG,
          " Falling back to a small synthetic demo dataset so this app shows",
          " something immediately - see the banner in the app itself once it",
          " loads. Build a real cache (see \"Before you start\" above) and",
          " reload this page to replace it with your actual MEDITS data.")
  demo_cache <- tryCatch(build_demo_cache(), error = function(e) {
    message("[pipeline_documentation_shiny] Building the synthetic demo dataset also failed (",
            conditionMessage(e), ") - falling back to the plain \"no data yet\" message instead.")
    NULL
  })

  if (!is.null(demo_cache)) {
    cache <- demo_cache
    IS_DEMO_DATA <- TRUE
    CACHE_AVAILABLE <- TRUE

    ALL_AREA_IDS <- sort(unique(cache$dt$AreaID))
    STRATA_AREA_BY_AREA_FULL <- cache$strata_area_by_area_demo
    YEAR_MIN <- min(cache$dt$Year, na.rm = TRUE)
    YEAR_MAX <- max(cache$dt$Year, na.rm = TRUE)
  } else {
    cache <- NULL
    ALL_AREA_IDS <- integer(0)
    STRATA_AREA_BY_AREA_FULL <- NULL
    YEAR_MIN <- 1990L
    YEAR_MAX <- 2023L
  }
}
```

:::: {.callout-warning}
## Before you start: this needs a live R session, and a one-time data cache

Unlike `pipeline_documentation.qmd`, this file is **not** a static document
you can just open in a browser. `server: shiny` in the YAML above means it
needs an actual R process running behind it - open it with `quarto preview
pipeline_documentation_shiny.qmd` (or `quarto serve`) from a machine with R,
Shiny, and this pipeline's packages installed, or deploy it to
shinyapps.io / Posit Connect / your own Shiny Server. Double-clicking a
rendered `.html` file will not work - there's no R behind a plain HTML file.

It also needs a **one-time data cache** before it can show your real
MEDITS results - but you should never see a blank page waiting on that.
**The moment you open this app with no real cache yet, it shows a small
synthetic demo dataset instead** - fabricated GSAs, strata, functional
groups and samples, built in-memory in milliseconds, no network or pCloud
access needed - so every control and figure below is already populated
and interactive on first load. A grey banner above the controls says
plainly when you're looking at that demo data rather than your real
results.

Building the real cache is a separate, deliberate step - it is **not**
attempted automatically by default, on purpose: a real `01_biomass.R` run
needs pCloud access and can take several minutes (GSA shapefile +
bathymetry downloads, MEDITS loading, FG matching), and running that
synchronously inside this document's own startup code would block the
page from rendering anything at all until it finishes - which looks
identical to the app being broken if pCloud isn't reachable from wherever
`quarto serve`/`quarto preview` is running, or the run is simply slow.
(If you'd rather have that automatic attempt back anyway - e.g. you know
pCloud is reachable here and are fine waiting - set `AUTO_BUILD_CACHE <-
TRUE` in this document's setup chunk.)

The manual steps to build a real cache yourself:

1. In `01_biomass.R`'s Configuration section, temporarily set
   `FILTER_AREAS` to every GSA you have MEDITS data for (not just your usual
   region) and `YEAR_ECOPATH`/`TS_YEARS` to your full available year range -
   the widest extent you want this app to be able to explore later.
2. Set `SAVE_SHINY_CACHE_PATH <- "shiny_cache/survey_density_raw.rds"`
   (relative to wherever `01_biomass.R` writes `out_dir`) and run the script
   once, in full.
3. Copy (or symlink) the resulting `shiny_cache/survey_density_raw.rds` to
   sit next to this `.qmd` file - `CACHE_PATH` in the setup chunk above
   expects it there.
4. Revert `FILTER_AREAS`/`YEAR_ECOPATH`/`TS_YEARS`/`SAVE_SHINY_CACHE_PATH`
   back to their normal values for `01_biomass.R`'s regular use - this cache
   is a frozen snapshot for THIS document only, not something the numbered
   pipeline scripts read back.

The cache is a snapshot either way: it does not update itself if the
underlying MEDITS data changes. Repeat steps 1-3 (or, with
`AUTO_BUILD_CACHE <- TRUE`, delete `shiny_cache/survey_density_raw.rds`
and reload to trigger another automatic build) whenever you want it
reflecting newer data.
::::

## What's actually live here, and what isn't

Changing the **year range** or the **GSAs included** below genuinely
re-runs the pipeline's own strata-weighting cascade
(`compute_fg_densities_by_stratum()` → `weight_by_strata()` →
`weight_by_area()`) against the cached per-sample data, and the four figures
re-render from that real result - not a lookup of a pre-rendered image.

Three of the static form's other options don't carry over as live toggles,
for reasons worth being upfront about rather than silently shipping a
control that doesn't really do anything:

* **`DROP_OUTLIERS`** - outlier removal already ran once, at cache-build
  time (`01_biomass.R`'s own Step, before this document's cache is saved).
  Toggling it here would need the raw pre-outlier-removal samples cached
  too, which this cache deliberately doesn't keep (it's already large
  enough as the full per-sample extent).
* **`STRATA`** - `weight_by_area()` (needed for the region-wide figure)
  only implements the strata-weighted path; the unstratified
  (`simple_area_density()`) path has no region-wide aggregation wired up in
  the shared library today. This app always uses `STRATA = TRUE`
  to keep the region-wide figure working; the by-area figure would be the
  only one affected by `STRATA = FALSE` in any case.
* **Custom bounding box** - re-deriving a custom polygon's strata *areas*
  live would mean a fresh bathymetry download per input change, which is
  far too slow for an interactive control. GSA selection is live because
  `STRATA_AREA_BY_AREA_FULL` above is precomputed once for every GSA in the
  cache; a live bbox would need that recomputed on the fly instead.

The **"Generate config block"** panel at the bottom still works exactly like
the static version - it always reflects your current picks above, ready to
paste into `01_biomass.R`, `DROP_OUTLIERS`/`STRATA`/bbox included, even
though only year range and GSA selection also drive the figures live.

```{r ui-inputs}
if (!CACHE_AVAILABLE) {
  div(
    style = paste(
      "padding: 1em 1.25em; margin: 0.5em 0 1.5em; border: 1px solid #f0ad4e;",
      "border-radius: 6px; background: #fcf8e3; color: #664d03;"
    ),
    strong("This app has no data to show yet."), br(),
    CACHE_MISSING_MSG
  )
} else {
  tagList(
    if (isTRUE(IS_DEMO_DATA)) {
      div(
        style = paste(
          "padding: 0.75em 1.25em; margin: 0.5em 0 1em; border: 1px solid #6c757d;",
          "border-radius: 6px; background: #f1f1f1; color: #333;"
        ),
        strong("You're viewing synthetic demo data, not real MEDITS data."), br(),
        "No real cache was found (and either the automatic build above hasn't finished or ",
        "it failed) - so every control and figure below is running against a small, ",
        "made-up dataset (fabricated GSAs, strata, functional groups and samples) just so ",
        "you can see how this app works before your own data is ready. Build a real cache ",
        "(see \"Before you start\" above) and reload this page to replace this with your ",
        "actual MEDITS-derived results."
      )
    },
    fluidRow(
    column(
      width = 4,
      sliderInput("year_range", "Ecosim time-series years (TS_YEARS)",
                  min = YEAR_MIN, max = YEAR_MAX,
                  value = c(YEAR_MIN, YEAR_MAX), step = 1, sep = ""),
      checkboxGroupInput("filter_areas", "GSA numbers to include (FILTER_AREAS)",
                          choices = ALL_AREA_IDS, selected = ALL_AREA_IDS, inline = TRUE),
      hr(),
      checkboxInput("drop_outliers", "DROP_OUTLIERS (config block only - see note above)", value = TRUE),
      checkboxInput("normalize_ts", "NORMALIZE_TS (config block only)", value = TRUE),
      checkboxInput("strata_config", "STRATA (config block only - figures always use TRUE)", value = TRUE),
      hr(),
      numericInput("year_ecopath_start", "Ecopath base year, start", value = 1994),
      numericInput("year_ecopath_end", "Ecopath base year, end", value = 1996),
      hr(),
      downloadButton("download_config", "Download config block (.R)")
    ),
    column(
      width = 8,
      leafletOutput("gsa_map", height = 320)
    )
    )
  )
}
```

```{r reactive-pipeline, context="server"}
## context="server" is required here - confirmed the hard way: without it,
## this chunk's reactives (selected_areas, filtered_dt, fg_index, ...) get
## created in Quarto's static-UI-build pass, a DIFFERENT environment from
## the one context="server" chunks (map-render, config-download, and now
## plot-render below) actually run in per session. The visible symptom was
## "Error in selected_areas: could not find function 'selected_areas'"
## from inside output$gsa_map's renderLeaflet() and its observe() block -
## the map/plot code could see cache/CACHE_AVAILABLE (those are plain
## objects from the setup chunk, same environment either way) but not
## these reactive()-wrapped functions, because they simply didn't exist in
## the environment those context="server" chunks execute in.
##
## req(CACHE_AVAILABLE) first in every reactive here - without it, a
## reactive further down the chain (e.g. filtered_dt(), which touches
## cache$dt) would be the first thing to notice the cache is missing,
## with a less clear error than the validate(need(...)) message each
## render*() block below already shows. This just makes the failure
## point explicit and identical everywhere.
selected_areas <- reactive({
  req(CACHE_AVAILABLE, input$filter_areas)
  as.numeric(input$filter_areas)
})

filtered_dt <- reactive({
  cache$dt[Year >= input$year_range[1] & Year <= input$year_range[2] &
             AreaID %in% selected_areas()]
})

per_group_fg <- reactive({
  compute_fg_densities_by_stratum(filtered_dt(), strata = TRUE)
})

n_samples_by_stratum <- reactive({
  filtered_dt()[!is.na(Stratum), .(n_samples = uniqueN(SampleID)), by = .(AreaID, Year, Stratum)]
})

strata_area_selected <- reactive({
  STRATA_AREA_BY_AREA_FULL[AreaID %in% selected_areas()]
})

fg_index <- reactive({
  weight_by_strata(per_group_fg(), n_samples_by_stratum(), strata_area_selected())
})

fg_index_regional <- reactive({
  weight_by_area(fg_index())
})
```

## Sample coverage

```{r plot-sample-map-ui}
## Bare plotOutput() (not a renderPlot() call) - a plain UI placeholder tag,
## the same way you'd declare it in ui.R. This chunk has NO context="server"
## on purpose: it only needs to exist as static HTML, not run per-session -
## the actual renderPlot() logic lives in plot-render below, in
## context="server", where the reactives from reactive-pipeline are
## visible. Splitting UI-placeholder from server-logic this way avoids
## relying on Quarto's implicit "a bare renderPlot() becomes an output"
## behavior, which is exactly what silently failed to update here before.
plotOutput("plot_sample_map")
```

## FG density by area

```{r plot-by-area-ui}
plotOutput("plot_by_area")
```

## FG density, region-wide

```{r plot-regional-ui}
plotOutput("plot_regional")
```

## Density by depth stratum

```{r plot-strata-ui}
plotOutput("plot_strata")
```

```{r plot-render, context="server"}
## All four figures' actual rendering logic, together in one
## context="server" chunk so they share the same live environment as
## reactive-pipeline above (selected_areas(), filtered_dt(), fg_index(),
## fg_index_regional() all need to be reachable here) - each output feeds
## the plotOutput() placeholder of the matching id declared just above its
## own section heading.
output$plot_sample_map <- renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_sample_map(filtered_dt(), cache$area_shp, cache$area_id_col,
                   selected_areas = selected_areas())
})

output$plot_by_area <- renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_by_area(fg_index())
})

output$plot_regional <- renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_fg_timeseries_regional(fg_index_regional())
})

output$plot_strata <- renderPlot({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  plot_strata_profile(fg_index(), cache$strata_def)
})
```

```{r map-render, context="server"}
## context="server" is required here, not optional - this is the actual
## fix for "Error: object 'output' not found" at this chunk. Quarto's
## Shiny engine evaluates every chunk WITHOUT context="server" during an
## initial render pass that only builds the static UI, where `output`
## (and `session`) don't exist yet - that pass is what raised the error.
## reactive()/renderPlot() calls elsewhere in this document survive that
## pass because they just build a promise object to return into the UI;
## an explicit `output$x <- ...` assignment needs `output` to actually
## exist, which only happens inside the live server function - i.e. in a
## context="server" chunk. (config-download below needs the same fix,
## for the same reason - it also assigns into `output`.)
##
## Also depends on reactive-pipeline above ALSO being context="server" -
## without that, selected_areas() below would fail with "could not find
## function 'selected_areas'" (confirmed live) even with this chunk's own
## context="server" fix in place, since a reactive created in a non-server
## chunk isn't visible from a context="server" chunk's environment.
##
## No leafletOutput("gsa_map") even exists in the UI when !CACHE_AVAILABLE
## (see ui-inputs above), so neither block below ever actually runs in
## that case - guarded anyway so this chunk fails the same explicit way
## as everything else if that ever changes, rather than silently.
output$gsa_map <- renderLeaflet({
  validate(need(CACHE_AVAILABLE, CACHE_MISSING_MSG))
  shp <- cache$area_shp
  leaflet(shp) |>
    addTiles() |>
    addPolygons(
      fillColor = ~ifelse(get(cache$area_id_col) %in% isolate(selected_areas()), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(cache$area_id_col))
    )
})

observe({
  req(CACHE_AVAILABLE)
  shp <- cache$area_shp
  leafletProxy("gsa_map", data = shp) |>
    clearShapes() |>
    addPolygons(
      fillColor = ~ifelse(get(cache$area_id_col) %in% selected_areas(), "#0d6efd", "#cccccc"),
      fillOpacity = 0.35, color = "#555555", weight = 1,
      label = ~paste("GSA", get(cache$area_id_col))
    )
})
```

```{r config-download, context="server"}
## context="server" required here too - this chunk assigns into
## output$download_config, same reason as map-render above.
##
## downloadButton("download_config") only exists in the UI when
## CACHE_AVAILABLE (see ui-inputs above), so this is unreachable
## otherwise - req() guard kept for the same reason as map-render.
config_text <- reactive({
  req(CACHE_AVAILABLE)
  paste(
    '## --- generated by "Live Explorer" - 01_biomass.R ---',
    paste0("FILTER_AREAS  <- c(", paste(selected_areas(), collapse = ", "), ")"),
    paste0("STRATA        <- ", if (input$strata_config) "TRUE" else "FALSE"),
    paste0("YEAR_ECOPATH  <- ", input$year_ecopath_start, ":", input$year_ecopath_end),
    paste0("TS_YEARS      <- ", input$year_range[1], ":", input$year_range[2]),
    paste0("DROP_OUTLIERS <- ", if (input$drop_outliers) "TRUE" else "FALSE"),
    paste0("NORMALIZE_TS  <- ", if (input$normalize_ts) "TRUE" else "FALSE"),
    sep = "\n"
  )
})

output$download_config <- downloadHandler(
  filename = function() "biomass_config_block.R",
  content = function(file) writeLines(config_text(), file)
)
```