---
title: "SAU Western Mediterranean Catch: Pipeline Reference & Gear × Species × Region Analysis"
subtitle: "Sea Around Us reconstructed catch, cross-checked against FAO official statistics"
author: "Andrea"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE, error = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(dirname(knitr::current_input(dir = TRUE))))

suppressPackageStartupMessages({
  library(httr)
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(knitr)
})

# ---------------------------------------------------------------------------
# This document is now SELF-CONTAINED: knitting it end-to-end downloads the
# SAU data itself (if it isn't already sitting next to this file) and then
# produces every table and plot below in one pass -- you do not need to run
# a separate script first. The extraction step is cached: once
# sau_raw_combined_west_med.csv and sau_west_med_summary.xlsx exist in this
# folder, re-knitting reuses them instantly instead of re-downloading. Set
# FORCE_REFRESH <- TRUE below (or delete those two files) to pull fresh data.
#
# Needs: internet access to api.seaaroundus.org (required for Part 0/A), and
# the 'fishstat' package (optional, for the FAO comparisons in Part B).
# ---------------------------------------------------------------------------

RAW_CSV        <- "sau_raw_combined_west_med.csv"
SUMMARY_XLSX   <- "sau_west_med_summary.xlsx"
FORCE_REFRESH  <- FALSE   # set TRUE to re-download even if the files above already exist

BASE_URL <- "https://api.seaaroundus.org/api/v1/"
UA <- httr::user_agent("sau-west-med-rmd/1.0")
MEASURE <- "tonnage"

N_TOP_SPECIES  <- 10
YEAR_MIN       <- 1994
YEAR_MAX       <- 2019

TARGET_COUNTRIES <- c("Spain", "France", "Italy", "Tunisia", "Algeria", "Morocco")
TARGET_ISO3 <- c(Spain = "ESP", France = "FRA", Italy = "ITA",
                  Tunisia = "TUN", Algeria = "DZA", Morocco = "MAR")

WEST_MED_EEZS <- list(
  "12"  = list(country = "Algeria", area = "Algeria"),
  "788" = list(country = "Tunisia", area = "Tunisia"),
  "947" = list(country = "Morocco", area = "Morocco (Mediterranean)"),
  "918" = list(country = "France", area = "France (Mediterranean)"),
  "899" = list(country = "France", area = "Corsica"),
  "380" = list(country = "Italy", area = "Italy (mainland)"),
  "902" = list(country = "Italy", area = "Sardinia"),
  "901" = list(country = "Italy", area = "Sicily"),
  "962" = list(country = "Spain", area = "Spain (mainland, Med + Gulf of Cadiz)"),
  "903" = list(country = "Spain", area = "Balearic Islands")
)

FAO_AREA_TITLE_PATTERN <- "mediterranean and black sea"

# ---- shared helpers (same logic as sau_west_med_full.R) --------------------

readr_or_base_read_csv <- function(path) {
  if (requireNamespace("readr", quietly = TRUE)) {
    readr::read_csv(path, show_col_types = FALSE, progress = FALSE)
  } else {
    utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
  }
}

readr_or_base_write_csv <- function(df, path) {
  if (requireNamespace("readr", quietly = TRUE)) {
    readr::write_csv(df, path)
  } else {
    utils::write.csv(df, path, row.names = FALSE)
  }
}

sanitize_raw_types <- function(df) {
  if ("year" %in% names(df)) df$year <- suppressWarnings(as.integer(as.character(df$year)))
  numeric_like <- grep("tonnes|value|catch|landed", names(df), ignore.case = TRUE, value = TRUE)
  for (col in numeric_like) df[[col]] <- suppressWarnings(as.numeric(as.character(df[[col]])))
  for (col in names(df)) {
    if (is.logical(df[[col]]) && all(is.na(df[[col]]))) df[[col]] <- as.character(df[[col]])
  }
  df
}

get_dimension_series <- function(region_type, region_id, dimension, measure = MEASURE,
                                  limit = 200, retries = 3) {
  url <- sprintf("%s%s/%s/%s/?format=json&limit=%d&sciname=false&region_id=%s",
                 BASE_URL, region_type, measure, dimension, limit, region_id)
  resp <- NULL
  for (attempt in seq_len(retries)) {
    resp <- tryCatch(httr::GET(url, UA, httr::timeout(60)), error = function(e) e)
    if (!inherits(resp, "error") && httr::status_code(resp) == 200) break
    if (attempt == retries) {
      message(sprintf("  ! failed %s for %s %s", dimension, region_type, region_id))
      return(tibble())
    }
    Sys.sleep(2 * attempt)
  }
  payload <- tryCatch(
    jsonlite::fromJSON(httr::content(resp, as = "text", encoding = "UTF-8"), simplifyVector = FALSE)$data,
    error = function(e) list()
  )
  rows <- list()
  for (series in payload) {
    key <- series$key
    for (pair in series$values) {
      year <- pair[[1]]; value <- pair[[2]]
      if (is.null(value)) next
      if (year >= YEAR_MIN && year <= YEAR_MAX) {
        rows[[length(rows) + 1]] <- tibble(!!dimension := key, year = as.integer(year), !!measure := as.numeric(value))
      }
    }
  }
  if (length(rows) == 0) return(tibble())
  bind_rows(rows)
}

get_raw_extract <- function(region_type, region_id, retries = 3, timeout_s = 180) {
  url <- sprintf("%s%s/%s/sector/?format=csv&limit=10&sciname=false&region_id=%s",
                 BASE_URL, region_type, MEASURE, region_id)
  resp <- NULL
  for (attempt in seq_len(retries)) {
    resp <- tryCatch(httr::GET(url, UA, httr::timeout(timeout_s)), error = function(e) e)
    if (!inherits(resp, "error") && httr::status_code(resp) == 200) break
    if (attempt == retries) {
      status_msg <- if (inherits(resp, "error")) conditionMessage(resp) else httr::status_code(resp)
      message(sprintf("  ! failed raw extract for %s %s (%s)", region_type, region_id, status_msg))
      return(tibble())
    }
    Sys.sleep(3 * attempt)
  }
  zip_path <- tempfile(fileext = ".zip")
  writeBin(httr::content(resp, as = "raw"), zip_path)
  csv_name <- tryCatch(utils::unzip(zip_path, list = TRUE)$Name[1], error = function(e) NA)
  if (is.na(csv_name)) { unlink(zip_path); return(tibble()) }
  exdir <- tempfile("sau_"); dir.create(exdir)
  utils::unzip(zip_path, files = csv_name, exdir = exdir)
  df <- suppressWarnings(readr_or_base_read_csv(file.path(exdir, csv_name)))
  unlink(zip_path); unlink(exdir, recursive = TRUE)
  sanitize_raw_types(df)
}

find_fao_area <- function() {
  resp <- httr::GET(paste0(BASE_URL, "fao/?format=json"), UA, httr::timeout(120))
  httr::stop_for_status(resp)
  j <- jsonlite::fromJSON(httr::content(resp, as = "text", encoding = "UTF-8"), simplifyVector = FALSE)
  areas <- map_dfr(j$data$features, function(f) tibble(region_id = f$properties$region_id, title = f$properties$title))
  write.csv(areas, "sau_fao_area_lookup.csv", row.names = FALSE)
  match <- areas %>% filter(grepl(FAO_AREA_TITLE_PATTERN, title, ignore.case = TRUE))
  if (nrow(match) == 0) {
    message("Couldn't auto-find an FAO area matching '", FAO_AREA_TITLE_PATTERN, "'.")
    return(NA_character_)
  }
  as.character(match$region_id[1])
}
```

# What this document is

This is a single, self-contained R Markdown report: knitting it (1) pulls
Sea Around Us (SAU) reconstructed catch data for the western
Mediterranean -- Spain, France, Italy, Tunisia, Algeria, Morocco -- broken
down by taxon, gear, sector, country and year, (2) documents exactly what
SAU and FAO expose and which region codes were used and why, and (3)
computes and **renders every table and plot inline** -- the top 10
species by tonnage; gear composition across those species and across
countries; SAU catch over time by sector and by gear, by country; a
country-level (region) SAU vs. FAO comparison standardized for
cross-country trend comparison; a SAU-total vs. SAU-reported-only vs.
FAO time series by region; and a species-level SAU vs. FAO comparison
(all matched top-10 species).

The extraction step (Part 0) is cached to disk (`r RAW_CSV` and
`r SUMMARY_XLSX`): the first knit downloads from SAU's API (needs
internet access to `api.seaaroundus.org`); every knit after that reuses
the cached files and is fast. Delete those two files, or set
`FORCE_REFRESH <- TRUE` in the setup chunk, to pull fresh data.

The FAO comparisons (Part B) additionally need the `fishstat` package
(`install.packages("fishstat")`) -- if it isn't installed, those sections
print a message and are skipped, but Part 0/A (the SAU side, which
matters most) is unaffected either way.

# The two data sources

## Sea Around Us (SAU)

SAU is a *reconstructed* catch database (it adjusts official landings for
unreported, discarded, and illegal catch -- deliberately not the same
thing as official reported statistics). Everything comes from one
JSON/CSV API:

```
https://api.seaaroundus.org/api/v1/
```

The API is organized around **region types** (the geographic unit you're
querying) and **dimensions** (how you want catch within that region
broken down), combined in the URL path:

```
{BASE_URL}{region_type}/{measure}/{dimension}/?format=json&region_id={id}
```

### Region types

| region_type | meaning | region_id used in this document |
|---|---|---|
| `eez` | Exclusive Economic Zone (a country, or a sub-unit when a country has more than one EEZ) | `962` Spain (mainland, Med + Gulf of Cadiz), `903` Balearic Islands, `918` France (Mediterranean), `899` Corsica, `380`/`902`/`901` Italy mainland/Sardinia/Sicily, `788` Tunisia, `12` Algeria, `947` Morocco (Mediterranean) |
| `fao` | FAO Major Fishing Area | `37` = Mediterranean and Black Sea (the only Med-related one -- see below) |
| `lme`, `me`, `rfmo`, `highseas`, `taxon`, `country`, `global` | other region types SAU exposes | not used here |

The live, current list of FAO areas SAU has is written to
`sau_fao_area_lookup.csv` every time Part 0 runs -- that file is the
source of truth, not the table above. There's no equivalent EEZ dump
yet; if you ever want the full EEZ list (e.g. to extend this to Greece,
Croatia, Libya, Egypt, Turkey), a one-line addition
(`GET {BASE_URL}eez/?format=json`) gets it.

### Dimensions

| dimension | what it is |
|---|---|
| `sector` | Industrial / Artisanal / Subsistence / Recreational |
| `gear` | trawl, purse seine, longline, gillnet, etc. |
| `taxon` | species/genus/family caught |
| `country` | which country's fleet caught it (meaningful inside a non-country region like an FAO area) |
| `commercial-group`, `functional-group`, `catch-type`, `reporting-status` | other available breakdowns, not used here |

**Measures:** `tonnage` (used throughout) or `value` (landed value).

### Two ways of querying

1. **Dimension/JSON aggregation** (`get_dimension_series()`) -- one
   dimension at a time, small and always reliable. Used for the
   FAO-Area-37 country totals.
2. **Raw row-level CSV extract** (`get_raw_extract()`) -- the same URL
   pattern with `format=csv`, returning a ZIP with every dimension as a
   column on one row (sector, gear, taxon, country, area, year, tonnes,
   landed value...). This builds the full cross-tab, but only reliably at
   EEZ scale -- the whole of FAO Area 37 (~20 countries × all years × all
   taxa) is too large and the endpoint fails, which is why extraction is
   done per-EEZ and concatenated rather than as one big query.

## FAO (via the `fishstat` R package)

`fishstat` (CRAN, mirrors github.com/sofia-taf/fishstat) ships FAO's
official Capture Production statistics as ready-to-use R data frames.

| table | key columns | purpose |
|---|---|---|
| `capture` | `species, year, area, country, value, measure, status` | the catch records |
| `country` | `country, country_name, iso3, ...` | country code lookup |
| `area` | `area, area_name, ...` | FAO area code lookup |
| `measure` | `measure, measure_name, unit, ...` | tonnes vs. number of individuals |
| `species` | `species, species_name, scientific, isscaap, major, ...` | species code lookup |

FAO's own classification only goes down to **FAO Major Fishing Area**
(Area 37 = "Mediterranean and Black Sea") -- there is no official FAO
subdivision into western/central/eastern Mediterranean.

# Why the raw extract is per-EEZ, and the Spain / Gulf of Cadiz problem

SAU's EEZ list doesn't split Spain's Mediterranean coast from its Gulf of
Cadiz (Atlantic) coast -- both sit inside EEZ `962`, and there's no
column in the raw extract that separates them (confirmed by inspecting
`area_name`/`area_type`, which are constant per EEZ). So Spain's numbers
in the EEZ-based cross-tab run slightly high.

FAO Major Area 37 excludes the Gulf of Cadiz by construction (that's FAO
Area 27), so querying SAU's own `fao` region type for Area 37's `country`
dimension gives a clean, Gulf-of-Cadiz-free country-year total -- at the
cost of no gear/taxon breakdown (that raw extract is too large to
retrieve, as noted above). Part 0 writes both totals side by side:
`ByCountryYear_EEZbased` (full detail, minor Spain-only scope leak) and
`ByCountryYear_FAOarea37` (clean, coarse).

**Italy caveat, the other direction:** Area 37 covers Italy's *entire*
Mediterranean + Black Sea coast, including the Adriatic and Ionian -- not
just the western basin. Algeria, Tunisia, Morocco, France, and Spain have
no Mediterranean coast outside the western basin, so Area 37 happens to
equal "western Mediterranean" for them. For Italy it doesn't -- its
Area-37 (and FAO comparison) figures legitimately run higher than a true
western-Med-only number would.

# Known limitations

- SAU's public reconstruction currently runs 1950-2019; FAO's `fishstat`
  runs through 2024. All comparisons here are capped at 2019.
- SAU is a *reconstruction*; FAO is *officially reported* landings -- a
  gap between them is not automatically an error in either one.
- Species matching between SAU (`scientific_name`) and FAO
  (`fishstat::species$scientific`) is done by case-insensitive,
  whitespace-trimmed string match, since the two databases share no
  common species code. Unmatched species are reported explicitly below,
  never silently dropped.
- Cite both sources: Sea Around Us
  ([seaaroundus.org/citation-policy](https://www.seaaroundus.org/citation-policy/))
  and FAO FishStat via
  [github.com/sofia-taf/fishstat](https://github.com/sofia-taf/fishstat).

---

# Part 0: Data extraction (cached)

```{r part0-extract}
need_extract <- FORCE_REFRESH || !file.exists(RAW_CSV) || !file.exists(SUMMARY_XLSX)

if (!need_extract) {

  message("Using cached ", RAW_CSV, " and ", SUMMARY_XLSX,
          " -- delete them (or set FORCE_REFRESH <- TRUE) to re-download.")
  extraction_ok <- TRUE

} else {

  extraction_ok <- tryCatch({

    message("Fetching raw per-EEZ extracts from SAU (this can take a minute)...")
    raw_frames <- list()
    for (region_id in names(WEST_MED_EEZS)) {
      meta <- WEST_MED_EEZS[[region_id]]
      message(sprintf("  EEZ %s (%s)", region_id, meta$area))
      df <- get_raw_extract("eez", region_id)
      if (nrow(df) > 0) { df$country <- meta$country; df$area <- meta$area; raw_frames[[length(raw_frames) + 1]] <- df }
    }
    raw_combined <- if (length(raw_frames) > 0) bind_rows(raw_frames) else tibble()
    if (nrow(raw_combined) == 0) stop("All per-EEZ raw extracts failed -- check internet access to api.seaaroundus.org and retry.")

    readr_or_base_write_csv(raw_combined, RAW_CSV)
    message(sprintf("Saved %s (%d rows).", RAW_CSV, nrow(raw_combined)))

    if ("year" %in% names(raw_combined)) raw_combined <- raw_combined %>% filter(year >= YEAR_MIN, year <= YEAR_MAX)

    candidates <- list(sector = c("fishing_sector", "sector"), gear = c("gear_type", "gear"),
                       taxon = c("common_name", "taxon_name", "scientific_name"), tonnes = c("tonnes", "catch_sum", "value"))
    resolved <- list()
    for (field in names(candidates)) for (opt in candidates[[field]]) if (opt %in% names(raw_combined)) { resolved[[field]] <- opt; break }

    cross_tab <- if (length(resolved) == 4) {
      raw_combined %>%
        rename(.sector = !!resolved$sector, .gear = !!resolved$gear, .taxon = !!resolved$taxon, .tonnes = !!resolved$tonnes) %>%
        group_by(country, area, year, sector = .sector, gear = .gear, taxon = .taxon) %>%
        summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")
    } else tibble()

    by_gear <- if (!is.null(resolved$gear)) raw_combined %>% rename(.gear = !!resolved$gear, .tonnes = !!resolved$tonnes) %>%
      group_by(country, year, gear = .gear) %>% summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") else tibble()
    by_taxon <- if (!is.null(resolved$taxon)) raw_combined %>% rename(.taxon = !!resolved$taxon, .tonnes = !!resolved$tonnes) %>%
      group_by(country, year, taxon = .taxon) %>% summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") else tibble()
    by_sector <- if (!is.null(resolved$sector)) raw_combined %>% rename(.sector = !!resolved$sector, .tonnes = !!resolved$tonnes) %>%
      group_by(country, year, sector = .sector) %>% summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") else tibble()
    by_country_year_eez <- raw_combined %>% rename(.tonnes = !!resolved$tonnes) %>%
      group_by(country, year) %>% summarise(tonnage = sum(.tonnes, na.rm = TRUE), .groups = "drop")

    message("Looking up SAU's FAO area list...")
    fao_area_id <- find_fao_area()
    by_country_year_fao <- tibble()
    if (!is.na(fao_area_id)) {
      message("Fetching the 'country' dimension for FAO Area 37 (Mediterranean and Black Sea)...")
      country_dim <- get_dimension_series("fao", fao_area_id, "country")
      if (nrow(country_dim) > 0) by_country_year_fao <- country_dim %>% filter(country %in% TARGET_COUNTRIES)
    }

    wb <- openxlsx::createWorkbook()
    add_sheet <- function(name, df) if (!is.null(df) && nrow(df) > 0) { openxlsx::addWorksheet(wb, name); openxlsx::writeData(wb, name, df) }
    add_sheet("ByCountryYear_EEZbased", by_country_year_eez)
    add_sheet("ByCountryYear_FAOarea37", by_country_year_fao)
    add_sheet("ByCountrySectorYear", by_sector)
    add_sheet("ByCountryGearYear", by_gear)
    add_sheet("ByCountryTaxonYear", by_taxon)
    add_sheet("FullCrossTab", cross_tab)
    openxlsx::saveWorkbook(wb, SUMMARY_XLSX, overwrite = TRUE)
    message(sprintf("Wrote %s.", SUMMARY_XLSX))

    TRUE
  }, error = function(e) {
    message("\n! Extraction failed: ", conditionMessage(e))
    message("  The rest of this document will show what it can from any existing cached files.")
    FALSE
  })
}

have_raw     <- file.exists(RAW_CSV)
have_summary <- file.exists(SUMMARY_XLSX)
```

---

# Part A: Top 10 species and gear × species × region {.tabset}

```{r part-a-prep}
raw <- readr_or_base_read_csv(RAW_CSV)

candidates <- list(
  gear     = c("gear_type", "gear"),
  sci_name = c("scientific_name"),
  com_name = c("common_name"),
  tonnes   = c("tonnes", "catch_sum", "value"),
  sector   = c("fishing_sector", "sector"),
  report   = c("reporting_status")
)
resolved <- list()
for (field in names(candidates)) {
  for (opt in candidates[[field]]) {
    if (opt %in% names(raw)) { resolved[[field]] <- opt; break }
  }
}
missing_cols <- setdiff(c("gear", "sci_name", "tonnes"), names(resolved))
if (length(missing_cols) > 0) {
  stop("Raw extract is missing required column(s): ", paste(missing_cols, collapse = ", "))
}
if (is.null(resolved$com_name)) resolved$com_name <- resolved$sci_name

have_sector_col <- !is.null(resolved$sector)
have_report_col <- !is.null(resolved$report)
if (!have_sector_col) message("No sector column found in the raw extract -- the by-sector time series below will be skipped.")
if (!have_report_col) message("No reporting-status column found in the raw extract -- the SAU-total-vs-reported-only comparison below will be skipped.")

raw <- raw %>%
  rename(.gear = !!resolved$gear, .sci = !!resolved$sci_name,
         .com = !!resolved$com_name, .tonnes = !!resolved$tonnes) %>%
  filter(!is.na(year), year >= YEAR_MIN, year <= YEAR_MAX)

if (have_sector_col) raw <- raw %>% rename(.sector = !!resolved$sector)
if (have_report_col) raw <- raw %>% rename(.report = !!resolved$report)

cat(sprintf("Loaded %d rows, %d-%d.\n", nrow(raw), YEAR_MIN, YEAR_MAX))
```

## Top 10 species

```{r part-a-top10}
species_totals <- raw %>%
  group_by(scientific_name = .sci, common_name = .com) %>%
  summarise(total_tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_tonnes))

top10 <- species_totals %>% slice_head(n = N_TOP_SPECIES) %>%
  mutate(rank = row_number()) %>%
  dplyr::select(rank, scientific_name, common_name, total_tonnes)

kable(top10, digits = 0, caption = paste0("Top ", N_TOP_SPECIES,
      " species by total tonnes, all 6 countries, ", YEAR_MIN, "-", YEAR_MAX))
```

```{r part-a-top10-plot, fig.width=8, fig.height=5}
barplot(rev(top10$total_tonnes), names.arg = rev(top10$common_name),
        horiz = TRUE, las = 1, col = "steelblue",
        xlab = "Total tonnes (1994-2019)", cex.names = 0.8,
        main = "Top 10 species by total catch")
```

## Gear × species × country (summed across years)

```{r part-a-gear-species}
raw_top10 <- raw %>% filter(.sci %in% top10$scientific_name)

gear_species_country_year <- raw_top10 %>%
  group_by(country, year, gear = .gear, scientific_name = .sci, common_name = .com) %>%
  summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  left_join(top10 %>% dplyr::select(scientific_name, rank), by = "scientific_name") %>%
  arrange(rank, country, year, desc(tonnes))

gear_species_country_allyears <- gear_species_country_year %>%
  group_by(country, gear, scientific_name, common_name, rank) %>%
  summarise(tonnes = sum(tonnes, na.rm = TRUE), .groups = "drop") %>%
  arrange(rank, country, desc(tonnes))

kable(head(gear_species_country_allyears, 40), digits = 0,
      caption = "Gear x top-10-species x country, tonnes summed 1994-2019 (first 40 rows -- full table in the .xlsx export below)")
```

## Gear × species, compared across regions (plots)

Two views of the same table: a heatmap (which gear catches which of the
top species, one panel per country) and a set of within-country bar
charts (gear composition for each top species, one panel per country) so
you can see how the same 10 species are fished differently from country
to country.

```{r part-a-gear-species-heatmap, fig.width=10, fig.height=8}
have_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

# Keep the plots readable: collapse to the top gears overall, lump the rest
# into "Other gear" (a long tail of rare gear/species combinations otherwise
# swamps the palette and the axis labels).
N_TOP_GEARS <- 8
top_gears <- gear_species_country_allyears %>%
  group_by(gear) %>% summarise(g_tot = sum(tonnes, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(g_tot)) %>% slice_head(n = N_TOP_GEARS) %>% pull(gear)

plot_df <- gear_species_country_allyears %>%
  mutate(gear_grp = ifelse(gear %in% top_gears, gear, "Other gear")) %>%
  group_by(country, gear_grp, scientific_name, common_name, rank) %>%
  summarise(tonnes = sum(tonnes, na.rm = TRUE), .groups = "drop") %>%
  mutate(species_label = sprintf("%02d. %s", rank, common_name))

if (have_ggplot2) {
  library(ggplot2)
  ggplot(plot_df, aes(x = gear_grp, y = reorder(species_label, -rank), fill = tonnes)) +
    geom_tile() +
    facet_wrap(~country) +
    scale_fill_viridis_c(name = "Tonnes", trans = "sqrt", option = "magma") +
    labs(title = "Gear x top-10-species tonnage, by country (region)",
         x = "Gear", y = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text = element_text(face = "bold"))
} else {
  message("Package 'ggplot2' not installed -- showing a simpler base-R heatmap instead ",
          "(install.packages('ggplot2') for the faceted-by-country version).")
  agg <- plot_df %>% group_by(gear_grp, species_label) %>%
    summarise(tonnes = sum(tonnes, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = gear_grp, values_from = tonnes, values_fill = 0)
  mat <- as.matrix(agg[,-1]); rownames(mat) <- agg$species_label
  image(x = seq_len(ncol(mat)), y = seq_len(nrow(mat)), z = t(sqrt(mat)),
        axes = FALSE, col = hcl.colors(20, "YlOrRd", rev = TRUE),
        xlab = "", ylab = "", main = "Gear x top-10-species tonnage (all countries combined)")
  axis(1, at = seq_len(ncol(mat)), labels = colnames(mat), las = 2, cex.axis = 0.7)
  axis(2, at = seq_len(nrow(mat)), labels = rownames(mat), las = 1, cex.axis = 0.7)
}
```

```{r part-a-gear-species-bars, fig.width=10, fig.height=8}
if (have_ggplot2) {
  ggplot(plot_df, aes(x = reorder(species_label, rank), y = tonnes, fill = gear_grp)) +
    geom_col(position = "fill") +
    coord_flip() +
    facet_wrap(~country) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_viridis_d(name = "Gear", option = "turbo") +
    labs(title = "Gear composition of each top-10 species, by country (region)",
         x = NULL, y = "Share of that species' catch") +
    theme_minimal(base_size = 10) +
    theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
} else {
  countries_here <- sort(unique(plot_df$country))
  ncol_grid <- ceiling(sqrt(length(countries_here))); nrow_grid <- ceiling(length(countries_here) / ncol_grid)
  par(mfrow = c(nrow_grid, ncol_grid), mar = c(4, 8, 3, 1))
  gear_levels <- sort(unique(plot_df$gear_grp))
  cols <- hcl.colors(length(gear_levels), "Dark 3")
  for (cty in countries_here) {
    d <- plot_df %>% filter(country == cty) %>%
      tidyr::pivot_wider(id_cols = species_label, names_from = gear_grp, values_from = tonnes,
                          values_fill = 0) %>% arrange(species_label)
    mat <- t(as.matrix(d[,-1])); colnames(mat) <- d$species_label
    mat <- prop.table(mat, margin = 2)
    barplot(mat, horiz = TRUE, las = 1, col = cols[match(rownames(mat), gear_levels)],
            main = cty, cex.names = 0.7, xlab = "Share of species' catch")
  }
  plot.new()
  legend("center", legend = gear_levels, fill = cols, bty = "n", cex = 0.8, title = "Gear")
}
```

## Catch over time by sector and by gear, by region

These use the FULL raw extract (all species, not just the top 10) so
they show the whole fishery's composition, not just the headline
species. Both are time series, faceted by country (region).

```{r part-a-sector-ts, fig.width=10, fig.height=8, eval=have_sector_col}
sector_ts <- raw %>%
  group_by(country, year, sector = .sector) %>%
  summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")

if (have_ggplot2) {
  ggplot(sector_ts, aes(x = year, y = tonnes, color = sector)) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~country, scales = "free_y") +
    scale_color_viridis_d(name = "Sector", option = "turbo") +
    labs(title = "SAU catch by sector over time, by country (region)",
         x = "Year", y = "Tonnes") +
    theme_minimal(base_size = 10) +
    theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
} else {
  countries_here_s <- sort(unique(sector_ts$country))
  ncol_grid_s <- ceiling(sqrt(length(countries_here_s))); nrow_grid_s <- ceiling(length(countries_here_s) / ncol_grid_s)
  sector_levels <- sort(unique(sector_ts$sector))
  cols_s <- hcl.colors(length(sector_levels), "Dark 3")
  par(mfrow = c(nrow_grid_s, ncol_grid_s), mar = c(4, 4, 3, 1))
  for (cty in countries_here_s) {
    d <- sector_ts %>% filter(country == cty)
    plot(range(d$year), range(d$tonnes, na.rm = TRUE), type = "n", xlab = "Year", ylab = "Tonnes", main = cty)
    for (sec in sector_levels) {
      dd <- d %>% filter(sector == sec) %>% arrange(year)
      lines(dd$year, dd$tonnes, col = cols_s[match(sec, sector_levels)], lwd = 2)
    }
  }
  plot.new(); legend("center", legend = sector_levels, col = cols_s, lwd = 2, bty = "n", cex = 0.8, title = "Sector")
  mtext("SAU catch by sector over time, by country (region)", outer = TRUE, line = -1.2, cex = 1)
}
```

```{r part-a-gear-ts, fig.width=10, fig.height=8}
gear_ts_top <- raw %>%
  group_by(gear = .gear) %>% summarise(g_tot = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(g_tot)) %>% slice_head(n = 8) %>% pull(gear)

gear_ts <- raw %>%
  mutate(gear_grp = ifelse(.gear %in% gear_ts_top, .gear, "Other gear")) %>%
  group_by(country, year, gear_grp) %>%
  summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")

if (have_ggplot2) {
  ggplot(gear_ts, aes(x = year, y = tonnes, color = gear_grp)) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~country, scales = "free_y") +
    scale_color_viridis_d(name = "Gear", option = "turbo") +
    labs(title = "SAU catch by gear over time, by country (region)",
         x = "Year", y = "Tonnes") +
    theme_minimal(base_size = 10) +
    theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
} else {
  countries_here_g <- sort(unique(gear_ts$country))
  ncol_grid_g <- ceiling(sqrt(length(countries_here_g))); nrow_grid_g <- ceiling(length(countries_here_g) / ncol_grid_g)
  gear_levels_g <- sort(unique(gear_ts$gear_grp))
  cols_g <- hcl.colors(length(gear_levels_g), "Dark 3")
  par(mfrow = c(nrow_grid_g, ncol_grid_g), mar = c(4, 4, 3, 1))
  for (cty in countries_here_g) {
    d <- gear_ts %>% filter(country == cty)
    plot(range(d$year), range(d$tonnes, na.rm = TRUE), type = "n", xlab = "Year", ylab = "Tonnes", main = cty)
    for (g in gear_levels_g) {
      dd <- d %>% filter(gear_grp == g) %>% arrange(year)
      lines(dd$year, dd$tonnes, col = cols_g[match(g, gear_levels_g)], lwd = 2)
    }
  }
  plot.new(); legend("center", legend = gear_levels_g, col = cols_g, lwd = 2, bty = "n", cex = 0.8, title = "Gear")
  mtext("SAU catch by gear over time, by country (region)", outer = TRUE, line = -1.2, cex = 1)
}
```

## Full sector and gear summary, by country (all sectors, all gears)

Unlike the time-series plots above (which lump rare gears into "Other
gear" for readability), these tables cover **every** sector and **every**
gear that appears in the data -- nothing truncated to a top-N list.

```{r part-a-sector-summary, eval=have_sector_col}
sector_summary <- raw %>%
  group_by(country, sector = .sector) %>%
  summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  left_join(raw %>% group_by(country) %>% summarise(country_total = sum(.tonnes, na.rm = TRUE), .groups = "drop"),
            by = "country") %>%
  mutate(pct_of_country_catch = 100 * tonnes / country_total) %>%
  dplyr::select(-country_total) %>%
  arrange(country, desc(tonnes))

kable(sector_summary, digits = 1,
      caption = paste0("Every sector's tonnage and % share of each country's total catch, ", YEAR_MIN, "-", YEAR_MAX))
```

```{r part-a-sector-summary-plot, eval=have_sector_col, fig.width=9, fig.height=5}
if (have_ggplot2) {
  ggplot(sector_summary, aes(x = country, y = pct_of_country_catch, fill = sector)) +
    geom_col(position = "stack") +
    scale_fill_viridis_d(name = "Sector", option = "turbo") +
    labs(title = "Sector share of total catch, by country (region)",
         x = NULL, y = "% of country's total catch") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
} else {
  wide <- sector_summary %>% tidyr::pivot_wider(id_cols = country, names_from = sector,
                                                 values_from = pct_of_country_catch, values_fill = 0)
  mat <- t(as.matrix(wide[,-1])); colnames(mat) <- wide$country
  barplot(mat, col = hcl.colors(nrow(mat), "Dark 3"), legend.text = rownames(mat),
          args.legend = list(x = "topright", cex = 0.7, bty = "n"),
          ylab = "% of country's total catch", main = "Sector share of total catch, by country")
}
```

```{r part-a-gear-summary-full}
gear_summary_full <- raw %>%
  group_by(country, gear = .gear) %>%
  summarise(tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  left_join(raw %>% group_by(country) %>% summarise(country_total = sum(.tonnes, na.rm = TRUE), .groups = "drop"),
            by = "country") %>%
  mutate(pct_of_country_catch = 100 * tonnes / country_total) %>%
  dplyr::select(-country_total) %>%
  arrange(country, desc(tonnes))

kable(head(gear_summary_full, 40), digits = 1,
      caption = paste0("Every gear's tonnage and % share of each country's total catch, ", YEAR_MIN, "-", YEAR_MAX,
                        " (first 40 rows -- full table in the .xlsx export below)"))
```

## Each top species' share of its country's total catch

```{r part-a-share}
country_totals <- raw %>%
  group_by(country) %>%
  summarise(country_total_tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")

species_country_share <- raw_top10 %>%
  group_by(country, scientific_name = .sci, common_name = .com) %>%
  summarise(species_tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop") %>%
  left_join(country_totals, by = "country") %>%
  mutate(pct_of_country_catch = 100 * species_tonnes / country_total_tonnes) %>%
  left_join(top10 %>% dplyr::select(scientific_name, rank), by = "scientific_name") %>%
  arrange(rank, desc(pct_of_country_catch))

kable(species_country_share, digits = 1,
      caption = "Top-10 species' share (%) of each country's total catch")
```

## Export to Excel

```{r part-a-export}
if (requireNamespace("openxlsx", quietly = TRUE)) {
  xlsx_path <- "sau_gear_species_region.xlsx"
  wb <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(wb, "Top10Species"); openxlsx::writeData(wb, "Top10Species", top10)
  openxlsx::addWorksheet(wb, "GearBySpeciesByCountry"); openxlsx::writeData(wb, "GearBySpeciesByCountry", gear_species_country_year)
  openxlsx::addWorksheet(wb, "GearBySpeciesByCountry_AllYears"); openxlsx::writeData(wb, "GearBySpeciesByCountry_AllYears", gear_species_country_allyears)
  openxlsx::addWorksheet(wb, "SpeciesByCountry_Share"); openxlsx::writeData(wb, "SpeciesByCountry_Share", species_country_share)
  if (have_sector_col) { openxlsx::addWorksheet(wb, "SectorByCountry_Full"); openxlsx::writeData(wb, "SectorByCountry_Full", sector_summary) }
  openxlsx::addWorksheet(wb, "GearByCountry_Full"); openxlsx::writeData(wb, "GearByCountry_Full", gear_summary_full)
  openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)
  cat("Wrote", xlsx_path, "\n")
} else {
  cat("Package 'openxlsx' not installed -- skipping the .xlsx export (tables above are still in this report).\n")
}
```

---

# Part B: SAU vs. FAO comparison, by region and by species (best-effort) {.tabset}

Two comparisons, both best-effort (they're skipped gracefully if
`fishstat` isn't installed, or Part 0's `ByCountryYear_FAOarea37` sheet
came back empty):

1. **By region (country)** -- standardized, so trend shape is comparable
   across countries of very different catch size.
2. **SAU total vs. SAU reported-only vs. FAO, by region** -- since FAO
   only captures reported landings while SAU's headline number also adds
   back unreported/IUU catch, this checks whether SAU's *reported*
   component alone tracks FAO more closely than SAU's total does.
3. **By species** -- the same idea pushed down to all matched top-10
   species (not a subset), using `fishstat::capture` filtered to FAO
   Area 37. FAO does not report catch by gear at all, so the
   species-level comparison stops at species × country × year -- it
   cannot be extended to gear the way Part A's SAU-only analysis can.

```{r part-b-check}
have_fishstat <- requireNamespace("fishstat", quietly = TRUE)
have_openxlsx <- requireNamespace("openxlsx", quietly = TRUE)
have_summary_xlsx <- have_openxlsx && have_summary
run_part_b <- have_fishstat
run_part_b_region <- run_part_b && have_summary_xlsx
if (!have_fishstat) {
  message("Package 'fishstat' isn't installed -- run install.packages('fishstat') ",
          "and re-knit this document for the FAO comparisons below.")
}
if (have_fishstat && !have_summary_xlsx) {
  message("Can't find a usable ", SUMMARY_XLSX, " -- the country/region-level standardized ",
          "comparison is skipped, but the species-level comparison further down does not need it.")
}
```

## Country-level (region) comparison, standardized

This is the comparison "by region" -- SAU vs FAO totals for each of the 6
countries, both bounded to FAO Major Area 37 so the two sides are
apples-to-apples (see "Why the raw extract is per-EEZ..." above). Two
views:

- **Raw tonnes** -- shows actual magnitude and where SAU and FAO
  diverge the most in absolute terms.
- **Standardized (indexed to 100 in the first shared year)** -- rescales
  both series so trends/shapes are comparable across countries whose
  catch sizes differ by an order of magnitude (e.g. Algeria vs. Italy);
  useful for asking "do SAU and FAO agree on the *trend*, even if not on
  the *level*?".

```{r part-b-region-prep, eval=run_part_b_region}
suppressPackageStartupMessages(library(fishstat))

med_area_ids <- (area %>% filter(grepl("mediterranean", area_name, ignore.case = TRUE)))$area
target_country_codes <- (country %>% filter(iso3 %in% TARGET_ISO3))$country
tonnes_measure_codes <- (measure %>% filter(grepl("ton", measure_name, ignore.case = TRUE) |
                                               unit %in% c("t", "tonnes")))$measure

sau_country_fao <- openxlsx::read.xlsx(SUMMARY_XLSX, sheet = "ByCountryYear_FAOarea37") %>%
  rename(sau_tonnes = tonnage) %>%
  mutate(year = as.integer(year))

fao_country <- capture %>%
  filter(area %in% med_area_ids,
         country %in% target_country_codes,
         measure %in% tonnes_measure_codes,
         year >= YEAR_MIN, year <= YEAR_MAX) %>%
  left_join(country %>% dplyr::select(country, country_name), by = "country") %>%
  group_by(country = country_name, year) %>%
  summarise(fao_tonnes = sum(value, na.rm = TRUE), .groups = "drop")

region_comparison <- full_join(fao_country, sau_country_fao, by = c("country", "year")) %>%
  arrange(country, year) %>%
  group_by(country) %>%
  mutate(
    diff = sau_tonnes - fao_tonnes,
    pct_diff = ifelse(!is.na(fao_tonnes) & fao_tonnes != 0, 100 * diff / fao_tonnes, NA),
    # standardize: index each series to 100 at the first year BOTH sides have data
    first_common_year = suppressWarnings(min(year[!is.na(fao_tonnes) & !is.na(sau_tonnes)])),
    fao_base = fao_tonnes[year == first_common_year][1],
    sau_base = sau_tonnes[year == first_common_year][1],
    fao_index = 100 * fao_tonnes / fao_base,
    sau_index = 100 * sau_tonnes / sau_base
  ) %>%
  ungroup()

kable(region_comparison %>%
        group_by(country) %>%
        summarise(n_years = sum(!is.na(fao_tonnes) & !is.na(sau_tonnes)),
                  mean_fao_tonnes = mean(fao_tonnes, na.rm = TRUE),
                  mean_sau_tonnes = mean(sau_tonnes, na.rm = TRUE),
                  mean_pct_diff = mean(pct_diff, na.rm = TRUE),
                  .groups = "drop"),
      digits = 1, caption = "SAU vs FAO by country/region, FAO Area 37, raw tonnes")
```

```{r part-b-region-plot-raw, eval=run_part_b_region, fig.width=10, fig.height=7}
countries_here <- sort(unique(region_comparison$country))
ncol_grid <- ceiling(sqrt(length(countries_here))); nrow_grid <- ceiling(length(countries_here) / ncol_grid)
par(mfrow = c(nrow_grid, ncol_grid), mar = c(4, 4, 3, 1))
for (cty in countries_here) {
  d <- region_comparison %>% filter(country == cty) %>% arrange(year)
  yrange <- suppressWarnings(range(c(d$fao_tonnes, d$sau_tonnes), na.rm = TRUE))
  plot(d$year, d$fao_tonnes, type = "l", col = "steelblue", lwd = 2,
       ylim = yrange, xlab = "Year", ylab = "Tonnes", main = cty)
  lines(d$year, d$sau_tonnes, col = "firebrick", lwd = 2, lty = 2)
  legend("topleft", legend = c("FAO (Area 37)", "SAU (Area 37)"),
         col = c("steelblue", "firebrick"), lty = c(1, 2), lwd = 2, bty = "n", cex = 0.7)
}
mtext("SAU vs FAO by country (region) -- raw tonnes, FAO Major Area 37", outer = TRUE, line = -1.2, cex = 1)
```

```{r part-b-region-plot-index, eval=run_part_b_region, fig.width=10, fig.height=7}
par(mfrow = c(nrow_grid, ncol_grid), mar = c(4, 4, 3, 1))
for (cty in countries_here) {
  d <- region_comparison %>% filter(country == cty) %>% arrange(year)
  yrange <- suppressWarnings(range(c(d$fao_index, d$sau_index), na.rm = TRUE))
  plot(d$year, d$fao_index, type = "l", col = "steelblue", lwd = 2,
       ylim = yrange, xlab = "Year", ylab = sprintf("Index (100 = %d)", d$first_common_year[1]),
       main = cty)
  lines(d$year, d$sau_index, col = "firebrick", lwd = 2, lty = 2)
  abline(h = 100, col = "grey70", lty = 3)
  legend("topleft", legend = c("FAO", "SAU"), col = c("steelblue", "firebrick"),
         lty = c(1, 2), lwd = 2, bty = "n", cex = 0.7)
}
mtext("SAU vs FAO by country (region) -- standardized (indexed) trend comparison", outer = TRUE, line = -1.2, cex = 1)
```

## SAU total vs. SAU reported-only vs. FAO, by region (time series)

FAO's official statistics only capture *reported* landings, while SAU's
headline number is a full reconstruction that also adds back unreported,
discarded, and illegal (IUU) catch -- so SAU's total is expected to run
above FAO by construction, not because either side is wrong. This plot
adds a third line, **SAU reported-only** (the subset of the raw extract
whose `reporting_status` is "Reported"), so you can see whether SAU's
*reported* component tracks FAO more closely than SAU's *total* does.

Note this uses the EEZ-based raw extract (carrying the small Spain/Gulf
of Cadiz scope leak described earlier), not the FAO-Area-37 query used
above -- reporting-status detail only exists in the raw per-EEZ extract.

```{r part-b-reported-ts, eval=have_report_col && run_part_b_region, fig.width=10, fig.height=7}
sau_reported_ts <- raw %>%
  group_by(country, year) %>%
  summarise(
    sau_total_tonnes = sum(.tonnes, na.rm = TRUE),
    sau_reported_tonnes = sum(.tonnes[grepl("^report", .report, ignore.case = TRUE) &
                                         !grepl("unreport", .report, ignore.case = TRUE)], na.rm = TRUE),
    .groups = "drop"
  )

reported_comparison <- sau_reported_ts %>%
  left_join(fao_country, by = c("country", "year")) %>%
  arrange(country, year)

countries_here_r <- sort(unique(reported_comparison$country))
ncol_grid_r <- ceiling(sqrt(length(countries_here_r))); nrow_grid_r <- ceiling(length(countries_here_r) / ncol_grid_r)
par(mfrow = c(nrow_grid_r, ncol_grid_r), mar = c(4, 4, 3, 1))
for (cty in countries_here_r) {
  d <- reported_comparison %>% filter(country == cty) %>% arrange(year)
  yrange <- suppressWarnings(range(c(d$sau_total_tonnes, d$sau_reported_tonnes, d$fao_tonnes), na.rm = TRUE))
  plot(d$year, d$sau_total_tonnes, type = "l", col = "firebrick", lwd = 2,
       ylim = yrange, xlab = "Year", ylab = "Tonnes", main = cty)
  lines(d$year, d$sau_reported_tonnes, col = "darkorange", lwd = 2, lty = 2)
  lines(d$year, d$fao_tonnes, col = "steelblue", lwd = 2, lty = 3)
  legend("topleft", legend = c("SAU total (EEZ-based)", "SAU reported-only (EEZ-based)", "FAO (Area 37)"),
         col = c("firebrick", "darkorange", "steelblue"), lty = c(1, 2, 3), lwd = 2, bty = "n", cex = 0.65)
}
mtext("SAU total vs SAU reported-only vs FAO, by country (region)", outer = TRUE, line = -1.2, cex = 1)
```

## Matching SAU species to FAO species codes

```{r part-b-match, eval=run_part_b}
suppressPackageStartupMessages(library(fishstat))

med_area <- area %>% filter(grepl("mediterranean", area_name, ignore.case = TRUE))
tonnes_measures <- measure %>% filter(grepl("ton", measure_name, ignore.case = TRUE) |
                                         unit %in% c("t", "tonnes"))
target_countries <- country %>% filter(iso3 %in% TARGET_ISO3)

norm <- function(x) trimws(tolower(x))

species_lookup <- fishstat::species %>%
  filter(!is.na(scientific)) %>%
  mutate(.sci_norm = norm(scientific))

top10_matched <- top10 %>%
  mutate(.sci_norm = norm(scientific_name)) %>%
  left_join(species_lookup %>% dplyr::select(species, .sci_norm, fao_species_name = species_name),
            by = ".sci_norm")

unmatched <- top10_matched %>% filter(is.na(species)) %>%
  dplyr::select(rank, scientific_name, common_name, total_tonnes)
matched <- top10_matched %>% filter(!is.na(species))

kable(top10_matched %>% dplyr::select(rank, scientific_name, common_name, fao_species_name),
      caption = "Top-10 species matched to FAO scientific names (NA = unmatched)")
```

```{r part-b-unmatched-note, eval=run_part_b, results='asis'}
if (nrow(unmatched) > 0) {
  cat("**", nrow(unmatched), "of the top", N_TOP_SPECIES,
      "species could not be matched to an FAO scientific name** (shown as NA above). ",
      "They're excluded from the comparison below but remain fully present in Part A.\n\n",
      sep = "")
} else {
  cat("All", N_TOP_SPECIES, "top species matched an FAO scientific name.\n\n")
}
```

## Species × country comparison

```{r part-b-compare, eval=run_part_b}
fao_species <- capture %>%
  filter(area %in% med_area$area,
         country %in% target_countries$country,
         species %in% matched$species,
         measure %in% tonnes_measures$measure,
         year >= YEAR_MIN, year <= YEAR_MAX) %>%
  left_join(target_countries %>% dplyr::select(country, country_name), by = "country") %>%
  left_join(matched %>% dplyr::select(species, scientific_name, common_name), by = "species") %>%
  group_by(country = country_name, scientific_name, common_name, year) %>%
  summarise(fao_tonnes = sum(value, na.rm = TRUE), .groups = "drop")

sau_species <- raw_top10 %>%
  group_by(country, scientific_name = .sci, year) %>%
  summarise(sau_tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")

comparison <- full_join(fao_species, sau_species, by = c("country", "scientific_name", "year")) %>%
  arrange(scientific_name, country, year) %>%
  mutate(diff = sau_tonnes - fao_tonnes,
         pct_diff = ifelse(!is.na(fao_tonnes) & fao_tonnes != 0, 100 * diff / fao_tonnes, NA))

species_summary <- comparison %>%
  filter(!is.na(fao_tonnes), !is.na(sau_tonnes)) %>%
  group_by(scientific_name, common_name, country) %>%
  summarise(n_years = n(),
            mean_fao_tonnes = mean(fao_tonnes, na.rm = TRUE),
            mean_sau_tonnes = mean(sau_tonnes, na.rm = TRUE),
            mean_pct_diff = mean(pct_diff, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(scientific_name, country)

kable(species_summary, digits = 1,
      caption = "SAU vs FAO by species x country (mean_pct_diff: positive = SAU higher)")
```

All matched top-10 species are plotted below (not just a sample) --
one panel per species, summed across the 6 countries. For a per-country
breakdown of any single species, filter `comparison` (written to the
`Comparison` sheet of the exported workbook below) by `country`.

```{r part-b-plot, eval=run_part_b, fig.width=11, fig.height=10}
species_for_plot <- unique(comparison$scientific_name)
plot_data <- comparison

if (nrow(plot_data) > 0 && length(species_for_plot) > 0) {
  ncol_grid_sp <- ceiling(sqrt(length(species_for_plot)))
  nrow_grid_sp <- ceiling(length(species_for_plot) / ncol_grid_sp)
  par(mfrow = c(nrow_grid_sp, ncol_grid_sp), mar = c(4, 4, 3, 1))
  for (sp in species_for_plot) {
    d <- plot_data %>% filter(scientific_name == sp) %>%
      group_by(year) %>%
      summarise(fao_tonnes = sum(fao_tonnes, na.rm = TRUE),
                sau_tonnes = sum(sau_tonnes, na.rm = TRUE), .groups = "drop") %>%
      arrange(year)
    common_lbl <- unique(plot_data$common_name[plot_data$scientific_name == sp])[1]
    yrange <- suppressWarnings(range(c(d$fao_tonnes, d$sau_tonnes), na.rm = TRUE))
    plot(d$year, d$fao_tonnes, type = "l", col = "steelblue", lwd = 2,
         ylim = yrange, xlab = "Year", ylab = "Tonnes",
         main = ifelse(is.na(common_lbl), sp, common_lbl), cex.main = 0.9)
    lines(d$year, d$sau_tonnes, col = "firebrick", lwd = 2, lty = 2)
    legend("topleft", legend = c("FAO", "SAU"), col = c("steelblue", "firebrick"),
           lty = c(1, 2), lwd = 2, bty = "n", cex = 0.7)
  }
}
```

## Export to Excel

```{r part-b-export, eval=run_part_b}
if (requireNamespace("openxlsx", quietly = TRUE)) {
  notes <- c(
    "FAO figures: fishstat R package, Capture Quantity table, FAO Major Fishing Area 37,",
    "tonnage-based measures only, matched species, 1994-2019.",
    "SAU figures: per-EEZ raw extracts (Part 0 of this document), aggregated to",
    "country x species x year.",
    "Species matched by scientific name (case-insensitive, trimmed) -- see Unmatched_species.",
    "FAO does not report catch by gear -- this comparison stops at species x country x year.",
    "No true Western-Mediterranean subdivision exists in FAO's own classification --",
    "Area 37 is the finest available and includes Italy's Adriatic and Ionian coasts.",
    "Data vintage: fishstat runs through 2024; SAU stops at 2019. Capped at 2019 here."
  )
  xlsx_path <- "sau_vs_fao_species_comparison.xlsx"
  wb <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(wb, "Comparison"); openxlsx::writeData(wb, "Comparison", comparison)
  openxlsx::addWorksheet(wb, "SpeciesSummary"); openxlsx::writeData(wb, "SpeciesSummary", species_summary)
  openxlsx::addWorksheet(wb, "Unmatched_species"); openxlsx::writeData(wb, "Unmatched_species", unmatched)
  openxlsx::addWorksheet(wb, "Notes"); openxlsx::writeData(wb, "Notes", data.frame(note = notes), colNames = FALSE)
  openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)
  cat("Wrote", xlsx_path, "\n")
} else {
  cat("Package 'openxlsx' not installed -- skipping the .xlsx export.\n")
}
```

---

# Part C: Executive summary across all dimensions

One table per country (region), pulling together every dimension covered
above -- total SAU catch, its top sector and top gear (with each one's
share), its top species, and the SAU-vs-FAO comparison at country level.
Use this as the index into the rest of the document: each row points
back to the detailed section for that dimension.

```{r part-c-summary}
total_by_country <- raw %>% group_by(country) %>% summarise(total_tonnes = sum(.tonnes, na.rm = TRUE), .groups = "drop")

top_sector_by_country <- if (have_sector_col) {
  sector_summary %>% group_by(country) %>% slice_max(tonnes, n = 1, with_ties = FALSE) %>%
    transmute(country, top_sector = sector, top_sector_pct = round(pct_of_country_catch, 1))
} else {
  tibble(country = total_by_country$country, top_sector = NA_character_, top_sector_pct = NA_real_)
}

top_gear_by_country <- gear_summary_full %>% group_by(country) %>% slice_max(tonnes, n = 1, with_ties = FALSE) %>%
  transmute(country, top_gear = gear, top_gear_pct = round(pct_of_country_catch, 1))

top_species_by_country <- species_country_share %>% group_by(country) %>% slice_max(species_tonnes, n = 1, with_ties = FALSE) %>%
  transmute(country, top_species = common_name, top_species_pct = round(pct_of_country_catch, 1))

fao_summary_by_country <- if (exists("region_comparison") && is.data.frame(region_comparison) && nrow(region_comparison) > 0) {
  region_comparison %>%
    filter(!is.na(fao_tonnes), !is.na(sau_tonnes)) %>%
    group_by(country) %>%
    summarise(mean_fao_tonnes = round(mean(fao_tonnes, na.rm = TRUE), 0),
              mean_sau_tonnes_area37 = round(mean(sau_tonnes, na.rm = TRUE), 0),
              mean_pct_diff_sau_vs_fao = round(mean(pct_diff, na.rm = TRUE), 1),
              .groups = "drop")
} else {
  message("SAU-vs-FAO country comparison isn't available (fishstat not installed, or the FAO-Area-37 ",
          "sheet came back empty) -- those columns will show as NA below.")
  tibble(country = total_by_country$country, mean_fao_tonnes = NA_real_,
         mean_sau_tonnes_area37 = NA_real_, mean_pct_diff_sau_vs_fao = NA_real_)
}

master_summary <- total_by_country %>%
  left_join(top_sector_by_country, by = "country") %>%
  left_join(top_gear_by_country, by = "country") %>%
  left_join(top_species_by_country, by = "country") %>%
  left_join(fao_summary_by_country, by = "country") %>%
  arrange(desc(total_tonnes))

kable(master_summary, digits = 1,
      caption = paste0("Executive summary by country (region): sector, gear, species, and FAO comparison, ",
                        YEAR_MIN, "-", YEAR_MAX,
                        " (top_sector/top_gear/top_species % = share of that country's total SAU catch; ",
                        "mean_pct_diff_sau_vs_fao: positive = SAU higher, both bounded to FAO Area 37)"))
```

```{r part-c-export}
if (requireNamespace("openxlsx", quietly = TRUE)) {
  xlsx_path <- "sau_full_analysis_summary.xlsx"
  wb <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(wb, "ExecutiveSummary"); openxlsx::writeData(wb, "ExecutiveSummary", master_summary)
  openxlsx::addWorksheet(wb, "Top10Species"); openxlsx::writeData(wb, "Top10Species", top10)
  if (have_sector_col) { openxlsx::addWorksheet(wb, "SectorByCountry"); openxlsx::writeData(wb, "SectorByCountry", sector_summary) }
  openxlsx::addWorksheet(wb, "GearByCountry"); openxlsx::writeData(wb, "GearByCountry", gear_summary_full)
  openxlsx::addWorksheet(wb, "SpeciesByCountry"); openxlsx::writeData(wb, "SpeciesByCountry", species_country_share)
  if (exists("region_comparison") && is.data.frame(region_comparison) && nrow(region_comparison) > 0) {
    openxlsx::addWorksheet(wb, "CountryVsFAO"); openxlsx::writeData(wb, "CountryVsFAO", region_comparison)
  }
  if (exists("species_summary") && is.data.frame(species_summary) && nrow(species_summary) > 0) {
    openxlsx::addWorksheet(wb, "SpeciesVsFAO"); openxlsx::writeData(wb, "SpeciesVsFAO", species_summary)
  }
  openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)
  cat("Wrote", xlsx_path, "-- one workbook with every dimension (sector, gear, species, country, FAO comparison) together.\n")
} else {
  cat("Package 'openxlsx' not installed -- skipping the combined .xlsx export (tables above are still in this report).\n")
}
```

---

# Citation

Sea Around Us: [seaaroundus.org/citation-policy](https://www.seaaroundus.org/citation-policy/)

FAO FishStat (via the `fishstat` R package): [github.com/sofia-taf/fishstat](https://github.com/sofia-taf/fishstat)